### 2012-03-03 thesis.tex
init thesis.tex guide.tex

\documentclass{article}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{upgreek}
\usepackage{epsfig}
\usepackage[bottom]{footmisc}

\begin{document}
\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
\author{Benjamin Blum (\textbf{bblum})}
\maketitle

\newcommand\true{\;\textit{true}}
\newcommand\false{\;\textit{false}}

\newcommand\alpher\alpha
\newcommand\beter\beta
\newcommand\gammer\gamma
\newcommand\delter\delta
\newcommand\zeter\zeta
\newcommand\Sigmer\Sigma

\newcommand\NN{\mathbb{N}}
\newcommand\QQ{\mathbb{Q}}
\newcommand\RR{\mathbb{R}}
\newcommand\ZZ{\mathbb{Z}}

\begin{abstract}

\end{abstract}

\section{Introduction}

\section{Related Work}

\section{Design and Implementation}
\subsection{Components of Landslide}
\subsection{On Simics}

\section{Evaluation}

\section{Conclusion}

\end{document}

### 2012-03-03 userguide/guide.tex
init thesis.tex guide.tex

--- 2012-03-03 thesis.tex
@@
 \captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
+\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space \\ User Guide}
 \author{Benjamin Blum (\textbf{bblum})}
@@
 
-\section{Related Work}
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-\subsection{On Simics}
-
-\section{Evaluation}
-
-\section{Conclusion}
+Race conditions are notoriously difficult to debug. Because of their nondeterministic nature, they often do not manifest at all during testing, and when they do manifest, it is difficult to reproduce
 

### 2012-03-04 userguide/guide.tex
update guide

--- 2012-03-03 thesis.tex
@@
 \captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
+\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space \\ (User Guide)}
 \author{Benjamin Blum (\textbf{bblum})}
@@
 \begin{abstract}
-
 \end{abstract}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Related Work}
+Race conditions are notoriously difficult to debug.
+Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
+In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
 
-\section{Design and Implementation}
-\subsection{Components of Landslide}
+Landslide is an effort to make easier the process of debugging kernel-space races.
+It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
+During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
+When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
+
+With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
+Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Landslide's View of the World}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{On Simics}
 
-\section{Evaluation}
+Landslide runs as a Simics module.
+When running the kernel, Simics calls into Landslide once every time the kernel executes an instruction or performs a memory read or write.
+Landslide uses this information to determine ``what the kernel is doing'' at each instruction, and triggers timer interrupts in certain patterns to cause a currently-running kernel thread to get preempted by another one.
 
-\section{Conclusion}
+Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the execution state at certain points in the execution of the test.
+This enables it to try multiple possibilities in terms of which kernel threads run when.
+
+Once Simics is run with a custom \texttt{config.simics}, the above is all done automatically.
+
+\subsection{Components of Landslide}
+
+Between getting called at every instruction and deciding when to trigger timer interrupts or checkpoint or restore execution state, Landslide needs to model the kernel's behaviour and the concurrency properties of the test case.
+Several different components help achieve this.
+
+\subsubsection{Kernel-specific Hooks}
+%TODO
+
+\subsubsection{Scheduling}
+\subsubsection{Decision Points}
+\subsubsection{The Decision Tree}
+
+Modelling the decisions and interleavings in this way presents the ``mysterious nondeterminism'' of the race condition in a new way: If a test case can race, then the resulting bug will appear in some, but not all, of the branches of the tree.
+Therefore, ``systematic dynamic race detection'' means to explore all branches of the tree and test for the bug in each one.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{What to Expect}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+When working with Landslide, please remember to keep realistic expectations about the nature of the tool.
+
+Landslide is not a stress tester or a fuzzer, like the \texttt{cho} family of test cases.
+Tests such as \texttt{cho} attempt to break the kernel by making many system calls in as many different calling patterns as possible, and hoping that some of them will fail.
+Landslide (when paired with a Landslide-friendly test case - see section~\ref{sec:testcase}) attempts to break the kernel by exploring all possible interleavings of kernel threads in one particular pattern of a few system calls.
+Systematic testing (Landslide) and stress testing (\texttt{cho}) are orthogonal testing strategies.
+\footnote{Also not that it is completely infeasible to run \texttt{cho} under Landslide.
+Because the test case does so much stuff on its own, the decision tree will be enormous, and Landslide will not be able to make any meaningful progress trying to explore it.}
+
+Landslide is not a ``race condition oracle''.
+If you run Landslide with a certain test case and a certain set of decision points, and it finishes exploring the decision tree and found no bugs, it does not mean that the system calls that got executed are free of race conditions.
+It does, however, mean that the interactions caused by the test case's pattern of system calls do not race at the granularity of the specific decision points chosen.
+\footnote{Note that if you chose every instruction to be a choice point, the granularity would be perfect, and then you could say there are no races in the specific test case.
+This would make the decision tree impossibly large, though.}
+
+Landslide is a {\em framework} that enables you, the programmer, to explore properties of a concurrent test case that you previously could not.
+It relies on you to configure it accurately in accordance with your goals: whether you are seeking to find a race that you suspect exists, or seeking to show that your kernel doesn't race on a particular test, it is your responsibility to choose a set of choice points that is relevant, granular enough to yield a meaningful exploration, yet minimal enough that Landslide can finish in a reasonable amount of time.
+More on this in section~\ref{sec:choice}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Running Landslide}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Kernel Requirements}
+% TODO
+
+\subsection{Instrumenting Your Kernel with Landslide}
+
+This is the hardest part of using Landslide. Please pay extra attention to this section.
+
+Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.
+The file \texttt{kernel\_specifics.c} contains several ``hook'' functions, which landslide uses to know what the guest kernel is doing, and which you need to implement for your kernel.
+
+One step towards doing this is to generate a header file containing \texttt{\#define}s for the addresses of certain functions and global variables in your kernel.
+We provide a script, \texttt{definegen.sh}, which makes it easy to extract such addresses from your kernel image, and which automatically generates the header file.
+% TODO
+
+\begin{enumerate}
+	\item
+% TODO
+\end{enumerate}
+
+\subsection{Editing Your Kernel}
+
+A design goal of the instrumentation process was that you should never need to change your kernel itself to allow it to run on Landslide. (After all, changing a program with bugs can change the bugs!)
+Nevertheless, some types of changes to the kernel may prove {\em very helpful} in making your job of using Landslide easier.
+
+\begin{itemize}
+	\item When implementing hooks that tell where in a function or where in the codebase a specific event happens, it is best to refactor code so that the event happens at one centralised place.
+		For example, when implementing \texttt{kern\_thread\_runnable} and \texttt{kern\_thread\_descheduling}, make sure there is one function in the entire kernel that is always called when adding or removing threads to the runqueue.
+		\footnote{This is especially relevant to \texttt{variable\_queue} users - using macros like \texttt{Q\_INSERT\_HEAD} in many places will cause you to have to instrument each place.
+Instead, write a one-line \texttt{add\_to\_runqueue} helper function that invokes the macro.}
+		\footnote{Also make sure when instrumenting helper functions such as these that they are {\em not} marked \texttt{static}!
+When using \texttt{gcc -O1} to build the kernel, \texttt{static} functions may get optimised in ways that violate standard calling convention, which Landslide relies on.}
+	\item Landslide's most reliable found-a-bug check is checking for \texttt{panic} getting called.
+		The more important invariants for which you have \texttt{assert}s in your kernel, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
+\end{itemize}
+
+\subsection{Userspace Test Cases}
+\label{sec:testcase}
+
+We ship Landslide with a suite of small test cases designed to expose several common bugs that students frequently encounter during 15-410 Project 3.
+
+\begin{itemize}
+	\item \texttt{vanish\_vanish} - Tests when a parent and child process \texttt{vanish()} simultaneously. (This test is identical to \texttt{fork\_test1} from the P3 hurdle suite.)
+	\item \texttt{fork\_wait} - Tests basic interaction of \texttt{fork()} and \texttt{wait()}. (This test is from the P3 hurdle suite.)
+	\item \texttt{wild\_test2} - Tests simultaneous thread death by unnatural causes. (This test is an upgrade of \texttt{wild\_test1} from the P3 hurdle suite.)
+	\item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child. (new)
+	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in the same process vanishing. (new)
+\end{itemize}
+
+It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space
+
+\subsection{Configuring Choice Points}
+\label{sec:choice}
+% TODO
+
+\subsection{Interpreting Landslide's Results}
+% TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Feedback}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO
 

### 2012-03-05 userguide/guide.tex
more guide

@@
 \begin{abstract}
+% TODO
 \end{abstract}
@@
 
-Landslide is an effort to make easier the process of debugging kernel-space races.
+Landslide
+\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+is an effort to make easier the process of debugging kernel-space races.
 It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
@@
 \subsubsection{Kernel-specific Hooks}
-%TODO
+
+The file \texttt{kernel\_specifics.c} contains a set of functions that inform Landslide about the current state of the kernel.
+These typically involve comparing \texttt{\%eip} against certain addresses (to know what function is being executed), and sometimes reading memory from the stack or other variables (to know the arguments passed to functions or the state of certain data structures).
+
+An example hook is the function \texttt{kern\_thread\_runnable}, which tells whether a thread is about to be added to the runqueue, and reads that thread's TID out of memory and returns it if so.
+
+These are different for every kernel, so you need to implement these. Specific details are in section~\ref{sec:hooks}.
 
 \subsubsection{Scheduling}
+
+From several of the hooks, Landslide is able to maintain a queue of threads that mirrors the kernel's runqueue, and hence know which threads are runnable at any given point during the execution.
+(See section~\ref{sec:schedfunc}.)
+
+At any point during the test (provided interrupts are on and preemption is enabled), Landslide may choose to preempt the currently-running thread and run somebody else.
+This is done by triggering successive timer interrupts to induce context switches until the desired thread is switched to.
+
 \subsubsection{Decision Points}
+
+A perfectly accurate race detection tool would need to consider every single instruction
+\footnote{actually, just every single access to shared memory while interrupts are on}
+as a potential point at which a preemption might expose a race condition.
+This would make enumerating all interleavings impossibly expensive, however, so a feasible race detection tool needs to have intelligent heuristics to judge which instructions are more likely to be places where preemptions will trigger race conditions.
+At these ``decision points'', then, Landslide will choose whether to continue running the current thread or to preempt and run a different one.
+
+Good decision points are most likely to be voluntary reschedules (i.e., calls to \texttt{yield} or similar - these are mandatory to have as decision points), calls to synchronisation primitives, accesses to shared data structures, and so on.
+
+You do not need to configure the set of decision points in order to use landslide, but doing so is certainly more likely than using the default set. More on this in section~\ref{sec:choice}.
+
 \subsubsection{The Decision Tree}
 
-Modelling the decisions and interleavings in this way presents the ``mysterious nondeterminism'' of the race condition in a new way: If a test case can race, then the resulting bug will appear in some, but not all, of the branches of the tree.
+After one execution of the test case, Landslide analyses each decision point that was encountered, and chooses from one of them a thread that was runnable but different from the one that was run.
+It then uses Simics bookmarks to ``time travel'' to that decision point, and causes timer interrupts in a pattern that causes the kernel to schedule the newly chosen thread, executes the test with this new interleaving, and repeats.
+In this way we see that the many possible interleavings, as defined by the set of decision points, form a tree, which represents all possible ways the threads can run concurrently.
+
+Modelling the decisions and interleavings as a tree presents the ``mysterious nondeterminism'' of the race condition in a new way: If a test case can race, then the resulting bug will appear in some, but not all, of the branches of the tree.
 Therefore, ``systematic dynamic race detection'' means to explore all branches of the tree and test for the bug in each one.
+
+You might imagine that if Landslide had to try each runnable thread at each decision point, exponential explosion would make most trees take forever to explore.
+Landslide uses an algorithm called {\em dynamic partial order reduction} to prune redundant branches out of the tree to speed up exploration. %TODO: reference flanagan here
+An intuitive explanation of the pruning: Imagine that at decision point A, threads 1 and 2 are runnable, and that they don't interfere with each other, so that running 1 followed by 2 yields the same state as running 2 followed by 1.
+Partial order reduction identifies redundancies resulting from such ``independent'' transitions, and eliminates them.
+You don't need to worry about partial order reduction itself, except to realise that using it means that adding extra decision points will increase the size of the tree, but usually not cause exponential explosion.
 
@@
 \subsection{Kernel Requirements}
-% TODO
+
+\subsubsection{Scheduler Functionality}
+\label{sec:schedfunc}
+
+In order to cause desired preemptions, Landslide needs to assume that the core of your scheduler works.
+In short, this means that timer interrupts will trigger context switches, and that if a thread is ``runnable'' (see section \ref{sec:hooks}), at any point where interrupts/preemption is enabled, a finite number of timer preemptions will eventually cause that thread to run.
+
+\subsubsection{VM}
+
+Landslide does not have much to do with the VM, but your kernel must direct-map all memory below \texttt{USER\_MEM\_START}.
+
+\subsubsection{System Calls}
+
+In order to run with Landslide, your kernel must be able to boot up to the shell prompt, receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes.
+This means you must have the following system calls implemented and working in at least a rudimentary way:
+\begin{itemize}
+	\item \texttt{fork}
+	\item \texttt{exec}
+	\item \texttt{vanish}
+	\item \texttt{wait}
+	\item \texttt{readline}
+\end{itemize}
+The following system calls are exercised in some, but not all, of the test cases we distribute, so will be helpful to have:
+\begin{itemize}
+	\item \texttt{thread\_fork}
+	\item \texttt{yield}
+\end{itemize}
+Apart from the system calls which init and shell execute, the rest are irrelevant and do not need to be implemented to use Landslide on your kernel.
 
 \subsection{Instrumenting Your Kernel with Landslide}
-
-This is the hardest part of using Landslide. Please pay extra attention to this section.
+\label{sec:hooks}
+{\large \em This is the hardest part of using Landslide: you will need to write code that tells Landslide what your kernel is doing when, and it is essential that your code is accurate.
+Please pay extra attention to this section.}
 
@@
 
-One step towards doing this is to generate a header file containing \texttt{\#define}s for the addresses of certain functions and global variables in your kernel.
+You need to implement the following functions in \texttt{kernel\_specifics.c}. There are other functions there too, but you should not need to change them.
+\begin{enumerate}
+	\item \texttt{kern\_get\_current\_tcb} - Returns the address of the TCB for the currently-running thread.
+	\item \texttt{kern\_get\_current\_tid} - Returns the TID of the currently-running thread.
+	\item \texttt{kern\_scheduler\_locked} - Tells whether or not the scheduler is ``locked''; i.e., whether a timer interrupt at the current instruction would {\em not} cause a preemption.
+(In kernels that use explicit scheduler locking, this is typically a global variable.)
+	\item \texttt{kern\_mutex\_locking} - 
+	\item \texttt{kern\_mutex\_blocking} - 
+	\item \texttt{kern\_mutex\_locking\_done} - 
+	\item \texttt{kern\_mutex\_unlocking} - 
+	\item \texttt{kern\_mutex\_unlocking\_done} - 
+	\item \texttt{kern\_thread\_runnable} - Is a thread about to be added to the runqueue? Also returns the TID of that thread.
+	\item \texttt{kern\_thread\_descheduling} - Is a thread about to be removed from the runqueue? Also returns the TID of that thread.
+	\item \texttt{kern\_address\_own\_kstack} - Is a memory access within the current thread's stack region? % TODO: remove the need for this at all
+	\item \texttt{kern\_address\_other\_kstack} - 
+	\item \texttt{kern\_get\_init\_tid} - Returns the TID assigned to init. Typically 1.
+	\item \texttt{kern\_get\_idle\_tid} - Returns the TID of idle, if the kernel uses an idle thread. (This will be called only if \texttt{kern\_has\_idle} returns true.)
+	\item \texttt{kern\_get\_shell\_tid} - Returns the TID assigned to the shell. Typically 2.
+	\item \texttt{kern\_get\_first\_tid} - Returns the TID of the first thread to run (typically init's TID).
+	\item \texttt{kern\_has\_idle} - Returns whether or not the kernel uses an explicit thread for the idle loop.
+	\item \texttt{kern\_init\_threads} - Given a callback function, calls it once per thread that exists on the system after the scheduler is initialised (also with flags for whether each of those threads are on the runqueue and/or in the middle of a context switch).
+	\item \texttt{kern\_fork\_returns\_to\_cs} - Returns whether or not a newly-forked thread returns through the context-switcher path the first time it runs.
+	\item \texttt{kern\_fork\_return\_spot} - Returns whether or not the current instruction pointer is at the location that a newly-forked thread returns to.
+(If \texttt{kern\_fork\_returns\_to\_cs} returns true, this function should always return false.)
+	\item \texttt{kern\_current\_extra\_runnable} - Returns whether or not the current thread is {\em not} on the runqueue, but is ``runnable`` anyway.
+(This might involve reading a flag out of the current TCB.) (For kernels that keep the current thread on the runqueue, this function should always return false.)
+\end{enumerate}
+One step towards implementing these functions (and towards making some already-implemented functions work properly) is to generate a header file containing \texttt{\#define}s for the addresses of certain functions and global variables in your kernel.
 We provide a script, \texttt{definegen.sh}, which makes it easy to extract such addresses from your kernel image, and which automatically generates the header file.
 % TODO
-
+% kern_timer_entering/exiting
+% kern_get_timer_wrap_begin
+% kern_context_switch_entering/exiting
+% kern_sched_init_done
+% kern_in_scheduler
+% kern_access_in_scheduler
+% kern_mutex_ignore
+% kern_forking/vanishing/sleeping/readline
 \begin{enumerate}
-	\item
-% TODO
+	\item % TODO - clean this script up a ton
 \end{enumerate}
+There are several macros / interfaces that are already defined for your implementing convenience.
+\footnote{Peruse \texttt{x86.h} for more.}
+\begin{itemize}
+	\item \texttt{GET\_CPU\_ATTR(cpu, name)} - returns the value of the cpu's specified register. (e.g., write \texttt{eax} for \texttt{name}).
+	\item \texttt{READ\_MEMORY(cpu, addr)} - retrieves the 4-byte value at the given address.
+	\item \texttt{READ\_BYTE(cpu, addr)} - retrieves the 1-byte value at the given address.
+	\item \texttt{READ\_STACK(cpu, slot)} - equivalent to \texttt{READ\_MEMORY(cpu, \%esp + 4*slot)}.
+		(useful, for example, if you're at the first instruction of a function; \texttt{READ\_STACK(cpu, N)} returns the \texttt{N}th argument.)
+\end{itemize}
+There are also several functions in the \texttt{definegen.sh} script to help you generate the header file.
+\begin{itemize}
+	\item \texttt{get\_sym x} - Gets the address of the global symbol \texttt{x}.
+	\item \texttt{get\_func x} - Gets the address of the function \texttt{x}.
+	\item \texttt{get\_func\_end x} - Gets the address of the {\em last} instruction (almost always \texttt{ret} or \texttt{iret}) in the function \texttt{x}.
+	\item \texttt{get\_last\_such\_instruction x y} - Gets the address of the last instruction in function \texttt{x} that matches pattern \texttt{y}.
+\end{itemize}
 
@@
 	\item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child. (new)
-	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in the same process vanishing. (new)
+	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in one process vanishing. (new)
+		\footnote{Really, the point of this test is to expose a use-after-free bug if your \texttt{thread\_fork} implementation has ``\texttt{return child->tid;}'' as its last line of code.}
+	\item \texttt{yield\_vanish} - Tests for interactions between \texttt{yield()} and \texttt{vanish()}. (new)
+		\footnote{You may wonder why some of these are written in assembly.
+This is because using your userspace thread library for thread creation (a) is not guaranteed to work, and (b) necessitates many more system calls, which dramatically increases the size of the decision tree.}
 \end{itemize}
 
-It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space
+It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space.
+They also run several, but not too many, threads at once, producing potentially interleavings.
+
+To warm up, we recommend you test \texttt{double\_thread\_fork} (and edit your kernel explicitly to make sure it has a bug that this will expose - see the footnote in the test's description above).
+After that, we recommend trying \texttt{vanish\_vanish}, which is the most likely to expose many different types of bugs in many different \texttt{vanish} implementations.
 

### 2012-03-09 thesis.tex
begin writing thesis intro; add todo comments everywhere

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 \usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
@@
 
+Race conditions are notoriously difficult to debug.
+Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
+In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
+
+% Write here a paragraph about userspace techniques. Write about kernel races.
+
+Landslide
+\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+is an effort to make easier the process of debugging kernel-space races.
+It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
+During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
+When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
+
+With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
+Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
+
+% ``The contributions of this work are as follows.''
+In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
+We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
+% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
+% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
+We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
+We show that Landslide is an effective tool for both purposes.
+
 \section{Related Work}
+% Look up 'nooks'? re kernel space.
+
+\section{Challenges of Kernel-space}
+% concurrency implementation itself
+% inter-thread communication is not idealised!
 
@@
 \subsection{Components of Landslide}
+
 \subsection{On Simics}
 
+\subsection{Techniques}
+% use-after-free, partial order, etc
+% talk about the all-ancestors vs induction thing
+% how to even schedule threads
+
+% Have a section or subsection to talk about the tool-user relationship.
+
 \section{Evaluation}
+% 410 - talk about current state/methods (for students and for TAS)
+% test suite presented
+% process of instrumenting
+
+\section{Future Work}
+% new techniques for landslide:
+% - hybrid with data-race/lockset/static-analysis; shm choices
+% - parallelism
+% - ICB as in chess
+% - garth's shaping
+% linux, device drivers
+% VM instead of simics
+% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
+% - how to time travel? instead of time travel, snapshot once and replay
+% embedded microcontroller operating systems, simulated
+% teaching tool, study thinking patterns
 
@@
 
+\section{Acknowledgements}
+% Save this for last :)
+
 \end{document}

### 2012-04-23 thesis.tex
add citations and abstract

@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
+\documentclass[10pt,twocolumn]{article}
+%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
+\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
 \usepackage{upgreek}
-\usepackage{epsfig}
+%\usepackage{times}
+\usepackage[charter]{mathdesign}
+\usepackage{hyperref}
+\usepackage{draftwatermark}
+%\usepackage{epsfig}
 \usepackage[bottom]{footmisc}
@@
 \begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
+%\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
 \title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
+\author{Benjamin Blum (\texttt{bblum@andrew.cmu.edu})}
 \maketitle
@@
 \begin{abstract}
+Systematic exploration is an approach to finding race conditions by deterministically executing every possible interleaving of thread transitions and identifying which ones expose bugs.
+Current systematic exploration techniques are suitable for testing user-space programs, but are inadequate for testing kernels, where the testing framework's control over concurrency is more complicated.
 
+We present Landslide, a systematic exploration tool for finding races in kernels.
+Landslide targets Pebbles, the kernel specification that students implement in the undergraduate Operating Systems course at Carnegie Mellon University (15-410).
+We discuss the techniques Landslide uses to address the general challenges of kernel-level concurrency, and we evaluate its effectiveness and usability as a debugging aid.
+We show that our techniques make systematic testing in kernel-space feasible, and that Landslide is a useful tool for doing so in the context of 15-410.
 \end{abstract}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
 
-% Write here a paragraph about userspace techniques. Write about kernel races.
+In kernel-space, race condition debugging becomes even more difficult.
+Many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
+
+% TODO: talk about systematic exploration
+%It provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
 
@@
 is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
+It is geared towards kernels that meet the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
 During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
@@
 
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
+% TODO: this paragraph is incongruous in light of the above three; what audience are you writing for, anyway?
+Many techniques exist for dynamic testing of concurrent systems for race conditions.
+Systematic exploration, the strategy that Landslide uses, involves modelling a concurrent test case as a series of {\em transitions} by each of a set of {\em agents}
+\footnote{In this work, we equate agents with a kernel's scheduler threads.}
+between {\em decision points} (i.e., actions which the tool and/or the user determine to be relevant to the concurrent behaviour of the test), and attempting to exercise all possible orderings of the transitions in hopes that one or more ordering will expose a bug.
+Within this strategy
+
+In this work, we present Landslide as a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
 We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Related Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Look up 'nooks'? re kernel space.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Challenges of Kernel-space}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % concurrency implementation itself
 % inter-thread communication is not idealised!
+% kernels are designed in different ways; what can we assume
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Design and Implementation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Components of Landslide}
+
+\subsubsection{Kernel Instrumentation}
+
+\subsubsection{Scheduling}
+
+\subsubsection{The Arbiter}
+
+\subsubsection{The Decision Tree}
 
@@
 
+Landslide is implemented as a module for Simics\cite{simics}
+
 \subsection{Techniques}
-% use-after-free, partial order, etc
+
+\subsubsection{Thread Scheduling}
+
+\subsubsection{Identifying Bugs}
+
+% A healthy dose of snark. Perhaps cite the formally-proved haskell kernel here.
+Without a formal specification of the internals of the guest kernel's implementation, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
+We list the metrics here in order of reliability.
+\begin{enumerate}
+	\item Kernel panic bugs - 
+	\item Use-after-free bugs - 
+	\item Deadlock bugs - 
+	\item Memory leak bugs - % this could fail if..
+	\item Infinite loop bugs - f
+\end{enumerate}
+Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
+If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed above).
+
+\subsubsection{Partial-Order Reduction}
 % talk about the all-ancestors vs induction thing
-% how to even schedule threads
+
+\subsubsection{Selective Ignorance}
+
+\subsection{Debugging Feedback}
+
+In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context-switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
+When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
+
+The decision trace explains each preemption or voluntary reschedule in the interleaving: which thread used to be running, which thread was chosen to run instead, and the stack trace of the former thread at the point from which it was switched away.
+
+Landslide is impleAdditional information, such as the number of instructions executed during each transition, and the shared memory accesses that conflicted with other transitions, can also be printed at the user's configuration options.
+% Should it also print conflicting memory accesses? Configurable by the user?
+% Make the user able to replay the choices.
+
+% Discuss combining this with a choice trace minimisation tool. Reference future work.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Evaluation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % 410 - talk about current state/methods (for students and for TAS)
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % new techniques for landslide:
@@
 % teaching tool, study thinking patterns
+% symbolic execution
+% theory things to study:
+% - prove the soundness of the every-evil-ancestor thing
+% - study the implications of backwards vs forward exploration
+% - nadim's backtrack depth
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Save this for last :)
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliography{citations}{}
+\bibliographystyle{alpha}
+
 \end{document}

### 2012-04-23 userguide/guide.tex
final update user guide

--- 2012-03-04 userguide/guide.tex
@@
 \usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
+\usepackage{multirow}
 \usepackage{upgreek}
@@
 
-\begin{abstract}
-\end{abstract}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Landslide is an effort to make easier the process of debugging kernel-space races.
+Landslide\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+is an effort to make easier the process of debugging kernel-space races.
 It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{How To Use This Document}
+
+If you want to start using Landslide as soon as possible, skip to appendix~\ref{sec:stepbystep} and follow the step-by-step instructions, referring back to other sections for explanation as referenced.
+
+Try not to get overwhelmed by the details here - you don't need to understand all the text in this document to use Landslide. Think of it like a reference manual.
+
+The slides for the lecture which accompanies this user study are available at \\ \texttt{http://www.cs.cmu.edu/{\textasciitilde}410/landslide-lecture.pdf}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Landslide's View of the World}
+\label{sec:anatomy}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Components of Landslide}
@@
 
-\subsubsection{Kernel-specific Hooks}
-%TODO
+\subsubsection{Kernel Instrumentation}
+
+Landslide needs to know when the kernel performs certain types of actions. The user provides this information in several different ways, as discussed more in section~\ref{sec:hooks}.
+
+The build script, before building Landslide itself, turns this information into a header file which contains \texttt{\#define}s for addresses of functions within the kernel.
+Internally, Landslide compares \texttt{\%eip} against these addresses to know when the kernel runs what function.
+
+Because every kernel is slightly different, the communication between Landslide and your kernel will be unique, so you will need to spend some thought on these. More on this later.
 
 \subsubsection{Scheduling}
+
+From several of the hooks, Landslide is able to maintain a queue of threads that mirrors the kernel's runqueue, and hence know which threads are runnable at any given point during the execution.
+(See section~\ref{sec:schedfunc}.)
+
+At any point during the test (provided interrupts are on and preemption is enabled), Landslide may choose to preempt the currently-running thread and run somebody else.
+This is done by triggering successive timer interrupts to induce context switches until the desired thread is switched to.
+
 \subsubsection{Decision Points}
+
+A perfectly accurate race detection tool would need to consider every single instruction\footnote{actually, just every single access to shared memory while interrupts are on}
+as a potential point at which a preemption might expose a race condition.
+This would make enumerating all interleavings impossibly expensive, however, so a feasible race detection tool needs to have intelligent heuristics to judge which instructions are more likely to be places where preemptions will trigger race conditions.
+At these ``decision points'', then, Landslide will choose whether to continue running the current thread or to preempt and run a different one.
+
+Good decision points are most likely to be voluntary reschedules (i.e., calls to \texttt{yield} or similar - Landslide provides these by default), calls to synchronization primitives, accesses to shared data structures, and so on.
+
+You do not need to configure the set of decision points in order to use Landslide, but doing so is certainly more likely than using the default set. More on this in section~\ref{sec:choice}.
+
 \subsubsection{The Decision Tree}
 
-Modelling the decisions and interleavings in this way presents the ``mysterious nondeterminism'' of the race condition in a new way: If a test case can race, then the resulting bug will appear in some, but not all, of the branches of the tree.
+After one execution of the test case, Landslide analyses each decision point that was encountered, and chooses from one of them a thread that was runnable but different from the one that was run.
+It then uses Simics bookmarks to backtrack to that decision point, and causes timer interrupts in a pattern that causes the kernel to schedule the newly chosen thread, executes the test with this new interleaving, and repeats.
+In this way we see that the many possible interleavings, as defined by the set of decision points, form a tree, which represents all possible ways the threads can run concurrently.
+
+Modelling the decisions and interleavings as a tree presents the ``mysterious nondeterminism'' of the race condition in a new way: If a test case can race, then the resulting bug will appear in some, but not all, of the branches of the tree.
 Therefore, ``systematic dynamic race detection'' means to explore all branches of the tree and test for the bug in each one.
 
+You might imagine that if Landslide had to try each runnable thread at each decision point, exponential explosion would make most trees take forever to explore.
+Landslide uses an algorithm called {\em dynamic partial order reduction} to prune redundant branches out of the tree to speed up exploration.
+An intuitive explanation of the pruning: Imagine that at decision point A, threads 1 and 2 are runnable, and that they don't interfere with each other, so that running 1 followed by 2 yields the same state as running 2 followed by 1.
+Partial order reduction identifies redundancies resulting from such ``independent'' transitions, and eliminates them.
+You don't need to worry about partial order reduction itself, except to realise that using it means that adding extra decision points will increase the size of the tree, but usually not cause exponential explosion.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{What to Expect}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:expect}
 
@@
 Landslide (when paired with a Landslide-friendly test case - see section~\ref{sec:testcase}) attempts to break the kernel by exploring all possible interleavings of kernel threads in one particular pattern of a few system calls.
-Systematic testing (Landslide) and stress testing (\texttt{cho}) are orthogonal testing strategies.
-\footnote{Also not that it is completely infeasible to run \texttt{cho} under Landslide.
+Systematic testing (Landslide) and stress testing (\texttt{cho}) are orthogonal testing strategies.\footnote{Also note that it is completely infeasible to run \texttt{cho} under Landslide.
 Because the test case does so much stuff on its own, the decision tree will be enormous, and Landslide will not be able to make any meaningful progress trying to explore it.}
@@
 If you run Landslide with a certain test case and a certain set of decision points, and it finishes exploring the decision tree and found no bugs, it does not mean that the system calls that got executed are free of race conditions.
-It does, however, mean that the interactions caused by the test case's pattern of system calls do not race at the granularity of the specific decision points chosen.
-\footnote{Note that if you chose every instruction to be a choice point, the granularity would be perfect, and then you could say there are no races in the specific test case.
+It does, however, mean that the interactions caused by the test case's pattern of system calls do not race at the granularity of the specific decision points chosen.\footnote{Note that if you chose every instruction to be a decision point, the granularity would be perfect, and then you could say there are no races in the specific test case.
 This would make the decision tree impossibly large, though.}
@@
 Landslide is a {\em framework} that enables you, the programmer, to explore properties of a concurrent test case that you previously could not.
-It relies on you to configure it accurately in accordance with your goals: whether you are seeking to find a race that you suspect exists, or seeking to show that your kernel doesn't race on a particular test, it is your responsibility to choose a set of choice points that is relevant, granular enough to yield a meaningful exploration, yet minimal enough that Landslide can finish in a reasonable amount of time.
+It relies on you to configure it accurately in accordance with your goals: whether you are seeking to find a race that you suspect exists, or seeking to show that your kernel doesn't race on a particular test, it is your responsibility to choose a set of decision points that is relevant, granular enough to yield a meaningful exploration, yet minimal enough that Landslide can finish in a reasonable amount of time.
 More on this in section~\ref{sec:choice}.
 
+One final thing: It may seem tempting at times to look for ``simpler'' ways of circumventing some of the more complicated parts of the instrumentation process.
+Please do not do this - it will make your instrumentation brittle, and quite likely to break in more mysterious ways.
+If you are confused about what some part of the process is asking you to do, do not guess randomly (it will lead to frustrating crashes!); ask Ben for help instead.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Running Landslide}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Kernel Requirements}
+\label{sec:requirements}
 
-\subsection{Kernel Requirements}
-% TODO
+\subsection{Scheduler Functionality}
+\label{sec:schedfunc}
 
-\subsection{Instrumenting Your Kernel with Landslide}
+In order to cause desired preemptions, Landslide needs to assume that the core of your scheduler works.
+In short, this means that timer interrupts will trigger context switches, and that if a thread is ``runnable'' (see section \ref{sec:runqueue}), at any point where interrupts/preemption is enabled, a finite number of timer preemptions will eventually cause that thread to run.
 
-This is the hardest part of using Landslide. Please pay extra attention to this section.
+For deadlock detection, if you have mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:tell}.
+If you have mutexes which explicitly deschedule blocking threads, no special annotations are needed.
+
+It is not okay to spin-wait anywhere, whether in mutexes or otherwise, especially in \texttt{readline} or \texttt{sleep}.
+Relatedly, the kernel must never run the idle loop when not truly idle, because Landslide uses this as a way of telling when all threads are wedged and/or when the test is done running.
+
+\subsection{VM}
+
+Landslide does not have much to do with the VM, but your kernel must direct-map most of kernel memory (i.e.
+below \texttt{USER\_MEM\_START}), including the heap, globals, and kernel thread stacks.
+Landslide needs this to read values out of memory and to identify memory conflicts.
+
+Your kernel should also use LMM (the \texttt{malloc} package in the base code) for dynamic allocation, with the default configuration that the base code provides.
+
+\subsection{System Calls}
+
+In order to run with Landslide, your kernel must be able to boot up to the shell prompt, receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes.
+This means you must have the following system calls implemented and working in at least a rudimentary way:
+\texttt{fork}, \texttt{exec}, \texttt{vanish}, \texttt{wait}, \texttt{readline}.
+
+The following system calls are exercised in some, but not all, of the test cases we distribute, so will be helpful to have:
+\texttt{thread\_fork}, \texttt{yield}.
+
+Apart from the system calls which init and shell execute, the rest are irrelevant and do not need to be implemented to use Landslide on your kernel.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Instrumenting Your Kernel with Landslide}
+\label{sec:hooks}
 
 Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.
-The file \texttt{kernel\_specifics.c} contains several ``hook'' functions, which landslide uses to know what the guest kernel is doing, and which you need to implement for your kernel.
 
-One step towards doing this is to generate a header file containing \texttt{\#define}s for the addresses of certain functions and global variables in your kernel.
-We provide a script, \texttt{definegen.sh}, which makes it easy to extract such addresses from your kernel image, and which automatically generates the header file.
-% TODO
+\subsection{Annotations}
+\label{sec:tell}
 
-\begin{enumerate}
-	\item
-% TODO
-\end{enumerate}
+We provide a set of functions that your kernel needs to call at certain points during its execution to communicate what interesting events are happening.
 
-\subsection{Editing Your Kernel}
+\subsubsection{Important Note about Runqueues}
+\label{sec:runqueue}
+Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the scheduler's runqueue at each point during execution.
+Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not your kernel keeps the current thread on the runqueue or off of it.
 
-A design goal of the instrumentation process was that you should never need to change your kernel itself to allow it to run on Landslide. (After all, changing a program with bugs can change the bugs!)
-Nevertheless, some types of changes to the kernel may prove {\em very helpful} in making your job of using Landslide easier.
+If your kernel does not keep the currently-running thread on the runqueue, you'll need to tell Landslide this separately, in \texttt{kern\_current\_extra\_runnable} - see section~\ref{sec:student}.
+
+\subsubsection{Required Annotations}
+\begin{itemize}
+	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in your context switcher, when a new thread is switched to.
+	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when your scheduler is done being initialized.
+Until this is called, Landslide will ignore all other annotations.
+(This corresponds to \texttt{starting\_threads} in section~\ref{sec:config}.)
+	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
+		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)
+
+		Call it ``just before'' the action which makes the new thread runnable - whether your kernel just adds it to the runqueue (in which case the next annotation called would be \texttt{thread\_on\_rq}) or begins running it immediately (in which case \texttt{thread\_switch}); Landslide handles both cases.\footnote{It may help to think of this as setting a ``currently forking'' state flag in the calling thread: when the next annotation is called, Landslide will see the flag is set and act accordingly.
+It is okay if interrupts are on between this call and the subsequent event.}
+	\item \texttt{tell\_landslide\_vanishing()} - Call this when a thread is about to go away.
+		Call it ``just before'' the relevant context switch; i.e., the one that should never return.
+	\item \texttt{tell\_landslide\_sleeping()} - Call this when a thread is about to go to sleep (that's \texttt{sleep()} the system call).
+		Call it ``just before'' the relevant context switch; i.e., the one that won't return for as long as the thread requested to sleep.
+	\item \texttt{tell\_landslide\_thread\_on\_rq(int tid)} - Call this when a thread is about to be added to the runqueue.\footnote{If there are multiple such places, which is likely if you use \texttt{variable\_queue}, we recommend wrapping them in a helper function which contains the annotation.}
+		(Make sure this call is done within whatever protection is also used for the runqueue modification itself.)
+	\item \texttt{tell\_landslide\_thread\_off\_rq(int tid)} - Call this when a thread is about to be removed from the runqueue. (Same protection clause as above applies.)
+	\item Mutex annotations - These are only necessary if your kernel uses mutexes that leave blocked threads on the runqueue, rather than explicitly descheduling them.
+	\begin{itemize}
+		\item \texttt{tell\_landslide\_mutex\_locking(void *mutex\_addr)} - Call this when a thread is about to start locking a mutex.
+Landslide needs the mutex address for deadlock detection, to track which mutex each thread is blocked on.
+		\item \texttt{tell\_landslide\_mutex\_locking\_done()} - Call this when a thread finishes locking a mutex (and now owns it).
+		\item \texttt{tell\_landslide\_mutex\_unlocking(void *mutex\_addr)} - Call this when a thread is about to start unlocking a mutex.
+		\item \texttt{tell\_landslide\_mutex\_unlocking\_done()} - Call this when a thread finishes unlocking a mutex (and no longer owns it).
+		\item \texttt{tell\_landslide\_mutex\_blocking(int owner\_tid)} - Call this when a thread is contending on a locked mutex, and hence is no longer logically ``runnable'' yet still on the runqueue.
+Landslide needs to know who owns the mutex for deadlock detection.
+	\end{itemize}
+\end{itemize}
+
+\subsubsection{Optional Annotations}
+\begin{itemize}
+	\item \texttt{tell\_landslide\_decide()} - Call this to define additional decision points. See section~\ref{sec:choice}.
+	\item \texttt{assert()} or \texttt{panic()} - You should be using these already, but they are extra useful in Landslide.
+		The more important invariants for which you have \texttt{assert}s in your kernel, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
+		\footnote{Relatedly, a useful trick is to use poison values for memory that should not be read from until it is written again, and to use canary members in data structures that should not get scribbled over.
+Functions that operate on such memory can assert that poison values don't appear and that canary values don't change.}
+\end{itemize}
+
+\subsection{Config File}
+\label{sec:config}
+
+There is also a config file you need to fill out.
+Some of the fields are required information, and some are tweaks that change the way Landslide behaves.
+The specific meaning of each field is explained in comments in the file; here we describe their purpose.
+
+\subsubsection{Required Fields}
+\begin{itemize}
+	\item \texttt{KERNEL\_IMG} - The build scripts use \texttt{objdump} on this file to extract code and data addresses.
+	\item \texttt{KERNEL\_SOURCE\_DIR} - Used for file and line number information in stack traces.
+	\item \texttt{TEST\_CASE} - What userspace program should Landslide explore the state space of?
+	\item \texttt{TIMER\_WRAPPER} - Used for controlling timer interrupts and tracking whether a thread is handling an interrupt.
+	\item \texttt{CONTEXT\_SWITCH} - Used for detecting voluntary reschedules.
+	\item \texttt{READLINE} - Used for detecting when the kernel is ready to run a test, and when the test finishes.
+	\item \texttt{INIT\_TID}, \texttt{SHELL\_TID}, \texttt{IDLE\_TID}, \texttt{FIRST\_TID} - Used for initialization, and tracking of test state.
+	\item \texttt{starting\_threads} - This lets you tell Landslide what threads exist at the point where \\ \texttt{tell\_landslide\_sched\_init\_done} will be called.
+You should probably use it for each of init and (if it exists in your kernel) idle (but probably not shell).
+\end{itemize}
+
+\subsubsection{Optimizations}
+
+These configuration options help Landslide identify potentially-conflicting shared memory accesses that it should ignore.
+
+The partial order reduction algorithm achieves state space pruning by identifying {\em independent} transitions between decision points - that is, they that have no conflicting memory accesses.
+Telling Landslide to ignore some common types of shared accesses results in more independence between transitions, and hence more reduction, and hence a faster-running test.
+Of course, ignoring a memory access means you likely won't find races involving it, so specify only those that you are confident about.
 
 \begin{itemize}
-	\item When implementing hooks that tell where in a function or where in the codebase a specific event happens, it is best to refactor code so that the event happens at one centralised place.
-		For example, when implementing \texttt{kern\_thread\_runnable} and \texttt{kern\_thread\_descheduling}, make sure there is one function in the entire kernel that is always called when adding or removing threads to the runqueue.
-		\footnote{This is especially relevant to \texttt{variable\_queue} users - using macros like \texttt{Q\_INSERT\_HEAD} in many places will cause you to have to instrument each place.
-Instead, write a one-line \texttt{add\_to\_runqueue} helper function that invokes the macro.}
-		\footnote{Also make sure when instrumenting helper functions such as these that they are {\em not} marked \texttt{static}!
-When using \texttt{gcc -O1} to build the kernel, \texttt{static} functions may get optimised in ways that violate standard calling convention, which Landslide relies on.}
-	\item Landslide's most reliable found-a-bug check is checking for \texttt{panic} getting called.
-		The more important invariants for which you have \texttt{assert}s in your kernel, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
+	\item \texttt{sched\_func} - When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them.
+Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.
+	\item \texttt{ignore\_sym} - You may wish to ignore memory accesses from other global data structures as well.
+For example, if you have a mutex which is locked very frequently, but irrelevant to the test case you're actually running, ignoring that mutex results in an independence relation that better reflects what you actually care about.
+	\item \texttt{within\_function} - If you configure decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than you need.
+(For example, if you're testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.) This configuration option allows you to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.
+	\item \texttt{without\_function} - Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
 \end{itemize}
 
-\subsection{Userspace Test Cases}
+\subsubsection{Behaviour Tweaks}
+\begin{itemize}
+	\item \texttt{BUG\_ON\_THREADS\_WEDGED} - Landslide's ``all threads wedged'' detection (i.e., nobody runnable, not necessarily mutex-based deadlock) is not perfect.
+The default is to let the kernel keep running, but if it seems like you have an all-threads-wedged bug, turn this on.
+	\item \texttt{EXPLORE\_BACKWARDS} - In which order to explore the branches?
+Backwards means with more preemptions first; forwards means tending to let the current thread keep running more often.
+Backwards tends to find bugs more quickly, but produces longer decision traces.
+	\item \texttt{DECISION\_INFO\_ONLY} - This option makes Landslide stop after one branch, and output the list of decision points it identified using the current config.
+Useful if you want to see all decision points, and tweak settings to get rid of frivolous ones.
+	\item \texttt{BREAK\_ON\_BUG} - If you want to see how long Landslide takes to explore, this should be off (i.e., Landslide will make simics quit when it's finished).
+If you want to use the simics debug prompt when a bug is found, turn this on.
+\end{itemize}
+
+\subsection{In-Landslide Functions}
+\label{sec:student}
+
+There are two functions that you need to implement in Landslide itself, to express certain kernel designs that might be more complicated than a simple true/false condition or integer.
+They are in \\ \texttt{work/modules/landslide/student.c}.
+
+You may need to read some of your kernel's global variables (pointers or state flags) to implement these functions.
+To get access to these, use \texttt{extra\_sym} in \texttt{config.landslide} (which is explained in the comments there).
+
+\begin{itemize}
+	\item \texttt{bool kern\_current\_extra\_runnable(conf\_object\_t *cpu)} - See section~\ref{sec:runqueue}.
+This says whether the current thread is ``logically runnable'' despite not being on the runqueue itself (that is, \texttt{tell\_landslide\_on\_rq} won't have been called for it).
+	\item \texttt{bool kern\_ready\_for\_timer\_interrupt(conf\_object\_t *cpu)} - This function expresses whenever a clock interrupt will trigger a prompt context switch.
+Landslide will avoid trying to preempt your kernel whenever this is false.
+\end{itemize}
+The codebase also provides the following macros to help you implement these. The \texttt{cpu} argument to the above functions is an opaque object which these macros need.
+\begin{itemize}
+	\item \texttt{GET\_CPU\_ATTR(cpu, name)} - returns the value of the cpu's specified register. (e.g., write \texttt{eax} for \texttt{name}).
+	\item \texttt{READ\_MEMORY(cpu, addr)} - retrieves the 4-byte value at the given address.
+	\item \texttt{READ\_BYTE(cpu, addr)} - retrieves the 1-byte value at the given address.
+	\item \texttt{GET\_ESP0(cpu)} - retrieves the value of the pseudo-register \texttt{\%esp0}.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Userspace Test Cases}
 \label{sec:testcase}
 
-We ship Landslide with a suite of small test cases designed to expose several common bugs that students frequently encounter during 15-410 Project 3.
+We ship Landslide with a suite of small test cases designed to expose several common races that students frequently encounter during 15-410 Project 3.\footnote{You may wonder why some of these are written in assembly.
+This is because using your userspace thread library for thread creation (a) is not guaranteed to work, (b) might forcibly serialise certain operations that would expose kernel bugs only if concurrent, and (c) necessitates many more system calls, which dramatically increases the size of the decision tree.}
 
@@
 	\item \texttt{fork\_wait} - Tests basic interaction of \texttt{fork()} and \texttt{wait()}. (This test is from the P3 hurdle suite.)
-	\item \texttt{wild\_test2} - Tests simultaneous thread death by unnatural causes. (This test is an upgrade of \texttt{wild\_test1} from the P3 hurdle suite.)
+	%\item \texttt{wild\_test2} - Tests simultaneous thread death by unnatural causes. (This test is an upgrade of \texttt{wild\_test1} from the P3 hurdle suite.)
 	\item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child. (new)
-	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in the same process vanishing. (new)
+	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in one process vanishing.
+(new)\footnote{This test tests for the use-after-free bug presented in lecture - if your \texttt{thread\_fork} implementation has ``\texttt{return child->tid;}'' as its last line of code.}
+	\item \texttt{yield\_vanish} - Tests for interactions between \texttt{yield()} and \texttt{vanish()}. (new)
 \end{itemize}
 
-It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space
+It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space.
+They also run several, but not too many, threads at once, producing potentially interesting interleavings.
 
-\subsection{Configuring Choice Points}
-\label{sec:choice}
-% TODO
-
-\subsection{Interpreting Landslide's Results}
-% TODO
+To warm up, we recommend you test \texttt{double\_thread\_fork} (and edit your kernel explicitly to introduce the associated bug - see the footnote in the test's description above).
+After that, we recommend trying \texttt{vanish\_vanish}, which is the most likely to expose many different types of bugs in many different kernel designs.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Configuring Decision Points}
+\label{sec:choice}
+
+Getting a good set of decision points is important to being able both to explore reasonably quickly and to produce meaningful interleavings that are likely to find bugs.
+After you get Landslide working with the default set of decision points, it's time to add some more.
+
+In no particular order, you should be thinking about the following things when building the set of decision points.
+
+\begin{itemize}
+	\item Use \texttt{tell\_landslide\_decide()} to indicate choice points.
+Recommended places are at the start of \texttt{mutex\_lock}, and if that doesn't find bugs, at the end of \texttt{mutex\_unlock} as well.
+But use your own intuition, depending on what system call(s) you're testing.
+	\item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.
+	\item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of choice points, and figure out which ones you want to get rid of.
+	\item Use \texttt{within\_func} and \texttt{without\_func} to make Landslide only pay attention to parts of your kernel that you care about.
+\end{itemize}
+
+Example tip: If you \texttt{tell\_landslide\_decide()} in \texttt{mutex\_lock}, you could surround the call with an if statement that makes sure the mutex is not some global mutex you don't care about.
+You could also (or instead) make Landslide ignore it by using \texttt{without\_func} on whatever function takes the mutex you don't care about.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Interpreting Landslide's Results}
+
+Landslide prints a lot of information. Most of it is only useful for development, but you may find some of it interesting anyway.
+
+The most important thing that Landslide prints is the decision trace, when it finds a bug. Landslide prints the following information:
+
+\begin{itemize}
+	\item Before proclaiming ``A bug was found!'', there should be a message in bright red explaining what the bug is.
+	\item The trace itself consists of a list of the decision points that were encountered in the buggy interleaving.
+For each decision point, Landslide says what thread used to be running, what thread it chose to run next, and the stack trace of the old thread at the point it was switched away from.
+	\item At the end, there will be a stack trace for the current point in time, which will be where the bug was detected, and not a decision point itself.
+	\item Also at the end there will be some statistics about the exploration: how many decision points were visited, how many backtracks were taken (i.e., how many interleavings were explored), the average number of decision points per branch, etc.
+\end{itemize}
+
+We ship Landslide with a somewhat high verbosity setting.
+If you want it to be quieter, you can edit it in \texttt{work/modules/landslide/common.h} - \texttt{CHOICE} is the recommended minimum level (\texttt{INFO} is highest, \texttt{DEV} is default).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Feedback}
+\label{sec:feedback}
+
+Ben needs feedback about your experience using Landslide to help him present his thesis. He wants to find answers to the following questions:
+
+\begin{itemize}
+	\item What ways can Landslide be improved to make it a better race-finding tool?
+(This means in terms of the high-level process of finding useful decision points, less so in terms of nasty implementation details - the latter would be good for development and bug reports, but has little research value.)
+	\item Does using Landslide make you think differently about race conditions, in a way that makes you better at debugging them, even at times when you're not using Landslide?
+	\item Is Landslide effective at finding bugs, when properly configured? Also, is it effective at saying it found no bug when using a reasonably-thorough set of decision points?
+\end{itemize}
+
+Of course, you shouldn't try to answer those questions directly. But keep them in mind as the goal of the study, when answering the following questions.
+
+\subsection{While using Landslide}
+
+While you're using Landslide, please try to take 5 minutes approximately every half-hour to give short updates. For convenience, you can print out appendix~\ref{sec:30mins} and fill it in as you go.
+
+\subsection{After using Landslide}
+
+After you've finished with Landslide, please write some conclusions about your experience:
+
+\begin{itemize}
+	\item Were you able to get Landslide to work with your kernel (i.e., run through a minimal exploration tree completely)?
+If so, how long did it take, from sitting down to getting it to work?
+If not, did your kernel do something that was incompatible with Landslide, or were you unable to get the instrumentation right for some other reason?
+	\item Did you find bugs while using Landslide, that you imagine would have been very hard to find otherwise? Did Landslide's output help you understand what caused them?
+	\item Were you able to get Landslide to say ``you survived!'' with custom decision points?
+Did you think the set of decision points you used for this provide a strong guarantee about the absence of races?
+	\item Describe your experience configuring the set of decision points. What was intuitive, obvious to do? Did you feel stuck at any point, not knowing where to go next?
+	\item Was there anything you wanted to make Landslide do that it didn't support?
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\appendix
+\section{Step-by-Step Setup}
+\label{sec:stepbystep}
+
+We recommend you work on the Andrew Linux machines, as they are supported by the 410 environment, and Landslide has only been tested with Simics installation on AFS.
+
+As a way of keeping the attack plan steps small, we recommend not adding/configuring custom decision points until after getting Landslide working with the automatic minimal set of decision points (i.e., it explores the tree to completion).
+
+\begin{itemize}
+	\item Get the repository: \texttt{http://www.cs.cmu.edu/{\textasciitilde}410/landslide.tar.bz2}
+		\footnote{All directory paths henceforth are relative to the base directory of the repository.}
+	\item Read section~\ref{sec:requirements} to make sure your kernel is Landslide-compatible. (If not, fix it and/or ask Ben for help.)
+	\item Build your kernel for Landslide
+	\begin{itemize}
+		\item Run \texttt{prepare-workspace.sh}
+		\item Copy your kernel sources/build directory into \texttt{pebsim/}.
+		\item Add \texttt{tell\_landslide.c} and \texttt{tell\_landslide.h} to your kernel, and add \texttt{tell\_landslide.o} to your kernel's \texttt{config.mk}.
+		\item Copy in the test program(s) you want to run, from \texttt{tests/}, and add them to your kernel's \texttt{config.mk}. (section~\ref{sec:testcase})
+		\item Add calls to the annotation functions in your kernel. (section~\ref{sec:tell})
+		\item Build your kernel, and put \texttt{bootfd.img} and \texttt{kernel} in \texttt{pebsim/}.
+	\end{itemize}
+	\item Fill out \texttt{config.landslide} in \texttt{pebsim/}. (section~\ref{sec:config})
+	\item Implement the functions in \texttt{student.c} in \texttt{work/modules/landslide}. (section~\ref{sec:student})
+	\item In \texttt{pebsim/}, run \texttt{./landslide}. This will build everything (if necessary), and then run Landslide.
+	\item Iterate reconfiguring Landslide until satisfied with the results. (section~\ref{sec:choice})
+	\item Fill out feedback. (section~\ref{sec:feedback})
+\end{itemize}
+
+% TODO A section for debuggng landslide problems
+
+\newpage
+\section{While-You-Work Feedback Form}
+\label{sec:30mins}
+
+Questions to be thinking about while you're working:
+\begin{itemize}
+	\item How much time did you spend on each phase of the process?
+(annotating your kernel, doing \texttt{config.landslide}, writing \texttt{student.c}, fixing instrumentation problems/crashes, customizing decision points for more efficient search, or trying to understand/solve a race that was found)
+	\item What problems were you fighting, getting things to work? (conceptual understanding, fixing instrumentation, finding good decision points)
+	\item Did you have to change your kernel at all to make it compatible?
+	\item Is there a feature you wish Landslide had to help you out, that it doesn't offer?
+	\item Did you realize anything interesting or surprising about Landslide, or about your kernel?
+\end{itemize}
+Please fill out a row after every 30-60 minutes spent. Approximate time estimates are totally fine.
+
+\addtolength{\oddsidemargin}{-.5in} 
+\newcommand\asdf[1]{\hspace{-0.05in}\small{#1}\hspace{-0.05in}}
+\begin{center}
+\begin{tabular}{|c||c|c|c|c|c|c||c|}
+	\hline
+	\multirow{2}{*}{Time} & \multicolumn{6}{|c||}{Minutes spent doing\dots} & Remarks \\
+	\cline{2-7}
+	& \asdf{annotate} & \asdf{config} & \asdf{student.c} & \asdf{fixing} & \asdf{customizing} & \asdf{found races} & \hspace{0.7in} (problems, realizations, features) \hspace{0.7in} \\
+	\hline \hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+	& & & & & & & \\
+	& & & & & & & \\
+	& & & & & & & \\
+	\hline
+\end{tabular}
+\end{center}
 

### 2012-04-24 thesis.tex
add some challenge text

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Look up 'nooks'? re kernel space.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% concurrency implementation itself
+
+\subsection{Causes of Concurrency}
+
+% TODO: cite dbug
+In user-space, a systematic exploration tool may use the underlying kernel to help control the concurrent behaviour of the system.
+Simple system call invocations can cause a particular thread to run at a particular time, or to block while another thread runs first.
+
+In kernel-space, however, the scheduler is part of the system being tested, and we can no longer always interrupt the execution of a test case to ask the scheduler to instantly start running a different thread.
+Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
+Also, a context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
+% TODO: why?
+don't overflow the stack, for one; for two, ignore runqueue conflicts
+Finally, the kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
+
+\subsection{Ad-hoc Thread Communication}
 % inter-thread communication is not idealised!
-% kernels are designed in different ways; what can we assume
+
+\subsection{Kernel Design}
+
+One challenge particular to the context of 15-410 is that the kernels Landslide must be able to test may all use slightly different implementations to achieve the same goals.
+For Landslide to be generally applicable to (almost) any kernel that students may submit, it must make some abstract assumptions about the kernel design which are compatible with many different implementations.
+
+\begin{itemize}
+% TODO
+	\item
+\end{itemize}
+
+% TODO: reference future work section on linux here
+In certain avenues of future work, remaining agnostic of most implementation details becomes less of an issue.
+For example, if we target Landslide to Linux, it is unlikely that core scheduler or interrupt-handling behaviour will change for as long as Landslide remains useful for finding bugs.
+In 15-410, however, repeated re-targetting is a necessity.
 
@@
 \subsubsection{Partial-Order Reduction}
+\label{sec:techniques-por}
 % talk about the all-ancestors vs induction thing
+% talk about ignoring runqueue accesses, etc
 

### 2012-04-24 acknowledgements.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
-
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
-
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
-
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
-
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-
-\subsection{On Simics}
-
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
-
-% Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
-\section{Conclusion}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Save this for last :)
-
-\end{document}

### 2012-04-24 challenges.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Challenges of Kernel-space}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
+\subsection{Causes of Concurrency}
 
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
+% TODO: cite dbug
+In user-space, a systematic exploration tool may use the underlying kernel to help control the concurrent behaviour of the system.
+Simple system call invocations can cause a particular thread to run at a particular time, or to block while another thread runs first.
 
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
+In kernel-space, however, the scheduler is part of the system being tested, and we can no longer always interrupt the execution of a test case to ask the scheduler to instantly start running a different thread.
+Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
+Also, a context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
+% TODO: why?
+don't overflow the stack, for one; for two, ignore runqueue conflicts
+Finally, the kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
 
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
+% TODO reference future work sections
+In future work, we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
 
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
+\subsection{Ad-hoc Thread Communication}
 % inter-thread communication is not idealised!
 
-\section{Design and Implementation}
-\subsection{Components of Landslide}
+\subsection{Kernel Design}
 
-\subsection{On Simics}
+One challenge particular to the context of 15-410 is that the kernels Landslide must be able to test may all use slightly different implementations to achieve the same goals.
+For Landslide to be generally applicable to (almost) any kernel that students may submit, it must make some abstract assumptions about the kernel design which are compatible with many different implementations.
 
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
+% TODO: reference future work section on linux here
+In certain avenues of future work, remaining agnostic of most implementation details becomes less of an issue.
+For example, if we target Landslide to Linux, it is unlikely that core scheduler or interrupt-handling behaviour will change for as long as Landslide remains useful for finding bugs.
+In 15-410, however, repeated re-targetting is a necessity.
 
-% Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
-\section{Conclusion}
-
-\section{Acknowledgements}
-% Save this for last :)
-
-\end{document}
+\begin{itemize}
+% TODO
+	\item
+\end{itemize}

### 2012-04-24 conclusion.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
-
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
-
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
-
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
-
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-
-\subsection{On Simics}
-
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
-
-% Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
-
-\section{Acknowledgements}
-% Save this for last :)
-
-\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-04-24 design.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Design and Implementation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Components of Landslide}
 
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
+\subsubsection{Kernel Instrumentation}
 
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
+\subsubsection{Scheduling}
 
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
+\subsubsection{The Arbiter}
 
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
+\subsubsection{The Decision Tree}
 
@@
 
+Landslide is implemented as a module for Simics\cite{simics}
+
 \subsection{Techniques}
-% use-after-free, partial order, etc
+
+\subsubsection{Thread Scheduling}
+
+\subsubsection{Identifying Bugs}
+
+% A healthy dose of snark. Perhaps cite the formally-proved haskell kernel here.
+Without a formal specification of the internals of the guest kernel's implementation, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
+We list the metrics here in order of reliability.
+\begin{enumerate}
+	\item Kernel panic bugs - 
+	\item Use-after-free bugs - 
+	\item Deadlock bugs - 
+	\item Memory leak bugs - % this could fail if..
+	\item Infinite loop bugs - f
+\end{enumerate}
+Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
+If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed above).
+
+\subsubsection{Partial-Order Reduction}
+\label{sec:techniques-por}
 % talk about the all-ancestors vs induction thing
-% how to even schedule threads
+% talk about ignoring runqueue accesses, etc
+
+\subsubsection{Selective Ignorance}
+
+\subsection{Debugging Feedback}
+
+In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context-switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
+When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
+
+The decision trace explains each preemption or voluntary reschedule in the interleaving: which thread used to be running, which thread was chosen to run instead, and the stack trace of the former thread at the point from which it was switched away.
+
+% TODO: this sentence lies somewhat
+Additional information, such as the number of instructions executed during each transition, and the shared memory accesses that conflicted with other transitions, can also be printed at the user's configuration options.
+% Should it also print conflicting memory accesses? Configurable by the user?
+% Make the user able to replay the choices.
+
+% Discuss combining this with a choice trace minimisation tool. Reference future work.
 
 % Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
-\section{Conclusion}
-
-\section{Acknowledgements}
-% Save this for last :)
-
-\end{document}

### 2012-04-24 evaluation.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
-
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
-
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
-
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
-
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-
-\subsection{On Simics}
-
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
-
-% Have a section or subsection to talk about the tool-user relationship.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Evaluation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % 410 - talk about current state/methods (for students and for TAS)
@@
 
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
-\section{Conclusion}
-
-\section{Acknowledgements}
-% Save this for last :)
-
-\end{document}
+% vim: ft=tex

### 2012-04-24 futurework.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:future}
 
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
-
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
-
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
-
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-
-\subsection{On Simics}
-
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
-
-% Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
+\subsection{New Techniques}
+\label{sec:future-new}
 % new techniques for landslide:
@@
 % - ICB as in chess
-% - garth's shaping
+% - hybrid with symbolic execution
+
+\subsection{Education}
+\label{sec:future-education}
+% teaching tool, study thinking patterns
+% 410 stuff
+
+\subsection{Production Kernels}
+\label{sec:future-linux}
 % linux, device drivers
+% embedded microcontroller operating systems, simulated
+
+\subsection{Performance}
+\label{sec:future-perf}
 % VM instead of simics
@@
 % - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
 
-\section{Conclusion}
+\subsection{Long-Running Testing Approaches}
+\label{sec:future-shaping}
+% - garth's shaping
 
-\section{Acknowledgements}
-% Save this for last :)
+\subsection{Theoretical Oddities}
+\label{sec:future-theory}
+% theory things to study:
+% - study the implications of backwards vs forward exploration
+% - nadim's backtrack depth
 
-\end{document}
+% vim: ft=tex

### 2012-04-24 related.tex
separate into more tex files

--- 2012-03-09 thesis.tex
@@
-\documentclass[twocolumn]{article}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{upgreek}
-\usepackage{epsfig}
-\usepackage[bottom]{footmisc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{document}
-\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\textbf{bblum})}
-\maketitle
-
-\newcommand\true{\;\textit{true}}
-\newcommand\false{\;\textit{false}}
-
-\newcommand\alpher\alpha
-\newcommand\beter\beta
-\newcommand\gammer\gamma
-\newcommand\delter\delta
-\newcommand\zeter\zeta
-\newcommand\Sigmer\Sigma
-
-\newcommand\NN{\mathbb{N}}
-\newcommand\QQ{\mathbb{Q}}
-\newcommand\RR{\mathbb{R}}
-\newcommand\ZZ{\mathbb{Z}}
-
-\begin{abstract}
-
-\end{abstract}
-
-\section{Introduction}
-
-Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
-In kernel-space, race condition debugging becomes even more difficult, as many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
-
-% Write here a paragraph about userspace techniques. Write about kernel races.
-
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, which students in Operating System Design and Implementation (15-410) at CMU implement, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
-
-% ``The contributions of this work are as follows.''
-In this work, we present Landslide, a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-\section{Related Work}
-% Look up 'nooks'? re kernel space.
-
-\section{Challenges of Kernel-space}
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-
-\section{Design and Implementation}
-\subsection{Components of Landslide}
-
-\subsection{On Simics}
-
-\subsection{Techniques}
-% use-after-free, partial order, etc
-% talk about the all-ancestors vs induction thing
-% how to even schedule threads
-
-% Have a section or subsection to talk about the tool-user relationship.
-
-\section{Evaluation}
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-\section{Future Work}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-
-\section{Conclusion}
-
-\section{Acknowledgements}
-% Save this for last :)
-
-\end{document}
+% vim: ft=tex

### 2012-04-24 thesis.tex
separate into more tex files

--- 2012-04-23 thesis.tex
@@
 \title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\texttt{bblum@andrew.cmu.edu})}
+\author{Benjamin Blum (\texttt{bblum@andrew.cmu.edu}) \and Garth Gibson (\texttt{garth@cs.cmu.edu})}
 \maketitle
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Look up 'nooks'? re kernel space.
+\input{related}
+\input{challenges}
+\input{design}
+\input{evaluation}
+\input{futurework}
+\input{acknowledgements}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Challenges of Kernel-space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% concurrency implementation itself
-% inter-thread communication is not idealised!
-% kernels are designed in different ways; what can we assume
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Design and Implementation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Components of Landslide}
-
-\subsubsection{Kernel Instrumentation}
-
-\subsubsection{Scheduling}
-
-\subsubsection{The Arbiter}
-
-\subsubsection{The Decision Tree}
-
-\subsection{On Simics}
-
-Landslide is implemented as a module for Simics\cite{simics}
-
-\subsection{Techniques}
-
-\subsubsection{Thread Scheduling}
-
-\subsubsection{Identifying Bugs}
-
-% A healthy dose of snark. Perhaps cite the formally-proved haskell kernel here.
-Without a formal specification of the internals of the guest kernel's implementation, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
-In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
-We list the metrics here in order of reliability.
-\begin{enumerate}
-	\item Kernel panic bugs - 
-	\item Use-after-free bugs - 
-	\item Deadlock bugs - 
-	\item Memory leak bugs - % this could fail if..
-	\item Infinite loop bugs - f
-\end{enumerate}
-Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
-If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed above).
-
-\subsubsection{Partial-Order Reduction}
-% talk about the all-ancestors vs induction thing
-
-\subsubsection{Selective Ignorance}
-
-\subsection{Debugging Feedback}
-
-In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context-switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
-When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
-
-The decision trace explains each preemption or voluntary reschedule in the interleaving: which thread used to be running, which thread was chosen to run instead, and the stack trace of the former thread at the point from which it was switched away.
-
-Landslide is impleAdditional information, such as the number of instructions executed during each transition, and the shared memory accesses that conflicted with other transitions, can also be printed at the user's configuration options.
-% Should it also print conflicting memory accesses? Configurable by the user?
-% Make the user able to replay the choices.
-
-% Discuss combining this with a choice trace minimisation tool. Reference future work.
-
-% Have a section or subsection to talk about the tool-user relationship.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Evaluation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - garth's shaping
-% linux, device drivers
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
-% embedded microcontroller operating systems, simulated
-% teaching tool, study thinking patterns
-% symbolic execution
-% theory things to study:
-% - prove the soundness of the every-evil-ancestor thing
-% - study the implications of backwards vs forward exploration
-% - nadim's backtrack depth
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Save this for last :)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{citations}{}

### 2012-04-24 challenges.tex
write more design

@@
 
-% TODO reference future work sections
-In future work, we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
+In future work (Section~\ref{sec:future-linux}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
 
@@
 % inter-thread communication is not idealised!
+% TODO
 
@@
 
-% TODO: reference future work section on linux here
-In certain avenues of future work, remaining agnostic of most implementation details becomes less of an issue.
+In certain avenues of future work (Section~\ref{sec:future-linux}), remaining agnostic of most implementation details becomes less of an issue.
 For example, if we target Landslide to Linux, it is unlikely that core scheduler or interrupt-handling behaviour will change for as long as Landslide remains useful for finding bugs.
@@
 
+We built Landslide to be compatible with multiple designs for the following major scheduling behaviours:
+
 \begin{itemize}
-% TODO
-	\item
+	\item {\bf Runqueues}: Does the kernel store the currently-running thread on the runqueue, or is it removed and stored separately?
+	\item {\bf Mutexes}: When a thread blocks on a mutex, is it left on the runqueue in a yield-loop, or is it explicitly descheduled?
+In yield-looping mutexes, when do blocked threads become ``unblocked'', notionally?
+(This might happen before the blocked thread runs next.)
+	\item {\bf Idling}: Does the kernel have an explicit idle thread, or is there an idle loop that runs on the stack of whatever thread was last running?
+Do explicit idle threads run their idle loop in userspace or in kernelspace?
+	\item {\bf Thread Creation}: When a thread is newly forked, is it placed onto the runqueue for later, or is it context-switched to immediately?
+Do just-forked threads begin life through the usual context-switch-return path, or is there a special path for that?
 \end{itemize}
+
+Conversely, Landslide needs to make some assumptions about scheduling behaviour, in order for its control over scheduling to work:
+
+\begin{itemize}
+	\item With the exception of a non-preemptible ``scheduler lock'', and yield-looping mutexes (both of which must be instrumented by the user), a thread's presence on the runqueue indicates that a finite number of timer interrupts in succession will eventually cause it to run.
+	\item The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
+Landslide uses the idle loop to detect when a test begins/ends and to detect when all threads in a test are wedged.
+\end{itemize}

### 2012-04-24 design.tex
write more design

@@
 
+Here we briefly describe each major component of Landslide. % TODO: provide a figure landslide.png
+
 \subsubsection{Kernel Instrumentation}
+% TODO
 
@@
 
+In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+% TODO
+
 \subsubsection{The Arbiter}
 
-\subsubsection{The Decision Tree}
+The arbiter identifies points during execution should count as decision points.
+The selection is mainly controlled by the user, during the annotation and configuration process, but it also automatically identifies {\em voluntary reschedules}, which comprise the ``minimal necessary set'' of decision points.
+
+A voluntary reschedule is identified when the previously-running thread entered the context switcher without first entering the timer interrupt handler, and any one of the following:
+\begin{itemize}
+	\item The currently-running thread was handling a timer interrupt, and exits the {\em timer interrupt handler}.
+	\item The currently-running thread was not handling a timer interrupt, and exits the {\em context switcher}.
+	\item The currently-running thread was {\em just forked} (i.e, it has never run before), and hence may never exit either the timer handler or context switcher.
+\end{itemize}
+
+It is always necessary to identify a decision point during a voluntary reschedule, to maintain the invariant that each transition between decision points is comprised of the execution of only one thread.
+
+In future work (Section~\ref{sec:future-new}), the arbiter may also automatically identify extra decision points, such as on conflicting shared memory accesses.
+
+\subsubsection{The Explorer}
+
+The explorer maintains a representation of the current branch of the decision tree.
+It is responsible for checkpointing at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.
 
@@
 
-Landslide is implemented as a module for Simics\cite{simics}
+Landslide is implemented as a module for Simics\cite{simics}.
+When running the kernel, Simics calls into Landslide once every time the kernel executes an instruction or performs a memory read or write.
+Landslide uses this information, in conjunction with the user-provided instrumentation, to maintain its internal representations of the state of the guest kernel.
+
+Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the execution state at decision points during the test.
 
@@
 \subsubsection{Thread Scheduling}
+% TODO
 
@@
 
-% A healthy dose of snark. Perhaps cite the formally-proved haskell kernel here.
-Without a formal specification of the internals of the guest kernel's implementation, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+% A healthy dose of snark.
+Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
 In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
@@
 \begin{enumerate}
-	\item Kernel panic bugs - 
+	\item Kernel panic bugs - % TODO
 	\item Use-after-free bugs - 
@@
 % talk about ignoring runqueue accesses, etc
+% TODO
 
 \subsubsection{Selective Ignorance}
+% TODO
 
@@
 
-% TODO: this sentence lies somewhat
-Additional information, such as the number of instructions executed during each transition, and the shared memory accesses that conflicted with other transitions, can also be printed at the user's configuration options.
-% Should it also print conflicting memory accesses? Configurable by the user?
-% Make the user able to replay the choices.
+Landslide also prints the stack trace of the currently-running thread at the point where the bug was found, and (optionally) drops the user into the Simics debugging prompt.
+Depending on the nature of the bug found, Landslide also provides more detailed information:
 
-% Discuss combining this with a choice trace minimisation tool. Reference future work.
+\begin{enumerate}
+	\item If the kernel has panicked, Landslide prints the message used in the panic/assertion.
+	\item If a use-after-free bug is found, Landslide prints information about the most-recently-freed chunk containing the accessed address: the stack trace and TID both for when it was allocated and freed.
+	\item If deadlock is detected (currently only using the annotations for yielding mutexes), Landslide prints the cycle of TIDs that are blocked on each other.
+	\item If a memory leak is detected, Landslide prints 
+\end{enumerate}
 
-% Have a section or subsection to talk about the tool-user relationship.
+Landslide can also provide other useful information, even in cases where it did not find bugs, to help with the user's process of configuring decision points.
+Instead of exploring alternative interleavings, it can stop execution after the first interleaving and print out the set of decision points that were identified, along with the shared memory conflicts and happens-before relations for pairs of transitions between the decision points.
+
+% TODO: Discuss combining this with a choice trace minimisation tool. Reference future work.

### 2012-04-25 design.tex
add more design writing

@@
 
-In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
 % TODO
@@
 \subsubsection{Thread Scheduling}
-% TODO
+
+In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+
+Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
+% TODO: make an algorithm listing for the schedule-in-flight alg
 
@@
 
-\subsubsection{Partial-Order Reduction}
+\subsection{Partial-Order Reduction}
 \label{sec:techniques-por}
-% talk about the all-ancestors vs induction thing
-% talk about ignoring runqueue accesses, etc
+
+We make use of Dynamic Partial-Order Reduction, the state-space pruning algorithm presented in \cite{dpor}. Here we discuss the specifics of implementing DPOR in our environment.
+
+\subsubsection{Happens-Before Relation}
 % TODO
 
-\subsubsection{Selective Ignorance}
-% TODO
+\subsubsection{Memory Independence Relation}
+
+The independence relation expresses which transitions do not read-and-write or write-and-write to the same shared memory addresses.
+% talk about ignoring runqueue accesses, etc TODO
+
+\subsubsection{Coarse Granularity}
+% talk about the all-ancestors vs induction thing TODO
 
@@
 	\item If a use-after-free bug is found, Landslide prints information about the most-recently-freed chunk containing the accessed address: the stack trace and TID both for when it was allocated and freed.
-	\item If deadlock is detected (currently only using the annotations for yielding mutexes), Landslide prints the cycle of TIDs that are blocked on each other.
-	\item If a memory leak is detected, Landslide prints 
+	\item If deadlock is detected, Landslide prints the cycle of TIDs that are blocked on each other.\footnote{This currently only works when using the annotations for yielding mutexes, though it is not difficult to implement more generally.}
+	\item If a memory leak is suspected, Landslide prints how many bytes bigger the heap is after the test ended than when the test began.\footnote{Future feature: Printing when-allocated stack traces for each suspicious heap chunk.}
+	\item If an infinite loop is suspected, depending on which heuristic was triggered, Landslide prints either the number of instructions since the last decision point (and the previous average) or the current branch depth (and the previous average).
 \end{enumerate}

### 2012-04-25 design.tex
add some memory to design

@@
 % TODO
+
+\subsubsection{Memory Tracking}
+
+Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point what memory ranges are allocated and what ranges used to be allocated but now are freed.
+This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
+
+Landslide also maintains a set of memory accesses. % TODO
 
@@
 \begin{enumerate}
-	\item Kernel panic bugs - % TODO
-	\item Use-after-free bugs - 
+	\item Kernel panic bugs - If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
+	\item Use-after-free bugs - Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range.
+If not, Landslide proclaims the access to be illegal.
 	\item Deadlock bugs -
@@
 	\item Infinite loop bugs - f
+	% TODO
 \end{enumerate}
@@
 % talk about the all-ancestors vs induction thing TODO
+
+\subsection{Freed Memory Poisoning}
+
+A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
+Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
+
+Hence, even if the second thread that freed the block never accesses the particular address that the first thread used, the second thread's \texttt{free} still logically conflicts with the first thread's access.
+Landslide addresses this by treating every call to \texttt{free} as a {\em write} access to every address within the freed block when computing shared memory conflicts.
 

### 2012-04-25 challenges.tex
rephrase challenges a bit

@@
 In kernel-space, however, the scheduler is part of the system being tested, and we can no longer always interrupt the execution of a test case to ask the scheduler to instantly start running a different thread.
-Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
-Also, a context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
+\begin{itemize}
+	\item Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
+	\item A context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
 % TODO: why?
 don't overflow the stack, for one; for two, ignore runqueue conflicts
-Finally, the kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
+	\item The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
+\end{itemize}
 
+In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts.
 In future work (Section~\ref{sec:future-linux}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.

### 2012-04-25 design.tex
add inflight alg listing

@@
 Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
-% TODO: make an algorithm listing for the schedule-in-flight alg
+Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.
+
+One alternative, simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
+This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
+We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
+
+\begin{algorithm*}[t!]
+	\label{alg:inflight}
+	\begin{algorithmic}
+		\State {\em Global scheduler state.}
+		\State bool schedule\_in\_flight;
+		\State int target\_tid;
+		\\
+		\State {\em Per-thread state. (Updated elsewhere.)}
+		\State bool in\_timer\_handler, in\_context\_switch;
+		\State int current\_tid;
+		\\
+		\Function{scheduler\_update}{int eip}
+			\If{schedule\_in\_flight}
+				\State {\sc assert}(in\_timer\_handler || in\_context\_switch);
+				\If { {\sc kernel\_exiting\_timer}(eip) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(eip))}
+					\State {\em Then the kernel has just finished rescheduling and is about to resume normal thread execution.}
+					\If {current\_tid != target\_tid}
+						\State {\em The kernel switched to an undesirable thread. Keep the schedule operation ``in-flight''.}
+						\State {\sc cause\_timer\_interrupt}();
+					\Else
+						\State {\em The in-flight schedule is ``landing''.}
+						\State schedule\_in\_flight $\gets$ false;
+					\EndIf
+				\EndIf
+			\ElsIf{{\sc need\_to\_preempt}()}
+				\State target\_tid $\gets$ {\sc choose\_new\_thread}();
+				\State schedule\_in\_flight $\gets$ true;
+				\State {\sc cause\_timer\_interrupt}();
+			\EndIf
+		\EndFunction
+	\end{algorithmic}
+	\caption{The scheduling algorithm.
+This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{eip} each time.
+The predicates on \texttt{eip} are part of the kernel instrumentation which tell what the guest kernel is doing at the current instruction.}
+\end{algorithm*}
 

### 2012-04-26 design.tex
update design

@@
 \subsubsection{Scheduling}
+\label{sec:components-sched}
 
@@
 \subsection{Techniques}
-
 \subsubsection{Thread Scheduling}
@@
 In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
-We list the metrics here in order of reliability.
+
+Certain metrics:
 \begin{enumerate}
-	\item Kernel panic bugs - If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
-	\item Use-after-free bugs - Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range.
-If not, Landslide proclaims the access to be illegal.
-	\item Deadlock bugs -
-	\item Memory leak bugs - % this could fail if..
-	\item Infinite loop bugs - f
-	% TODO
+	\item {\bf Kernel panic bugs}: If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
+	\item {\bf Use-after-free bugs}: Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range.
+If not, Landslide proclaims the access to be illegal.\footnote{In the same way, Landslide also detects calls to \texttt{free} on blocks that were already freed or never allocated at all.}
+	\item {\bf Deadlock bugs}: If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
+\end{enumerate}
+Uncertain metrics:
+\begin{enumerate}
+	\item {\bf Memory leak bugs}: Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
+If memory allocated during the test was not freed, Landslide assumes that it was leaked.
+		(Some kernel designs may legitimately behave this way, so this bug-check may be disabled when testing such kernels.)\footnote{There is, of course, much room for improvement in this metric, but it is not part of the research contribution.}
+	\item {\bf Infinite loop bugs}: Landslide judges whether the kernel has entered an infinite loop by comparing the current state of the decision tree to past executions of the same test case.
+	\begin{itemize}
+		\item While exploring the decision tree, Landslide computes the average number of instructions executed between two consecutive decision points.
+If at any point the current number of instructions executed since the most recent decision point exceeds this average times a constant factor (arbitrarily chosen to be 2000), Landslide assumes the kernel must have gotten stuck in an infinite loop.
+		\item Landslide also computes the average number of decision points in each branch of the decision tree (the average ``branch depth'').
+If the depth of the current branch ever exceeds this average times a constant factor (arbitrarily chosen to be 20), Landslide assumes the kernel must have gotten stuck in an infinite loop.
+	\end{itemize}
 \end{enumerate}

### 2012-04-26 design.tex
update design

@@
 \subsubsection{Kernel Instrumentation}
+\label{sec:components-kern}
 % TODO
@@
 
-Landslide also maintains a set of memory accesses. % TODO
+Landslide also maintains a set of shared memory accesses made since the last decision point.
+Whenever the guest kernel accesses memory in its heap or in its global data regions, Landslide adds the address of the access to the set, with a flag indicating whether it was a read or a write.
+(If the address was already present, and the recorded access was a read and the current access is a write, we upgrade the recorded access to a write.
+Otherwise if the address was already present, we do nothing.)
+
+Landslide ignores shared memory accesses from the kernel's dynamic allocator itself, and it also ignores shared memory accesses from the components of the kernel's scheduler which run every transition (Section~\ref{sec:por-independence}).
+
+When Landslide reaches a decision point, this accumulated set is copied in the decision tree (Section~\ref{sec:components-explore}), and reset to empty before continuing execution.
 
@@
 It is responsible for checkpointing at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.
+
+At each decision point, Landslide creates a new node in the decision tree.
+It stores the TID of the thread that was chosen, the state of the scheduler (what threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.
 
@@
 				\If { {\sc kernel\_exiting\_timer}(eip) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(eip))}
-					\State {\em Then the kernel has just finished rescheduling and is about to resume normal thread execution.}
+					\State {\em The kernel has just finished rescheduling and is about to resume normal thread execution.}
 					\If {current\_tid != target\_tid}
@@
 This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{eip} each time.
-The predicates on \texttt{eip} are part of the kernel instrumentation which tell what the guest kernel is doing at the current instruction.}
+The predicates on \texttt{eip} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
 \end{algorithm*}
@@
 \subsubsection{Memory Independence Relation}
-
+\label{sec:por-independence}
 The independence relation expresses which transitions do not read-and-write or write-and-write to the same shared memory addresses.
@@
 \subsection{Debugging Feedback}
+
+% TODO: provide a figure
 
@@
 Instead of exploring alternative interleavings, it can stop execution after the first interleaving and print out the set of decision points that were identified, along with the shared memory conflicts and happens-before relations for pairs of transitions between the decision points.
-
-% TODO: Discuss combining this with a choice trace minimisation tool. Reference future work.

### 2012-04-26 related.tex
update design
add related

@@
 
+Systematic exploration is a relatively young approach to concurrency verification.
+Recent work has focused on theoretical aspects of the technique, on distributed systems, and in userspace in general.
+Other efforts have focused on kernel verification apart from systematic exploration, using techniques such as data race detection and formal model verification.
+
+Landslide distinguishes itself from related systematic testing tools in its field as more of an exploratory work, presenting techniques to make systematic testing compatible with kernel-level code; less so as a development on the core testing approach itself.
+
+\subsection{Systematic Exploration}
+
+Dynamic Partial Order Reduction (DPOR) \cite{dpor,sdpor,dbug-retreat} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
+Landslide makes generous use of the DPOR algorithm.
+
+dBug\cite{dbug-ssv} is a systematic testing tool which makes use of DPOR.
+It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \texttt{libc} library calls.
+It uses a message-passing model for inter-thread communication and for establishing a happens-before relationship between transitions.
+The dBug project heavily inspired the development of Landslide.
+
+CHESS\cite{chess} is another userspace systematic testing tool.
+They explore the insight that many races require very few forced preemptions to uncover, and develop a search strategy which prioritises thread interleavings with fewer preemptions.
+Landslide does not (yet) make use of this strategy, though we did find a related insight which we discuss in Section~\ref{sec:future-backwards}.
+
+DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
+Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) - it provides the user with options to constrain decision point identification to certain subsets of the kernel.
+
+\subsection{Kernel-level Testing}
+
+% TODO
+
+%\subsection{Orthogonal Testing Techniques}
+% TODO: talk about DMT systems?
+
 % vim: ft=tex

### 2012-04-27 related.tex
a little more related
more related: orthogonal

@@
 Recent work has focused on theoretical aspects of the technique, on distributed systems, and in userspace in general.
-Other efforts have focused on kernel verification apart from systematic exploration, using techniques such as data race detection and formal model verification.
+Other work has focused on kernel verification apart from systematic exploration, using techniques such as data race detection and formal model verification.
+Still other efforts are for testing and verification techniques that are orthogonal to systematic exploration and kernel-level testing entirely.
+Here we discuss related work in all three categories.
 
@@
 
+There are also several tools, such as MaceMC\cite{macemc} and MODIST\cite{modist}, which provide systematic testing frameworks for networked and distributed applications.
+
 DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
@@
 
-\subsection{Kernel-level Testing}
+\subsection{Kernel-level Verification}
 
-% TODO
+\cite{racepro}
+\cite{sel4}
+\cite{datacollider}
+\cite{carburizer}
 
-%\subsection{Orthogonal Testing Techniques}
-% TODO: talk about DMT systems?
+\subsection{Orthogonal Testing Techniques}
+
+Research in dynamic verification has also seen other techniques apart from systematic exploration.
+
+Symbolic execution\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpretating symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
+Tools such as KLEE\cite{klee} and projects such as Automated Exploit Generation\cite{aeg} demonstrate the effectiveness of symbolic execution in dynamic verification.
+We discuss the possibility of integrating symbolic execution into a testing framework such as Landslide in Section~\ref{sec:future-new}.
+
+Deterministic multithreading is a different technique for dealing with concurrency bugs which focuses on avoiding them when running production code rather than trying to expose them during testing.
+Deterministic multithreading tools, such as Kendo\cite{kendo}, PEREGRINE\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
+Deterministic multithreading serves a different purpose than systematic exploration; it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.
 

### 2012-04-27 related.tex
text-complete related work

@@
 
-DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
+RacePRO\cite{racepro} is a tool based in kernel-space which uncovers ``process races'' using systematic exploration of multi-process interactions in user-space.
+They use system calls, such as filesystem accesses, as points to explore different interleavings of concurrent code, and detect bugs which can corrupt persistent system resources.
+
+Finally, DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
 Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) - it provides the user with options to constrain decision point identification to certain subsets of the kernel.
@@
 
-\cite{racepro}
-\cite{sel4}
-\cite{datacollider}
-\cite{carburizer}
+DataCollider\cite{datacollider} is a tool for detecting data races, a special kind of concurrency bug which involve concurrent unprotected accesses to memory that can interleave unsafely on a fine-grained level.
+DataCollider uses memory access sampling, and remains largely agnostic of synchronisation operations, to find data races in the Windows 7 kernel.
+In Section~\ref{sec:future-new} we discuss how systematic execution might be enhanced by integrating data race detection techniques.
+
+Carburizer\cite{carburizer} is an effort to make imperfect kernel-level drivers work more reliably in the presence of faulty peripheral devices.
+They use static analysis to identify variables ``tainted'' by device input and unsafe uses thereof.
+Carburizer is able to transform such badly-written code to be more robust.
+Kernel device drivers are a ripe breeding ground for bugs, because it is frequently written by less careful developers, and the sheer volume of code makes it difficult to rigorously review.
+Due to the interrupt-driven nature of many drivers, they are also prone to concurrency bugs, which Carburizer does not address.
+In Section~\ref{sec:future-linux} we discuss the potential applicability of systematic exploration to concurrency bugs in kernel device drivers.
+
+On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4\cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
+Because Haskell is type-safe (strict constraints on legal program behaviour that make it easier to reason about) and pure (most code cannot have stateful side-effects), the developers of seL4 were able to formally prove the specification's correctness and the implementation's adherence to the spec.
+Developing kernel code in languages and environments that enable formal complete verification, even in small components of a larger codebase, represents an admirable step towards writing correct code to begin with and avoiding many classes of bugs entirely.
+However, such steps have little bearing on code that exists in the world today, and it is also important to find ways to verify such sloppily-written code that already has bugs.
 
@@
 Deterministic multithreading is a different technique for dealing with concurrency bugs which focuses on avoiding them when running production code rather than trying to expose them during testing.
-Deterministic multithreading tools, such as Kendo\cite{kendo}, PEREGRINE\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
+Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
 Deterministic multithreading serves a different purpose than systematic exploration; it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.

### 2012-04-27 terminology.tex
add terminology

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Terminology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{sec:terminology}

In this section we define certain terms that have specific meanings in the context of Landslide.

Basic terminology:
\begin{enumerate}
	\item {\bf Guest kernel}:
		The kernel which is being tested for concurrency bugs by Landslide.
	\item {\bf Test case}:
		Abstractly, the set of inputs under which the guest kernel is tested. Practically, a user-space program which runs on top of the guest kernel to execute a particular set of system calls, while Landslide performs systematic exploration
\suspend{enumerate}

Scheduling terminology:
\resume{enumerate}
	\item {\bf Thread}:
		One participating agent in a concurrent program, which executes code sequentially, with the potential to be interleaved with the execution of other threads in the system.
		Each thread has a unique numeric identifier (TID).
	\item {\bf Involuntary preemption}:
		A context switch from one thread to another caused by a nondeterministic event, such as a timer tick or device interrupt. The systematic testing framework must control all sources of nondeterminism. In this work, we focus only on timer-driven thread switches.
	\item {\bf Voluntary reschedule}:
		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.\footnote{Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not). Involuntary preemptions are never necessary for correct kernel behaviour.}
		See Section~\ref{sec:components-arbiter}.
\suspend{enumerate}

Systematic exploration terminology:
\resume{enumerate}
	\item {\bf Decision point} (or {\bf Choice point}):
		A point during execution (between two consecutive instructions, precisely) which is deemed ``interesting'' in terms of the likelihood that thread switches at that point will cause concurrency bugs to arise. All thread switches occur at a decision point; some are voluntary reschedules caused by the kernel and others are involuntary preemptions caused by Landslide.
	\item {\bf Transition}:
		A sequence of instructions executed by a single thread between two decision points. At a given decision point, Landslide chooses which thread will run (and causes it to do so), and that thread executes a transition, after which point Landslide will have identified a subsequent decision point
	\item {\bf Decision tree} (or {\bf Execution tree}):
		The tree defined by the possibility at each decision point of causing any runnable thread to execute its own transition. Abstractly, this tree comprises all possible states reachable by any interleaving of threads.
	\item {\bf Branch}:
		One particular execution of the test case; a set of decision points with exactly one transition in between each of them, characterising a single interleaving of threads.
	\item {\bf Interleaving}:
		A less precise / more abstract term for a branch, or for a subset of transitions which make up a branch.
	\item {\bf Bug}: An execution state which Landslide identifies as incorrect, which is indicative of illegal behaviour. Each branch of the tree may or may not have a bug in it. See Section~\ref{sec:techniques-bugs}.
	\item {\bf Backtracking}:
		Performed at the end of each branch of the tree, when the test case has finished running. Landslide identifies a decision point from the current branch at which it wants to have chosen a different thread to run, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
	\item {\bf Decision trace}:
		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause. Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace may also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
\end{enumerate}

% vim: ft=tex

### 2012-04-28 futurework.tex
write more design
add inflight alg listing
a little more related
add a bunch of future work

@@
 
-\subsection{New Techniques}
-\label{sec:future-new}
-% new techniques for landslide:
-% - hybrid with data-race/lockset/static-analysis; shm choices
-% - parallelism
-% - ICB as in chess
-% - hybrid with symbolic execution
+As Landslide is an experimental foray into the world of systematic exploration in kernel-space, it has opened up many avenues for potential future improvements.
+
+\subsection{Interface Improvements}
+\label{sec:future-interface}
+
+Through our evaluation, we found many ways in which Landslide's user interface is lacking.
+These improvements would not be research developments, but would enhance the user experience in ways necessary for continuing work in the context of 15-410.
+
+\begin{itemize}
+	\item Landslide currently does not support ability to replay a particular choice sequence.
+If the user finds a bug, and wants to re-execute the interleaving that led to it, they have no choice but to re-explore the tree.
+	\item Landslide currently does not support being interrupted, so the user can use the Simics debug prompt arbitrarily and resume Landslide when ready.
+This is because the current implementation uses a wrapper script around the Simics prompt which Landslide communicates with to perform backtracking; it would be difficult to redesign, but could also be worked around.
+	\item Landslide could print its decision traces in a better format.
+The current format is basically a ``raw data dump''.
+Raw text is probably the wrong interface; a graphical/tabular format with distinct columns for the execution of each different thread would be easier to understand.
+\end{itemize}
 
@@
 % 410 stuff
+% TODO
+
+15-410 currently teaches students concurrency debugging skills ``the hard way'': by immersing them in environments where races will arise, and letting them find debugging tactics to use in conjunction with conventional stress testing on their own.
+
+\subsection{New Techniques}
+\label{sec:future-new}
+
+Related research in dynamic verification has introduced testing techniques orthogonal to systematic exploration, which could augment its effectiveness if combined in one tool.
+We discuss the potential for using other techniques in Landslide or in any other systematic testing framework.
+
+\subsubsection{Data Race Detection and Static Analysis}
+
+Landslide currently never alters its set of decision points once it begins executing.
+The user must configure the set of decision points in advance, and Landslide follows them to the letter when exploring the tree.
+This is useful for investigating the size of trees generated by certain sets of decision points, but future work should do better than leaving it up to the user to stumble across a set of decision points that exposes a bug.
+
+Most notably among the information Landslide has at its disposal is the collection of conflicting shared memory accesses among transitions.
+Theoretically, to identify a decision point immediately before and after each such access would generate an execution tree with perfect granularity (i.e., exactly one shared memory access per transition), and thence find every possible race condition - this is, of course, the opposite extreme to the current setup.
+
+Landslide could make use of data race detection techniques\cite{datacollider}, however, to strike a middle ground in which shared memory accesses it chooses.
+First, it should be aware of the types and interfaces associated with the kernel's synchronisation primitives (mutexes, condvars, semaphores, etc), and be able to treat operations on those as ``guaranteed to work'' (similar to Section~\ref{sec:por-independence}).
+Next, it should recognise acquire and release (or sleep and wake) operations on synchronisation primitives, and use that information to track lockset-type information, and use that to identify which memory accesses are raciest, and add decision points around those.
+
+There may also exist static analysis techniques that would identify the most "at-risk" memory accesses, which could also be used in this manner.
+
+Next is the question of when during the testing process these decision points should be introduced:
+
+\begin{itemize}
+	\item One possibility would be to begin exploration with a small set of decision points, explore the resulting tree, analyse the tree post-hoc to identify additional decision points, and iterate exploring with the new set.
+	\item Another possibility would be to identify decision points along the way, analysing each branch of the tree after executing it, to generate the decision points for that very branch, and thence use them to find which branch to explore next.
+\end{itemize}
+
+With either method, in order to mitigate exponential explosion of tree size, it would probably be best to only allow the decision point set size to increase by one each iteration.
+An iteration framework could explore using different combinations/subsets of such, using a workqueue model, and using heuristics to identify which decision points are most useful and should be used first.
+
+\subsubsection{Parallelism}
+
+Landslide's tree exploration is implemented sequentially.
+However, because DPOR's approach of tagging which sibling branches should be explored next generally follows a workqueue-based depth-first-search structure, it should not be too difficult to parallelise.
+Prior work exists for this technique\cite{distributed-dpor}, so it would not be a research contribution, but would substantially improve Landslide's effectiveness regardless.
+
+\subsubsection{Symbolic Execution}
+% TODO
+symbolic execution\cite{klee,dawson}
+
+\subsubsection{Trace Minimization}
+% TODO
+% \ref{sec:future-backwards}
 
@@
 % theory things to study:
+
+% TODO: make visualisations for both of these
+\subsubsection{``Backwards'' Exploration}
+\label{sec:future-backwards}
 % - study the implications of backwards vs forward exploration
+% - ICB as in chess
+
+\subsubsection{Exploration Tree Structure}
 % - nadim's backtrack depth
-
-% vim: ft=tex

### 2012-04-28 challenges.tex
text-complete challenges

@@
 
-% TODO: cite dbug
-In user-space, a systematic exploration tool may use the underlying kernel to help control the concurrent behaviour of the system.
+In user-space, a systematic exploration tool may use the underlying kernel to help control the concurrent behaviour of the system\cite{dbug-ssv}.
 Simple system call invocations can cause a particular thread to run at a particular time, or to block while another thread runs first.
@@
 	\item A context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
-% TODO: why?
-don't overflow the stack, for one; for two, ignore runqueue conflicts
+A tool must be aware of the context switch process both to avoid flooding the kernel with interrupt frames that would overflow the stack, and to be able to ignore shared memory conflicts common to all thread transitions (Section~\ref{sec:por-independence}).
 	\item The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
@@
 \subsection{Ad-hoc Thread Communication}
-% inter-thread communication is not idealised!
-% TODO
+
+Inter-thread communication can be much more ad-hoc in kernel-space than in user-space\cite{datacollider}, and a systematic testing framework needs to be aware of all the kernel's synchronisation idioms.
+
+\begin{itemize}
+	\item Some systems for user-space systematic testing require that threads communicate only by message-passing, to better track the concurrency relationships between thread transitions\cite{dbug-ssv}.
+With few exceptions\cite{barrelfish}, kernels do not rely on message-passing as a primary communication mechanism, and to be compatible with the kernels of today, a testing framework must allow for less idealised state-sharing (Section~\ref{sec:components-memory}).
+	\item It is difficult to detect when threads are ready to do useful work.
+Even if notionally runnable, they may be waiting in a yield loop (Sections~\ref{sec:using-annotations} and \ref{sec:future-linux}).
+A tool that did not recognise this might decide to attempt to run that thread over and over, and get stuck because the other threads would never be allowed to make progress.
+	\item It is also difficult to demarcate threads' lifecycles: when a new thread gets created, when is it available to run?
+and when a thread is exiting, at what point does it stop running code (Section~\ref{sec:using-annotations})?
+In user-space, these boundaries are defined by the system call instructions, but in kernel-space, the boundary is fuzzier.
+	\item The kernel's complete control over the machine's virtual memory system means a system for tracking memory accesses must be aware of changing virtual memory mappings (Section~\ref{sec:using-requirements}).
+\end{itemize}
 
@@
 Do just-forked threads begin life through the usual context-switch-return path, or is there a special path for that?
+	\item {\bf Test Lifecycle}: Even apart from running on a particular input that the user wishes to test, the kernel may need to perform extra work irrelevant to the test, such as initialisation (e.g.
+the bootup process) and housekeeping (e.g.
+cleaning up dead processes).
+Which parts of this work should be included or excluded in the test?
 \end{itemize}

### 2012-04-28 related.tex
text-complete challenges

@@
 
-Dynamic Partial Order Reduction (DPOR) \cite{dpor,sdpor,dbug-retreat} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
+Dynamic Partial Order Reduction (DPOR) \cite{dpor,sdpor,dbug-retreat,distributed-dpor} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
 Landslide makes generous use of the DPOR algorithm.
@@
 
+\subsubsection{New Kernel Architectures}
+
 On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4\cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
@@
 Developing kernel code in languages and environments that enable formal complete verification, even in small components of a larger codebase, represents an admirable step towards writing correct code to begin with and avoiding many classes of bugs entirely.
-However, such steps have little bearing on code that exists in the world today, and it is also important to find ways to verify such sloppily-written code that already has bugs.
+
+Barrelfish\cite{barrelfish} is another novel kernel architecture designed with concurrency in mind.
+The barrelfish kernel relies exclusively on message-passing for inter-thread communication, for increased performance on massively multi-core systems.
+While not a verification project, the strict limitation of inter-thread communication makes the concurrent properties of the system easy to reason about in ways that modern production kernels do not provide.
+We believe this work, like seL4, is a step in the right direction for bringing better concurrent code into the world.
+
+However, in both cases, such steps have little bearing on code that exists in the world today, and it is also important to find ways to verify such sloppily-written code that already has bugs, which is the focus of our work.
 

### 2012-04-28 design.tex
more related: orthogonal
minor adds
text-complete challenges
buncha minor formatting changes

@@
 \subsubsection{Memory Tracking}
+\label{sec:components-memory}
 
@@
 \subsubsection{The Arbiter}
+\label{sec:components-arbiter}
 
@@
 
-A voluntary reschedule is identified when the previously-running thread entered the context switcher without first entering the timer interrupt handler, and any one of the following:
+A voluntary reschedule is identified when:
+\begin{itemize}
+	\item The previously-running thread entered the context switcher without first entering the timer interrupt handler,
+\end{itemize}
+
+and {\em any one} of the following:
 \begin{itemize}
@@
 \subsubsection{The Explorer}
+\label{sec:components-explore}
 
@@
 It stores the TID of the thread that was chosen, the state of the scheduler (what threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.
+
+\subsubsection{Test Lifecycle}
+\label{sec:components-test}
+% Talk about the test lifecycle - pre-readline; pre-test-start; etc.
+% Talk about the population tricks.
+% Talk about how we want idle/shell's wait() to be part of the test.
 
@@
 
-One alternative, simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
+One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
 This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
@@
 \subsubsection{Identifying Bugs}
+\label{sec:techniques-bugs}
 
@@
 	\item {\bf Deadlock bugs}: If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
-\end{enumerate}
+\suspend{enumerate}
 Uncertain metrics:
-\begin{enumerate}
+\resume{enumerate}
 	\item {\bf Memory leak bugs}: Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
@@
 If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed above).
+
+\subsubsection{Freed Memory Poisoning}
+
+A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
+Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
+
+Hence, even if the second thread that freed the block never accesses the particular address that the first thread used, the second thread's \texttt{free} still logically conflicts with the first thread's access.
+Landslide addresses this by treating every call to \texttt{free} as a {\em write} access to every address within the freed block when computing shared memory conflicts.
 
@@
 % talk about the all-ancestors vs induction thing TODO
-
-\subsection{Freed Memory Poisoning}
-
-A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
-Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
-
-Hence, even if the second thread that freed the block never accesses the particular address that the first thread used, the second thread's \texttt{free} still logically conflicts with the first thread's access.
-Landslide addresses this by treating every call to \texttt{free} as a {\em write} access to every address within the freed block when computing shared memory conflicts.
 

### 2012-04-28 futurework.tex
text-complete challenges
buncha minor formatting changes

@@
 Landslide could make use of data race detection techniques\cite{datacollider}, however, to strike a middle ground in which shared memory accesses it chooses.
-First, it should be aware of the types and interfaces associated with the kernel's synchronisation primitives (mutexes, condvars, semaphores, etc), and be able to treat operations on those as ``guaranteed to work'' (similar to Section~\ref{sec:por-independence}).
-Next, it should recognise acquire and release (or sleep and wake) operations on synchronisation primitives, and use that information to track lockset-type information, and use that to identify which memory accesses are raciest, and add decision points around those.
+\begin{enumerate}
+	\item It should be aware of the types and interfaces associated with the kernel's synchronisation primitives (mutexes, condvars, semaphores, etc), and be able to treat operations on those as ``guaranteed to work'' (similar to Section~\ref{sec:por-independence}).
+	\item Next, it should recognise acquire and release (or sleep and wake) operations on synchronisation primitives, and use that information to track lockset-type information.
+	\item Ultimately, it should use the lockset information in conjunction with the happens-before relation to identify which memory accesses are ``raciest'', and add decision points around those.
+\end{enumerate}
 
@@
 % linux, device drivers
+% open-coded sync dances: UGH (ad-hoc yield loops would make landslide get stuck)
 % embedded microcontroller operating systems, simulated

### 2012-04-29 evaluation.tex
add eval skeleton
add tables to eval

@@
 
+\subsection{User Experience}
+
+%\subsubsection{
+
+\subsection{Bug Case Studies}
+\label{sec:eval-casestudy}
+
+% TODO: intro text goes here
+
+\subsubsection{Numbers}
+
+Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
+All numbers represent the average of 5 trials.
+``N/A ($X$)'' indicates the bug was not found with a given test setup (and instead, Landslide took $X$ time to explore the entire decision tree).
+``$infty$'' indicates the bug was not found with stress testing.
+% TODO: after how long?
+what timeout?
+All Landslide trial times include the Simics start-up and kernel boot-up time (time between issuing the command and the test case beginning to run), roughly 15 seconds.
+
+All Landslide trials were run on the Gates-Hillman cluster machines (2.6 GHz Xeon).
+% TODO: say specs of crash machine
+
+\begin{figure*}[t!]
+	\begin{tabular}{|l|l||c|c|c|c||c|}
+		\hline
+		\multirow{2}{*}{\bf Kernel and test case} & \multirow{2}{*}{\bf Bug description} & \multicolumn{4}{c||}{\bf Custom decision points} & \multirow{2}{*}{\bf Stress test} \\
+		\cline{3-6}
+		& & \bf default & \bf lock & \bf unlock & \bf both & \\
+		\hline\hline
+		POBBLES vanish/vanish (a) & Deadlock & & & & & \\
+		\hline
+		POBBLES vanish/vanish (b) & Data race (panic) & & & & & \\
+		\hline
+		POBBLES wait/wait & Data structure bug (panic) & & 38.6s & & & \\
+		\hline
+		POBBLES thread\_fork/vanish & Use-after-free & & & & & \\
+		\hline
+		LudicrOS vanish/vanish & Data race (infinite loop) & & & & & \\
+		\hline
+		LudicrOS yield/vanish & Use-after-free & & & & & \\
+		\hline
+	\end{tabular}
+	\caption{Comparison of time taken to find bugs using Landslide and using conventional stress testing.
+	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.}
+	\label{fig:numbers}
+\end{figure*}
+
+All Landslide trials were run with ``backwards exploration'' enabled (Section~\ref{sec:using-search}).
+All trials were also run with Landslide configured to pay attention to only the relevant system calls (using \texttt{within\_function}; Section~\ref{sec:using-decision}).
+
+We believe it is reasonable to test for these bugs in this way - using the minimal set of system calls to be paid attention to as necessary to find the bug - because it follows the recommended workflow of using Landslide, which is to start with what the user judges to be the ``smallest relevant set'' of decision points.
+The configuration using \texttt{within\_function} was as follows.
+
+\begin{itemize}
+	\item POBBLES vanish/vanish(a): \texttt{within\_function vanish}.
+	\item POBBLES vanish/vanish(b): \texttt{within\_function vanish}.
+	\item POBBLES wait/wait: \texttt{within\_function wait}.
+	\item POBBLES thread\_fork/vanish: none was needed, but \texttt{within\_function thread\_fork} and \texttt{within\_function vanish} both were used for the sake of demonstration.
+	\item LudicrOS vanish/vanish: \texttt{within\_function vanish}.
+	\item LudicrOS yield/vanish: \texttt{within\_function yield}.
+\end{itemize}
+
+%%%%
+
+Table~\ref{fig:trees} presents more detailed information about the decision trees that Landslide explored when finding these bugs
+For each set of decision points on each bug, we give the total number of decision points in the tree, the total number of backtracks (i.e.
+branches explored before the bug was found), the average branch depth (i.e.
+number of decision points in each branch), and the average time taken to explore each branch (i.e.
+the total test time divided by the backtrack count)\footnote{
+It is not clear how meaningful the average time per branch is, considering the Simics and kernel startup time are included in the test time, and Landslide is not exploring then.}.
+
+\newcommand\nobug[1]{N/A {\em \small (#1)}}
+\begin{figure*}[t!]
+	\begin{tabular}{|l|l||c|c|c|c|}
+		\hline
+		\multirow{2}{*}{\bf Kernel and test case} & \multirow{2}{*}{\bf Property of tree} & \multicolumn{4}{|c|}{\bf Custom decision points} \\
+		\cline{3-6}
+		& & \bf default & \bf lock & \bf unlock & \bf both \\
+		\hline\hline
+		\multirow{4}{*}{POBBLES vanish/vanish (a)} & Decision points & \nobug{56} & 1296 & & \\
+		& Backtracks   & \nobug{16} & 376 & & \\
+		& Branch depth & \nobug{5} & 19 & & \\
+		& Time/branch  & & & & \\
+		\hline
+		\multirow{4}{*}{POBBLES vanish/vanish (b)} & Decision points & & & & \\
+		& Backtracks   & & & & \\
+		& Branch depth & & & & \\
+		& Time/branch  & & & & \\
+		\hline
+		\multirow{4}{*}{POBBLES wait/wait} & Decision points & 23 & 102 & 74 & 378 \\
+		& Backtracks   & 4 & 17 & 12 & 56 \\
+		& Branch depth & 6 & 10 & 9 & 14 \\
+		& Time/branch  & & & & \\
+		\hline
+		\multirow{4}{*}{POBBLES thread\_fork/vanish} & Decision points & 24 & 394 & 273 & 2269 \\
+		& Backtracks   & 5 & 70 & 56 & 410 \\
+		& Branch depth & 5 & 16 & 14 & 23 \\
+		& Time/branch  & & & & \\
+		\hline
+		\multirow{4}{*}{LudicrOS vanish/vanish} & Decision points & & & & \\
+		& Backtracks   & & & & \\
+		& Branch depth & & & & \\
+		& Time/branch  & & & & \\
+		\hline
+		\multirow{4}{*}{LudicrOS yield/vanish} & Decision points & & & & \\
+		& Backtracks   & & & & \\
+		& Branch depth & & & & \\
+		& Time/branch  & & & & \\
+		\hline
+	\end{tabular}
+	\caption{Information about the decision trees explored when finding bugs.
+As in the previous table, each test case was run with the four different sets of decision points.
+``\nobug{X}'' means the tree was completely explored because Landslide did not find a bug in that configuration.}
+	\label{fig:trees}
+\end{figure*}
+
+\subsection{Summary of Bugs Found}
+
 % vim: ft=tex

### 2012-04-29 futurework.tex
add tables to eval

@@
 Through our evaluation, we found many ways in which Landslide's user interface is lacking.
-These improvements would not be research developments, but would enhance the user experience in ways necessary for continuing work in the context of 15-410.
+These improvements would not be research developments, but would enhance the user experience in ways necessary for continuing work in the context of 15-410.\footnote{
+Landslide is also missing functionality to automatically identify decision points when newly-forked threads become runnable but not immediately switched to.
+This was not necessary to find the \texttt{double\_thread\_fork} bug in POBBLES (Section~\ref{sec:eval-casestudy}) only because POBBLES automatically switched to the child thread immediately.}
 

### 2012-04-29 evaluation.tex
update eval

--- 2012-04-24 evaluation.tex
@@
 
+\subsection{User Experience}
+
+% TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Bug Case Studies}
+\label{sec:eval-casestudy}
+
+\subsubsection{Bug Descriptions}
+
+% TODO: intro text goes here
+
+\subsubsection{Numbers}
+
+\newcommand\bugnum[2]{\textcolor{BrickRed}{{\bf #1} {\small \em (#2)}}}
+\newcommand\nobugnum[2]{\textcolor{Blue}{{\em #1} {\small \em (#2)}}}
+
+\begin{figure*}[t!]
+	\begin{center}
+	\begin{tabular}{|l||c|c|c|c||c|}
+		\hline
+		\multirow{2}{*}{\bf Kernel and test case} & \multicolumn{4}{c||}{\bf Custom decision points} & \multirow{2}{*}{\bf Stress test} \\
+		\cline{2-5}
+		& \bf default & \bf lock & \bf unlock & \bf both & \\
+		\hline\hline
+		POBBLES vanish/vanish (a) & \nobugnum{31.8}{0.6} & \bugnum{57.1}{1.7} & & & \\
+		\hline
+		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & & & \\
+		\hline
+		POBBLES wait/wait & \bugnum{23.3}{0.7} & \bugnum{27.9}{0.8} & \bugnum{27.9}{2.1} & \bugnum{41.6}{1.4} & \\
+		\hline
+		POBBLES thread\_fork/vanish & \bugnum{22}{0.6} & \bugnum{37.4}{1.1} & \bugnum{27.6}{0.5} & \bugnum{72}{2.6} & \\
+		\hline
+		LudicrOS vanish/vanish & \nobugnum{13.2}{0.2} & \bugnum{13.7}{0.7} & \bugnum{34.6}{1.1} & \bugnum{17.1}{0.3} & \\
+		\hline
+		LudicrOS yield/vanish & \nobugnum{12.3}{0.3} & \bugnum{11.4}{0.4} & \nobugnum{27.4}{0.8} & \bugnum{11.7}{0.4} & \\
+		\hline
+	\end{tabular}
+	\end{center}
+	\caption{Comparison of time taken (in seconds) to find bugs using Landslide and using conventional stress testing.
+	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
+	All numbers represent the average elapsed time in seconds from 5 trials, and the standard deviation is shown in parentheses.
+``\nobugnum{$X$}{$s$}'' indicates the bug was not found with a given test setup (and instead, Landslide took $X$ time to explore the entire decision tree).
+``$\infty$'' indicates the bug was not found with stress testing.
+% TODO: after how long?
+what timeout?
+	}
+	\label{fig:numbers}
+\end{figure*}
+
+% TODO: have a graph
+
+Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
+The experimental set-up is as follows:
+
+\begin{itemize}
+	\item All Landslide trial times include the Simics start-up and kernel boot-up time (time between issuing the command and the test case beginning to run), roughly 15 seconds for POBBLES and 10 seconds for LudicrOS.
+	\item All Landslide trials were run on the Gates-Hillman cluster machines (2.6 GHz Xeon).
+	% TODO: say specs of crash machine
+	\item All Landslide trials were run with ``backwards exploration'' enabled (Section~\ref{sec:using-search}).
+	\item All trials were also run with Landslide configured to pay attention to only the relevant system calls (using \texttt{within\_function}; Section~\ref{sec:using-decision}).
+
+	We believe it is reasonable to test for these bugs in this way - using the minimal set of system calls to be paid attention to as necessary to find the bug - because it follows the recommended workflow of using Landslide, which is to start with what the user judges to be the ``smallest relevant set'' of decision points.
+The configuration using \texttt{within\_function} was as follows.
+	\begin{itemize}
+		\item POBBLES vanish/vanish(a): \texttt{within\_function vanish}
+		\item POBBLES vanish/vanish(b): \texttt{within\_function vanish}
+		\item POBBLES wait/wait: \texttt{within\_function wait}
+		\item POBBLES thread\_fork/vanish: \texttt{within\_function thread\_fork} and \texttt{within\_function vanish}
+		\item LudicrOS vanish/vanish: \texttt{within\_function vanish}
+		\item LudicrOS yield/vanish: \texttt{within\_function yield} and \texttt{within\_function vanish}
+	\end{itemize}
+\end{itemize}
+
+%%%%
+
+\newcommand\bugtree[1]{\textcolor{BrickRed}{\bf #1}}
+\newcommand\nobugtree[1]{\textcolor{Blue}{\em #1}}
+\begin{figure*}[t!]
+	\begin{center}
+	\begin{tabular}{|l|l||c|c|c|c|}
+		\hline
+		\multirow{2}{*}{\bf Kernel and test case} & \multirow{2}{*}{\bf Property of tree} & \multicolumn{4}{|c|}{\bf Custom decision points} \\
+		\cline{3-6}
+		& & \bf default & \bf lock & \bf unlock & \bf both \\
+		\hline\hline
+		\multirow{4}{*}{POBBLES vanish/vanish (a)} & Decision points & \nobugtree{56} & \bugtree{1296} & & \\
+		& Total backtracks   & \nobugtree{16} & \bugtree{376} & & \\
+		& Average branch depth & \nobugtree{5} & \bugtree{19} & & \\
+		\hline
+		\multirow{4}{*}{POBBLES vanish/vanish (b)} & Decision points & \nobugtree{56} & \bugtree{1295} & \nobugtree{382071} & \\
+		& Total backtracks   & \nobugtree{16} & \bugtree{376} & \nobugtree{112706} & \\
+		& Average branch depth & \nobugtree{5} & \bugtree{17} & \nobugtree{16} & \\
+		\hline
+		\multirow{4}{*}{POBBLES wait/wait} & Decision points & \bugtree{23} & \bugtree{102} & \bugtree{74} & \bugtree{378} \\
+		& Total backtracks   & \bugtree{4} & \bugtree{17} & \bugtree{12} & \bugtree{56} \\
+		& Average branch depth & \bugtree{6} & \bugtree{10} & \bugtree{9} & \bugtree{14} \\
+		\hline
+		\multirow{4}{*}{POBBLES thread\_fork/vanish} & Decision points & \bugtree{24} & \bugtree{394} & \bugtree{273} & \bugtree{2269} \\
+		& Total backtracks   & \bugtree{5} & \bugtree{70} & \bugtree{56} & \bugtree{410} \\
+		& Average branch depth & \bugtree{5} & \bugtree{16} & \bugtree{14} & \bugtree{23} \\
+		\hline
+		\multirow{4}{*}{LudicrOS vanish/vanish} & Decision points & \nobugtree{10} & \bugtree{16} & \bugtree{141} & \bugtree{42} \\
+		& Total backtracks   & \nobugtree{2} & \bugtree{3} & \bugtree{48} & \bugtree{10} \\
+		& Average branch depth & \nobugtree{2} & \bugtree{7} & \bugtree{9} & \bugtree{14} \\
+		\hline
+		\multirow{4}{*}{LudicrOS yield/vanish} & Decision points & \nobugtree{8} & \bugtree{5} & \nobugtree{149} & \bugtree{7} \\
+		& Total backtracks   & \nobugtree{1} & \bugtree{0} & \nobugtree{43} & \bugtree{0} \\
+		& Average branch depth & \nobugtree{2} & \bugtree{0} & \nobugtree{9} & \bugtree{0} \\
+		\hline
+	\end{tabular}
+	\end{center}
+	\caption{Information about the decision trees explored when finding bugs.
+As in the previous table, each test case was run with the four different sets of decision points.
+``\nobugtree{X}'' means the tree was completely explored because Landslide did not find a bug in that configuration.
+``\bugtree{X}'' reflects the portion of the tree that was explored before a bug was found.}
+	\label{fig:trees}
+\end{figure*}
+
+Table~\ref{fig:trees} presents more detailed information about the decision trees that Landslide explored when finding these bugs.
+For each set of decision points on each bug, we give the total number of decision points in the tree, the total number of backtracks (i.e.
+branches explored before the bug was found), and the average branch depth (i.e.
+number of decision points in each branch).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Summary of Bugs Found}
+
+% TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Discussion}
+
+\subsubsection{Invariants}
+
+While evaluating Landslide on these bugs, we determined two invariants that should hold for multiple explorations on the same test case.
+
+\begin{enumerate}
+	\item {\bf Ordering invariant}: For a given set of decision points, exploring the tree in multiple different orders (``forwards''/``backwards'') should produce the same result in terms of whether a bug was found or not.
+The bugs found may be different, and the number of branches explored may be different, but it should never be that one ordering finds a bug while another ordering of the same tree finds no bug.
+	\item {\bf Superset invariant}: For a given set of decision points, if an exploration of the resulting tree finds a bug, using a superset of that set of decision points should also find a bug.
+This is because the first tree will be a sub-tree of the second, as shown by never preempting at a decision point that appears in the second set but not the first.
+\end{enumerate}
+
+In short, even though Landslide may give false negatives from using imperfect sets of decision points, the exploration itself must be sound (i.e.
+not missing any bugs that exist in the resulting tree).\footnote{
+When running LudicrOS yield/vanish with decision points on \texttt{mutex\_unlock} but not on \texttt{mutex\_lock}, we found that the ordering invariant failed - ``backwards'' exploration found no bug, while ``forwards'' exploration did.
+We attribute this to a bug in Landslide itself, and present the results for the backwards exploration as usual, in which Landslide found no bug.}
+
+\subsubsection{Recommended Testing Strategies}
+
+% TODO
+
 % vim: ft=tex

### 2012-04-29 design.tex
minor alg update in design

@@
 		\\
-		\Function{scheduler\_update}{int eip}
+		\Function{scheduler\_update}{int pc}
 			\If{schedule\_in\_flight}
 				\State {\sc assert}(in\_timer\_handler || in\_context\_switch);
-				\If { {\sc kernel\_exiting\_timer}(eip) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(eip))}
+				\If { {\sc kernel\_exiting\_timer}(pc) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(pc))}
 					\State {\em The kernel has just finished rescheduling and is about to resume normal thread execution.}
@@
 	\caption{The scheduling algorithm.
-This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{eip} each time.
-The predicates on \texttt{eip} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
+This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{pc} (the program counter) each time.
+The predicates on \texttt{pc} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
 \end{algorithm*}

### 2012-04-29 challenges.tex
change back to 1col and fix up bullet point text

@@
 \begin{itemize}
-	\item Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
-	\item A context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
+	\item {\bf Non-preemptibility.} Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
+	\item {\bf The context switcher.} A context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
 A tool must be aware of the context switch process both to avoid flooding the kernel with interrupt frames that would overflow the stack, and to be able to ignore shared memory conflicts common to all thread transitions (Section~\ref{sec:por-independence}).
-	\item The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
+	\item {\bf Runqueue tracking.} The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
 \end{itemize}
@@
 \begin{itemize}
-	\item Some systems for user-space systematic testing require that threads communicate only by message-passing, to better track the concurrency relationships between thread transitions\cite{dbug-ssv}.
+	\item {\bf Avoiding reliance on message-passing.} Some systems for user-space systematic testing require that threads communicate only by message-passing, to better track the concurrency relationships between thread transitions\cite{dbug-ssv}.
 With few exceptions\cite{barrelfish}, kernels do not rely on message-passing as a primary communication mechanism, and to be compatible with the kernels of today, a testing framework must allow for less idealised state-sharing (Section~\ref{sec:components-memory}).
-	\item It is difficult to detect when threads are ready to do useful work.
+	\item {\bf Recognising blocked threads.} It is difficult to detect when threads are ready to do useful work.
 Even if notionally runnable, they may be waiting in a yield loop (Sections~\ref{sec:using-annotations} and \ref{sec:future-linux}).
 A tool that did not recognise this might decide to attempt to run that thread over and over, and get stuck because the other threads would never be allowed to make progress.
-	\item It is also difficult to demarcate threads' lifecycles: when a new thread gets created, when is it available to run?
+	\item {\bf Thread lifecycle tracking.} It is also difficult to demarcate threads' lifecycles: when a new thread gets created, when is it available to run?
 and when a thread is exiting, at what point does it stop running code (Section~\ref{sec:using-annotations})?
 In user-space, these boundaries are defined by the system call instructions, but in kernel-space, the boundary is fuzzier.
-	\item The kernel's complete control over the machine's virtual memory system means a system for tracking memory accesses must be aware of changing virtual memory mappings (Section~\ref{sec:using-requirements}).
+	\item {\bf Use of virtual memory.} The kernel's complete control over the machine's virtual memory system means a system for tracking memory accesses must be aware of changing virtual memory mappings (Section~\ref{sec:using-requirements}).
 \end{itemize}
@@
 \begin{itemize}
-	\item With the exception of a non-preemptible ``scheduler lock'', and yield-looping mutexes (both of which must be instrumented by the user), a thread's presence on the runqueue indicates that a finite number of timer interrupts in succession will eventually cause it to run.
-	\item The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
+	\item {\bf Timer ticks control ``runnable'' threads.} With the exception of a non-preemptible ``scheduler lock'', and yield-looping mutexes (both of which must be instrumented by the user), a thread's presence on the runqueue indicates that a finite number of timer interrupts in succession will eventually cause it to run.
+	\item {\bf No idling when progress can be made.} The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
 Landslide uses the idle loop to detect when a test begins/ends and to detect when all threads in a test are wedged.

### 2012-04-29 design.tex
change back to 1col and fix up bullet point text

@@
 
-\subsection{Techniques}
-\subsubsection{Thread Scheduling}
-
-In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
-
-Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
-Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.
-
-One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
-This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
-We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
+\subsection{Thread Scheduling}
 
@@
 
-\subsubsection{Identifying Bugs}
+In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+
+Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
+Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.
+
+One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
+This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
+We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
+
+\subsection{Identifying Bugs}
 \label{sec:techniques-bugs}
@@
 
-Certain metrics:
+Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
+If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed below).
+
+\subsubsection{Accurate Bug-Detection Conditions}
+
 \begin{enumerate}
-	\item {\bf Kernel panic bugs}: If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
-	\item {\bf Use-after-free bugs}: Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range.
+	\item {\bf Kernel panic bugs.} If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
+	\item {\bf Use-after-free bugs.} Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range.
 If not, Landslide proclaims the access to be illegal.\footnote{In the same way, Landslide also detects calls to \texttt{free} on blocks that were already freed or never allocated at all.}
-	\item {\bf Deadlock bugs}: If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
+	\item {\bf Deadlock bugs.} If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
 \suspend{enumerate}
-Uncertain metrics:
+
+\subsubsection{Uncertain Bug-Detection Conditions}
+
 \resume{enumerate}
-	\item {\bf Memory leak bugs}: Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
+	\item {\bf Memory leak bugs.} Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
 If memory allocated during the test was not freed, Landslide assumes that it was leaked.
 		(Some kernel designs may legitimately behave this way, so this bug-check may be disabled when testing such kernels.)\footnote{There is, of course, much room for improvement in this metric, but it is not part of the research contribution.}
-	\item {\bf Infinite loop bugs}: Landslide judges whether the kernel has entered an infinite loop by comparing the current state of the decision tree to past executions of the same test case.
+	\item {\bf Infinite loop bugs.} Landslide judges whether the kernel has entered an infinite loop by comparing the current state of the decision tree to past executions of the same test case.
 	\begin{itemize}
@@
 \end{enumerate}
-Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
-If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed above).
-
-\subsubsection{Freed Memory Poisoning}
-
-A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
-Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
-
-Hence, even if the second thread that freed the block never accesses the particular address that the first thread used, the second thread's \texttt{free} still logically conflicts with the first thread's access.
-Landslide addresses this by treating every call to \texttt{free} as a {\em write} access to every address within the freed block when computing shared memory conflicts.
 
@@
 The independence relation expresses which transitions do not read-and-write or write-and-write to the same shared memory addresses.
-% talk about ignoring runqueue accesses, etc TODO
+
+\begin{itemize}
+	\item {\bf Always-accessed memory locations.} % talk about ignoring runqueue accesses, etc TODO
+	\item {\bf Freed memory poisoning.}
+		A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
+Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
+
+		Hence, even if the second thread that freed the block never accesses the particular address that the first thread used, the second thread's \texttt{free} still logically conflicts with the first thread's access.
+Landslide addresses this by treating every call to \texttt{free} as a {\em write} access to every address within the freed block when computing shared memory conflicts.
+\end{itemize}
 

### 2012-04-29 evaluation.tex
change back to 1col and fix up bullet point text

@@
 		\hline\hline
-		POBBLES vanish/vanish (a) & \nobugnum{31.8}{0.6} & \bugnum{57.1}{1.7} & & & \\
+		POBBLES vanish/vanish (a) & \nobugnum{31.8}{0.6} & \bugnum{57.1}{1.7} & $\infty$ & $\infty$ & \\
 		\hline
-		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & & & \\
+		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & & $\infty$ & \\
 		\hline
@@
 ``\nobugnum{$X$}{$s$}'' indicates the bug was not found with a given test setup (and instead, Landslide took $X$ time to explore the entire decision tree).
-``$\infty$'' indicates the bug was not found with stress testing.
+``$\infty$'' indicates the bug was not found with stress testing (after 1 hour), or that Landslide's search did not finish (after 8 hours).
 % TODO: after how long?
@@
 		\hline\hline
-		\multirow{4}{*}{POBBLES vanish/vanish (a)} & Decision points & \nobugtree{56} & \bugtree{1296} & & \\
-		& Total backtracks   & \nobugtree{16} & \bugtree{376} & & \\
-		& Average branch depth & \nobugtree{5} & \bugtree{19} & & \\
+		\multirow{4}{*}{POBBLES vanish/vanish (a)} & Decision points & \nobugtree{56} & \bugtree{1296} & N/A & N/A \\
+		& Total backtracks   & \nobugtree{16} & \bugtree{376} & N/A & N/A \\
+		& Average branch depth & \nobugtree{5} & \bugtree{19} & N/A & N/A \\
 		\hline
-		\multirow{4}{*}{POBBLES vanish/vanish (b)} & Decision points & \nobugtree{56} & \bugtree{1295} & \nobugtree{382071} & \\
-		& Total backtracks   & \nobugtree{16} & \bugtree{376} & \nobugtree{112706} & \\
-		& Average branch depth & \nobugtree{5} & \bugtree{17} & \nobugtree{16} & \\
+		\multirow{4}{*}{POBBLES vanish/vanish (b)} & Decision points & \nobugtree{56} & \bugtree{1295} & \nobugtree{382071} & N/A \\
+		& Total backtracks   & \nobugtree{16} & \bugtree{376} & \nobugtree{112706} & N/A \\
+		& Average branch depth & \nobugtree{5} & \bugtree{17} & \nobugtree{16} & N/A \\
 		\hline
@@
 \begin{enumerate}
-	\item {\bf Ordering invariant}: For a given set of decision points, exploring the tree in multiple different orders (``forwards''/``backwards'') should produce the same result in terms of whether a bug was found or not.
+	\item {\bf Ordering invariant.} For a given set of decision points, exploring the tree in multiple different orders (``forwards''/``backwards'') should produce the same result in terms of whether a bug was found or not.
 The bugs found may be different, and the number of branches explored may be different, but it should never be that one ordering finds a bug while another ordering of the same tree finds no bug.
-	\item {\bf Superset invariant}: For a given set of decision points, if an exploration of the resulting tree finds a bug, using a superset of that set of decision points should also find a bug.
+	\item {\bf Superset invariant.} For a given set of decision points, if an exploration of the resulting tree finds a bug, using a superset of that set of decision points should also find a bug.
 This is because the first tree will be a sub-tree of the second, as shown by never preempting at a decision point that appears in the second set but not the first.

### 2012-04-29 evaluation.tex
update eval

@@
 		\hline
-		\multirow{2}{*}{\bf Kernel and test case} & \multicolumn{4}{c||}{\bf Custom decision points} & \multirow{2}{*}{\bf Stress test} \\
+		\multirow{2}{*}{\bf Kernel and test case} & \multicolumn{4}{c||}{\bf Decision points used} & \multirow{2}{*}{\bf Stress test} \\
 		\cline{2-5}
@@
 		\hline
-		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & & $\infty$ & \\
+		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & \nobugnum{8057.9}{336.9} & $\infty$ & \\
 		\hline
@@
 	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
-	All numbers represent the average elapsed time in seconds from 5 trials, and the standard deviation is shown in parentheses.
+	All numbers represent the average from 5 trials, with the standard deviation shown in parentheses.
 ``\nobugnum{$X$}{$s$}'' indicates the bug was not found with a given test setup (and instead, Landslide took $X$ time to explore the entire decision tree).
@@
 		\hline
-		\multirow{2}{*}{\bf Kernel and test case} & \multirow{2}{*}{\bf Property of tree} & \multicolumn{4}{|c|}{\bf Custom decision points} \\
+		\multirow{2}{*}{\bf Kernel and test case} & \multirow{2}{*}{\bf Property of tree} & \multicolumn{4}{|c|}{\bf Decision points used} \\
 		\cline{3-6}
@@
 \subsubsection{Recommended Testing Strategies}
+\label{sec:discussion-strategies}
 

### 2012-04-30 futurework.tex
Add images

@@
 % TODO
+
+\begin{figure*}[h]
+	\begin{center}
+	\includegraphics[width=0.75\textwidth]{wont-modify-vail.jpg}
+	\end{center}
+	\caption{Ex-members of 15-410 course staff lament the difficulty involved in teaching students concurrency debugging skills.}
+\end{figure*}
 
@@
 
-With either method, in order to mitigate exponential explosion of tree size, it would probably be best to only allow the decision point set size to increase by one each iteration.
-An iteration framework could explore using different combinations/subsets of such, using a workqueue model, and using heuristics to identify which decision points are most useful and should be used first.
+In Section~\ref{sec:discussion-strategies}, we recommend a strategy for an iteration framework to effectively explore trees with multiple different sets of decision points, while not knowing ahead of time which, if any, may contain a bug.
 

### 2012-04-30 evaluation.tex
write more eval

@@
 \subsubsection{Numbers}
+\label{sec:eval-numbers}
 
-\newcommand\bugnum[2]{\textcolor{BrickRed}{{\bf #1} {\small \em (#2)}}}
-\newcommand\nobugnum[2]{\textcolor{Blue}{{\em #1} {\small \em (#2)}}}
+\newcommand\bugnum[2]{\textcolor{BrickRed}{{\bf #1} {\scriptsize \em (#2)}}}
+\newcommand\nobugnum[2]{\textcolor{Blue}{{\em #1} {\scriptsize \em (#2)}}}
 
@@
 	\begin{center}
+	\small
 	\begin{tabular}{|l||c|c|c|c||c|}
@@
 		\hline
+		\multicolumn{6}{c}{Key: \bugnum{seconds}{stddev} indicates bug found; \nobugnum{seconds}{stddev} indicates whole tree explored with no bug.} \\
 	\end{tabular}
@@
 	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
-	All numbers represent the average from 5 trials, with the standard deviation shown in parentheses.
-``\nobugnum{$X$}{$s$}'' indicates the bug was not found with a given test setup (and instead, Landslide took $X$ time to explore the entire decision tree).
+	All numbers represent the average from 5 trials, with the standard deviation given in parentheses.
 ``$\infty$'' indicates the bug was not found with stress testing (after 1 hour), or that Landslide's search did not finish (after 8 hours).
@@
 	\begin{center}
+	\small
 	\begin{tabular}{|l|l||c|c|c|c|}
@@
 
-% TODO
+We make several observations about the experimental results from Section~\ref{sec:eval-numbers}.
+
+\begin{enumerate}
+	\item {\bf Fewer is faster.} While it is theoretically possible that a tree built of finer-grained interleavings might encounter a bug after fewer overall backtracks, we found that this did not happen in practice.\footnote{
+		The one suspicious case is the LudicrOS vanish/vanish bug - exploring the ``both'' tree found the bug faster than exploring the ``unlock'' tree, despite the latter's decision points being a subset of the former's.
+However, we see that exploring the ``lock'' tree found the bug faster than either, so overall the ``both'' tree did not outperform the fastest of the smaller trees.}
+		In general, for two sets of decision points that both find the same bug, the one that results in shorter {\em branches} will result in fewer {\em backtracks} needed to expose the bug, and hence less overall time.
+	\item {\bf Different decision points are differently likely to expose different bugs.} We see that even though the ``lock'' and ``unlock'' trees tended to be about the same size, they were each sometimes better than the other at finding bugs.
+The ``lock'' tree did better on the vanish/vanish bugs and the yield/vanish bug, while the ``unlock'' tree did better on the wait/wait bug and the thread\_fork/vanish bug.\footnote{
+	The latter two bugs needed no more than the default set to uncover at all, but we claim this still demonstrates the principle in general.}
+	\item {\bf Finding a bug is fast, if it exists.} As especially exhibited in the POBBLES vanish/vanish (b) case, if two sets of decision points generate trees of approximately equal size, but one tree contains a bug and the other doesn't, searching the bugful tree will terminate much more quickly.
+Of course, it is always possible that a bug may only exist in the very last branch of a tree, but we found that in general bugs show up early during exploration.
+\end{enumerate}
+
+% TODO: Investigate the tree size that bugs did exist in. Investigate how many buggy branches there were.
+
+In light of these, we recommend several principles to govern an overall strategy to automatically iterate through different sets of decision points in search of a bug.\footnote{
+We now say ``Landslide'' here to refer to a hypothetical test framework to embody these strategies, although of course it would not have to be named that.}
+
+\begin{enumerate}
+	\item {\bf Iterate exploring, starting with smaller decision sets.} To properly test for a bug in a particular test case, Landslide should try as many different decision sets as possible.
+	The first exploration should always be just the default decision set, because that tends to complete quickly, and can help identify new decision points (Section~\ref{sec:future-analysis}).
+	When Landslide has multiple decision sets as candidates for the next exploration, it should prefer to explore ones that result in shorter average branch depth.
+	In this way, Landslide will tend to find bugs with the minimal decision set needed to expose them, in accordance with observation 1 above.
+	\item {\bf Run multiple explorations at once.}
+	As per observation 2, if Landslide has two decision sets that result in roughly equal average branch depth, it cannot know in advance whether either one will find a bug and/or finish significantly faster than the other.
+As such, it should try to run both explorations in parallel, wait for either one to finish, and continue iterating as appropriate even if the other one has not finished.
+	\item {\bf De-prioritise longer-lasting test configurations.}
+	With finite resources for parallelisation, Landslide should attempt to load-balance whatever searches are running in parallel according to each one's likelihood of finding a bug.
+	In light of observation 3, if a particular search is running abnormally long for its average branch depth (the POBBLES vanish/vanish (b) bug with the ``unlock'' tree is a prime example), Landslide could judge that it is less likely to end soon with a positive result, and prioritise searches with other decision sets.
+\end{enumerate}
 

### 2012-04-30 futurework.tex
write more eval

@@
 
-As Landslide is an experimental foray into the world of systematic exploration in kernel-space, it has opened up many avenues for potential future improvements.
+As Landslide is an experimental foray into the world of systematic exploration in kernel-space, and a detailed study of systematic exploration in general, it has opened up many avenues for potential future improvements.
 
@@
 	\end{center}
-	\caption{Ex-members of 15-410 course staff lament the difficulty involved in teaching students concurrency debugging skills.}
+	\caption{Ex-members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
 \end{figure*}
@@
 \subsubsection{Data Race Detection and Static Analysis}
+\label{sec:future-analysis}
 
@@
 
-Next is the question of when during the testing process these decision points should be introduced:
-
-\begin{itemize}
-	\item One possibility would be to begin exploration with a small set of decision points, explore the resulting tree, analyse the tree post-hoc to identify additional decision points, and iterate exploring with the new set.
-	\item Another possibility would be to identify decision points along the way, analysing each branch of the tree after executing it, to generate the decision points for that very branch, and thence use them to find which branch to explore next.
-\end{itemize}
-
-In Section~\ref{sec:discussion-strategies}, we recommend a strategy for an iteration framework to effectively explore trees with multiple different sets of decision points, while not knowing ahead of time which, if any, may contain a bug.
-
 \subsubsection{Parallelism}
@@
 
-\subsubsection{Trace Minimization}
+\subsubsection{Trace Minimisation}
 % TODO
@@
 \label{sec:future-shaping}
-% - garth's shaping
+
+% TODO more
+\subsubsection{Iterating Decision Sets}
+
+In Section~\ref{sec:discussion-strategies}, we recommend an iteration strategy to effectively explore trees with multiple different sets of decision points, while not knowing ahead of time which, if any, may contain a bug.
+This strategy would give rise to a test framework which could explore multiple different types and granularities of interleavings at once, heuristically judge which are more likely to uncover bugs, and prioritise resource allocation and search direction accordingly.
+
+There is also the question of when during the testing process new decision points should be introduced.
+One possibility would be to begin exploration with a small set of decision points, explore the resulting tree, analyse the tree post-hoc to identify additional decision points, and iterate exploring with the new set.
+Another possibility would be to identify decision points along the way, analysing each branch of the tree after executing it, to generate the decision points for that very branch, and thence use them to find which branch to explore next.
+It remains to be seen which approach would be more effective.
 

### 2012-04-30 terminology.tex
text-complete challenges
change back to 1col and fix up bullet point text
write more eval

@@
 
-Basic terminology:
+\subsection{Basic Terms}
+
 \begin{enumerate}
@@
 
-Scheduling terminology:
+\subsection{Scheduling Terms}
+
 \resume{enumerate}
@@
 
-Systematic exploration terminology:
+\subsection{Systematic Exploration Terms}
+
+% TODO: have a diagram of the decision tree
 \resume{enumerate}
-	\item {\bf Decision point} (or {\bf Choice point}):
+	\item {\bf Decision point}:
 		A point during execution (between two consecutive instructions, precisely) which is deemed ``interesting'' in terms of the likelihood that thread switches at that point will cause concurrency bugs to arise.
@@
 At a given decision point, Landslide chooses which thread will run (and causes it to do so), and that thread executes a transition, after which point Landslide will have identified a subsequent decision point
+	\item {\bf Decision point set} (or {\bf Decision set}):
+		One or more predicates on the execution state of the guest kernel which identify whether any current state should be a decision point.
+(For example, when we say ``the decision set including all calls to \texttt{mutex\_lock}'', the corresponding predicate is simply ``did the kernel just invoke \texttt{mutex\_lock}?''.)
 	\item {\bf Decision tree} (or {\bf Execution tree}):
@@
 Abstractly, this tree comprises all possible states reachable by any interleaving of threads.
+		A decision tree is created/discovered during an exploration using a particular decision point set; the set is what the user configures, and the tree is what arises as a result.
 	\item {\bf Branch}:

### 2012-04-30 challenges.tex
Add images
add blurb; resize fonts and fix spacing

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure*}[h]
+	\begin{center}
+	\includegraphics[width=0.75\textwidth]{vip1066720.jpg}
+	\end{center}
+	\caption{{\em !`Cuidado! !`Las llamas son muy peligrosas!}\cite{de0u}}
+\end{figure*}
 
@@
 
-Inter-thread communication can be much more ad-hoc in kernel-space than in user-space\cite{datacollider}, and a systematic testing framework needs to be aware of all the kernel's synchronisation idioms.
+Inter-thread communication can be much more ad-hoc in kernel-space than in user-space\hspace{0in}\cite{datacollider}, and a systematic testing framework needs to be aware of all the kernel's synchronisation idioms.
 

### 2012-04-30 related.tex
buncha minor formatting changes
add blurb; resize fonts and fix spacing

@@
 
-Dynamic Partial Order Reduction (DPOR) \cite{dpor,sdpor,dbug-retreat,distributed-dpor} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
+Dynamic Partial Order Reduction (DPOR)\hspace{0in}\cite{dpor,sdpor,dbug-retreat,distributed-dpor} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
 Landslide makes generous use of the DPOR algorithm.
@@
 
-On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4\cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
+On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4\hspace{0in}\cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
 Because Haskell is type-safe (strict constraints on legal program behaviour that make it easier to reason about) and pure (most code cannot have stateful side-effects), the developers of seL4 were able to formally prove the specification's correctness and the implementation's adherence to the spec.
@@
 While not a verification project, the strict limitation of inter-thread communication makes the concurrent properties of the system easy to reason about in ways that modern production kernels do not provide.
-We believe this work, like seL4, is a step in the right direction for bringing better concurrent code into the world.
+We believe Barrelfish, like seL4, is a step in the right direction for bringing better concurrent code into the world.
 
@@
 Deterministic multithreading is a different technique for dealing with concurrency bugs which focuses on avoiding them when running production code rather than trying to expose them during testing.
-Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
+Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\hspace{0in}\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
 Deterministic multithreading serves a different purpose than systematic exploration; it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.

### 2012-04-30 thesis.tex
write more design
add more design writing
add inflight alg listing
a little more related
add terminology
text-complete challenges
add tables to eval
add usepackage
change back to 1col and fix up bullet point text
add blurb; resize fonts and fix spacing

@@
-\documentclass[10pt,twocolumn]{article}
+\documentclass[12pt]{article}
 %\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
@@
 \usepackage{draftwatermark}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{multirow}
+\usepackage[usenames,dvipsnames]{xcolor}
 %\usepackage{epsfig}
 \usepackage[bottom]{footmisc}
+\usepackage{setspace}
+\setstretch{1.05}
 
@@
 \title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\texttt{bblum@andrew.cmu.edu}) \and Garth Gibson (\texttt{garth@cs.cmu.edu})}
+\author{Benjamin Blum\thanks{Carnegie Mellon University; MS student; \texttt{bblum@andrew.cmu.edu}.} \and Garth Gibson \thanks{Carnegie Mellon University; Professor; \texttt{garth@cs.cmu.edu}.}}
 \maketitle
@@
 
-\input{related}
-\input{challenges}
+% TODO: contribution and organisation sections
+
+\input{related} % COMPLETE!
+\input{challenges} % COMPLETE!
+\input{terminology} % almost complete
 \input{design}
+\input{using}
 \input{evaluation}
 \input{futurework}
+\input{conclusion}
 \input{acknowledgements}
+
+% TODO: read presubmit.txt and check everything off
 
@@
 
+\appendix
+\section{Note}
+
+There is a wondrous fractal nature to all aspects of the universe.
+
+At the time of writing this document, my life is focused on this one project.
+There is a vast amount of thought in this thesis, and many more rich worlds lie hidden in each of the open avenues for future work.
+
+Yet it pales in comparison to the scope of a Ph.D. thesis, and even more so next to all the details of a person's entire life.
+After I finish this project, it will become but a fond memory, a mere building block in a city of even grander things.
+
+Human nature is beautiful in its ability to exist at any level of perspective, to hold in awareness a story of any size, and to zoom in and out at will to see how multiple parts fit together.
+Tearing ourselves up over minutiae is part of the fun, and contemplating overarching directions beyond our ability to change is also part of the fun.
+
+Our true home, though, lies somewhere in between; a place of potential, and of peace. May we always remember to return to it.
+
 \end{document}

### 2012-04-30 evaluation.tex
minor fmt changes evaluation
minor changes in eval; futurework

@@
 		\hline
-		POBBLES vanish/vanish (b) & \nobugnum{32}{0.4} & \bugnum{51.5}{2.1} & \nobugnum{8057.9}{336.9} & $\infty$ & \\
+		POBBLES vanish/vanish (b) & \nobugnum{32.0}{0.4} & \bugnum{51.5}{2.1} & \nobugnum{8057.9}{336.9} & $\infty$ & \\
 		\hline
@@
 		\hline
-		POBBLES thread\_fork/vanish & \bugnum{22}{0.6} & \bugnum{37.4}{1.1} & \bugnum{27.6}{0.5} & \bugnum{72}{2.6} & \\
+		POBBLES thread\_fork/vanish & \bugnum{22.0}{0.6} & \bugnum{37.4}{1.1} & \bugnum{27.6}{0.5} & \bugnum{72.0}{2.6} & \\
 		\hline
@@
 	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
-	All numbers represent the average from 5 trials, with the standard deviation given in parentheses.
+	All numbers represent the average from 5 trials, with the standard deviations given in parentheses.
 ``$\infty$'' indicates the bug was not found with stress testing (after 1 hour), or that Landslide's search did not finish (after 8 hours).
@@
 The configuration using \texttt{within\_function} was as follows.
+	{\small
 	\begin{itemize}
@@
 	\end{itemize}
+	}
 \end{itemize}
@@
 	The latter two bugs needed no more than the default set to uncover at all, but we claim this still demonstrates the principle in general.}
-	\item {\bf Finding a bug is fast, if it exists.} As especially exhibited in the POBBLES vanish/vanish (b) case, if two sets of decision points generate trees of approximately equal size, but one tree contains a bug and the other doesn't, searching the bugful tree will terminate much more quickly.
+	\item {\bf Finding a bug is fast, if it exists.} As especially exhibited in the POBBLES vanish/vanish (b) case, if two sets of decision points generate trees of approximately equal size, but one tree contains a bug and the other doesn't, then searching the bugful tree will likely terminate much more quickly.
 	Of course, it is always possible that a bug may only exist in the very last branch of a tree, but we found that in general bugs show up early during exploration.

### 2012-05-01 futurework.tex
minor changes in eval; futurework
buncha minor edits

@@
 
-Through our evaluation, we found many ways in which Landslide's user interface is lacking.
-These improvements would not be research developments, but would enhance the user experience in ways necessary for continuing work in the context of 15-410.\footnote{
+In Section~\ref{sec:eval-feedback}, we describe many ways in which Landslide's user interface is lacking.
+These improvements would not necessarily be research developments, but would enhance the user experience in ways necessary for continuing work in the context of 15-410.\footnote{
 Landslide is also missing functionality to automatically identify decision points when newly-forked threads become runnable but not immediately switched to.
@@
 \label{sec:future-shaping}
-
 % TODO more
+\footnote{Garth coined the name ``shaping'' to refer to this testing approach.}
 \subsubsection{Iterating Decision Sets}

### 2012-05-01 evaluation.tex
add lots of user study stuff

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 410 - talk about current state/methods (for students and for TAS)
-% test suite presented
-% process of instrumenting
+
+We evaluated Landslide in two ways.
+First, we conducted a user study with present students of 15-410, to understand the eases and difficulties of Landslide's user interface process, to understand how to polish it into a useful debugging tool.
+Second, we did a case study of six bugs in two kernels, investigating them in-depth to understand how systematic testing might best be harnessed to find bugs without in advance knowing the decision set necessary to expose them.
 
@@
 
+To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project, and had them use Landslide with their own kernels.
+
+In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.
+The lecture slides are available at \url{http://www.cs.cmu.edu/~410-s12/lectures/L30_Landslide.pdf}.\footnote{
+Also available at \url{http://bblum.net/landslide-lecture.pdf} and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-lecture.pdf}.}
+
+We wished to measure which phases of using Landslide are most time-consuming, so we asked students approximately every half-hour to record the amount of time they had spent doing each phase:
+
+\begin{enumerate}
+	\item Annotating their kernel (Section~\ref{sec:using-annotations}).
+	\item Implementing instrumentation within Landslide (Section~\ref{sec:using-student-c}).
+	\item Fixing problems encountered getting Landslide to run, such as editing their kernel to meet Landslide's requirements and debugging incorrect instrumentation.
+	\item Customising Landslide's search parameters such as decision points (Section~\ref{sec:using-customise}).
+	\item Analysing the output from bugs that Landslide found (Section~\ref{sec:using-interpreting}).
+\end{enumerate}
+
+We also wanted descriptive feedback, to evaluate what specific parts of the process of using Landslide need improved.
+We asked students to write brief remarks whenever they recorded amounts of time spent, and also after they were finished to answer the following questions.
+
+\begin{itemize}
+	\item Were you able to get Landslide to work with your kernel (i.e., run through a minimal exploration tree completely)?
+If so, how long did it take, from sitting down to getting it to work?
+If not, did your kernel do something that was incompatible with Landslide, or were you unable to get the instrumentation right for some other reason?
+	\item Did you find bugs while using Landslide, that you imagine would have been very hard to find otherwise? Did Landslide's output help you understand what caused them?
+	\item Were you able to get Landslide to say ``you survived!'' with custom decision points?
+Did you think the set of decision points you used for this provide a strong guarantee about the absence of races?
+	\item Describe your experience configuring the set of decision points. What was intuitive, obvious to do? Did you feel stuck at any point, not knowing where to go next?
+	\item Was there anything you wanted to make Landslide do that it didn't support?
+\end{itemize}
+
+Finally, we briefly studied each bug that the students found, which we describe below.
+
+\subsubsection{Time Breakdown}
+
 % TODO
+
+\subsubsection{Descriptive Feedback}
+\label{sec:eval-feedback}
+
+The students provided specific remarks which fell into two categories: issues with Landslide's user interface, and needing to change something in one's kernel to make it work with Landslide.
+
+The following feedback pertains to the user interface.
+For Landslide to be appealing as a debugging tool, we should improve the user interface (or the infrastructure, in some cases) to address these issues.
+
+\begin{itemize}
+	\item {\bf Difficulty of debugging incorrect instrumentation.} When the user-provided annotations/instrumentation are incorrect, Landslide's error messages are difficult to understand, and often don't point to the real problem.
+	\item {\bf Interpreting debugging output.} Some students reported the decision trace representation could be better.
+One student reported the output was easy to follow by ``tracing the life of the buggy thread'', which is an insight into the learning process we should capitalise on.
+	\item {\bf Ease of configuring decision sets.} Most students seemed to find the process of adding more decision points intuitive. One reported, ``quick and effective at detecting basic bugs/races.''
+	\item {\bf General feature wishlist.} The following features, of varying difficulty to implement, would have improved user experience overall.
+		\begin{itemize}
+			\item Being able to continue exploration after interrupting Simics to work with the debug prompt, instead of having to start over.
+			\item Support for multiprocessor kernels.
+			\item Cutting down overall simulation time.
+		\end{itemize}
+\end{itemize}
+
+Students also described compatibility issues they faced between their kernels and Landslide, which required them to make certain changes to their kernels.
+% FIXME: Say how we will address these issues?
+
+\begin{itemize}
+	\item {\bf Special-case context switch behaviour.} One kernel had a special context-switcher for the case when a thread was vanishing, which had to be special-cased in the instrumentation.
+		Some kernels had different context switch designs for just-forked threads (as discussed in the ``Thread Creation'' bullet in Section~\ref{sec:challenges-design}), and Landslide's code had to be modified during the study to support different designs.
+	\item {\bf Unexpected compiler optimisations.} Listing scheduler functions in \texttt{config.landslide} could be confusing if the compiler inlined them.
+	\item {\bf Driver thread interfering with test lifecycle tracking.} One kernel (Pathos, the 15-410 reference kernel) had to be modified to ensure the keyboard thread ran once before the test started, which was not mandated by the kernel design.
+(If the thread didn't run, it would conflict with the population-tracking described in Section~\ref{sec:components-test}, and Landslide would not be able to detect when the test was over.)
+	\item {\bf Kernel behaviour requirement violations.}
+		The majority of groups needed to change their scheduling around the idle thread in order to meet the requirement that the kernel never run idle when progress could otherwise be made (Sections~\ref{sec:challenges-design} and \ref{sec:using-requirements-sched}).
+		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the provided allocator.
+\end{itemize}
+
+Additionally, two enterprising users\footnote{The second case here was an ex-TA, not a student.} found ways to avoid implementing the functions in \texttt{student.c} (Section~\ref{sec:using-student-c}).
+It is worth considering integrating these approaches into the recommended approach to eliminate the need for \texttt{student.c} entirely.
+\begin{itemize}
+	\item One studnt bypassed the \texttt{kern\_ready\_for\_timer\_interrupt} annotation by disabling their scheduler's preemption-disabling mechanism and replacing it with explicit disabling of interrupts, thereby making every timer interrupt cause a context switch.
+	\item One ex-TA was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
+\end{itemize}
+
+\subsubsection{Landslide Victories}
+% TODO
+% nadim - wait bug (claimed easy to spot with code review)
+% margaret/echo - ds assertion bug
+% margaret/echo - vanish c_s return idle bug; helped find when idle was running when shouldn't be
+% tpassaro found double free, but deterministic
+% apodolsk found no bugs because no hello world
+% pjumde's group found no bugs..?
 
@@
 
-% TODO: intro text goes here
+While building Landslide, we worked primarily with two kernels: one, the kernel the author wrote as a student in 15-410 (Fall 2008), and another, a kernel the author graded as a TA for 15-410 (Fall 2011).
+We refer to the first kernel as ``POBBLES'' and the second as ``LudicrOS''.
+
+With Landslide, we found 4 bugs in POBBLES and 2 bugs in LudicrOS, and investigated them more deeply.
+
+% TODO
 
@@
 ``$\infty$'' indicates the bug was not found with stress testing (after 1 hour), or that Landslide's search did not finish (after 8 hours).
-% TODO: after how long?
-what timeout?
+% TODO: what timeout for stress testing?
 	}
@@
 Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
-The experimental set-up is as follows:
+
+The Landslide experimental set-up is as follows:
 
@@
 \end{itemize}
+
+%The stress test experimental set-up is as follows:
+%
+%\begin{itemize}
+%	\item A wrapper program runs 1024 simultaneous copies of the same test program that Landslide uses. Whenever one of them exits, the wrapper starts a new one.
+%	\item All stress test trials are on on the ``412 lab crashbox'' (3.2 GHz Pentium D).
+%	\item The kernel and test case are configured to not print any messages during the test
+%\end{itemize}
 

### 2012-05-01 evaluation.tex
add dag-mining and tsan to related; some minor updates

@@
 To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project, and had them use Landslide with their own kernels.
-
-In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.
-The lecture slides are available at \url{http://www.cs.cmu.edu/~410-s12/lectures/L30_Landslide.pdf}.\footnote{
-Also available at \url{http://bblum.net/landslide-lecture.pdf} and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-lecture.pdf}.}
+In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.\footnote{
+The lecture slides are available at \url{http://www.cs.cmu.edu/~410-s12/lectures/L30_Landslide.pdf}, and at \url{http://bblum.net/landslide-lecture.pdf}, and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-lecture.pdf}.}
 

### 2012-05-01 related.tex
add todo in related
buncha minor edits
add dag-mining and tsan to related; some minor updates

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% FIXME: Give another pass for dad's feedback - another sentence about each thing saying what landslide does better
 
@@
 
+Data race detection is the canonical runtime concurrency verification technique.
+Tools such as ThreadSanitizer\cite{tsan} track synchronisation operations and happens-before relationships to identify unsafe concurrent memory accesses.
+Such accesses, called ``data races'', might not necessarily cause incorrect behaviour, but are usually warning indicators of such, and represent bad coding practice in any case.
+
 Symbolic execution\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpretating symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
 Tools such as KLEE\cite{klee} and projects such as Automated Exploit Generation\cite{aeg} demonstrate the effectiveness of symbolic execution in dynamic verification.
-We discuss the possibility of integrating symbolic execution into a testing framework such as Landslide in Section~\ref{sec:future-new}.
+
+There has also been recent research in searching for which conditions, among a known set of conditions associated with a buggy behaviour, are actually necessary to cause bugs to arise\cite{dag-mining}.
+Such strategies are helpful because even once a bug is exposed, it can be difficult to determine what specifically went wrong.
+
+We discuss the possibility of integrating each of these orthogonal techniques into a testing framework such as Landslide in Section~\ref{sec:future-new}.
 

### 2012-05-01 titlepage.tex
add title page and convert to book format

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{\bf Landslide: \\
  Systematic Dynamic Race Detection in Kernel-space}
\author{Ben Blum}
\date{May 2012}
\Year{2012}
\trnumber{CMU-CS-12-118}

\committee{
Garth Gibson, Chair \\
Dave Eckhardt
}

\support{This research was sponsored by the U.S. Army Research Office under
grant number W911NF0910273.}

\disclaimer{The views and conclusions contained in
this document are those of the author and should not be interpreted as
representing the official policies, either expressed or implied, of
any sponsoring institution, the U.S. government or any other entity.}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{FOO, BAR, Computers, Science}

\maketitle

\begin{dedication}
%There is a wondrous fractal nature to all aspects of the universe.

At the time of writing, my life is focused entirely on this one project. There is a vast amount of thought in this thesis, and many more rich worlds lie hidden in each of the open avenues for future work.

Yet it pales in comparison to the scope of a Ph.D. thesis, and even more so next to all the details of a person's entire life.
After I finish this project (whatever that may mean), it will become but a fond memory, a mere building block in a city of even grander things.

Human nature is beautiful in its ability to exist at any level of perspective, to hold in awareness a story of any size, and to zoom in and out at will to see how multiple parts fit together.
Tearing ourselves up over minutiae is part of the fun, and contemplating overarching directions beyond our ability to change is also part of the fun.

Our true home, though, lies somewhere in between; a place of potential, and of peace. May we always remember to return to it.
\end{dedication}

\pagestyle{plain} % for toc, was empty

### 2012-05-01 titlepage.tex
update blurb and keywords

@@
 
-\keywords{FOO, BAR, Computers, Science}
+\keywords{concurrency, verification, kernels}
 
@@
 \begin{dedication}
-%There is a wondrous fractal nature to all aspects of the universe.
+There is a fractal nature to all aspects of life.
 
-At the time of writing, my life is focused entirely on this one project.
-There is a vast amount of thought in this thesis, and many more rich worlds lie hidden in each of the open avenues for future work.
+At the time of writing, I am focused entirely on this one project. There is a vast amount of thought in this thesis, and many more rich worlds lie hidden in each of the open avenues for future work.
 
 Yet it pales in comparison to the scope of a Ph.D. thesis, and even more so next to all the details of a person's entire life.
-After I finish this project (whatever that may mean), it will become but a fond memory, a mere building block in a city of even grander things.
+After I finish this project, it will become but a fond memory, a mere building block in a city of even grander things.
 

### 2012-05-01 challenges.tex
buncha minor edits
add title page and convert to book format
move alpaca 1 page
add a ton of future work

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Challenges of Kernel-space}
+\chapter{Challenges of Kernel-space}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{figure*}[h]
-	\begin{center}
-	\includegraphics[width=0.75\textwidth]{vip1066720.jpg}
-	\end{center}
-	\caption{{\em !`Cuidado! !`Las llamas son muy peligrosas!}\cite{de0u}}
-\end{figure*}
-
-\subsection{Causes of Concurrency}
+\section{Causes of Concurrency}
 
@@
 In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts.
-In future work (Section~\ref{sec:future-linux}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
+In future work (Section~\ref{sec:future-drivers}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
 
-\subsection{Ad-hoc Thread Communication}
+\begin{figure*}[t]
+	\begin{center}
+	\includegraphics[width=0.75\textwidth]{vip1066720.jpg}
+	\end{center}
+	\caption{{\em !`Cuidado! !`Las llamas son muy peligrosas!}\cite{de0u}}
+\end{figure*}
+
+\section{Ad-hoc Thread Communication}
 
@@
 
-\subsection{Kernel Design}
+\section{Kernel Design}
+\label{sec:challenges-design}
 

### 2012-05-01 futurework.tex
add dag-mining and tsan to related; some minor updates
add title page and convert to book format
add a ton of future work

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work}
+\chapter{Future Work}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Interface Improvements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Interface Improvements}
 \label{sec:future-interface}
@@
 If the user finds a bug, and wants to re-execute the interleaving that led to it, they have no choice but to re-explore the tree.
-	\item Landslide currently does not support being interrupted, so the user can use the Simics debug prompt arbitrarily and resume Landslide when ready.
-This is because the current implementation uses a wrapper script around the Simics prompt which Landslide communicates with to perform backtracking; it would be difficult to redesign, but could also be worked around.
+	\item Although Simics can pause execution and give the user a debug prompt, Landslide currently does not support being interrupted during exploration.
+		This is because the current implementation uses a wrapper script around the Simics prompt which Landslide communicates with to perform backtracking.
+		Though it would be difficult to redesign, it would be good to work around, to enable the user to interrupt, use the Simics debug prompt arbitrarily, and resume Landslide when ready.
 	\item Landslide could print its decision traces in a better format.
@@
 Raw text is probably the wrong interface; a graphical/tabular format with distinct columns for the execution of each different thread would be easier to understand.
+	\item Landslide could attempt to detect when something has gone wrong that is not its fault (such as incorrect instrumentation, or the kernel violating some of its requirements), and print a useful error message rather than behaving mysteriously.
 \end{itemize}
 
-\subsection{Education}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Education}
 \label{sec:future-education}
-% teaching tool, study thinking patterns
-% 410 stuff
-% TODO
+
+\subsection{Landslide as a Teaching Tool}
 
@@
 
-15-410 currently teaches students concurrency debugging skills ``the hard way'': by immersing them in environments where races will arise, and letting them find debugging tactics to use in conjunction with conventional stress testing on their own.
+15-410 currently teaches students concurrency debugging skills ``the hard way'': by immersing them in environments where races will arise, and letting them find debugging tactics on their own to use in conjunction with conventional stress testing.
 
-\subsection{New Techniques}
+We believe that a tool such as Landslide could change the way students learn such skills for the better, beyond simply increasing their likelihood of finding bugs during testing.
+The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving''.
+Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.
+
+\subsection{Landslide as a Grading Tool}
+
+The 15-410 grading infrastructure currently uses a program called Fritz, which is a stress testing wrapper around a suite of test programs.
+Some of these tests are Landslide-friendly (Section~\ref{sec:using-landslide-friendly-tests}), and some are themselves stress tests.
+The 15-410 course staff also hand-grade every kernel, in part because Fritz is known for only catching race conditions at random and infrequently.
+
+While far from being able to replace talented humans for grading, we believe that, after some improvement, Landslide could augment or replace Fritz as a tool for automatically finding several common patterns of concurrency errors that students have (especially races involving exiting threads, which we have shown Landslide is effective at finding).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{New Techniques}
 \label{sec:future-new}
@@
 
-\subsubsection{Data Race Detection and Static Analysis}
+\subsection{Data Race Detection}
 \label{sec:future-analysis}
@@
 
-There may also exist static analysis techniques that would identify the most "at-risk" memory accesses, which could also be used in this manner.
-
-\subsubsection{Parallelism}
+\subsection{Parallelism}
 
@@
 
-\subsubsection{Symbolic Execution}
-% TODO
-symbolic execution\cite{klee,dawson}
+\subsection{Symbolic Execution}
 
-\subsubsection{Trace Minimisation}
-% TODO
-% \ref{sec:future-backwards}
+In Section~\ref{sec:related-orthogonal}, we mentioned the technique of symbolic execution for dynamic verification.
+We believe that symbolic execution could be combined with systematic explorationto help find races in certain obscure code paths that systematic exploration by itself might not even execute.
 
-\subsection{Production Kernels}
+One interesting phenomenon that would occur from combining symbolic execution and systematic exploration would be a new notion of decision points.
+Symbolic execution involves its own state space exploration, in which control flow branch points correspond to decision points; hence, to combine it with systematic exploration would create a hybrid state space in which a ``decision point'' could be a place where either different threads could preempt each other or multiple control flow paths could be taken\cite{dawson}.
+
+\subsection{Trace Minimisation}
+\label{sec:future-trace-minimisation}
+
+Also in Section~\ref{sec:related-orthogonal}, we mentioned related research that attempts to {\em minimise} the set of conditions that are known to be associated with a bug\cite{dag-mining}.
+In the context of Landslide, this corresponds to how easy the decision trace is to understand.
+Even if Landslide finds a bug and prints the list of thread switches that were made, it may still be very difficult for the user to understand if the list includes many unnecessary switches that were completely unrelated to the bug.
+
+Landslide could continue to be of use even after it has found a bug, by attempting to find a ``minimal decision trace''.
+It could search different subsets of the known-buggy trace in an attempt to find shorter traces which result in the same bug.
+In this way, it would be able to print decision traces that automatically give the programmer much more insight into the true nature of each race condition.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Linux}
 \label{sec:future-linux}
-% linux, device drivers
-% open-coded sync dances: UGH (ad-hoc yield loops would make landslide get stuck)
-% embedded microcontroller operating systems, simulated
+
+The Linux kernel is a logical next kernel architecture to target with Landslide.
+
+One advantage of targetting Linux is that a testing framework will be able to make assumptions about the scheduler design, in ways that Landslide was not able to (Section~\ref{sec:challenges-design}), because the kernel instrumentation will not need to be repeatedly re-implemented (as it did for each 15-410 student kernel that we tested).
+
+However, Linux is a much more complicated kernel architecture than Pebbles, and we would face several challenges.
+
+\subsection{Multi-Processor Support}
+
+Landslide's way of modelling concurrency as a tree of thread interleavings directly expresses the way concurrency arises on uniprocessor systems.
+On multiprocessor systems such as Linux, however, there is ``true concurrency'': different threads may be running at the same time on different processors, rather than just interleaved.
+Hence, race conditions that were previously impossible in a uniprocessor environment may be commonplace in SMP.
+
+In order to support multiprocessor kernels, Landslide would need to refine its representation of the decision tree to express the potential for threads interleaving either on the same processor or on different processors.
+As a starting point, we note that on SMP, at each decision point, in addition to choosing any of $N$ threads to run, we must choose from among $M$ processors for it to run on.
 
 \subsection{Performance}
+
+Linux is much bigger than a Pebbles kernel (in terms of amount of code that runs during startup, during scheduling, etc), and hence would be much more expensive to test in a simulated environment.
+We suspect that in order to practically test Linux, we would need to implement systematic exploration in a virtualised environment, which we discuss more in Section~\ref{sec:future-perf}.
+
+\subsection{Complicated Synchronisation Patterns}
+
+Many components of Linux contain ad-hoc synchronisation which would be difficult to reason about.
+One common pattern is attempting some operation, checking later if it got interfered with, and rewinding and trying again if so.
+A straightforward approach to building a decision tree representing this might result in an infinitely deep branch: if the thread that gets interfered with keeps getting selected for scheduling, it would never make any progress.
+The testing tool would need to recognise these cases, and understand the invisible dependency on ``somebody else'' running first.
+
+\subsection{Device Drivers}
+\label{sec:future-drivers}
+
+There is much more device driver code in Linux than there is for its core components, and the device driver code also does not undergo as rigorous code review.
+Because of this, device drivers are much more prone to concurrency errors than the core components, and hence a systematic exploration framework for Linux should focus on testing them.
+
+However, the concurrency model of device drivers is much more complicated than the simple timer-based scheduling model we used in this work.
+In addition to timer-driven preemptions, code may also be interleaved as caused by device interrupts and input.
+Like support for SMP, being able to systematically test for device driver races will also require a more sophisticated representation of the decision tree.
+
+In addition to controlling timer interrupts, a systematic testing tool would also need to control device interrupts and device input, both of which are also nondeterministic inputs to the kernel.
+Furthermore, it would need to recognise common components of device drivers to properly cause relevant scheduling patterns, such as the ``top half'' and ``bottom half'' of interrupt handlers and dedicated device driver threads.
+
+These components would all be candidates for being scheduled at a decision point, although it more complicated still than simply choosing from among several threads to run.
+For example, the bottom half of an interrupt handler would depend on a top half having already executed, and interrupt handlers would run on the same stack as other thread rather than on their own stack.
+
+% FIXME
+%\subsection{Other Kernel Environments}
+
+%Some other production kernels would also make good targets for systematic exploration.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Performance}
 \label{sec:future-perf}
+% TODO
 % VM instead of simics
@@
 
-\subsection{Long-Running Testing Approaches}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Long-Running Test Shaping}
 \label{sec:future-shaping}
-% TODO more
-\footnote{Garth coined the name ``shaping'' to refer to this testing approach.}
-\subsubsection{Iterating Decision Sets}
 
+Another area to investigate is that of systems which could be tested continuously.
+Instead of the goal of finishing exploring a test case without finding a bug, a long-running testing approach would focus on ``shaping'' the test configuration to more and more refined configurations in hopes of finding a bug eventually.
+
+%\subsection{Iterating Decision Sets}
 In Section~\ref{sec:discussion-strategies}, we recommend an iteration strategy to effectively explore trees with multiple different sets of decision points, while not knowing ahead of time which, if any, may contain a bug.
@@
 
-\subsection{Theoretical Oddities}
+% FIXME
+%In Section~\ref{Sec:future-nadim}, we discuss an interesting property of certain decision trees, and suggest a heuristic
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Theoretical Oddities}
 \label{sec:future-theory}
-% theory things to study:
+
+During our in-depth case studies, we noticed two interesting properties about certain decision trees.
+We believe these warrant further theoretical study, to better understand the nature of these decision trees.
+This may in turn give rise to heuristics for making systematic testing more effective.
 
 % TODO: make visualisations for both of these
-\subsubsection{``Backwards'' Exploration}
+\subsection{``Backwards'' Exploration}
 \label{sec:future-backwards}
 % - study the implications of backwards vs forward exploration
+% \ref{sec:future-trace-minimisation}
 % - ICB as in chess
 
-\subsubsection{Exploration Tree Structure}
+\subsection{Exploration Tree Structure}
+\label{sec:future-nadim}
 % - nadim's backtrack depth

### 2012-05-01 evaluation.tex
add title page and convert to book format
random shit

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Evaluation}
+\chapter{Evaluation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{User Experience}
+\section{User Experience}
 
@@
 
-\subsubsection{Time Breakdown}
+\subsection{Time Breakdown}
 
@@
 
-\subsubsection{Descriptive Feedback}
+\subsection{Descriptive Feedback}
 \label{sec:eval-feedback}
@@
 
-The following feedback pertains to the user interface.
-For Landslide to be appealing as a debugging tool, we should improve the user interface (or the infrastructure, in some cases) to address these issues.
+\subsubsection{User Interface Feedback}
+
+For Landslide to be appealing as a debugging tool, we should improve the user interface (or the infrastructure, in some cases) to address the following issues.
 
@@
 \end{itemize}
+
+\subsubsection{Technical Feedback}
 
@@
 
-\subsubsection{Landslide Victories}
+\subsection{Landslide Victories}
 % TODO
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Bug Case Studies}
+\section{Bug Case Studies}
 \label{sec:eval-casestudy}
 
-\subsubsection{Bug Descriptions}
+\subsection{Bug Descriptions}
 
@@
 
-\subsubsection{Numbers}
+\subsection{Numbers}
 \label{sec:eval-numbers}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Summary of Bugs Found}
+\section{Summary of Bugs Found}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Discussion}
+\section{Discussion}
 
-\subsubsection{Invariants}
+\subsection{Invariants}
 
@@
 
-\subsubsection{Recommended Testing Strategies}
+\subsection{Recommended Testing Strategies}
 \label{sec:discussion-strategies}

### 2012-05-01 terminology.tex
add title page and convert to book format
random shit

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Terminology}
+\chapter{Terminology}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In this section we define certain terms that have specific meanings in the context of Landslide.
+In this chapter we define certain terms that have specific meanings in the context of Landslide.
 
-\subsection{Basic Terms}
+\section{Basic Terms}
 
@@
 		Abstractly, the set of inputs under which the guest kernel is tested.
-Practically, a user-space program which runs on top of the guest kernel to execute a particular set of system calls, while Landslide performs systematic exploration
+Practically, a user-space program which runs on top of the guest kernel to execute a particular set of system calls, while Landslide performs systematic exploration.
 \suspend{enumerate}
 
-\subsection{Scheduling Terms}
+\section{Scheduling Terms}
 
@@
 
-\subsection{Systematic Exploration Terms}
+\section{Systematic Exploration Terms}
 

### 2012-05-02 design.tex
fix spacing in design alg
footnotesize alg in design
buncha minor edits
add title page and convert to book format
add a ton more design

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Design and Implementation}
+\chapter{Design and Implementation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Components of Landslide}
 
-Here we briefly describe each major component of Landslide. % TODO: provide a figure landslide.png
+\section{Components of Landslide}
 
-\subsubsection{Kernel Instrumentation}
+In this section we briefly describe each major component of Landslide. % TODO: provide a figure landslide.png
+
+\subsection{Kernel Instrumentation}
 \label{sec:components-kern}
@@
 
-\subsubsection{Scheduling}
+\subsection{Scheduling}
 \label{sec:components-sched}
@@
 
-\subsubsection{Memory Tracking}
+\subsection{Memory Tracking}
 \label{sec:components-memory}
@@
 
-\subsubsection{The Arbiter}
+\subsection{The Arbiter}
 \label{sec:components-arbiter}
@@
 
-\subsubsection{The Explorer}
+\subsection{The Explorer}
 \label{sec:components-explore}
@@
 
-\subsubsection{Test Lifecycle}
+\subsection{Test Lifecycle}
 \label{sec:components-test}
-% Talk about the test lifecycle - pre-readline; pre-test-start; etc.
-% Talk about the population tricks.
-% Talk about how we want idle/shell's wait() to be part of the test.
 
-\subsection{On Simics}
+\begin{figure}[t]
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{teststate.png}
+	\end{center}
+	\caption{The test lifecycle state machine.}
+	\label{fig:teststate}
+\end{figure}
+Landslide knows the state of the test case through a simple state machine, which is updated with information about the number of threads currently on the system, the number of threads on the runqueue, the state of the idle thread (if it exists), and whether or not the shell is waiting for keyboard input.
+
+The state machine relies on certain information provided by the scheduler (Section~\ref{sec:components-sched}).
+
+\begin{itemize}
+	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
+		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process could have yet to reap a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well.
+For these cases an additional check is needed.
+	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.\footnote{
+		If the kernel is idling and the shell is not waiting for keyboard input, it is likely that the test case has deadlocked.}
+	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
+\end{itemize}
+
+The states in the test lifecycle, and the rules for transitioning between them are as follows.
+
+\begin{enumerate}
+	\item {\bf Bootup.} Initial state. When the kernel is idling and the shell is waiting for keyboard input, advance state.
+	\item {\bf Keyboard input.} Record the current thread count (the ``start population''). Generate keystrokes to cause the desired test case to run.
+		When the current thread count is greater than the start population, advance state.
+	\item {\bf Test running.} Begin constructing the decision tree here (since it is pointless to backtrack to before the test begins).
+When the current thread count equals the start population, advance state.
+	\item {\bf Test cleanup.} The test processes have exited, but some housekeeping may yet remain. When the the kernel is idling and the shell is waiting for keyboard input, advance state.
+	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack.
+\end{enumerate}
+
+Figure~\ref{fig:teststate} depicts the states and transitions visually.
+
+\subsection{Stack Tracing}
+% TODO: decide whether to do this or not
+
+\section{On Simics}
 
@@
 
-\subsection{Thread Scheduling}
+\section{Thread Scheduling}
 
-\begin{algorithm*}[t!]
-	\label{alg:inflight}
+\begin{algorithm}[t]
+	\footnotesize
 	\begin{algorithmic}
@@
 The predicates on \texttt{pc} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
-\end{algorithm*}
+	\label{alg:inflight}
+\end{algorithm}
 
@@
 
-\subsection{Identifying Bugs}
+\section{Identifying Bugs}
 \label{sec:techniques-bugs}
@@
 
-\subsubsection{Accurate Bug-Detection Conditions}
+\subsection{Accurate Bug-Detection Conditions}
 
@@
 
-\subsubsection{Uncertain Bug-Detection Conditions}
+\subsection{Uncertain Bug-Detection Conditions}
 
@@
 
-\subsection{Partial-Order Reduction}
+\section{Partial-Order Reduction}
 \label{sec:techniques-por}
 
-We make use of Dynamic Partial-Order Reduction, the state-space pruning algorithm presented in \cite{dpor}. Here we discuss the specifics of implementing DPOR in our environment.
+We make use of Dynamic Partial-Order Reduction (DPOR), the state-space pruning algorithm presented in \cite{dpor}.
+DPOR requires two sets to be computed that describe the concurrency relationship between transitions: the happens-before relation, and the memory independence relation.
+Here we discuss the specifics of implementing these in our environment.
 
-\subsubsection{Happens-Before Relation}
-% TODO
+\subsection{Happens-Before Relation}
 
-\subsubsection{Memory Independence Relation}
+In order to establish a happens-before relation between transitions, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point.
+
+Unfortunately, a thread's presence or absence on the scheduler runqueue does not necessarily correspond to whether it is {\em runnable} at any point.
+We identify three exceptions, as foreshadowed in Section~\ref{sec:challenges-design}:
+
+\begin{itemize}
+	\item {\bf Current thread not on runqueue.} If the currently-running thread is not stored on the runqueue, we identify it as runnable anyway (except in the special case of idle, as described below).
+	\item {\bf Idle thread.} Some kernels may have an explicit idle thread and store it on the runqueue, with explicit code to skip over it if other threads are runnable.
+Hence, if the idle thread is on the runqueue (and/or the current thread), it is only runnable if no other threads are runnable.
+	\item {\bf Yielding mutexes.} In kernels whose mutexes (or other synchronisation idioms) leave threads on the runqueue when they are notionally blocked, such threads are not runnable.
+\end{itemize}
+
+Using this notion of runnability, the happens-before relation is established as follows.
+Let $X$ be a transition and $Y$ be a subsequent transition in a branch that was just explored, and $T_Y$ be the thread associated with $Y$.
+We say that $X$ {\em enables} $Y$ if $T_Y$ was not runnable immediately before $X$, but was runnable immediately after.
+Then, a transition $A$ happens-before a transition $B$ if $A$ is the same thread as $B$, if $A$ enables $B$, or if $A$ happens-before some transition $C$ that enables $B$.
+
+\subsection{Memory Independence Relation}
 \label{sec:por-independence}
 The independence relation expresses which transitions do not read-and-write or write-and-write to the same shared memory addresses.
+Landslide computes this relation using the set of shared memory accesses described in Section~\ref{sec:components-memory}.
+We encountered two challenges pertaining to memory independence:
 
 \begin{itemize}
-	\item {\bf Always-accessed memory locations.} % talk about ignoring runqueue accesses, etc TODO
+	\item {\bf Always-accessed memory locations.}
+		In kernel-space, every thread switch goes through common scheduler and context-switcher routines. These routines inevitably access scheduler data structures, such as the runqueue.\footnote{
+		Another example is the timer tick counter: this global variable is not necessarily accessed every transition, because voluntary reschedules do not involve timer interrupts, but every involuntary preemption will cause a write to this counter.}
+		If we include such memory accesses in the independence relation, it will result that all transitions conflict, and DPOR won't be able to achieve any reduction.
+To allow for state space reduction, we sacrifice our ability to find races involving these particular accesses by ignoring them.
+Currently we require the user to identify these locations as part of the instrumentation process, as described in Section~\ref{sec:using-config-landslide}.
 	\item {\bf Freed memory poisoning.}
@@
 
-\subsubsection{Coarse Granularity}
-% talk about the all-ancestors vs induction thing TODO
+\subsection{Coarse Granularity}
 
-\subsection{Debugging Feedback}
+In \cite{dpor}, the ``Avoiding Stack Traversals'' section discusses a potential optimisation in which only the first ``evil ancestor'' (defined in \cite{dbug-retreat}) of each transition need be considered when identifying which alternate interleavings need to be explored.
 
-% TODO: provide a figure
+\begin{figure}[h]
+	\centering
+	\begin{tabular}{cc}
+	\subfloat[Race-free interleaving.]{
+		\footnotesize
+		\begin{tabular}{l|c|c|c|}
+			\cline{2-4}
+			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
+			\cline{2-4}
+			$C$ & & & \texttt{x = 42;} \\
+			$B$ & & \texttt{y = 7;} & \\
+			\multirow{2}{*}{$A_1$} & \texttt{if (y == 7)} & & \\
+			& \texttt{assert(x == 42);} & & \\
+			\cline{2-4}
+		\end{tabular}
+	} &
+	\subfloat[Alternate interleaving explored with DPOR.]{
+		\footnotesize
+		\begin{tabular}{l|c|c|c|}
+			\cline{2-4}
+			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
+			\cline{2-4}
+			$C$ & & & \texttt{x = 42;} \\
+			$A_2$ & \texttt{if (y == 7)} & & \\
+			$B$ & & \texttt{y = 7;} & \\
+			\cline{2-4}
+		\end{tabular}
+	} \\
+	\multicolumn{2}{c}{
+	\subfloat[Buggy interleaving.]{
+		\footnotesize
+		\begin{tabular}{l|c|c|c|}
+			\cline{2-4}
+			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
+			\cline{2-4}
+			$B$ & & \texttt{y = 7;} & \\
+			\multirow{2}{*}{$A_3$} & \texttt{if (y == 7)} & & \\
+			& \texttt{assert(x == 42);} & & \\
+			$C$ & & & \texttt{x = 42;} \\
+			\cline{2-4}
+		\end{tabular}
+	}}
+	\end{tabular}
+	\caption{This sample code defeats the ``Avoiding Stack Traversals'' optimisation if a decision point is not defined between the two lines of thread 1's code.
+The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next.
+In the second interleaving, $C$ is no longer a second evil ancestor of $A_2$, so the third interleaving is never explored, and the assertion failure is missed.}
+	\label{fig:evil-ancestors}
+\end{figure}
+
+We found that this optimisation was not sound in our setup, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
+Because Landslide builds its decision trees with coarser granularity, i.e.
+multiple different shared memory conflicts may happen in each transition, Landslide must consider every evil ancestor of each transition when performing DPOR.
+
+Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of this optimisation if transitions are allowed to have multiple different memory conflicts.
+
+\section{Debugging Feedback}
+
+\begin{figure}[t]
+	\begin{center}
+		\includegraphics[width=0.95\textwidth]{found_a_bug.png}
+	\end{center}
+	\caption{A sample decision trace (with extraneous text trimmed) that Landslide generated using the \texttt{double\_thread\_fork} test case (Section~\ref{sec:using-tests}).}
+	\label{fig:found_a_bug}
+\end{figure}
 
@@
 When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
+Figure~\ref{fig:found_a_bug} depicts a simple decision trace for a use-after-free bug.
 

### 2012-05-02 design.tex
write about the scheduler

@@
 
-% TODO
+The scheduler is responsible for keeping track of which threads exist in the guest kernel: which are runnable at any given time, and when they are created and destroyed.
+
+It maintains a ``mirror image'' of the guest kernel's scheduler state in the form of three queues, a pointer to the currently-running thread, and a pointer to the previously-running thread.
+The queues are the {\em runqueue}, containing the runnable threads, the {\em sleep queue}, containing threads which become runnable after a certain number of timer ticks\footnote{
+Landslide treats the sleep queue no differently from the runqueue when deciding which threads are runnable.
+The guest kernel will treat it differently, but since sleeping for a predetermined amount of time is never an appropriate way to solve race conditions, Landslide treats sleeping threads as notionally runnable.
+This fits directly into its model that runnable threads are ones that can be caused to run with a finite number of timer ticks in succession.},
+and the {\em deschedule queue}, which might not correspond to one data structure in the guest kernel, but contains all other threads that exist on the system, which are not runnable for whatever reason.
+
+The scheduler also tracks which important actions each thread is performing.
+These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing, and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
+There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}
+
+\subsubsection{Thread Lifecycle Tracking}
+
+\begin{algorithm}[t]
+	\footnotesize
+	\begin{algorithmic}
+		\State {\em Per-thread state.}
+		\State bool forking, sleeping, vanishing, in\_timer\_handler;
+		\State int current\_tid;
+		\\
+		\Function{handle\_tell\_landslide\_forking}{}
+			\State forking $\gets$ true;
+		\EndFunction
+		\\
+		\Function{handle\_tell\_landslide\_sleeping}{}
+			\State sleeping $\gets$ true;
+		\EndFunction
+		\\
+		\Function{handle\_tell\_landslide\_vanishing}{}
+			\State vanishing $\gets$ true;
+		\EndFunction
+		\\
+		\Function{handle\_tell\_landslide\_thread\_switch}{int new\_tid}
+			\If {!in\_timer\_handler}
+				\If{sleeping}
+					\State {\sc add\_to\_sleep\_queue}(current\_thread);
+					\State sleeping $\gets$ false;
+				\ElsIf{vanishing}
+					\State {\sc destroy\_thread}(current\_tid);
+					\State vanishing $\gets$ false;
+				\ElsIf{forking}
+					\State {\sc create\_thread}(new\_tid);
+					\State forking $\gets$ false;
+				\EndIf
+			\EndIf
+			\State {\sc update\_current\_thread}(new\_tid);
+		\EndFunction
+	\end{algorithmic}
+	\caption{The scheduler's routines for tracking thread lifecycles. These routines are invoked each time the guest kernel calls one of the \texttt{tell\_landslide} annotations.}
+	\label{alg:tell-landslide}
+\end{algorithm}
+
+The scheduler relies on the guest kernel informing it about certain important events in the thread lifecycle.
+The guest kernel does this by means of annotations (which are described in Section~\ref{sec:using-annotations}).
+Algorithm~\ref{alg:tell-landslide} shows how landslide handles these annotations internally.\footnote{
+There are also cases for \texttt{tell\_landslide\_on\_rq} and \texttt{tell\_landslide\_off\_rq}, which we omit for brevity.
+One important note is that the routine for \texttt{tell\_landslide\_on\_rq} must also check the ``forking'' flag, in case the kernel's \texttt{fork} implementation does not immediately switch to the new thread but instead adds it to the runqueue for later.}
+
+\subsubsection{Thread Scheduling}
+\label{sec:components-inflight}
+
+Finally, the scheduler is responsible for managing involuntary preemptions, causing arbitrary threads to begin running in place of the (as chosen at decision points and the ends of branches by the explorer, Section~\ref{sec:components-explore}).
+
+\begin{algorithm}[t]
+	\footnotesize
+	\begin{algorithmic}
+		\State {\em Global scheduler state.}
+		\State bool schedule\_in\_flight;
+		\State int target\_tid;
+		\\
+		\State {\em Per-thread state. (Updated elsewhere.)}
+		\State bool in\_timer\_handler, in\_context\_switch;
+		\State int current\_tid;
+		\\
+		\Function{scheduler\_update}{int pc}
+			\If{schedule\_in\_flight}
+				\State {\sc assert}(in\_timer\_handler || in\_context\_switch);
+				\If { {\sc kernel\_exiting\_timer}(pc) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(pc))}
+					\State {\em The kernel has just finished rescheduling and is about to resume normal thread execution.}
+					\If {current\_tid != target\_tid}
+						\State {\em The kernel switched to an undesirable thread. Keep the schedule operation ``in-flight''.}
+						\State {\sc cause\_timer\_interrupt}();
+					\Else
+						\State {\em The in-flight schedule is ``landing''.}
+						\State schedule\_in\_flight $\gets$ false;
+					\EndIf
+				\EndIf
+			\ElsIf{{\sc need\_to\_preempt}()}
+				\State target\_tid $\gets$ {\sc choose\_new\_thread}();
+				\State schedule\_in\_flight $\gets$ true;
+				\State {\sc cause\_timer\_interrupt}();
+			\EndIf
+		\EndFunction
+	\end{algorithmic}
+	\caption{The scheduling algorithm.
+This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{pc} (the program counter) each time.
+The predicates on \texttt{pc} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
+	\label{alg:inflight}
+\end{algorithm}
+
+In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+
+Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
+Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.
+
+One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
+This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
+We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
 
@@
 
-\section{Thread Scheduling}
-
-\begin{algorithm}[t]
-	\footnotesize
-	\begin{algorithmic}
-		\State {\em Global scheduler state.}
-		\State bool schedule\_in\_flight;
-		\State int target\_tid;
-		\\
-		\State {\em Per-thread state. (Updated elsewhere.)}
-		\State bool in\_timer\_handler, in\_context\_switch;
-		\State int current\_tid;
-		\\
-		\Function{scheduler\_update}{int pc}
-			\If{schedule\_in\_flight}
-				\State {\sc assert}(in\_timer\_handler || in\_context\_switch);
-				\If { {\sc kernel\_exiting\_timer}(pc) || (!handling\_timer \&\& {\sc kernel\_exiting\_context\_switch}(pc))}
-					\State {\em The kernel has just finished rescheduling and is about to resume normal thread execution.}
-					\If {current\_tid != target\_tid}
-						\State {\em The kernel switched to an undesirable thread. Keep the schedule operation ``in-flight''.}
-						\State {\sc cause\_timer\_interrupt}();
-					\Else
-						\State {\em The in-flight schedule is ``landing''.}
-						\State schedule\_in\_flight $\gets$ false;
-					\EndIf
-				\EndIf
-			\ElsIf{{\sc need\_to\_preempt}()}
-				\State target\_tid $\gets$ {\sc choose\_new\_thread}();
-				\State schedule\_in\_flight $\gets$ true;
-				\State {\sc cause\_timer\_interrupt}();
-			\EndIf
-		\EndFunction
-	\end{algorithmic}
-	\caption{The scheduling algorithm.
-This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{pc} (the program counter) each time.
-The predicates on \texttt{pc} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
-	\label{alg:inflight}
-\end{algorithm}
-
-In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
-
-Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
-Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.
-
-One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
-This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
-We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
-
 \section{Identifying Bugs}
@@
 
+% TODO: fix this
 In \cite{dpor}, the ``Avoiding Stack Traversals'' section discusses a potential optimisation in which only the first ``evil ancestor'' (defined in \cite{dbug-retreat}) of each transition need be considered when identifying which alternate interleavings need to be explored.

### 2012-05-02 design.tex
write kernel instrumentation section

@@
 
-In this section we briefly describe each major component of Landslide. % TODO: provide a figure landslide.png
+In this section we briefly describe each major component of Landslide.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.75\textwidth]{landslide.png}
+	\caption{A simple visualisation of Landslide and its components, and how they interact with the guest kernel.}
+	\label{fig:landslide}
+\end{figure}
 
@@
 \label{sec:components-kern}
-% TODO
+
+The kernel instrumentation serves as the glue between what the guest kernel is doing and Landslide's understanding of the guest kernel's state.
+
+\begin{itemize}
+	\item {\bf User-provided instrumentation.} Some parts of the kernel may be written in any number of ways, and hence require the user's assistance for Landslide to understand.
+The user-provided instrumentation, described in detail in Section~\ref{sec:using-instrumenting}, informs Landslide about thread-related lifecycle and scheduling events and the anatomy of the kernel's scheduler.
+	\item {\bf Automatic instrumentation.} All Pebbles kernels have some things in common, and Landslide's build system is able to automatically instrument certain parts of the kernel.
+This instrumentation informs Landslide about the dynamic memory allocator, common library functions such as \texttt{panic}, and certain aspects of the kernel's executable format.
+\end{itemize}
 

### 2012-05-02 using.tex
add using.tex
minor adds
text-complete challenges
add tables to eval
buncha minor edits
add title page and convert to book format
add a ton of future work
add a ton more design
write kernel instrumentation section
write using chapter

--- 2012-04-24 related.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work}
+\chapter{Using Landslide}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:using}
 
-% vim: ft=tex
+In this chapter we discuss how to use Landslide with a kernel that meets 15-410's Pebbles specification.
+We list some design requirements that a Pebbles kernel must meet, document the instrumentation process and how to customise Landslide's search behaviour, and tell how to interpret Landslide's results.
+
+In preparation for conducting the student user study (Section~\ref{sec:eval-studence}), we distributed a user guide roughly similar to this chapter.\footnote{
+The user guide itself is available at \url{http://bblum.net/landslide-guide.pdf}, and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide.pdf}.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Kernel Requirements}
+\label{sec:using-requirements}
+
+\subsection{Scheduler Functionality}
+\label{sec:using-requirements-sched}
+
+In order to cause desired preemptions, Landslide needs to assume that the core of the scheduler works.
+In short, this means that timer interrupts will trigger context switches, and that if a thread is ``runnable'', at any point where interrupts/preemption is enabled, a finite number of timer preemptions will eventually cause that thread to run.
+
+For deadlock detection, if the kernel has mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:using-annotations}.
+If mutexes explicitly deschedule blocking threads, no special annotations are needed.
+
+Landslide does not support kernels that spin-wait anywhere, whether in mutexes or otherwise, especially when waiting for keyboard input or in \texttt{sleep}.
+Relatedly, the kernel must never run the idle loop when not truly idle, because Landslide uses this as a way of telling when all threads are wedged and/or when the test is done running.
+
+\subsection{VM}
+
+Landslide does not have much to do with the VM, but the kernel must direct-map most of kernel memory, including the heap, globals, and kernel thread stacks.
+Landslide needs this to read values out of memory and to identify memory conflicts.\footnote{
+It would be simple to support more advanced virtual memory set-ups, in which the same virtual address has different physical memory mappings at different times, but this is not implemented.}
+
+The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation.
+
+\subsection{System Calls}
+
+In order to run with Landslide, the kernel must be able to boot up to the shell prompt\footnote{
+In Pebbles, the bootup sequence is usually fairly simple, involving the init process spawning the shell.},
+receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes.
+The kernel must have the following system calls implemented and working in at least a rudimentary way:
+\texttt{fork}, \texttt{exec}, \texttt{vanish}, \texttt{wait}, \texttt{readline}.
+
+The following system calls are exercised in some, but not all, of the test cases we distribute, so are helpful to have, but not necessary:
+\texttt{thread\_fork}, \texttt{yield}.
+
+Apart from the system calls which init and shell execute, the rest are irrelevant and do not need to be implemented to use Landslide on a Pebbles kernel.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Instrumenting Kernels with Landslide}
+\label{sec:using-instrumenting}
+
+Here we describe the interface by which Landslide understands the execution of the guest kernel.
+Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.
+
+\subsection{In-Kernel Annotations}
+\label{sec:using-annotations}
+
+We provide a set of functions that the kernel needs to call at certain points during its execution to communicate what interesting events are happening.
+
+\subsubsection{Important Note about Runqueues}
+\label{sec:using-runqueue}
+Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the scheduler's runqueue at each point during execution.
+Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not the kernel keeps the current thread on the runqueue or off of it.
+
+If the kernel does not keep the currently-running thread on the runqueue, the user must also use \texttt{kern\_current\_extra\_runnable} (Section~\ref{sec:using-student-c}).
+
+\subsubsection{Required Annotations}
+\begin{itemize}
+	\small
+	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in the context switcher, when a new thread is switched to.
+	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when the scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations.
+	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
+		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)
+
+		Call it ``just before'' the action which makes the new thread runnable - whether the kernel just adds it to the runqueue (in which case the next annotation called would be \texttt{thread\_on\_rq}) or begins running it immediately (in which case \texttt{thread\_switch}); Landslide handles both cases.
+	\item \texttt{tell\_landslide\_vanishing()} - Call this when a thread is about to go away.
+		Call it ``just before'' the relevant context switch; i.e., the one that should never return.
+	\item \texttt{tell\_landslide\_sleeping()} - Call this when a thread is about to go to sleep (that's \texttt{sleep()} the system call).
+		Call it ``just before'' the relevant context switch; i.e., the one that won't return for as long as the thread requested to sleep.
+	\item \texttt{tell\_landslide\_thread\_on\_rq(int tid)} - Call this when a thread is about to be added to the runqueue.
+		(Make sure this call is done within whatever protection is also used for the runqueue modification itself.)
+	\item \texttt{tell\_landslide\_thread\_off\_rq(int tid)} - Call this when a thread is about to be removed from the runqueue. (Same protection clause as above applies.)
+	\item Mutex annotations - These are only necessary if the kernel uses mutexes that leave blocked threads on the runqueue, rather than explicitly descheduling them.
+	\begin{itemize}
+		\item \texttt{tell\_landslide\_mutex\_locking(void *mutex\_addr)} - Call this when a thread is about to start locking a mutex.
+Landslide needs the mutex address for deadlock detection, to track which mutex each thread is blocked on.
+		\item \texttt{tell\_landslide\_mutex\_locking\_done()} - Call this when a thread finishes locking a mutex (and now owns it).
+		\item \texttt{tell\_landslide\_mutex\_unlocking(void *mutex\_addr)} - Call this when a thread is about to start unlocking a mutex.
+		\item \texttt{tell\_landslide\_mutex\_unlocking\_done()} - Call this when a thread finishes unlocking a mutex (and no longer owns it).
+		\item \texttt{tell\_landslide\_mutex\_blocking(int owner\_tid)} - Call this when a thread is contending on a locked mutex, and hence is no longer logically ``runnable'' yet still on the runqueue.
+Landslide needs to know who owns the mutex for deadlock detection.
+	\end{itemize}
+\end{itemize}
+
+\subsubsection{Optional Annotations}
+\begin{itemize}
+	\small
+	\item \texttt{tell\_landslide\_decide()} - Call this to define additional decision points. See section~\ref{sec:using-decision}.
+	\item \texttt{assert()} or \texttt{panic()} - You should be using these already, but they are extra useful in Landslide.
+		The more important invariants for which a kernel has \texttt{assert}s, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
+\end{itemize}
+
+\subsection{Configuration File}
+\label{sec:using-config-landslide}
+
+There is also a config file the user needs to fill out, called \texttt{config.landslide}.
+Some of the fields are required information, and some are tweaks that change the way Landslide behaves.
+We omit discussion of the required fields, and explain the optimisations and behaviour tweaks that Landslide supports in Section~\ref{sec:using-search}.
+
+\subsection{Instrumenting Within Landslide}
+\label{sec:using-student-c}
+
+There are two functions that the user needs to implement in Landslide itself, in a file called \texttt{student.c}, to express certain kernel designs that might be more complicated than a simple true/false condition or integer.
+
+\begin{itemize}
+	\small
+        \item \texttt{bool kern\_current\_extra\_runnable(conf\_object\_t *cpu)} - See Section~\ref{sec:using-runqueue}.
+This says whether the current thread is ``logically runnable'' despite not being on the runqueue itself (that is, \texttt{tell\_landslide\_on\_rq} won't have been called for it).
+        \item \texttt{bool kern\_ready\_for\_timer\_interrupt(conf\_object\_t *cpu)} - This function can be used to express when the scheduler is ``locked'' in a way not involving disabled interrupts.
+Landslide will avoid trying to preempt the kernel whenever this is false.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Configuring Landslide's Behaviour}
+\label{sec:using-customise}
+\subsection{Decision Points}
+\label{sec:using-decision}
+
+Getting a good set of decision points is important to being able both to explore reasonably quickly and to produce meaningful interleavings that are likely to find bugs.
+After getting Landslide working with the default set of decision points (voluntary reschedules only), the next step is to add some more.
+We list here the general steps in this process.
+
+\begin{itemize}
+        \item Use \texttt{tell\_landslide\_decide()} to indicate decision points.\footnote{
+	We recommend at the start of \texttt{mutex\_lock} and at the end of \texttt{mutex\_unlock}, though it is also important to use one's own intuition, depending on what system call(s) are being tested.}
+        \item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.
+        \item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of choice points, and figure out which ones are irrelevant.
+        \item Use \texttt{within\_func} and \texttt{without\_func} to make Landslide only use relevant decision points.
+\end{itemize}
+
+\subsection{Search Parameters}
+\label{sec:using-search}
+
+Landslide's configuration file supports several options for changing the way it explores the decision tree.
+
+\subsubsection{Optimizations}
+
+These configuration options help Landslide identify potentially-conflicting shared memory accesses that it should ignore (Section~\ref{sec:por-independence}).
+
+\begin{itemize}
+	\small
+	\item \texttt{sched\_func} - When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them.
+Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.
+	\item \texttt{ignore\_sym} - The user may wish to ignore memory accesses from other global data structures as well.
+For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that better reflects what is actually being tested for.
+	\item \texttt{within\_function} - If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{
+For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.}
+This configuration option allows the user to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.
+	\item \texttt{without\_function} - Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
+\end{itemize}
+
+\subsubsection{Behaviour Tweaks}
+\begin{itemize}
+	\small
+	\item \texttt{EXPLORE\_BACKWARDS} - In which order to explore the branches?
+Backwards means with more preemptions first; forwards means tending to let the current thread keep running more often.
+Backwards tends to find bugs more quickly, but produces longer decision traces.
+(Section~\ref{sec:future-backwards})
+	\item \texttt{DECISION\_INFO\_ONLY} - This option makes Landslide stop after one branch, and output the list of decision points it identified using the current config.
+Useful to see all decision points, and tweak settings to get rid of frivolous ones.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Test Cases}
+\label{sec:using-tests}
+
+We ship Landslide with a suite of small test cases designed to expose several common races that students frequently encounter during 15-410 Project 3.
+We list their code in Appendix~\ref{sec:test-code}.
+
+\begin{itemize}
+        \item \texttt{vanish\_vanish} - Tests when a parent and child process \texttt{vanish()} simultaneously.
+        \item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child.
+        \item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in one process vanishing.
+        \item \texttt{yield\_vanish} - Tests for interactions between \texttt{yield()} and \texttt{vanish()}.
+\end{itemize}
+
+\subsection{Landslide-Friendly Test Characteristics}
+\label{sec:using-landslide-friendly-tests}
+
+It is important to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state-space.
+They also run several, but not too many, threads at once, producing potentially interesting interleavings.

### 2012-05-03 evaluation.tex
write using chapter
write summaries of bugs

@@
 \section{User Experience}
+\label{sec:eval-studence}
 
@@
 	\item Customising Landslide's search parameters such as decision points (Section~\ref{sec:using-customise}).
-	\item Analysing the output from bugs that Landslide found (Section~\ref{sec:using-interpreting}).
+	\item Analysing the output from bugs that Landslide found.
 \end{enumerate}
@@
 
+This section will be written after the draft is submitted. Hang on tight, my friends.
 % TODO
@@
 \subsection{Landslide Victories}
-% TODO
-% nadim - wait bug (claimed easy to spot with code review)
-% margaret/echo - ds assertion bug
-% margaret/echo - vanish c_s return idle bug; helped find when idle was running when shouldn't be
-% tpassaro found double free, but deterministic
-% apodolsk found no bugs because no hello world
-% pjumde's group found no bugs..?
+
+Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree.
+Of those groups, two found race conditions in their kernel.
+Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
+
+We list how much progress each group made, and whether or not they found any bugs.
+
+\begin{enumerate}
+	\item % ntaha/jmchow
+		Group 1 got Landslide to explore a minimal decision tree, and also explored \texttt{double\_wait} and \texttt{vanish\_vanish} with several combinations of different decision points around \texttt{mutex\_lock} and \texttt{mutex\_unlock}.
+They found one race and a problem with idle.
+		After fixing this bug, they were able to completely explore both test cases with decision points on both \texttt{mutex\_lock} and \texttt{mutex\_unlock}, which we consider a strong negative result (i.e., no bug found with fine-grained interleavings).
+	\item % margaret/echo
+		Group 2 got Landslide to explore a minimal decision tree.
+They did not configure any extra decision points.
+They found a problem with idle, and after fixing it, found a vanish-related race condition in the fix.
+They also found a bug in an assertion which they added during the process, but that bug was deterministic, so we consider this part of the ``fixing incorrect instrumentation'' process.
+	\item % tpassaro/?
+		Group 3 got Landslide to explore a minimal decision tree, and also configured extra decision points for \texttt{vanish\_vanish}. They found only a deterministic use-after-free bug.
+	\item % pjumde/?
+		Group 4 got Landslide to explore a minimal decision tree. They found only a problem with idle.
+	\item % apodolsk
+		Group 5 did not get Landslide to explore a minimal decision tree.
+\end{enumerate}
+
+\subsubsection{Race Conditions Found}
+
+Two groups found race conditions using Landslide. We describe them here.
+
+\begin{itemize}
+	\item {\bf Too many waiters allowed.} Using the \texttt{double\_wait} test case, group 1 found a bug in which more threads invoking \texttt{wait} would be allowed to block than the number of child processes that could ever be reaped, and some of the waiters would end up blocking forever.
+The fix for this was two lines of code, and it could have been found with an in-depth code review.
+	\item {\bf Accidental return to a vanished thread.} Group 2 had a problem with the scheduler running idle when it was not supposed to; when they fixed it, however, their implementation would return to a thread that had already vanished in the case when no thread was left to run.
+It is unclear if this race would have been exposed without Landslide.
+Fixing it required adding an extra condition to the code that checked whether the kernel should idle.
+\end{itemize}
+
+% TODO confirm this
+% We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a  [...]
+% This bug is similar to the LudicrOS vanish/vanish bug studied in Section~\ref{sec:eval-casestudy}, and would have required a complicated redesign of the \texttt{vanish} implementation.
+
+% We consider this a mixed blessing: Landslide was capable of finding a deep concurrency bug in a student kernel, but the instrumentation process was expensive enough that they were not motivated to continue working.
+We believe that improving Landslide's interface (Section~\ref{sec:future-interface}) will improve the success rate in such cases.
+
+\subsubsection{Deterministic Bugs Found}
+
+We also claim that Landslide is useful in its ability to find certain types of non-concurrency-related bugs.
+Though its primary purpose is to find race conditions, we found that it caught two main types of bugs that the students were previously unaware their kernels had.
+
+\begin{itemize}
+	\item {\bf Inappropriate idling.} Two groups had errors in their keyboard handling code, where when a newline was received when the kernel was idle, a thread blocked on \texttt{readline} would be added to the runqueue and but not switched directly to, so the kernel would continue idling even if there was useful work to do.
+Another group left idle on the scheduler runqueue, and had no special-case code to skip over it, so would interleave idle arbitrarily with other threads.
+We consider both of these performance problems.
+	\item {\bf Deterministic use-after-free.} One group had a deterministic use-after-free bug in their \texttt{wait} implementation, in which they freed something and immediately dereferenced it.
+This hadn't presented a problem because the block was never reallocated during the critical window, which would have taken a complicated and fine-grained interleaving, but Landslide detected the use-after-free immediately, without the need for any interleaving.
+\end{itemize}
 
@@
 
-\subsection{Bug Descriptions}
-
 While building Landslide, we worked primarily with two kernels: one, the kernel the author wrote as a student in 15-410 (Fall 2008), and another, a kernel the author graded as a TA for 15-410 (Fall 2011).
@@
 
-With Landslide, we found 4 bugs in POBBLES and 2 bugs in LudicrOS, and investigated them more deeply.
+With Landslide, we found four bugs in POBBLES and two bugs in LudicrOS, and investigated them more deeply.
 
-% TODO
+\subsection{Bug Descriptions}
 
-\subsection{Numbers}
+Here we describe each of the six bugs that we used as case studies. We rate their complexity and the estimated ease of fixing them.
+
+\subsubsection{POBBLES vanish/vanish (a)}
+
+We introduced this bug into POBBLES's vanish implementation.
+When a parent and child invoke vanish concurrently, the parent needs to modify the child's parent pointer to reparent it to the init process, and the child needs to read its parent pointer to signal to its parent that it can be reaped.
+
+In version (a) of this bug, we introduce a mutex for each process's parent pointer.
+When reparenting, the parent takes its own process control block (PCB) lock, iterates over its child-list, and takes its children's parent pointer locks.
+When signalling, the child takes its parent pointer lock, then takes the parent's PCB lock to ensure safe wake-up of waiters.
+This results in deadlock.
+
+This bug is exposed with the \texttt{vanish\_vanish} test case. It requires a very specific interleaving sequence to uncover.
+The fix is nontrivial, because the circular data structure access still needs to happen, but be protected in a way that doesn't involve circular wait.
+Vanish races are notorious among 15-410 course staff as being difficult to find with the naked eye, and also difficult to figure out how to fix.
+
+\subsubsection{POBBLES vanish/vanish (b)}
+
+We also introduced this bug into POBBLES's vanish.
+Version (b) is similar to version (a), except we removed the parent pointer mutex entirely.
+This results in a data race around the parent pointer, in which the child can end up accessing the wrong parent either before or after it is actually moved to init's child-list.
+
+This bug is exposed with the \texttt{vanish\_vanish} test case. The complexity and difficulty of fixing are the same as in version (a).
+
+\subsubsection{POBBLES wait/wait}
+
+Finding this bug was somewhat of a surprise.
+When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{cond\_queue\_next} pointer.
+However, when woken, the condition variable code did not clear the value of this pointer.
+Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
+If not for this assert, extremely mysterious behaviour could have arisen - if the woken thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
+Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}
+
+This bug is exposed with the \texttt{double\_wait} test case.
+The complexity of this bug is moderate, since it requires both waiting threads to block before the child process exits.
+The fix is simple; in two places in the condition variable code a pointer needed to be set to \texttt{NULL}.
+
+We note that the author never knew of its existence between submitting the kernel as a student three years ago and uncovering it before the user study three weeks ago, and it was not spotted by the TA who graded it.\footnote{
+We would also note that this bug was never uncovered by any stress test, but the \texttt{double\_wait} test case was not distributed as part of the 15-410 test suite, and it was necessary to generate the right sequence of system calls.}
+
+\subsubsection{POBBLES thread\_fork/wait}
+\label{sec:eval-thread-fork}
+
+This bug was presented in the lecture, and exhibited as a decision tree in Figure~\ref{fig:threadfork} and as a trace in Figure~\ref{fig:found_a_bug} and as source code in Figure~\ref{fig:tell-landslide}.
+When a new thread is forked and made runnable, and the forking thread subsequently dereferences its TID to use as a return value, it's possible for the dereference to be a use-after-free.
+This could result in a garbage value being returned.
+
+This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple\footnote{
+Although many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the bug may never be noticed when using such libraries.
+(This caused the author some trouble.)},
+and the fix is only two lines long; the dereference should simply occur before making the child thread runnable.
+
+Nevertheless, this bug was present in the author's student kernel and was not noticed by the TA who graded it.
+
+\subsubsection{LudicrOS vanish/vanish}
+\label{sec:eval-ludicros-vanish}
+
+This bug was found by the author when grading this kernel.
+It took roughly an hour of manual inspection and hard thinking to ascertain the bug's nature, from the time when the author first suspected a bug might be present.
+
+The parent pointer problem is the same as in the POBBLES vanish/vanish bugs, except the locking pattern is slightly different.
+When the parent reparents children, it drops its own mutex while it changes the parent pointer using the child's parent pointer mutex, which happens before the parent moves the child to init's child-list.
+
+Many problems can arise from this error.
+The one Landslide discovered was as follows: During this gap, the child can take its own parent pointer mutex (after the pointer was changed to init), and attempt to move itself from init's ``live child-list'' to the ``dead child-list'' before the parent moved the child to init's child-list at all.
+This results in the child instead moving the {\em shell process} off of init's ``live child-list''.
+Init gets signalled, reaps the reparented child, does \texttt{wait} again (expecting the shell to be alive), which fails.
+This results in init infinitely looping around failing calls to \texttt{wait}.
+To catch this bug, we added an assert statement in the \texttt{wait} failure case that the current thread was not init (because init should always have children to reap or wait for), which Landslide then caused to trip.
+
+This was the most complicated bug we studied. The fix would be nontrivial, like in POBBLES vanish/vanish.
+
+\subsubsection{LudicrOS yield/vanish}
+
+This bug was also found during grading, and took substantially less time to verify by hand than the previous bug.
+When a thread does \texttt{yield} to a specific other TID, the other thread's thread control block (TCB)is not protected between lookup-by-TID and yield-to.
+If that thread exits during that time, a use-after-free (and worse, a completely invalid schedule to a nonexistent thread) results.
+
+This bug is moderately complicated, requiring a preemption during that specific window of the first thread's execution during which the second thread must exit entirely and have its TCB freed.
+The fix is moderately complicated - some protection mechanism must be designed so that threads cannot exit after being looked-up by TID.
+
+\subsection{Performance}
 \label{sec:eval-numbers}
+
+Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
 
@@
 
-Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
-
 The Landslide experimental set-up is as follows:
@@
 	\item All Landslide trials were run on the Gates-Hillman cluster machines (2.6 GHz Xeon).
-	% TODO: say specs of crash machine
 	\item All Landslide trials were run with ``backwards exploration'' enabled (Section~\ref{sec:using-search}).
@@
 %\end{itemize}
+% TODO: say specs of crash machine
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary of Bugs Found}
+%\section{Summary of Bugs Found}
 
-% TODO
+% FIXME
 

### 2012-05-03 design.tex
fix coarse granularity section

@@
 The guest kernel does this by means of annotations (which are described in Section~\ref{sec:using-annotations}).
-Algorithm~\ref{alg:tell-landslide} shows how landslide handles these annotations internally.\footnote{
+Algorithm~\ref{alg:tell-landslide} shows how Landslide handles these annotations internally.\footnote{
 There are also cases for \texttt{tell\_landslide\_on\_rq} and \texttt{tell\_landslide\_off\_rq}, which we omit for brevity.
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 	\begin{center}
@@
 
-\subsection{Stack Tracing}
-% TODO: decide whether to do this or not
+%\subsection{Stack Tracing}
+% FIXME
 
@@
 
-% TODO: fix this
-In \cite{dpor}, the ``Avoiding Stack Traversals'' section discusses a potential optimisation in which only the first ``evil ancestor'' (defined in \cite{dbug-retreat}) of each transition need be considered when identifying which alternate interleavings need to be explored.
+\cite{dpor} and \cite{dbug-retreat} present a partial order reduction algorithm in which only the first ``evil ancestor'' (term defined in \cite{dbug-retreat}) of each transition need be considered when identifying which alternate interleavings need to be explored.
 
@@
 	\end{tabular}
-	\caption{This sample code defeats the ``Avoiding Stack Traversals'' optimisation if a decision point is not defined between the two lines of thread 1's code.
+	\caption{This sample code defeats a DPOR implementation that only uses the first evil ancestor if a decision point is not defined between the two lines of thread 1's code.
 The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next.
@@
 
-We found that this optimisation was not sound in our setup, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
+We found that implementing DPOR in this way was not sound in our setup, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
 Because Landslide builds its decision trees with coarser granularity, i.e.
@@
 Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of this optimisation if transitions are allowed to have multiple different memory conflicts.
+
+While DPOR is a proved-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment.
+We justify the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave proving soundness to future work (and to the theoreticians).
 

### 2012-05-03 futurework.tex
write virt section

@@
 Linux is much bigger than a Pebbles kernel (in terms of amount of code that runs during startup, during scheduling, etc), and hence would be much more expensive to test in a simulated environment.
-We suspect that in order to practically test Linux, we would need to implement systematic exploration in a virtualised environment, which we discuss more in Section~\ref{sec:future-perf}.
+We suspect that in order to practically test Linux, we would need to implement systematic exploration in a virtualised environment, which we discuss more in Section~\ref{sec:future-virt}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Performance}
-\label{sec:future-perf}
-% TODO
-% VM instead of simics
-% - how to interpose? single-step, or unmap heap, and/or annotate (for hooks) with hypercalls
-% - how to time travel? instead of time travel, snapshot once and replay
+\section{Virtualisation}
+\label{sec:future-virt}
+
+Compared to executing the code being tested on real hardware (which tools such as \cite{dbug-ssv} are able to do for user-space programs), Simics is {\em slow}, and even slower with Landslide hooked up to it.
+
+Major benefit could be achieved by redesigning Landslide to work in a virtualised environment instead of a simulated one.
+This would pose several challenges, because Landslide would no longer be running in ``god mode'', having immediate access to every single instruction and memory access executed.
+It would, however, still be able to modify the kernel's memory and address space, which would be important to harness for increased control over the kernel's execution.
+
+We speculate on some challenges and potential solutions that virtualised execution might have to offer.
+We assume that our users will be talented kernel developers rather than undergraduate students, and hence allow for more complicated interface requirements.
+
+\subsection{Interposition}
+
+\subsubsection{Event Tracking}
+
+The event-based nature of the \texttt{tell\_landslide} annotations can still be harnessed in a virtualised environment - they would need to be converted into hypercalls, which would briefly transfer control to Landslide at key execution points to enable it to update its state machine and potentially influence the kernel's execution.
+
+We would add additional \texttt{tells\_landslide} for the kernel components which are presently automatically instrumented, such as the dynamic memory allocator and the \texttt{panic} routines, and also for certain components that are currently instrumented with \texttt{config.landslide}, such as identifying when the context switcher returns or when the timer interrupt handler is invoked.
+
+Rather than using the schedule-in-flight algorithm, Landslide could simplify its job by asking the kernel to run a particular thread directly, as described in Section~\ref{sec:components-inflight}.
+
+\subsubsection{Memory Tracking}
+
+More challenging than tracking important kernel events will be tracking all accesses to shared memory.
+We could exploit the virtualised virtual memory system, and write-protect (possibly also read-protect, i.e., unmap) all pages that contain shared data, such as the global data section and the dynamic allocation heap.
+All accesses would then page-fault, transferring control to Landslide.
+
+\subsection{Backtracking}
+
+Simics's convenient bookmarking mechanism will be no more in virtual-machine-land.
+One possibility for backtracking would be to record-and-replay the events of each branch (up to the point at which Landslide chooses to have decided differently), restarting the system's execution from the beginning each time.
+Another option might be to snapshot the state of the kernel at each decision point, and use those to resume without re-executing everything from the beginning (kernel boot-up and initialisation is much more expensive than individual system calls, after all).
+
+\subsection{Control over Non-Determinism}
+
+Finally, Landslide will need to find new ways to control non-deterministic events in the system.
+Since virtualised kernels use the same timer interrupt as the host they run on, Landslide is unlikely to be able to intercept it.
+Instead, we may replace a guest kernel's timer-driven context switcher with a ``Landslide-driven'' context switcher, leaving the timer handler only to do other work.
+
+Likewise, Landslide will need to control other non-deterministic inputs to the kernel, including device interrupts/input and even the system clock.
 

### 2012-05-03 evaluation.tex
add a footnote to eval

@@
 Of those groups, two found race conditions in their kernel.
-Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
+Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.\footnote{
+It is lamely ironic that only one-half of the ``Landslide victory'' pun actually applies.}
 
@@
 
-% TODO confirm this
-% We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a  [...]
-% This bug is similar to the LudicrOS vanish/vanish bug studied in Section~\ref{sec:eval-casestudy}, and would have required a complicated redesign of the \texttt{vanish} implementation.
+We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a subtle race in the reparenting section of their \texttt{vanish} code using the \texttt{vanish\_vanish} test.\footnote{
+We have not confirmed this with Landslide. We will before submitting this thesis. % TODO
+}
+This bug is similar to the LudicrOS vanish/vanish bug studied in Section~\ref{sec:eval-casestudy}, and would have required a complicated redesign of the \texttt{vanish} implementation.
 
-% We consider this a mixed blessing: Landslide was capable of finding a deep concurrency bug in a student kernel, but the instrumentation process was expensive enough that they were not motivated to continue working.
+We consider this a mixed blessing: Landslide was capable of finding a deep concurrency bug in a student kernel, but the instrumentation process was expensive enough that they were not motivated to continue working.
 We believe that improving Landslide's interface (Section~\ref{sec:future-interface}) will improve the success rate in such cases.

### 2012-05-03 futurework.tex
add note about later transitions

@@
 In this way, it would be able to print decision traces that automatically give the programmer much more insight into the true nature of each race condition.
+
+While studying various bugs found by Landslide, we found that in general, when there were frivolous thread switches in the decision trace, users most benefitted from starting to read the trace at the end; i.e., considering the most recent transition made by each thread.
+Hence, when building a search algorithm for trace minimisation, it would be good to more heavily prioritise decisions that occurred later in a buggy branch.
 

### 2012-05-03 conclusion.tex
add title page and convert to book format
write a fake conclusion
improve conclusion

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\chapter{Conclusion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\s}{\fontfamily{jkpvos}\selectfont{}s}
+
+% TODO write this for real
+And lo, the Author did{\s}t pre{\s}ent Land{\s}lide to yon re{\s}earch Community, and to the {\s}tudents of XV-CDX as well.
+Verily, the Kernel developers young and olde found Bugges of Concurren{\s}y in their code.
+The {\s}oftware then{\s}eforth worked as Intended, cra{\s}hing nevermore, and there was much rejoicing.

### 2012-05-03 evaluation.tex
remove footnote evaluation

@@
 Of those groups, two found race conditions in their kernel.
-Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.\footnote{
-It is lamely ironic that only one-half of the ``Landslide victory'' pun actually applies.}
+Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
 

### 2012-05-03 evaluation.tex
confirm group 2's vanish bug

@@
 
-We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a subtle race in the reparenting section of their \texttt{vanish} code using the \texttt{vanish\_vanish} test.\footnote{
-We have not confirmed this with Landslide. We will before submitting this thesis. % TODO
-}
-This bug is similar to the LudicrOS vanish/vanish bug studied in Section~\ref{sec:eval-casestudy}, and would have required a complicated redesign of the \texttt{vanish} implementation.
+We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a subtle race in the reparenting section of their \texttt{vanish} code using the \texttt{vanish\_vanish} test.
+We verified this by having an ex-TA re-instrument their kernel and add the decision points: from the time of exploring a minimal decision tree to finding the bug with the new decision set took half an hour of work, and fully understanding the nature of the bug took an hour on top of that.
+
+This bug is similar to the LudicrOS vanish/vanish bug studied in Section~\ref{sec:eval-casestudy}, and would have required a complicated redesign of the \texttt{vanish} implementation to fix.
 

### 2012-05-03 design.tex
some copy editing

@@
 			\cline{2-4}
-			$C$ & & & \texttt{x = 42;} \\
-			$B$ & & \texttt{y = 7;} & \\
-			\multirow{2}{*}{$A_1$} & \texttt{if (y == 7)} & & \\
-			& \texttt{assert(x == 42);} & & \\
+			$C$ & & & \texttt{x=42;} \\
+			$B$ & & \texttt{y=31337;} & \\
+			\multirow{2}{*}{$A_1$} & \texttt{if (y==31337)} & & \\
+			& \texttt{assert(x==42);} & & \\
 			\cline{2-4}
@@
 			\cline{2-4}
-			$C$ & & & \texttt{x = 42;} \\
-			$A_2$ & \texttt{if (y == 7)} & & \\
-			$B$ & & \texttt{y = 7;} & \\
+			$C$ & & & \texttt{x=42;} \\
+			$A_2$ & \texttt{if (y==31337)} & & \\
+			$B$ & & \texttt{y=31337;} & \\
 			\cline{2-4}
@@
 			\cline{2-4}
-			$B$ & & \texttt{y = 7;} & \\
-			\multirow{2}{*}{$A_3$} & \texttt{if (y == 7)} & & \\
-			& \texttt{assert(x == 42);} & & \\
-			$C$ & & & \texttt{x = 42;} \\
+			$B$ & & \texttt{y=31337;} & \\
+			\multirow{2}{*}{$A_3$} & \texttt{if (y==31337)} & & \\
+			& \texttt{assert(x==42);} & & \\
+			$C$ & & & \texttt{x=42;} \\
 			\cline{2-4}
@@
 
-Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of this optimisation if transitions are allowed to have multiple different memory conflicts.
+Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of only identifying one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.
 

### 2012-05-03 evaluation.tex
some copy editing

@@
 
-We consider this a mixed blessing: Landslide was capable of finding a deep concurrency bug in a student kernel, but the instrumentation process was expensive enough that they were not motivated to continue working.
+We consider this a mixed blessing: Landslide was capable of finding a deep concurrency bug in a student kernel, but the instrumentation process was expensive enough that they were not motivated to continue working, and stopped just short.
 We believe that improving Landslide's interface (Section~\ref{sec:future-interface}) will improve the success rate in such cases.
@@
 Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
-If not for this assert, extremely mysterious behaviour could have arisen - if the woken thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
+Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing - if the woken thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
 Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}

### 2012-05-03 terminology.tex
write summaries of bugs
some copy editing

@@
 
-% TODO: have a diagram of the decision tree
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.75\textwidth]{threadfork.png}
+	% TODO: make some other subfigs
+	\caption{A simple decision tree, showing possible interleavings of a buggy \texttt{thread\_fork} implementation. (This describes a bug studied in Section~\ref{sec:eval-thread-fork}.)}
+	\label{fig:threadfork}
+\end{figure}
+
 \resume{enumerate}
@@
 At a given decision point, Landslide chooses which thread will run (and causes it to do so), and that thread executes a transition, after which point Landslide will have identified a subsequent decision point
-	\item {\bf Decision point set} (or {\bf Decision set}):
+	\item {\bf Decision set} (or {\bf Set of decision points}):
 		One or more predicates on the execution state of the guest kernel which identify whether any current state should be a decision point.

### 2012-05-03 using.tex
write summaries of bugs
some copy editing

@@
 
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{tell_landslide.png}
+	\caption{Example annotated code.
+In theory, the use of \texttt{tell\_landslide\_decide()} in the place where a preemption needs to occur is not necessary for Landslide to know to preempt then; it is shown for the sake of demonstration.}
+	\label{fig:tell-landslide}
+\end{figure}
+
 We provide a set of functions that the kernel needs to call at certain points during its execution to communicate what interesting events are happening.
+Figure~\ref{fig:tell-landslide} shows some example code annotated for Landslide.
 
@@
         \item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.
-        \item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of choice points, and figure out which ones are irrelevant.
+        \item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of decision points, and figure out which ones are irrelevant.
         \item Use \texttt{within\_func} and \texttt{without\_func} to make Landslide only use relevant decision points.

### 2012-05-03 testcases.tex
add testcases

\chapter{Code for Provided Test Cases}
\label{sec:test-code}

The test cases that use the \texttt{thread\_fork} system call are written in assembly, because thread creation can only safely be done from C using a userspace thread library. However, using a userspace thread library for thread creation might forcibly serialise certain operations that would expose kernel bugs only if concurrent, and also necessitates many more system calls, which dramatically increases the size of the decision tree.

\section{\texttt{vanish\_vanish.c}}
%\lstinputlisting[language=C]{vanish_vanish.c}
{\small
\begin{verbatim}
#include <syscall.h>

void main()
{
    fork();
    vanish(); /* parent and child process exit simultaneously */
}
\end{verbatim}
}
\section{\texttt{double\_wait.S}}
%\lstinputlisting[language={[x86masm]Assembler}]{double_wait.S}
{\small
\begin{verbatim}
#include <syscall_int.h>

.global main
main:
    int $FORK_INT
    cmp $0x0,%eax
    jne parent
    int $VANISH_INT
parent:
    int $THREAD_FORK_INT
    mov $0x0,%esi
    int $WAIT_INT        # two parent threads do wait(NULL);
    int $VANISH_INT
\end{verbatim}
}
\section{\texttt{yield\_vanish.S}}
%\lstinputlisting[language={[x86masm]Assembler}]{yield_vanish.S}
{\small
\begin{verbatim}
#include <syscall_int.h>

.global main
main:
    int $THREAD_FORK_INT
    cmp $0x0,%eax
    jne parent
    int $VANISH_INT
parent:
    mov %eax,%esi
    int $YIELD_INT       # yield(child_tid);
    int $VANISH_INT
\end{verbatim}
}
\section{\texttt{double\_thread\_fork.S}}
%\lstinputlisting[language={[x86masm]Assembler}]{double_thread_fork.S}
{\small
\begin{verbatim}
#include <syscall_int.h>

.global main
main:
    int $THREAD_FORK_INT # fork 1st thread
    cmp $0x0,%eax
    je first_child
parent:
    int $VANISH_INT
first_child:
    int $THREAD_FORK_INT # fork 2nd thread
    int $VANISH_INT      # both threads vanish
\end{verbatim}
}

### 2012-05-03 futurework.tex
add todo in future
write oddities

@@
 
+% TODO talk about heuristic-guided tree construction
+
 % FIXME
@@
 \label{sec:future-backwards}
-% - study the implications of backwards vs forward exploration
-% \ref{sec:future-trace-minimisation}
-% - ICB as in chess
+
+We configured Landslide to be able to explore trees in two different orderings, and we found that the two orderings behaved substantially differently in terms of time taken to find bugs and length of decision traces produced.
+
+\begin{itemize}
+	\item {\bf Forwards exploration.} When exploring the tree ``forwards'', at each decision point the explorer would choose the currently-running thread if it hadn't already been explored.
+This resulted in preference for branches of the tree with fewer forced preemptions; the first branch to be explored would have no preemptions at all, and subsequent branches would have generally (but not monotonically) increasing numbers of preemptions.
+
+		Compared to backwards exploration, we found that forwards exploration tended to produce much shorter decision traces for the same bugs.
+These shorter traces would have fewer thread switches that were irrelevant to the bug itself, and hence be much easier to understand.
+	\item {\bf Backwards exploration.} When exploring ``backwards'', at each decision point the explorer would prefer to choose a thread that required a forced preemption.
+This resulted in preference for branches with more forced preemptions, and the first branch to be explored would have a preemption at every decision point.\footnote{
+		This property of backwards exploration is what resulted in the LudicrOS yield/vanish bug being found without any backtracks.}
+
+		Compared to forwards exploration, we found that backwards exploration tended to find bugs much more quickly, with fewer backtracks needed before the buggy branch was chosen.
+\end{itemize}
+% TODO: provide results here
+% FIXME: study diffs in trees with no bugs at all
+
+Compared to Iterative Context Bounding\cite{chess}, which is based on the insight that bugs need fewer forced preemptions to expose, we believe that our insight is orthogonal.
+The ICB insight results in smaller decision trees overall needing to be explored before bugs appear, while our insight determines exploration ordering once the tree is constructed.
+
+We are not sure if combining these two insights is possible.
+In future work we might compare them and investigate this possibility, to determine a heuristic for what exploration orderings are most likely to expose bugs with the fewest needed preemptions.
 
@@
 \label{sec:future-nadim}
-% - nadim's backtrack depth
+
+One member of group 1 from the user study fully explored decision trees from \texttt{double\_wait} and \texttt{vanish\_vanish} (Section~\ref{sec:eval-victory}), and found an oddity in the exploration time for a particular decision set.
+
+When using decision points on just \texttt{mutex\_lock} or \texttt{mutex\_unlock}, the explorations for each test case took similar time.
+However, when using decision points on both mutex functions, \texttt{double\_wait} took substantially longer, despite the tree structure suggesting they should have been comparable, with \texttt{double\_wait} even being perhaps slightly faster.
+Figure~\ref{fig:nadim-stats} shows a comparison of the tree anatomy for these two test cases.
+
+\begin{figure}[h]
+	\small
+	\centering
+	\begin{tabular}{|l||c|c|}
+		\hline
+		& \texttt{double\_wait} & \texttt{vanish\_vanish} \\
+		\hline\hline
+		Exploration time & {\bf 30 minutes} & {\bf 5 minutes} \\
+		\hline
+		Backtracks & {\bf 1,511} & {\bf 1,401} \\
+		\hline
+		Decision points & 10,355 & 5,690 \\
+		\hline
+		Average branch depth & 18.34 & 16.00 \\
+		\hline
+		Average instructions/branch& 1,975,033 & 1,984,249 \\
+		\hline
+		Total instructions & 206,757,601 & 24,907,516 \\
+		\hline
+		Backtrack distance minimum & {\bf 6} & {\bf 2} \\
+		\hline
+		Backtrack distance maximum & 17 & 16 \\
+		\hline
+		Backtrack distance average & {\bf 7.38}  & {\bf 3.75} \\
+		\hline
+		% FIXME: add stddev?
+	\end{tabular}
+	\caption{Comparison of expensive \texttt{double\_wait} test and cheap \texttt{vanish\_vanish} test.
+The number of branches and depth of each branch are similar, but \texttt{double\_wait}'s backtracks were significantly longer, causing more total instructions to be executed.}
+	\label{fig:nadim-stats}
+\end{figure}
+
+Despite the two trees having nearly equal numbers of branches and instructions executed per branch, \texttt{double\_wait} took significantly more time.
+This is because Landslide's backtracks during \texttt{double\_wait} were significantly {\em longer} than the backtracks in \texttt{vanish\_vanish} on average; i.e., the backtracks were targetting an earlier point during the test's execution.
+Hence, Landslide had to execute more total instructions overall to explore the longer branches in \texttt{double\_wait}.\footnote{
+Of note, if backtracking had been implemented by restarting the test each time; i.e., executing the full depth of every branch from the root of the tree, the repeated work would have caused the execution times to be equal.
+The fact that Simics's backtracking avoids the cost of execution that happened before the backtrack target is to credit for this discrepancy.}
+% as visualised in figure\ldots TODO
+
+This suggests that the ``interesting'' parts of \texttt{double\_wait} were earlier in the test case, whereas in \texttt{vanish\_vanish} they are more towards the end.
+We conjecture that, with theoretical study of these structural properties, we could develop search heuristics for focusing on the ``interesting'' parts of the trees, and avoid repeating the irrelevant instructions at the end of each branch.

### 2012-05-08 defence/defence.tex
add defence slides prelim

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{ulem}
\usepackage{marvosym}
\usepackage{skull}
\usepackage{proof}
\usepackage{array}
\setbeamertemplate{navigation symbols}{}

\title[Landslide]{{\bf Landslide: Systematic Dynamic Race Detection in Kernel-space}}
\subtitle[]{ {\em showing that Pebbles are less stable than you thought since 2011.}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[CMU]{Carnegie Mellon University}
\date[]{2012, May 10}

\setbeamertemplate{footline}{\hspace*{.5cm}\scriptsize{\insertauthor\hspace*{50pt} \hfill\insertframenumber\hspace*{.5cm}}}

\usecolortheme{seahorse}
\usecolortheme{rose}
\useoutertheme{infolines}

\usecolortheme[named=ForestGreen]{structure}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\begin{document}
\normalem
\begin{frame}
	\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\linegap{\vspace{0.2in}}
\newcommand\breakslide[1]{\begin{frame}{} \begin{center} #1 \end{center} \end{frame}}
\newcommand\related[1]{\textsuperscript{\em #1}}

% slide: outline
\begin{frame}{Outline}
	\textbf{Motivation: Race condition debugging}
	\begin{itemize}
		\item Systematic exploration versus stress testing
		\item Challenges of kernel-space
	\end{itemize}
	{\bf Landslide: Making it happen}
	\begin{itemize}
		\item Design and interface 
		\item Addressing challenges
	\end{itemize}
	{\bf Evaluation}
	\begin{itemize}
		\item Student user study
		\item Case study
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivation}

\subsection{Race Conditions and the Finding Thereof}

\begin{frame}[fragile]{Case Study}
	\begin{center}
	\begin{verbatim}
	int thread_fork()
	{
	    thread_t *child = spawn_new_thread();
	    add_to_runqueue(child);
	    return child->tid;
	}
	\end{verbatim}
	\end{center}

	\begin{itemize}
		\item On vanish, child's TCB is freed
		\item Forking thread does use-after-free
		\item Might return garbage instead of TID
	\end{itemize}
\end{frame}

\begin{frame}{Testing Techniques} % slide: sys ex vs stress testing
	\textbf{Stress Testing}: Common testing approach
	\begin{itemize}
		\item Attempting to exercise as many interleavings as possible
		\item Exposes race conditions at random
		\begin{itemize}
			\item ``If a preemption occurs at {\em just the right time}\dots''
		\end{itemize}
		\item Cryptic panic messages or machine reboots
	\end{itemize}
	\linegap

	{\bf Systematic Exploration}\related{Flanagan '05}
	\begin{itemize}
		\item Make educated guesses about when to preempt
		\item Run {\em every single} interleaving
		\item Provide better debugging information
	\end{itemize}
\end{frame}

\begin{frame}{Execution Tree} % slide: decision tree
	\begin{center}
		\includegraphics[width=\textwidth]{tree.png}
	\end{center}
\end{frame}

% slide: 15-410 6-week project why we need it in kernels
\begin{frame}{Finding Races in Kernel-Space}
	% cite datacollider etc?
\end{frame}

\subsection{Challenges of Kernel-space}

\begin{frame}{Causes of Concurrency} % slide: causes of concurrency, thread scheduling
	\textbf{Kernels contain their own concurrency implementation.}

	\linegap
	Context switcher
	\begin{itemize}
		\item Interleaving threads is no longer ``instantaneous''.
	\end{itemize}
	Runqueue tracking
	\begin{itemize}
		\item Which threads are runnable?
	\end{itemize}
	Thread lifecycle tracking
	\begin{itemize}
		\item When do threads created/destroyed?
	\end{itemize}
\end{frame}

% slide: ad-hoc communication, memory conflicts (freeing, runqueue)
\begin{frame}{Ad-Hoc Thread Communication}
	No message-passing % TODO this slide sucks

	User-space systematic exploration relies on it\related{Simsa '10}; we cannot.

	Kernel threads share state via globals, heap.
\end{frame}

\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
	\includegraphics[width=\textwidth]{pebbles.png}
	% Say: "we don't want to have to explore every possible interaction between each of these components"
\end{frame}

\begin{frame}{Contribution} % slide: summary: "sys ex is a tool \ldots``
	Systematic exploration is a {\em tool} for testing concurrent systems.
	\begin{itemize}
		\item How can it be applied in kernel-space?
		\item What should/must we ask of the user to make it possible?
	\end{itemize}
	\linegap

	{\bf Landslide} is an effort to answer these questions.
	% say: landslide is an exploratory work into finding out what lies behind these questions.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Landslide}

% slide: outline

\subsection{Anatomy}
% slide: pebbles, 15-410, simics
% slide: visualisation (here address challenge #1)

\subsection{Design}
% slide: using -- annotations & in-landslide instrumentation
% slide: POR -- mostly invisible, but (transition)

\subsection{Addressing Challenges} % rename
% slide: POR, modified - with runqueue conflicts, will get no pruning
% slide: ignoring shared memory accesses - venn diagram
% slide: other_prob: decide on mutex_lock as opposed to everywhere, but...
% slide: manual shaping - have *user* specify what to focus on (automating intuition)
% slide: exciting teaser "i think this is not too far off!"

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation}

% slide: outline

\subsection{User Study}

\begin{frame}{Working with Students} % slide: worked with students - give some time breakdown
	Solicited students in 15-410 to use Landslide on their kernels.
	\begin{itemize}
		\item Can Landslide find bugs ``in the wild''?
		\item How much time does it take to get it working?
	\end{itemize}
	Five groups participated; four got it to work.
	\begin{itemize}
		\item Average instrumentation time 100 minutes (min 50 max 150)
		\item Average customisation time 25 minutes (min 0 max 60)
	\end{itemize}
\end{frame}

\begin{frame}{Found Bugs} % slide: anecdotes: two deterministic bugs & two races
	All four groups who got it to work found bugs.

	Race conditions
	\begin{itemize}
		\item Too many waiters allowed
		\begin{itemize}
			\item One parent thread might block forever
		\end{itemize}
		\item Accidental return to an exited thread
		% Also talk about their vanish/vanish bug
	\end{itemize}
	Deterministic bugs
	\begin{itemize}
		\item Inappropriate idling
		\begin{itemize}
			\item Waiting for a timer tick with no useful work to do
		\end{itemize}
		\item Use-after-free
		\begin{itemize}
			\item Required memory tracking to detect
		\end{itemize}
	\end{itemize}
\end{frame}

\subsection{Case Study}

\begin{frame}{Studying Bugs In-Depth} % slide: "I have built a software smarter than me!" 30-40sec to find
	Worked with two kernels: my own, and one I graded

	``I have made something smarter than me.''
	\begin{itemize}
		\item \texttt{double\_wait} - two parent threads \texttt{wait()} on one exiting child
		\item Expected: One thread reaps child, the other returns failure
		\item Got: \texttt{failed assertion `nobe->cond\_queue\_next == NULL'}
	\end{itemize}
\end{frame}

\begin{frame}{Studying Bugs In-Depth} % slide: visual explanation of bug
	% graphic want to be colored TODO
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

% slide: recap challenges and how we dealt with them
% slide: conclusion
\begin{frame}{Conclusion}
	\textbf{Systematic exploration in kernel-space}
	% TODO
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\subsection{Introduction}

\subsection{Race Conditions}

% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
% race condition that some of you might have, for the sake of example.

\begin{frame}{Decision Points (``good'' case)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		{\bf Thread 1} & {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread) \\
		\cline{1-2}
		\texttt{return child->tid} &&  \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

\begin{frame}{Decision Points (race condition)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		{\bf Thread 1} & {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread + preempted) \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
		\texttt{return child->tid} && (bad!) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

\breakslide{\Large A different way of looking at race conditions\ldots}

\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork0.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork05.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork1.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork2.png}
	\end{center}
\end{frame}

\begin{frame}{Decision Points} % TODO: duplicate slide with later
	A {\bf decision point} is\ldots

	\linegap
	A code location where being preempted causes different behaviour.
	\begin{itemize}
		\item Intuitively: Somewhere that interesting interleavings can happen around.
	\end{itemize}
	Examples:
	\begin{itemize}
		\item A new thread becomes runnable
		\item Voluntary reschedule (e.g. \texttt{yield}, \texttt{cond\_wait})
		\item Synchronization primitives
	\end{itemize}
	% Say: We'll revisit this later.
\end{frame}

% Say: Obviously we want to be able to always run the third path. But knowing
% that ahead of time is impossible.
% Say: Okay, now what if we could build a tool that could ``systematically''
% explore this tree, and find the branch with the bug in it? What would such a
% tool need to be able to do?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Systematic Testing}

\breakslide{\Large Systematic Testing}

\begin{frame}{Systematic Testing}
	\textbf{Systematic testing} is:
	\begin{itemize}
		\item Systematically enumerating different interleavings
		\begin{itemize}
			\item Intuitively: Generate many ``tabular execution traces''
		\end{itemize}
		\item Exploring all branches in these trees
		\begin{itemize}
			\item (by controlling scheduling decisions at decision points)
		\end{itemize}
		\item In practice: Depth-first search of branches
	\end{itemize}
\end{frame}

\subsection{Requirements}

\begin{frame}{Execution Tree Exploration}
	Important point: When does a branch end?
	\begin{itemize}
		\item All threads run to completion, or
		\item A bug is detected % Say: more on this later.
	\end{itemize}
	{\bf Backtracking}:
	\begin{itemize}
		\item Identify a decision point to choose differently
		\item Reset machine state and start over
		\item Replay test from the beginning, with different choices
	\end{itemize}
\end{frame}

\begin{frame}{More on Decision Points}
	Important point: What does ``all possible interleavings'' mean?

	\linegap
	One extreme: Decide at every instruction
	\begin{itemize}
		\item Good news: Will find every possible race condition.
		\item Bad news: Runtime of test will be impossibly large.
	\end{itemize}
	\linegap

	Other extreme: Nothing is a decision point
	\begin{itemize}
		\item Good news: Test will finish quickly.
		\item Bad news: Only one execution was checked for bugginess.
		\item Bad news: No alternative interleavings explored.
		\begin{itemize}
			\item Makes ``no race found'' a weak claim.
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{More on Decision Points}
	Sweet spot: Insert a thread switch everywhere it ``might matter''.

	\linegap
	When do we fear being preempted?
	\begin{itemize}
		\item New threads becoming runnable (\texttt{fork}, \texttt{cond\_signal}, etc)
			\begin{itemize}
				\item Preemptions may cause it to run before we're ready
			\end{itemize}
		\item Synchronization primitives (\texttt{mutex\_lock}, etc)
			\begin{itemize}
				\item If used improperly\ldots
			\end{itemize}
		\item Unprotected shared memory accesses % Say: more on this later.
			\begin{itemize}
				\item May result in data structure corruption
			\end{itemize}
	\end{itemize}
	\linegap

	Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
\end{frame}

\subsection{Challenges}

\begin{frame}{Controlling Scheduling Decisions}
	Control over sources of nondeterminism
	\begin{itemize}
		\item Device interrupts/input
			\begin{itemize}
				\item Disk drivers: when disk reads finish
				\item Ethernet drivers: when packets arrive
			\end{itemize}
		\item To control thread switches in a 410 kernel, vary when clock ticks happen.
	\end{itemize}
\end{frame}

\begin{frame}{Memory Interposition}
	In order to find use-after-free, need to know:
	\begin{itemize}
		\item When objects are \texttt{free()}d
		\item When threads access shared memory in the heap
	\end{itemize}
	\linegap

	Solution: Keep track of all memory events
	\begin{itemize}
		\item All calls to \texttt{malloc}/\texttt{free}
		\item All shared memory reads/writes
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{State Space Explosion}

\begin{frame}{State Space Explosion}
	State spaces grow exponentially
	\begin{itemize}
		\item With $d$ decision points, $k$ runnable threads, size $d^k$.
		\item Threatens our ability to explore everything.
		\item Fortunately, some sequences result in identical states.
	\end{itemize}
	\linegap

	{\bf Partial Order Reduction} can help.
	\begin{itemize}
		\item Complicated algorithm; ask me later for details.
		\item Intuitive explanation follows.
	\end{itemize}
\end{frame}

\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{undiamond0.png}
	\end{center}
\end{frame}
\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{undiamond1.png}
	\end{center}
\end{frame}
\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{diamond1.png}
	\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Landslide}

\breakslide{\Huge Landslide}

\begin{frame}{About The Project}
	5th year MS since June 2011 % Say: planning to finish this may, hopefully

	\linegap
	Working with Garth Gibson, Jiri Simsa

	\linegap
	{\bf Landslide}: Shows that your Pebbles are not as stable as you thought.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{How Landslide Works}

\begin{frame}{Landslide in Simics}
	As a Simics module, Landslide knows:
	\begin{itemize}
		\item Every instruction the kernel executes
		\item Every memory address the kernel reads/writes
	\end{itemize}
	\linegap
	Artificially causes timer interrupts

	\linegap
	Checkpointing/backtracking via Simics bookmarks
\end{frame}

\begin{frame}{Anatomy}
	\begin{center}
	%\includegraphics[width=0.6\textwidth]{landslide.png}
	\end{center}
\end{frame}

\begin{frame}{Identifying Bugs}
	Landslide can {\em definitely discover}:
	\begin{itemize}
		\item Kernel panics % Say: note: the better your asserts are..!
		\item Deadlock
		\item Use-after-free / double-free
	\end{itemize}
	Landslide can {\em reasonably suspect}:
	\begin{itemize}
		\item Memory leak
		\item Progress sense (halting problem)
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Using Landslide}

\breakslide{\Large Using Landslide}

\begin{frame}{In Which Ben Offers Help}
	% Say: So, this wasn't just a research talk
	This is something you can try!
	\linegap

	Mutual benefit
	\begin{itemize}
		\item Landslide may help you find bugs
		\item You may help Ben evaluate his thesis project
	\end{itemize}
\end{frame}

\begin{frame}{Keeping It Real}
	Finding race conditions is hard for humans.

	\linegap
	It is hard for computer programs too.

	\linegap
	Landslide is not an oracle.
\end{frame}

\begin{frame}{Annotating Your Kernel}
	Step 1
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{tell_landslide.png}
	\end{center}
	Your kernel needs to say when certain events happen:
	\begin{itemize}
		\item When do threads become runnable / descheduled?
		\item When does the scheduler switch threads?
	\end{itemize}
	\linegap

	\framebox{Time estimate: 40 minutes}
\end{frame}

\begin{frame}{Configuring Landslide}
	Step 2
	\begin{center}
		%\includegraphics[width=0.8\textwidth]{config-landslide.png}
	\end{center}
	Edit \texttt{config.landslide} with some details and tweaks

	Fill in two implementation-dependent C functions in Landslide ($\le$10 lines)
	% Say: Which depend on your kernel's specific implementation
	\linegap

	\framebox{Time estimate: 60 minutes}
\end{frame}

\begin{frame}{Configuring Decision Points}
	Landslide automatically identifies a minimal set of decision points.
	\begin{itemize}
		\item It might find bugs.
		\item It might overlook more fine-grained interleavings.
	\end{itemize}
	With help from you, it could find more.
	\begin{itemize}
		\item Optional annotation: \texttt{tell\_landslide\_decide()}
		\item Hints to where a context switch should be forced.
		\item Inside every call to \texttt{mutex\_lock}\ldots
	\end{itemize}
\end{frame}

\breakslide{\Large Quick Demo}

% TODO more of these with screenshots

%\begin{frame}{Decision Points (noticing a theme here?)}
%	Landslide is {\bf false negative}-oriented.
%	\begin{itemize}
%		\item If Landslide reports a bug, there is a bug for sure.
%			% Say: Well, except progress sense might be wrong.
%		\item If not,
%			\begin{itemize}
%				\item Maybe there was no race, or
%				\item Maybe your decision points were not granular enough.
%			\end{itemize}
%	\end{itemize}
%\end{frame}
%\begin{frame}{Decision Points (noticing a theme here?)}
%	Sometimes the default set will work.
%
%	Sometimes you need to tweak\ldots
%	\begin{itemize}
%		\item Ignore certain global mutexes
%		\item Whitelist or blacklist entire syscalls
%		\item (get creative)
%	\end{itemize}
%	Intuitively: focusing Landslide on relevant parts of the kernel.
%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{User Study}

\begin{frame}{In Which Ben Offers Help - Warning}
	\textbf{If you are already struggling, this will not ``save'' you.}
	\begin{itemize}
		\item False-negatives: not guaranteed to find races at all
		\item Research-quality: possibly difficult to integrate with your kernel
		\item Finishing the kernel project is more important.
	\end{itemize}
	You should:
	\begin{itemize}
		\item Be expecting an A or B
		\item Be able to turn in without late days (if you really had to)
		\item Be looking for\ldots
			\begin{itemize}
				\item That ``one pesky race''
				\item A race that stress-testing missed
				\item Or just familiarity with a new technique
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{In Which Ben Offers Help}
	Your kernel
	\begin{itemize}
		\item Must load the shell and run programs
		\item \texttt{fork},
			\texttt{exec},
			\texttt{vanish},
			\texttt{wait},
			\texttt{readline}
		\item Must never spin-wait (see hurdle form!)
		\item Should \texttt{assert()} important invariants
		\begin{itemize}
			\item Think of \texttt{panic()} as \texttt{tell\_landslide\_bug()}
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{In Which Ben Offers Help}
	User study next week, starting Monday

	\linegap
	Expect to spend:
	\begin{itemize}
		\item Up to 4 hours, just to try it out.
		\item 6-8 hours, if you find a bug and track it down.
		\item More, for multiple bugs or the truly dedicated\ldots
	\end{itemize}
	\linegap

	Give feedback (intuitive? frustrating? found bugs?)

	\linegap
	{\bf Watch \texttt{410.announce} for details!}
\end{frame}

\subsection{End}
\breakslide{\Large Questions?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \section{Conclusion}
% \subsection{Conclusion}
% 
% \begin{frame}{Summary}
% 	\begin{itemize}
% 		\item
% 	\end{itemize}
% \end{frame}

\end{document}
%\begin{frame}{Partial Order Reduction}
%	\begin{center}
%	The next 6 slides won't be on the exam.
%
%		\Huge \dbend
%	\end{center}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction}
%	A {\bf transition} is\ldots
%	\begin{itemize}
%		\item The sequence of instructions executed between choice points.
%	\end{itemize}
%	To express concurrency between transitions:
%	% Say: As a prerequisite for this algorithm, we need to build two sets to express the concurrency relations between the transitions
%	\begin{itemize}
%		\item Independence relation
%		\item Happens-before relation
%	\end{itemize}
%	% Say: This is hard, and I don't expect you to understand it during this lecture; you won't be tested on it or anything. My goal in presenting it is to convey how intensely cool the algorithm is, and hopefully you will keep thinking about it later. I'd be happy to answer questions about it later.
%\end{frame}
%
%\begin{frame}[fragile]{Partial Order Reduction - Independence Relation}
%	\begin{columns}[l]
%		\column{0.5\textwidth}
%		\begin{center}
%		\begin{verbatim}
%		void thread_A() {
%		    foo = 42;
%		}
%		void thread_B() {
%		    if (bar == 1337)
%		        print "Hello"
%		}
%		\end{verbatim}
%		\end{center}
%		\linegap
%
%		$A$ and $B$ are {\bf independent}.
%		\column{0.5\textwidth}
%		\begin{center}
%		\begin{verbatim}
%		void thread_A() {
%		    foo = 42;
%		}
%		void thread_C() {
%		    if (foo == 1337)
%		        print "Hello"
%		}
%		\end{verbatim}
%		\end{center}
%		\linegap
%
%		$A$ and $C$ {\bf conflict}.
%	\end{columns}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Independence Relation}
%	Consider a trace of all shared memory reads/writes
%
%	\linegap
%	Two transitions (instruction sequences) $A$ and $B$ are independent iff:
%	\begin{itemize}
%		\item $Writes(A) \cap Writes(B) = \emptyset$
%		\item $Writes(A) \cap Reads(B) = \emptyset$
%		\item $Reads(A) \cap Writes(B) = \emptyset$
%	\end{itemize}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Happens-before Relation}
%	\begin{columns}[l]
%		\column{0.5\textwidth}
%		%\includegraphics[width=\textwidth]{hb.png}
%		\column{0.5\textwidth}
%		Is a transition ``enabled'' by a previous one?
%		\begin{itemize}
%			\item $A$ enables $B$
%			\item $B$ enables $C$ (same thread)
%			\item $B$ enables $D$
%		\end{itemize}
%%		Transitive closure
%%		\begin{itemize}
%%			\item $HB^* = \{AB, AC, AD, BC, BD\}$
%%		\end{itemize}
%		$C$ and $D$ are {\bf concurrent}!
%	\end{columns}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Tagging}
%	Run POR at the end of each branch
%	\begin{itemize}
%		\item Identify which branches to explore next
%		\item Tag only a subset of all branches (hopefully)
%	\end{itemize}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Tagging}
%	For each transition $T$, find an {\bf evil ancestor} $E$:
%	\begin{itemize}
%		\item $E$ and $T$ are concurrent.
%		\item $E$ and $T$ are not independent.
%		\item Intuitively: $T$ and $E$ could be reordered, and might behave differently.
%	\end{itemize}
%	Find a {\bf good sibling} $G$ of $E$:
%	\begin{itemize}
%		\item $G$ is the same thread as $T$
%		% Say: (if no such sibling, all siblings)
%		\item Intuitively: Find a branch in which $T$ runs before $E$.
%	\end{itemize}
%	Explore tagged branches depth-first.
%\end{frame}
%
%\begin{frame}{Partial Order Reduction}
%	\begin{center}
%		{\Huge \dbend}
%		\linegap
%		\linegap
%		\linegap
%
%		Feel free to ask me about this later.
%	\end{center}
%\end{frame}

% TODO: run through an example

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subsection{Other Dynamic Techniques}
%
%\begin{frame}{Heuristic-Based Searching}
%	% Say: Note that this is beyond the scope of my personal project.
%	Maybe the tree is too big to search anyway?
%
%	\linegap
%	Search {\em most likely} branches first.
%	\begin{itemize}
%		\item ``Iterative Context Bounding''
%		\footnote{Microsoft Research - CHESS - Musuvathi et al.}
%		\begin{itemize}
%			\item Insight: Races tend to show up with few forced context switches.
%			\item Search branches with fewer preemptions first.
%		\end{itemize}
%		\item Find behaviours that are typical of buggy software.
%		\footnote{Stanford University - Dawson Engler}
%		\begin{itemize}
%			\item Different behaviour in different branches.
%				% Say: different all the time: no. different rarely: yes.
%			\item Use suspicious memory accesses as future choice points.
%		\end{itemize}
%	\end{itemize}
%	\linegap
%	Open research field, possibly my future work.
%\end{frame}
%
%% Say: I'm going to take a break from talking about systematic exploration to talk about other types of dynamic race detection - FIXME: if these slides fit in 60min.
%\begin{frame}{Data Race Detection}
%	Requires happens-before set and ``lock set''
%	\linegap
%
%	Two memory accesses $m_1$ and $m_2$ are a data race iff:
%	\begin{itemize}
%		\item Read/write or write/write to the same address, and
%		\item $(m_1,m_2) \not\in HB$, and
%		\item $Lock(m_1) \cap Lock(m_2) = \emptyset$
%	\end{itemize}
%\end{frame}
%\begin{frame}{Data Race Detection}
%	Data race detection vs systematic exploration
%	\begin{itemize}
%		\item More false positives
%			\begin{itemize}
%				\item (but still good style-checking!)
%			\end{itemize}
%		\item No false negatives
%		\item Still unable to find some types of bugs
%	\end{itemize}
%	\linegap
%	Hybrid approach? (future research)
%\end{frame}

% vim: foldmethod=indent

### 2012-05-09 defence/defence.tex
minor update defence
update presentation

@@
 
-\usecolortheme[named=ForestGreen]{structure}
+\usecolortheme[named=RoyalBlue]{structure}
 
@@
 \begin{frame}{Outline}
-	\textbf{Motivation: Race condition debugging}
+	\textbf{Motivation: Concurrency debugging}
 	\begin{itemize}
@@
 	\end{itemize}
-	{\bf Landslide: Making it happen}
+	\linegap
+
+	{\bf Tool: Landslide}
 	\begin{itemize}
@@
 	\end{itemize}
-	{\bf Evaluation}
+	\linegap
+
+	{\bf Evaluation: Finding Races}
 	\begin{itemize}
@@
 \section{Motivation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Race Conditions and the Finding Thereof}
@@
 \begin{frame}{Testing Techniques} % slide: sys ex vs stress testing
-	\textbf{Stress Testing}: Common testing approach
+	\textbf{Stress Testing:} Common testing approach
 	\begin{itemize}
@@
 
-	{\bf Systematic Exploration}\related{Flanagan '05}
+	{\bf Systematic Exploration} % \related{Flanagan '05}
 	\begin{itemize}
@@
 
-% slide: 15-410 6-week project why we need it in kernels
-\begin{frame}{Finding Races in Kernel-Space}
-	% cite datacollider etc?
+\begin{frame}{Finding Races in Kernel-Space} % slide: 15-410 6-week project why we need it in kernels
+	15-410: Operating System Design and Implementation
+	\begin{itemize}
+		\item Project 3: students write a kernel in 6 weeks
+		\item ``Pebbles'': UNIX-like, fully preemptible
+	\end{itemize}
+	\linegap
+
+	Previous work in kernel-space has not used systematic exploration.
+	\begin{itemize}
+		\item Data race detection\related{Erickson '10}
+		\item Failure resiliency\related{Kadav '09}
+	\end{itemize}
 \end{frame}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Challenges of Kernel-space}
+
+\breakslide{Why is this environment different from all other environments?}
 
@@
 	\begin{itemize}
-		\item When do threads created/destroyed?
+		\item When are threads created/destroyed?
 	\end{itemize}
@@
 % slide: ad-hoc communication, memory conflicts (freeing, runqueue)
-\begin{frame}{Ad-Hoc Thread Communication}
-	No message-passing % TODO this slide sucks
-
-	User-space systematic exploration relies on it\related{Simsa '10}; we cannot.
-
-	Kernel threads share state via globals, heap.
-\end{frame}
+%\begin{frame}{Ad-Hoc Thread Communication}
+%	No message-passing % TODO this slide sucks
+%	
+%	User-space systematic exploration relies on it\related{Simsa '10}; we cannot.
+%	
+%	Kernel threads share state via globals, heap.
+%\end{frame}
 
 \begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
+	\begin{itemize}
+		\item ``Everything interleaves with everything else''?
+		\item What is a ``bug'', anyway?\related{Baumann '09}
+	\end{itemize}
+	\begin{center}
 	\includegraphics[width=\textwidth]{pebbles.png}
+	\end{center}
 	% Say: "we don't want to have to explore every possible interaction between each of these components"
@@
 		\item How can it be applied in kernel-space?
-		\item What should/must we ask of the user to make it possible?
+		\item What can we ask of the user to make it possible?
 	\end{itemize}
@@
 \section{Landslide}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\subsection{Anatomy}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Design}
+
 % slide: pebbles, 15-410, simics
+\begin{frame}{Environment}
+\end{frame}
+
 % slide: visualisation (here address challenge #1)
+\begin{frame}{Anatomy of Landslide}
+	\begin{center}
+	\includegraphics[width=0.7\textwidth]{landslide.png}
+	\end{center}
+\end{frame}
 
-\subsection{Design}
 % slide: using -- annotations & in-landslide instrumentation
+\begin{frame}{Using Landslide}
+\end{frame}
+
 % slide: POR -- mostly invisible, but (transition)
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond1.png}
+	\end{center}
+\end{frame}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Addressing Challenges} % rename
-% slide: POR, modified - with runqueue conflicts, will get no pruning
-% slide: ignoring shared memory accesses - venn diagram
+
+\begin{frame}{Shared Memory Conflicts} % slide: ignoring shared memory accesses - venn diagram
+	\begin{center}
+	\includegraphics[width=\textwidth]{shm-conflict0.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Shared Memory Conflicts}
+	\begin{center}
+	\includegraphics[width=\textwidth]{shm-conflict1.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Shared Memory Conflicts}
+	\begin{center}
+	\includegraphics[width=\textwidth]{shm-conflict2.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Shared Memory Conflicts}
+	\textbf{Problem:} Many conflicts we don't care about
+	\begin{itemize}
+		\item Every transition accesses scheduler state
+		\item Many transitions access ``unrelated'' global objects
+	\end{itemize}
+	\linegap
+
+	{\bf Solution:} Only consider ``relevant'' memory conflicts
+	\begin{itemize}
+		\item Ignore scheduler, global objects ``known to be correct''
+		\begin{itemize}
+			\item Sacrifice ability to test these to efficiently test everything else.
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+% slide: identifying bugs
+\begin{frame}{Identifying Bugs}
+	\textbf{How do we know we've found a bug?}
+\end{frame}
 % slide: other_prob: decide on mutex_lock as opposed to everywhere, but...
 % slide: manual shaping - have *user* specify what to focus on (automating intuition)
+\begin{frame}{Limiting Decision Points}
+\end{frame}
 % slide: exciting teaser "i think this is not too far off!"
+\begin{frame}{Teaser}
+\end{frame}
 
@@
 	\end{itemize}
+	\linegap
+
 	Five groups participated; four got it to work.
 	\begin{itemize}
-		\item Average instrumentation time 100 minutes (min 50 max 150)
-		\item Average customisation time 25 minutes (min 0 max 60)
+		\item Average instrumentation time 100 minutes (50 to 150)
+		\item Average customisation time 25 minutes (0 to 60)
 	\end{itemize}
@@
 
-\begin{frame}{Found Bugs} % slide: anecdotes: two deterministic bugs & two races
+\begin{frame}{Landslide Victories} % slide: anecdotes: two deterministic bugs & two races
 	All four groups who got it to work found bugs.
+	\linegap
 
@@
 	Worked with two kernels: my own, and one I graded
+	\linegap
 
 	``I have made something smarter than me.''
+	% Heh, if it were really smarter than me, I wouldn't have had to debug it.
 	\begin{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
+\section{Conclusion}
 
-% slide: recap challenges and how we dealt with them
-% slide: conclusion
-\begin{frame}{Conclusion}
-	\textbf{Systematic exploration in kernel-space}
-	% TODO
+\subsection{Future Work}
+\begin{frame}{Promising Directions}
+	Top 15-410 students, effectively supervised, can find hard bugs faster
+	\begin{itemize}
+		\item Use as a debugging tool may help students learn more
+	\end{itemize}
+	\linegap
+
+	After Pebbles\dots
+	\begin{itemize}
+		\item Potential application to production kernels
+		\begin{itemize}
+			\item General purpose OSes, device drivers, embedded kernels
+			% Be ready to field questions about SMP.
+		\end{itemize}
+	\end{itemize}
+	\linegap
+
+	Emphasis on ``steering'' of test parameters by experts
+	\begin{itemize}
+		\item Perhaps we can automate this steering?
+	\end{itemize}
+	\linegap
 \end{frame}
 
+\subsection{Summary}
+\begin{frame}{Summary}
+	\textbf{Systematic exploration in kernel-space}
+	\begin{itemize}
+		\item Use internal kernel abstractions to understand concurrency behaviour
+		\item Relying on user's knowledge makes testing easier
+	\end{itemize}
+	\linegap
+	\textbf{Landslide}
+	\begin{itemize}
+		\item After polishing, a useful tool for 15-410
+		\item A first step towards sophisticated kernel debugging techniques
+	\end{itemize}
+
+	\linegap
+	\linegap
+	\begin{center}
+	{\scriptsize \em Please hold on. This transit is approaching the Landslide terminal, \\baggage claim, ground transportation, and ticketing.}
+	\end{center}
+\end{frame}
+
+% slide: bonus slides
+
+\end{document}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\end{document}
 %\begin{frame}{Partial Order Reduction}

### 2012-05-09 evaluation.tex
write oddities
fix typos in thesis

@@
 
-Additionally, two enterprising users\footnote{The second case here was an ex-TA, not a student.} found ways to avoid implementing the functions in \texttt{student.c} (Section~\ref{sec:using-student-c}).
+Additionally, two enterprising users found ways to avoid implementing the functions in \texttt{student.c} (Section~\ref{sec:using-student-c}).
 It is worth considering integrating these approaches into the recommended approach to eliminate the need for \texttt{student.c} entirely.
 \begin{itemize}
-	\item One studnt bypassed the \texttt{kern\_ready\_for\_timer\_interrupt} annotation by disabling their scheduler's preemption-disabling mechanism and replacing it with explicit disabling of interrupts, thereby making every timer interrupt cause a context switch.
+	\item One student bypassed the \texttt{kern\_ready\_for\_timer\_interrupt} annotation by disabling their scheduler's preemption-disabling mechanism and replacing it with explicit disabling of interrupts, thereby making every timer interrupt cause a context switch.
 	\item One ex-TA was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
@@
 \subsection{Landslide Victories}
+\label{sec:eval-victory}
 

### 2012-05-09 defence/defence.tex
add fw/concl to outline
add more stuff

@@
 	\end{itemize}
+	\linegap
+
+	{\bf Future Work and Conclusion}
 \end{frame}
@@
 \begin{frame}{Finding Races in Kernel-Space} % slide: 15-410 6-week project why we need it in kernels
-	15-410: Operating System Design and Implementation
-	\begin{itemize}
-		\item Project 3: students write a kernel in 6 weeks
-		\item ``Pebbles'': UNIX-like, fully preemptible
+	\textbf{15-410:} Operating System Design and Implementation
+	\begin{itemize}
+		\item Project 3: students write a kernel in 6 weeks.
+		\item ``Pebbles'' is a UNIX-like, preemptible kernel specification.
 	\end{itemize}
@@
 	\linegap
-	Context switcher
-	\begin{itemize}
-		\item Interleaving threads is no longer ``instantaneous''.
+	Context switching
+	\begin{itemize}
+		\item Non-deterministic timer-driven thread scheduling
 	\end{itemize}
@@
 % slide: outline
+\begin{frame}{Outline}
+\end{frame}
 
@@
 \begin{frame}{Environment}
+	\textbf{Simics:} a full-system x86 simulator
+	\begin{itemize}
+		\item Students in 15-410 use Simics to test their kernels.
+		\item Landslide is a transparently-embedded Simics module.
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Using Landslide}
-\end{frame}
-
-% slide: POR -- mostly invisible, but (transition)
+	\begin{center}
+	\includegraphics[width=0.9\textwidth]{tell_landslide.png}
+	\end{center}
+	User must tell Landslide when certain kernel events happen:
+	\begin{itemize}
+		\item When do threads become runnable / descheduled?
+		\item When does the scheduler switch threads?
+	\end{itemize}
+	\linegap
+	Some complicated conditions require in-Landslide instrumentation.
+	\begin{itemize}
+		\item Scheduler design details
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Decision Trace}
+	\includegraphics[width=\textwidth]{found_a_bug.png}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Addressing Challenges} % rename
+
 \begin{frame}{State Space Reduction}
@@
 \end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Addressing Challenges} % rename
 
@@
 	\begin{itemize}
-		\item Every transition accesses scheduler state
-		\item Many transitions access ``unrelated'' global objects
-	\end{itemize}
-	\linegap
-
-	{\bf Solution:} Only consider ``relevant'' memory conflicts
+		\item Every transition accesses scheduler state.
+		\item Many transitions access ``unrelated'' global objects.
+	\end{itemize}
+	\linegap
+
+	{\bf Solution:} Only consider ``relevant'' memory conflicts.
 	\begin{itemize}
@@
 	\textbf{How do we know we've found a bug?}
+	\linegap
+
+	% TODO: does this slide want to be multicolumn, with an infinite loop diagram?
+	Accurate conditions
+	\begin{itemize}
+		\item Kernel panic
+		\item Use-after-free
+		\item Deadlock
+	\end{itemize}
+	Uncertain conditions
+	\begin{itemize}
+		\item Memory leak
+		\item Infinite loop
+		\begin{itemize}
+			\item Use structure of execution tree to judge stuckness.
+		\end{itemize}
+	\end{itemize}
 \end{frame}
@@
 \section{Conclusion}
-
-\subsection{Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Outline}
+\end{frame}
+
+%\subsection{Future Work}
 \begin{frame}{Promising Directions}
-	Top 15-410 students, effectively supervised, can find hard bugs faster
-	\begin{itemize}
-		\item Use as a debugging tool may help students learn more
+	Top 15-410 students, effectively supervised, can find hard bugs faster.
+	\begin{itemize}
+		\item Use as a debugging tool may help students learn more.
 	\end{itemize}
@@
 
-\subsection{Summary}
+%\subsection{Summary}
 \begin{frame}{Summary}
@@
 	\begin{itemize}
-		\item Use internal kernel abstractions to understand concurrency behaviour
-		\item Relying on user's knowledge makes testing easier
+		\item Use internal kernel abstractions to understand concurrency behaviour.
+		\item Relying on user's knowledge makes testing easier.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item After polishing, a useful tool for 15-410
-		\item A first step towards sophisticated kernel debugging techniques
+		\item After polishing, a useful tool for 15-410.
+		\item A first step towards sophisticated kernel debugging techniques.
 	\end{itemize}

### 2012-05-09 defence/defence.tex
rearrange comments
add outline slides and a few minor edits

@@
 \newcommand\breakslide[1]{\begin{frame}{} \begin{center} #1 \end{center} \end{frame}}
-\newcommand\related[1]{\textsuperscript{\em #1}}
-
-% slide: outline
+\newcommand\related[1]{\textsuperscript{\em [#1]}}
+
 \begin{frame}{Outline}
@@
 
-	{\bf Systematic Exploration} % \related{Flanagan '05}
+	{\bf Systematic Exploration} \related{Simsa '10}
 	\begin{itemize}
@@
 
-% slide: outline
+\newcommand\done[1]{\textcolor{gray}{\em \small #1}}
+
 \begin{frame}{Outline}
+	\done{Motivation: Concurrency debugging}
+	\begin{itemize}
+		\item \done{Systematic exploration versus stress testing}
+		\item \done{Challenges of kernel-space}
+	\end{itemize}
+	\linegap
+
+	{\bf Tool: Landslide}
+	\begin{itemize}
+		\item {\bf Design and interface}
+		\item {\bf Addressing challenges}
+	\end{itemize}
+	\linegap
+
+	Evaluation: Finding Races
+	\begin{itemize}
+		\item Student user study
+		\item Case study
+	\end{itemize}
+	\linegap
+
+	Future Work and Conclusion
 \end{frame}
@@
 
-% slide: pebbles, 15-410, simics
 \begin{frame}{Environment}
@@
 
-% slide: visualisation (here address challenge #1)
 \begin{frame}{Anatomy of Landslide}
@@
 	\end{center}
-\end{frame}
-
-% slide: using -- annotations & in-landslide instrumentation
+	% Remember to talk about solving the thread scheduling challenge
+\end{frame}
+
+\subsection{Interface}
+
 \begin{frame}{Using Landslide}
@@
 \begin{frame}{Decision Trace}
+	% TODO make the stack trace coloured
 	\includegraphics[width=\textwidth]{found_a_bug.png}
@@
 
+\begin{frame}{State Space Reduction}
+	State spaces grow exponentially.
+	\begin{itemize}
+		\item Fortunately, some sequences result in identical states.
+		\item Dynamic Partial Order Reduction\related{Flanagan '05}
+		\begin{itemize}
+			\item Requires ``memory independence relation'' between transitions.
+		\end{itemize}
+	\end{itemize}
+\end{frame}
 \begin{frame}{State Space Reduction}
@@
 		\end{itemize}
-	\end{itemize}
-\end{frame}
-% slide: identifying bugs
+		\item User must specify what to ignore.
+	\end{itemize}
+\end{frame}
+
 \begin{frame}{Identifying Bugs}
@@
 
-% slide: outline
 \begin{frame}{Outline}
+	\done{Motivation: Concurrency debugging}
+	\begin{itemize}
+		\item \done{Systematic exploration versus stress testing}
+		\item \done{Challenges of kernel-space}
+	\end{itemize}
+	\linegap
+
+	\done{Tool: Landslide}
+	\begin{itemize}
+		\item \done{Design and interface}
+		\item \done{Addressing challenges}
+	\end{itemize}
+	\linegap
+
+	{\bf Evaluation: Finding Races}
+	\begin{itemize}
+		\item {\bf Student user study}
+		\item {\bf Case study}
+	\end{itemize}
+	\linegap
+
+	Future Work and Conclusion
 \end{frame}
@@
 	\end{itemize}
+	\linegap
+
 	Deterministic bugs
@@
 		\item \texttt{double\_wait} - two parent threads \texttt{wait()} on one exiting child
+		\begin{itemize}
 			\item Expected: One thread reaps child, the other returns failure
@@
 		\end{itemize}
+		\item Previously unknown, not found by stress testing or by TA
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Outline}
+	\done{Motivation: Concurrency debugging}
+	\begin{itemize}
+		\item \done{Systematic exploration versus stress testing}
+		\item \done{Challenges of kernel-space}
+	\end{itemize}
+	\linegap
+
+	\done{Tool: Landslide}
+	\begin{itemize}
+		\item \done{Design and interface}
+		\item \done{Addressing challenges}
+	\end{itemize}
+	\linegap
+
+	\done{Evaluation: Finding Races}
+	\begin{itemize}
+		\item \done{Student user study}
+		\item \done{Case study}
+	\end{itemize}
+	\linegap
+
+	{\bf Future Work and Conclusion}
 \end{frame}
@@
 
-\begin{frame}{Decision Points} % TODO: duplicate slide with later
+\begin{frame}{Decision Points} % : duplicate slide with later
 	A {\bf decision point} is\ldots
@@
 \breakslide{\Large Quick Demo}
-
-% TODO more of these with screenshots
 
@@
 
-% TODO: run through an example
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-05-09 defence/defence.tex
update defence for carlo

@@
 	\begin{itemize}
-		\item On vanish, child's TCB is freed
+		\item On exit, child's TCB is freed
 		\item Forking thread does use-after-free
@@
 \begin{frame}{Finding Races in Kernel-Space} % slide: 15-410 6-week project why we need it in kernels
+	Previous work in kernel-space has not used systematic exploration.
+	\linegap
+
 	\textbf{15-410:} Operating System Design and Implementation
@@
 	\end{itemize}
-	\linegap
-
-	Previous work in kernel-space has not used systematic exploration.
-	\begin{itemize}
-		\item Data race detection\related{Erickson '10}
-		\item Failure resiliency\related{Kadav '09}
-	\end{itemize}
+	%\begin{itemize}
+	%	\item Data race detection\related{Erickson '10}
+	%	\item Failure resiliency\related{Kadav '09}
+	%\end{itemize}
 \end{frame}
@@
 \begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
+	\begin{center}
+	\includegraphics[width=\textwidth]{pebbles.png}
+	\end{center}
+\end{frame}
+\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
 	\begin{itemize}
@@
 	\end{itemize}
-	\begin{center}
-	\includegraphics[width=\textwidth]{pebbles.png}
-	\end{center}
 	% Say: "we don't want to have to explore every possible interaction between each of these components"
@@
 
-\begin{frame}{Outline}
-	\done{Motivation: Concurrency debugging}
-	\begin{itemize}
-		\item \done{Systematic exploration versus stress testing}
-		\item \done{Challenges of kernel-space}
-	\end{itemize}
-	\linegap
-
-	{\bf Tool: Landslide}
-	\begin{itemize}
-		\item {\bf Design and interface}
-		\item {\bf Addressing challenges}
-	\end{itemize}
-	\linegap
-
-	Evaluation: Finding Races
-	\begin{itemize}
-		\item Student user study
-		\item Case study
-	\end{itemize}
-	\linegap
-
-	Future Work and Conclusion
-\end{frame}
+%\begin{frame}{Outline}
+%	\done{Motivation: Concurrency debugging}
+%	\begin{itemize}
+%		\item \done{Systematic exploration versus stress testing}
+%		\item \done{Challenges of kernel-space}
+%	\end{itemize}
+%	\linegap
+%
+%	{\bf Tool: Landslide}
+%	\begin{itemize}
+%		\item {\bf Design and interface}
+%		\item {\bf Addressing challenges}
+%	\end{itemize}
+%	\linegap
+%
+%	Evaluation: Finding Races
+%	\begin{itemize}
+%		\item Student user study
+%		\item Case study
+%	\end{itemize}
+%	\linegap
+%
+%	Future Work and Conclusion
+%\end{frame}
 
@@
 	\linegap
+	% TODO
 	Some complicated conditions require in-Landslide instrumentation.
@@
 \begin{frame}{Decision Trace}
-	% TODO make the stack trace coloured
+	% FIXME make the stack trace coloured
 	\includegraphics[width=\textwidth]{found_a_bug.png}
@@
 
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond1.png}
+	\end{center}
+\end{frame}
 \begin{frame}{State Space Reduction}
@@
 \end{frame}
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond1.png}
-	\end{center}
-\end{frame}
 
@@
 \begin{frame}{Shared Memory Conflicts}
-	\textbf{Problem:} Many conflicts we don't care about
-	\begin{itemize}
-		\item Every transition accesses scheduler state.
-		\item Many transitions access ``unrelated'' global objects.
-	\end{itemize}
-	\linegap
-
-	{\bf Solution:} Only consider ``relevant'' memory conflicts.
-	\begin{itemize}
-		\item Ignore scheduler, global objects ``known to be correct''
+	%\textbf{Problem:} Many conflicts we don't care about
+	%\begin{itemize}
+	%	\item Every transition accesses scheduler state.
+	%	\item Many transitions access ``unrelated'' global objects.
+	%\end{itemize}
+	%\linegap
+	\textbf{Solution:} Only consider ``relevant'' memory conflicts.
+	\begin{itemize}
+		\item Ignore scheduler, global objects assumed to be correct
 		\begin{itemize}
@@
 	% TODO: does this slide want to be multicolumn, with an infinite loop diagram?
-	Accurate conditions
-	\begin{itemize}
-		\item Kernel panic
+	Definite bugs
+	\begin{itemize}
+		\item Kernel panics / assertion failures
 		\item Use-after-free
@@
 	\end{itemize}
-	Uncertain conditions
-	\begin{itemize}
-		\item Memory leak
+	Probable bugs
+	\begin{itemize}
 		\item Infinite loop
@@
 
-\begin{frame}{Outline}
-	\done{Motivation: Concurrency debugging}
-	\begin{itemize}
-		\item \done{Systematic exploration versus stress testing}
-		\item \done{Challenges of kernel-space}
-	\end{itemize}
-	\linegap
-
-	\done{Tool: Landslide}
-	\begin{itemize}
-		\item \done{Design and interface}
-		\item \done{Addressing challenges}
-	\end{itemize}
-	\linegap
-
-	{\bf Evaluation: Finding Races}
-	\begin{itemize}
-		\item {\bf Student user study}
-		\item {\bf Case study}
-	\end{itemize}
-	\linegap
-
-	Future Work and Conclusion
-\end{frame}
+%\begin{frame}{Outline}
+%	\done{Motivation: Concurrency debugging}
+%	\begin{itemize}
+%		\item \done{Systematic exploration versus stress testing}
+%		\item \done{Challenges of kernel-space}
+%	\end{itemize}
+%	\linegap
+%
+%	\done{Tool: Landslide}
+%	\begin{itemize}
+%		\item \done{Design and interface}
+%		\item \done{Addressing challenges}
+%	\end{itemize}
+%	\linegap
+%
+%	{\bf Evaluation: Finding Races}
+%	\begin{itemize}
+%		\item {\bf Student user study}
+%		\item {\bf Case study}
+%	\end{itemize}
+%	\linegap
+%
+%	Future Work and Conclusion
+%\end{frame}
 
@@
 		\item Average instrumentation time 100 minutes (50 to 150)
-		\item Average customisation time 25 minutes (0 to 60)
+		%\item Average customisation time 25 minutes (0 to 60)
 	\end{itemize}
@@
 
-\begin{frame}{Outline}
-	\done{Motivation: Concurrency debugging}
-	\begin{itemize}
-		\item \done{Systematic exploration versus stress testing}
-		\item \done{Challenges of kernel-space}
-	\end{itemize}
-	\linegap
-
-	\done{Tool: Landslide}
-	\begin{itemize}
-		\item \done{Design and interface}
-		\item \done{Addressing challenges}
-	\end{itemize}
-	\linegap
-
-	\done{Evaluation: Finding Races}
-	\begin{itemize}
-		\item \done{Student user study}
-		\item \done{Case study}
-	\end{itemize}
-	\linegap
-
-	{\bf Future Work and Conclusion}
-\end{frame}
+%\begin{frame}{Outline}
+%	\done{Motivation: Concurrency debugging}
+%	\begin{itemize}
+%		\item \done{Systematic exploration versus stress testing}
+%		\item \done{Challenges of kernel-space}
+%	\end{itemize}
+%	\linegap
+%
+%	\done{Tool: Landslide}
+%	\begin{itemize}
+%		\item \done{Design and interface}
+%		\item \done{Addressing challenges}
+%	\end{itemize}
+%	\linegap
+%
+%	\done{Evaluation: Finding Races}
+%	\begin{itemize}
+%		\item \done{Student user study}
+%		\item \done{Case study}
+%	\end{itemize}
+%	\linegap
+%
+%	{\bf Future Work and Conclusion}
+%\end{frame}
 
@@
 		\begin{itemize}
-			\item General purpose OSes, device drivers, embedded kernels
+			\item General purpose OSes, device drivers
 			% Be ready to field questions about SMP.
@@
 	\begin{itemize}
-		\item After polishing, a useful tool for 15-410.
+		\item Has potential to be a useful tool for 15-410.
 		\item A first step towards sophisticated kernel debugging techniques.

### 2012-05-10 defence/defence.tex
update defence for jiri; add livelock visual

@@
 	\begin{itemize}
-		\item Systematic exploration versus stress testing
+		\item Systematic testing versus stress testing
 		\item Challenges of kernel-space
@@
 	\begin{itemize}
-		\item On exit, child's TCB is freed
+		\item On exit, child's state is freed
 		\item Forking thread does use-after-free
-		\item Might return garbage instead of TID
+		\item Might return garbage instead of thread ID
 	\end{itemize}
@@
 
-	{\bf Systematic Exploration} \related{Simsa '10}
+	{\bf Systematic Testing} \related{Musuvathi '08}
 	\begin{itemize}
@@
 		\item Run {\em every single} interleaving
-		\item Provide better debugging information
+		\item Provide better debugging information, reproducibility
 	\end{itemize}
@@
 \begin{frame}{Finding Races in Kernel-Space} % slide: 15-410 6-week project why we need it in kernels
-	Previous work in kernel-space has not used systematic exploration.
+	Previous work in kernel-space has not used systematic testing.
 	\linegap
@@
 \begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
+	% say: it's also important to understand the kernel's internals instead of just viewing it as one ``black box''
 	\begin{center}
@@
 	\end{center}
+	% now that we see these internal interfaces, how do we apply systematic testing to it?
 \end{frame}
@@
 \begin{frame}{Contribution} % slide: summary: "sys ex is a tool \ldots``
-	Systematic exploration is a {\em tool} for testing concurrent systems.
+	Systematic testing is a {\em tool} for testing concurrent systems.
 	\begin{itemize}
 		\item How can it be applied in kernel-space?
-		\item What can we ask of the user to make it possible?
+		\item What simple things can we ask of the user to make it possible?
 	\end{itemize}
@@
 	\linegap
-	% TODO
+	% Make sure to expound on what these are
 	Some complicated conditions require in-Landslide instrumentation.
@@
 \begin{frame}{State Space Reduction}
+	% Say: ``combinatorial explosion''
 	\begin{center}
@@
 
-\begin{frame}{Identifying Bugs}
+\begin{frame}[fragile]{Identifying Bugs}
+	\begin{columns}[l]
+		\column{0.75\textwidth}
 		\textbf{How do we know we've found a bug?}
@@
 
-	% TODO: does this slide want to be multicolumn, with an infinite loop diagram?
 		Definite bugs
@@
 			\begin{itemize}
-			\item Use structure of execution tree to judge stuckness.
-		\end{itemize}
-	\end{itemize}
+				\item Use structure of execution tree to judge progress.
+			\end{itemize}
+		\end{itemize}
+		\column{0.25\textwidth}
+		\includegraphics[width=\textwidth]{blank-livelock.png}
+	\end{columns}
+\end{frame}
+\begin{frame}[fragile]{Identifying Bugs}
+	\begin{columns}[l]
+		\column{0.75\textwidth}
+		\textbf{How do we know we've found a bug?}
+		\linegap
+
+		Definite bugs
+		\begin{itemize}
+			\item Kernel panics / assertion failures
+			\item Use-after-free
+			\item Deadlock
+		\end{itemize}
+		Probable bugs
+		\begin{itemize}
+			\item Infinite loop
+			\begin{itemize}
+				\item Use structure of execution tree to judge progress.
+			\end{itemize}
+		\end{itemize}
+		\column{0.25\textwidth}
+		\includegraphics[width=\textwidth]{livelock.png}
+	\end{columns}
 \end{frame}
@@
 
-\begin{frame}{Studying Bugs In-Depth} % slide: visual explanation of bug
-	% graphic want to be colored TODO
-\end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-	Emphasis on ``steering'' of test parameters by experts
-	\begin{itemize}
+	Emphasis on users ``steering'' test by changing parameters
+	\begin{itemize}
+		\item Start with coarse granularity, iterate refining decision points
 		\item Perhaps we can automate this steering?
@@
 \begin{frame}{Summary}
-	\textbf{Systematic exploration in kernel-space}
+	\textbf{Systematic testing in kernel-space}
 	\begin{itemize}

### 2012-05-10 defence/defence.tex
minor update
write focusing slides

@@
 \begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
-	\begin{itemize}
-		\item ``Everything interleaves with everything else''?
-		\item What is a ``bug'', anyway?\related{Baumann '09}
-	\end{itemize}
+	``Everything interleaves with everything else''?
+	\begin{itemize}
 		% Say: "we don't want to have to explore every possible interaction between each of these components"
+		\item Impractical to explore resulting state spaces
+	\end{itemize}
+	\linegap
+
+	What is a ``bug'', anyway?\related{Baumann '09}
 \end{frame}
@@
 	\includegraphics[width=\textwidth]{found_a_bug.png}
+	% Say: So landslide had to explore 5 different interleavings to find the bug. But this was a pretty simple test case, and 
 \end{frame}
@@
 % slide: manual shaping - have *user* specify what to focus on (automating intuition)
-\begin{frame}{Limiting Decision Points}
+\begin{frame}{Focusing the Search Space}
+	Recommended decision points: \texttt{mutex\_lock()} and \texttt{mutex\_unlock()}
+	\linegap
+
+	{\bf Problem:} Kernel uses mutexes {\em everywhere}.
+	\begin{itemize}
+		\item Good: \texttt{exit()} calls \texttt{mutex\_lock()}
+		\item Bad: \texttt{exit()} calls \texttt{destroy\_address\_space()} calls \texttt{mutex\_lock()}
+	\end{itemize}
+	\linegap
+	{\bf Solution:} User specifies what modules of the kernel to pay attention to.
+	\begin{itemize}
+		\item \texttt{within\_function exit}
+		\item \texttt{without\_function destroy\_address\_space}
+	\end{itemize}
 \end{frame}
 % slide: exciting teaser "i think this is not too far off!"
-\begin{frame}{Teaser}
+\begin{frame}{Teaser: (Semi-)Automated Search Focusing}
+	Searching for bugs given a stack trace
+	\begin{itemize}
+		\item User knows {\em where} the bug is, but not {\em what}.
+		\item Testing framework can use stack trace to conduct a more focused search.
+	\end{itemize}
 \end{frame}
@@
 \section{Evaluation}
+
+\breakslide{Evaluating Landslide}
 

### 2012-05-12 challenges.tex
begin copyediting

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\chapter{Challenges of Kernel-space}
+\chapter{Challenges of Kernel Space}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In user-space, a systematic exploration tool may use the underlying kernel to help control the concurrent behaviour of the system\cite{dbug-ssv}.
+In user space, a systematic exploration tool may cooperate with the underlying kernel to help control the concurrent behaviour of the system\cite{dbug-ssv}.
 Simple system call invocations can cause a particular thread to run at a particular time, or to block while another thread runs first.
 
-In kernel-space, however, the scheduler is part of the system being tested, and we can no longer always interrupt the execution of a test case to ask the scheduler to instantly start running a different thread.
+In kernel space, however, the scheduler is part of the system being tested, and we can no longer always interrupt the execution of a test case to ask the scheduler to instantly start running a different thread.
 \begin{itemize}
 	\item {\bf Non-preemptibility.} Certain regions of code may be non-preemptible, so a testing tool must know when it is legal to preempt a kernel thread.
-	\item {\bf The context switcher.} A context switch between threads is no longer "instantaneous"; many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this "intermediate state".
-A tool must be aware of the context switch process both to avoid flooding the kernel with interrupt frames that would overflow the stack, and to be able to ignore shared memory conflicts common to all thread transitions (Section~\ref{sec:por-independence}).
-	\item {\bf Runqueue tracking.} The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
+	\item {\bf The context switcher.} A context switch between threads is no longer ``instantaneous'': many instructions must be run between when we decide to preempt and when the next thread begins running, and the tool must be aware of this ``intermediate state''.
+A tool must be aware of the context switch process both to avoid flooding the kernel with interrupt frames that would overflow the stack, and to be able to ignore shared memory conflicts inherent to all thread transitions (Section~\ref{sec:por-independence}).
+	\item {\bf Run-queue tracking.} The kernel's scheduler and the tool must cooperate in some way so that the tool can both know what threads are runnable at every point during execution and cause any given runnable thread to begin running in place of the current one.
 \end{itemize}

### 2012-05-12 related.tex
add title page and convert to book format
add a ton of future work
begin copyediting

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work}
+\chapter{Related Work}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Systematic Exploration}
+\section{Systematic Exploration}
 
@@
 dBug\cite{dbug-ssv} is a systematic testing tool which makes use of DPOR.
-It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \texttt{libc} library calls.
+It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \texttt{libc} and \texttt{pthreads} library calls.
 It uses a message-passing model for inter-thread communication and for establishing a happens-before relationship between transitions.
-The dBug project heavily inspired the development of Landslide.
+The dBug project inspired the development of Landslide.
 
 CHESS\cite{chess} is another userspace systematic testing tool.
-They explore the insight that many races require very few forced preemptions to uncover, and develop a search strategy which prioritises thread interleavings with fewer preemptions.
+The authors explore the insight that many races require very few forced preemptions to uncover, and develop a search strategy which prioritises thread interleavings with fewer preemptions.
 Landslide does not (yet) make use of this strategy, though we did find a related insight which we discuss in Section~\ref{sec:future-backwards}.
@@
 
-RacePRO\cite{racepro} is a tool based in kernel-space which uncovers ``process races'' using systematic exploration of multi-process interactions in user-space.
+RacePRO\cite{racepro} is a tool based in kernel space which uncovers ``process races'' (races involving interleaving system call patterns between multiple processes) using systematic exploration of multi-process interactions in user-space.
 They use system calls, such as filesystem accesses, as points to explore different interleavings of concurrent code, and detect bugs which can corrupt persistent system resources.
@@
 Finally, DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
-Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) - it provides the user with options to constrain decision point identification to certain subsets of the kernel.
+Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) --- it provides the user with options to constrain decision point identification to certain subsets of the kernel.
 
-\subsection{Kernel-level Verification}
+\section{Kernel-level Verification}
 
@@
 Carburizer is able to transform such badly-written code to be more robust.
-Kernel device drivers are a ripe breeding ground for bugs, because it is frequently written by less careful developers, and the sheer volume of code makes it difficult to rigorously review.
+Kernel device drivers are a ripe breeding ground for bugs, because they are frequently written by less careful developers, and the sheer volume of code makes it difficult to rigorously review.
 Due to the interrupt-driven nature of many drivers, they are also prone to concurrency bugs, which Carburizer does not address.
-In Section~\ref{sec:future-linux} we discuss the potential applicability of systematic exploration to concurrency bugs in kernel device drivers.
+In Section~\ref{sec:future-drivers} we discuss the potential applicability of systematic exploration to concurrency bugs in kernel device drivers.
 
-\subsubsection{New Kernel Architectures}
+\subsection{New Kernel Architectures}
 
@@
 Barrelfish\cite{barrelfish} is another novel kernel architecture designed with concurrency in mind.
-The barrelfish kernel relies exclusively on message-passing for inter-thread communication, for increased performance on massively multi-core systems.
+The Barrelfish kernel relies exclusively on message-passing for inter-thread communication, for increased performance on massively multi-core systems.
 While not a verification project, the strict limitation of inter-thread communication makes the concurrent properties of the system easy to reason about in ways that modern production kernels do not provide.
@@
 
-However, in both cases, such steps have little bearing on code that exists in the world today, and it is also important to find ways to verify such sloppily-written code that already has bugs, which is the focus of our work.
+However, in both cases, such steps have little bearing on code that exists in the world today, and it is also important to find ways to verify more imprecisely-written code that already has bugs, which is the focus of our work.
 
-\subsection{Orthogonal Testing Techniques}
+\section{Orthogonal Testing Techniques}
+\label{sec:related-orthogonal}
 
@@
 
-Symbolic execution\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpretating symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
+Symbolic execution\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpreting symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
 Tools such as KLEE\cite{klee} and projects such as Automated Exploit Generation\cite{aeg} demonstrate the effectiveness of symbolic execution in dynamic verification.
@@
 Deterministic multithreading is a different technique for dealing with concurrency bugs which focuses on avoiding them when running production code rather than trying to expose them during testing.
-Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\hspace{0in}\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system and compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
-Deterministic multithreading serves a different purpose than systematic exploration; it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.
+Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\hspace{0in}\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system, compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
+Deterministic multithreading serves a different purpose than systematic exploration: it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.
 

### 2012-05-12 thesis.tex
add title page and convert to book format
add a ton more design
write using chapter
confirm group 2's vanish bug
uncomment testcase appendix
fix minor typoes
begin copyediting

--- 2012-04-24 thesis.tex
@@
-\documentclass[10pt,twocolumn]{article}
+\documentclass[12pt]{book}
 %\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
@@
 \usepackage{draftwatermark}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{multirow}
+\usepackage[usenames,dvipsnames]{xcolor}
 %\usepackage{epsfig}
 \usepackage[bottom]{footmisc}
+\usepackage{mjz-titlepage}
+\usepackage{setspace}
+\usepackage{subfig}
+\setstretch{1.05}
 
@@
 %\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
-\title{\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space}
-\author{Benjamin Blum (\texttt{bblum@andrew.cmu.edu}) \and Garth Gibson (\texttt{garth@cs.cmu.edu})}
-\maketitle
+
+\input{titlepage}
 
@@
 
+\input{acknowledgements}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\chapter{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Race conditions are notoriously difficult to debug.
-Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debug.
+Because of their nondeterministic nature, they frequently do not manifest at all during testing, and when they do manifest, it is difficult to reproduce them reliably enough to collect enough information to help debugging.
 
-In kernel-space, race condition debugging becomes even more difficult.
+In kernel space, race condition debugging becomes even more difficult.
 Many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
@@
 
-Landslide
-\footnote{Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+% TODO: make sure to talk about TAs
+
+Landslide\footnote{
+Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
 is an effort to make easier the process of debugging kernel-space races.
-It is geared towards kernels that meet the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Simics, the x86 simulator that students use to run their Pebbles kernels.
+It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}, the x86 simulator that students use to run their Pebbles kernels.
 During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-When Landslide finds a bug (determined with a set of various checks and heuristics), it stops execution and prints information about the sequence of interleavings that caused the bug to show up.
+When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.
 
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the userspace program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
+With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the user-space program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
 Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
@@
 \footnote{In this work, we equate agents with a kernel's scheduler threads.}
-between {\em decision points} (i.e., actions which the tool and/or the user determine to be relevant to the concurrent behaviour of the test), and attempting to exercise all possible orderings of the transitions in hopes that one or more ordering will expose a bug.
-Within this strategy
+between {\em decision points} (i.e., actions which the tool and/or the user determine to be relevant to the concurrent behaviour of the test), and attempting to exercise all possible orderings of the transitions in hope that one or more ordering will expose a bug.
 
@@
 
-\input{related}
-\input{challenges}
+% TODO: contribution and organisation sections
+
+\input{related} % COMPLETE!
+\input{challenges} % COMPLETE!
+\input{terminology} % almost complete
 \input{design}
+\input{using}
 \input{evaluation}
 \input{futurework}
-\input{acknowledgements}
+\input{conclusion}
+
+% TODO: read presubmit.txt and check everything off
 
@@
 
+\appendix
+\input{testcases}
+
 \end{document}

### 2012-05-12 titlepage.tex
fix typos in thesis
begin copyediting

@@
 \title{\bf Landslide: \\
-  Systematic Dynamic Race Detection in Kernel-space}
+  Systematic Dynamic Race Detection in Kernel Space}
 \author{Ben Blum}
@@
 Garth Gibson, Chair \\
-Dave Eckhardt
+David Eckhardt
 }
@@
 \begin{dedication}
+\begin{tabular}{p{6in}}
+\centering
 There is a fractal nature to all aspects of life.
@@
 
-Our true home, though, lies somewhere in between; a place of potential, and of peace. May we always remember to return to it.
+Our true home, though, lies somewhere in between: a place of potential, and of peace. May we always remember to return to it.
+\end{tabular}
 \end{dedication}

### 2012-05-12 design.tex
copyedit

@@
 
-\subsection{Scheduling}
+\subsection{Landslide Scheduling}
 \label{sec:components-sched}
 
-The scheduler is responsible for keeping track of which threads exist in the guest kernel: which are runnable at any given time, and when they are created and destroyed.
+The Landslide scheduler is responsible for keeping track of which threads exist in the guest kernel: which are runnable at any given time, and when they are created and destroyed.
 
@@
 This fits directly into its model that runnable threads are ones that can be caused to run with a finite number of timer ticks in succession.},
-and the {\em deschedule queue}, which might not correspond to one data structure in the guest kernel, but contains all other threads that exist on the system, which are not runnable for whatever reason.
+and the {\em deschedule queue}, which might not correspond to a data structure in the guest kernel, but contains all other threads that exist on the system, which are not runnable for whatever reason.
 
-The scheduler also tracks which important actions each thread is performing.
+The Landslide scheduler also tracks which important actions each thread is performing.
 These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing, and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
@@
 	\end{algorithmic}
-	\caption{The scheduler's routines for tracking thread lifecycles. These routines are invoked each time the guest kernel calls one of the \texttt{tell\_landslide} annotations.}
+	\caption{Landslide's scheduler's routines for tracking thread lifecycles. These routines are invoked each time the guest kernel calls one of the \texttt{tell\_landslide} annotations.}
 	\label{alg:tell-landslide}
@@
 
-The scheduler relies on the guest kernel informing it about certain important events in the thread lifecycle.
+The Landslide scheduler relies on the guest kernel informing it about certain important events in the thread lifecycle.
 The guest kernel does this by means of annotations (which are described in Section~\ref{sec:using-annotations}).
 Algorithm~\ref{alg:tell-landslide} shows how Landslide handles these annotations internally.\footnote{
-There are also cases for \texttt{tell\_landslide\_on\_rq} and \texttt{tell\_landslide\_off\_rq}, which we omit for brevity.
+There are also cases for the runqueue annotations, \texttt{tell\_landslide\_on\_rq} and \texttt{tell\_landslide\_off\_rq}, which we omit for brevity.
 One important note is that the routine for \texttt{tell\_landslide\_on\_rq} must also check the ``forking'' flag, in case the kernel's \texttt{fork} implementation does not immediately switch to the new thread but instead adds it to the runqueue for later.}
@@
 
-Finally, the scheduler is responsible for managing involuntary preemptions, causing arbitrary threads to begin running in place of the (as chosen at decision points and the ends of branches by the explorer, Section~\ref{sec:components-explore}).
+Finally, the Landslide scheduler is responsible for managing involuntary preemptions, causing arbitrary threads to begin running in place of the current one (as chosen at decision points and the ends of branches by the explorer, Section~\ref{sec:components-explore}).
 
@@
 	\end{algorithmic}
-	\caption{The scheduling algorithm.
-This procedure for updating the scheduler's state is executed once per instruction, with a corresponding value for \texttt{pc} (the program counter) each time.
+	\caption{Landslide's scheduling algorithm.
+This procedure for updating Landslide's state is executed once per instruction, with a corresponding value for \texttt{pc} (the program counter) each time.
 The predicates on \texttt{pc} are part of the kernel instrumentation (Section~\ref{sec:components-kern}).}
@@
 
-In Landslide's view of the world, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+In Landslide's current model, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+(We discuss expanding this limited model in Section~\ref{sec:future-drivers}.) However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
 
@@
 
-Landslide also maintains a set of shared memory accesses made since the last decision point.
+Landslide also maintains a set of shared memory accesses made since the last decision point, for use with Partial Order Reduction (Section~\ref{sec:por-independence}).
 Whenever the guest kernel accesses memory in its heap or in its global data regions, Landslide adds the address of the access to the set, with a flag indicating whether it was a read or a write.

### 2012-05-12 design.tex
continue copyediting

@@
 The user-provided instrumentation, described in detail in Section~\ref{sec:using-instrumenting}, informs Landslide about thread-related lifecycle and scheduling events and the anatomy of the kernel's scheduler.
-	\item {\bf Automatic instrumentation.} All Pebbles kernels have some things in common, and Landslide's build system is able to automatically instrument certain parts of the kernel.
+	\item {\bf Automatic instrumentation.} All Pebbles kernels have some things in common (largely due to the common starter code provided for the class project), and Landslide's build system is able to automatically instrument certain parts of the kernel.
 This instrumentation informs Landslide about the dynamic memory allocator, common library functions such as \texttt{panic}, and certain aspects of the kernel's executable format.
@@
 
-Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point what memory ranges are allocated and what ranges used to be allocated but now are freed.
+Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point which memory ranges are allocated and which ranges used to be allocated but now are freed.
 This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
@@
 
-The arbiter identifies points during execution should count as decision points.
-The selection is mainly controlled by the user, during the annotation and configuration process, but it also automatically identifies {\em voluntary reschedules}, which comprise the ``minimal necessary set'' of decision points.
+The arbiter identifies points during execution that should count as decision points.
+The selection is mainly controlled by the user, during the annotation and configuration process.
+In addition, the arbiter also automatically identifies {\em voluntary reschedules}, which comprise the ``minimal necessary set'' of decision points.
 
@@
 The explorer maintains a representation of the current branch of the decision tree.
-It is responsible for checkpointing at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.
+It is responsible for checkpointing the state of both Landslide and the guest kernel at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.
 
@@
 
-Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the execution state at decision points during the test.
+Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the guest kernel's execution state at decision points during the test.
 
@@
 % A healthy dose of snark.
-Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
 In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
@@
 
-In order to establish a happens-before relation between transitions, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point.
+The happens-before relation expresses for each pair of transitions whether executing the first one is required to ``enable'' the second one.
+In order to establish the relation, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point (Section~\ref{sec:components-explore}).
 

### 2012-05-12 evaluation.tex
continue copyediting

@@
 We evaluated Landslide in two ways.
-First, we conducted a user study with present students of 15-410, to understand the eases and difficulties of Landslide's user interface process, to understand how to polish it into a useful debugging tool.
+First, we conducted a user study with students of 15-410 during the spring 2012 semester, to understand the ease and difficulty of Landslide's user interface process, to understand how to polish it into a more useful debugging tool.
 Second, we did a case study of six bugs in two kernels, investigating them in-depth to understand how systematic testing might best be harnessed to find bugs without in advance knowing the decision set necessary to expose them.
@@
 
-To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project, and had them use Landslide with their own kernels.
+To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project in spring 2012, and had them use Landslide with their own kernels.
 In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.\footnote{
@@
 		The majority of groups needed to change their scheduling around the idle thread in order to meet the requirement that the kernel never run idle when progress could otherwise be made (Sections~\ref{sec:challenges-design} and \ref{sec:using-requirements-sched}).
-		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the provided allocator.
+		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the allocator that 15-410 provides.
 \end{itemize}
@@
 Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree.
-Of those groups, two found race conditions in their kernel.
+(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.) Of those groups, two found race conditions in their kernel.
 Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.

### 2012-05-12 futurework.tex
fix minor typoes
continue copyediting

@@
 	\end{center}
-	\caption{Ex-members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
+	\caption{Former members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
 \end{figure*}
@@
 The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving''.
-Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.
+Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.\footnote{
+In 15-410, before the kernel project, students also implement a user-space thread library.
+Just as Landslide could help the students during the kernel project, so too could a user-space systematic exploration tool help during the thread library project.}
 
@@
 In Section~\ref{sec:related-orthogonal}, we mentioned the technique of symbolic execution for dynamic verification.
-We believe that symbolic execution could be combined with systematic explorationto help find races in certain obscure code paths that systematic exploration by itself might not even execute.
+We believe that symbolic execution could be combined with systematic exploration to help find races in certain obscure code paths that systematic exploration by itself might not even execute.
 
@@
 		\hline
-		Backtracks & {\bf 1,511} & {\bf 1,401} \\
+		Backtracks & {\bf 1,401} & {\bf 1,511} \\
 		\hline

### 2012-05-12 terminology.tex
continue copyediting

@@
 \section{Scheduling Terms}
+
+The following terms refer to parts of the guest kernel's execution.
 
@@
 
+The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.
+
 \resume{enumerate}
@@
 		A point during execution (between two consecutive instructions, precisely) which is deemed ``interesting'' in terms of the likelihood that thread switches at that point will cause concurrency bugs to arise.
-All thread switches occur at a decision point; some are voluntary reschedules caused by the kernel and others are involuntary preemptions caused by Landslide.
+Every thread switch occurs at a decision point; some are voluntary reschedules caused by the kernel and others are involuntary preemptions caused by Landslide.
 	\item {\bf Transition}:
@@
 	\item {\bf Branch}:
-		One particular execution of the test case; a set of decision points with exactly one transition in between each of them, characterising a single interleaving of threads.
+		One particular execution of the test case; a set of decision points with exactly one transition between each of them, characterising a single interleaving of threads.
 	\item {\bf Interleaving}:

### 2012-05-12 using.tex
continue copyediting

@@
 
-\subsection{VM}
+\subsection{Virtual Memory}
 
-Landslide does not have much to do with the VM, but the kernel must direct-map most of kernel memory, including the heap, globals, and kernel thread stacks.
+Landslide does not have much to do with the virtual memory system, but the kernel must direct-map most of kernel memory, including the heap, globals, and kernel thread stacks.
 Landslide needs this to read values out of memory and to identify memory conflicts.\footnote{
@@
 \label{sec:using-runqueue}
-Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the scheduler's runqueue at each point during execution.
+Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the guest kernel's scheduler runqueue at each point during execution.
 Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not the kernel keeps the current thread on the runqueue or off of it.

### 2012-05-12 design.tex
more

@@
 
-In future work (Section~\ref{sec:future-new}), the arbiter may also automatically identify extra decision points, such as on conflicting shared memory accesses.
+In future work (Section~\ref{sec:future-new}), the arbiter may also automatically identify extra decision points, such as conflicting shared memory accesses.
 
@@
 At each decision point, Landslide creates a new node in the decision tree.
-It stores the TID of the thread that was chosen, the state of the scheduler (what threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.
+It stores the TID of the thread that was chosen, the state of the scheduler (which threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.
 
@@
 	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
-		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process could have yet to reap a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well.
+		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process might not yet have reaped a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well.
 For these cases an additional check is needed.
@@
 
-The states in the test lifecycle, and the rules for transitioning between them are as follows.
+The states in the test lifecycle, and the rules for transitioning between them, are as follows.
 
@@
 	\item {\bf Test cleanup.} The test processes have exited, but some housekeeping may yet remain. When the the kernel is idling and the shell is waiting for keyboard input, advance state.
-	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack.
+	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack (to step 3).
 \end{enumerate}
@@
 If not, Landslide proclaims the access to be illegal.\footnote{In the same way, Landslide also detects calls to \texttt{free} on blocks that were already freed or never allocated at all.}
-	\item {\bf Deadlock bugs.} If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
+	\item {\bf Deadlock bugs.} If Landslide finds no runnable threads on the runqueue,\footnote{Except for idle, if it exists.} or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
 \suspend{enumerate}
@@
 Hence, if the idle thread is on the runqueue (and/or the current thread), it is only runnable if no other threads are runnable.
-	\item {\bf Yielding mutexes.} In kernels whose mutexes (or other synchronisation idioms) leave threads on the runqueue when they are notionally blocked, such threads are not runnable.
+	\item {\bf Yielding mutexes.} In kernels whose mutexes (or other synchronisation idioms) leave threads on the runqueue when they are notionally blocked, Landslide classifies such threads as not runnable.
 \end{itemize}
@@
 We say that $X$ {\em enables} $Y$ if $T_Y$ was not runnable immediately before $X$, but was runnable immediately after.
-Then, a transition $A$ happens-before a transition $B$ if $A$ is the same thread as $B$, if $A$ enables $B$, or if $A$ happens-before some transition $C$ that enables $B$.
+Then, a transition $A$ happens-before a transition $B$ if $T_A$ is the same thread as $T_B$, if $A$ enables $B$, or if $A$ happens-before some transition $C$ that enables $B$.
 
@@
 	\end{tabular}
-	\caption{This sample code defeats a DPOR implementation that only uses the first evil ancestor if a decision point is not defined between the two lines of thread 1's code.
+	\caption{This sample code defeats a DPOR implementation that uses only the first evil ancestor if a decision point is not defined between the two lines of thread 1's code.
 The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next.
@@
 
-Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of only identifying one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.
+Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.
 
-While DPOR is a proved-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment.
-We justify the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave proving soundness to future work (and to the theoreticians).
+While DPOR is a proven-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment.
+We advocate the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave investigatin soundness to future work.
 
@@
 
-In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context-switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
+In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
 When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.

### 2012-05-12 using.tex
more

@@
 For deadlock detection, if the kernel has mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:using-annotations}.
-If mutexes explicitly deschedule blocking threads, no special annotations are needed.
+If mutexes explicitly deschedule blocking threads, no special annotations are needed, because the other annotations do the job.
 
@@
 
-The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation.
+The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation, to avoid having to manually instrument a custom allocator.
 
@@
 
-In order to run with Landslide, the kernel must be able to boot up to the shell prompt\footnote{
-In Pebbles, the bootup sequence is usually fairly simple, involving the init process spawning the shell.},
+In order to run with Landslide, the kernel must be able to boot up to the shell prompt,\footnote{
+In Pebbles, the bootup sequence is usually fairly simple, involving the init process spawning the shell.}
 receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes.
@@
 
-Here we describe the interface by which Landslide understands the execution of the guest kernel.
+Here we document the interface by which Landslide understands the execution of the guest kernel.
 Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.
@@
 \subsubsection{Required Annotations}
+
+Some of these annotations may need to be used in ``sensitive'' regions of scheduler code, where it may not necessarily be clear at what point during the operation the annotation should go.
+Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range..
+
 \begin{itemize}

### 2012-05-13 using.tex
more

@@
 Some of these annotations may need to be used in ``sensitive'' regions of scheduler code, where it may not necessarily be clear at what point during the operation the annotation should go.
-Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range..
+Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range.
 
@@
 	\item \texttt{tell\_landslide\_decide()} - Call this to define additional decision points. See section~\ref{sec:using-decision}.
-	\item \texttt{assert()} or \texttt{panic()} - You should be using these already, but they are extra useful in Landslide.
+	\item \texttt{assert()} or \texttt{panic()} -
 		The more important invariants for which a kernel has \texttt{assert}s, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
@@
 Some of the fields are required information, and some are tweaks that change the way Landslide behaves.
-We omit discussion of the required fields, and explain the optimisations and behaviour tweaks that Landslide supports in Section~\ref{sec:using-search}.
+We omit discussion of the required fields, and explain the optimisations and behaviour tweaks that Landslide supports using this file in Section~\ref{sec:using-search}.
 
@@
 
+Both of these functions are predicates on the guest kernel's execution state (usually on the contents of its scheduler's data structures).
+Example implementations are given in Figure~\ref{fig:student-c}.
+
+\begin{figure}[h]
+	% FIXME: replace with an image
+	\centering
+	\begin{framed}
+	{\small \begin{verbatim}
+	bool kern_current_extra_runnable(conf_object_t *cpu)
+	{
+	    int tcb = READ_MEMORY(cpu, GUEST_CURRENT_TCB_POINTER);
+	    int state_flag = READ_MEMORY(cpu, tcb + GUEST_TCB_STATE_FLAG_OFFSET);
+	    return state_flag == GUEST_TCB_RUNNABLE;
+	}
+	bool kern_ready_for_timer_interrupt(conf_object_t *cpu)
+	{
+	    int x = READ_MEMORY(cpu, GUEST_PREEMPTION_FLAG);
+	    return x == GUEST_PREEMPTION_ENABLED;
+	}
+	\end{verbatim}
+	}
+	\end{framed}
+	\caption{Example implementations of the two functions in \texttt{student.c}.}
+	\label{fig:student-c}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
         \item Use \texttt{tell\_landslide\_decide()} to indicate decision points.\footnote{
-	We recommend at the start of \texttt{mutex\_lock} and at the end of \texttt{mutex\_unlock}, though it is also important to use one's own intuition, depending on what system call(s) are being tested.}
+	We recommend at the start of \texttt{mutex\_lock} and at the end of \texttt{mutex\_unlock}, though it is also important to use one's own intuition, depending on which system call(s) are being tested.}
         \item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.

### 2012-05-13 design.tex
add changebars

@@
 
-The happens-before relation expresses for each pair of transitions whether executing the first one is required to ``enable'' the second one.
-In order to establish the relation, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point (Section~\ref{sec:components-explore}).
+\revision{The happens-before relation expresses for each pair of transitions whether executing the first one is required to ``enable'' the second one.
+In order to establish the relation,} Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point (Section~\ref{sec:components-explore}).
 

### 2012-05-13 evaluation.tex
add changebars

@@
 Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree.
-(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.) Of those groups, two found race conditions in their kernel.
+\revision{(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.)} Of those groups, two found race conditions in their kernel.
 Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.

### 2012-05-13 futurework.tex
add changebars

@@
 Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.\footnote{
-In 15-410, before the kernel project, students also implement a user-space thread library.
-Just as Landslide could help the students during the kernel project, so too could a user-space systematic exploration tool help during the thread library project.}
+\revision{In 15-410, before the kernel project, students also implement a user-space thread library.
+Just as Landslide could help the students during the kernel project, so too could a user-space systematic exploration tool help during the thread library project.}}
 

### 2012-05-13 related.tex
add changebars

@@
 dBug\cite{dbug-ssv} is a systematic testing tool which makes use of DPOR.
-It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \texttt{libc} and \texttt{pthreads} library calls.
+It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \revision{\texttt{libc} and \texttt{pthreads}} library calls.
 It uses a message-passing model for inter-thread communication and for establishing a happens-before relationship between transitions.
@@
 
-RacePRO\cite{racepro} is a tool based in kernel space which uncovers ``process races'' (races involving interleaving system call patterns between multiple processes) using systematic exploration of multi-process interactions in user-space.
+RacePRO\cite{racepro} is a tool based in kernel space which uncovers ``process races'' \revision{(races involving interleaving system call patterns between multiple processes)} using systematic exploration of multi-process interactions in user-space.
 They use system calls, such as filesystem accesses, as points to explore different interleavings of concurrent code, and detect bugs which can corrupt persistent system resources.

### 2012-05-13 terminology.tex
add changebars

@@
 
-The following terms refer to parts of the guest kernel's execution.
+\revision{The following terms refer to parts of the guest kernel's execution.}
 
@@
 
-The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.
+\revision{The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.}
 

### 2012-05-13 using.tex
add changebars

@@
 For deadlock detection, if the kernel has mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:using-annotations}.
-If mutexes explicitly deschedule blocking threads, no special annotations are needed, because the other annotations do the job.
+\revision{If mutexes explicitly deschedule blocking threads, no special annotations are needed, because the other annotations do the job.}
 
@@
 
-The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation, to avoid having to manually instrument a custom allocator.
+The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation, \revision{to avoid having to manually instrument a custom allocator.}
 
@@
 
-Some of these annotations may need to be used in ``sensitive'' regions of scheduler code, where it may not necessarily be clear at what point during the operation the annotation should go.
-Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range.
+\revision{Some of these annotations may need to be used in ``sensitive'' regions of scheduler code, where it may not necessarily be clear at what point during the operation the annotation should go.
+Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range.}
 
@@
 
-Both of these functions are predicates on the guest kernel's execution state (usually on the contents of its scheduler's data structures).
-Example implementations are given in Figure~\ref{fig:student-c}.
+\revision{Both of these functions are predicates on the guest kernel's execution state (usually on the contents of its scheduler's data structures).
+Example implementations are given in Figure~\ref{fig:student-c}.}
 
@@
 	\item \texttt{ignore\_sym} - The user may wish to ignore memory accesses from other global data structures as well.
-For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that better reflects what is actually being tested for.
+For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
 	\item \texttt{within\_function} - If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{

### 2012-05-13 evaluation.tex
more

@@
 
-We evaluated Landslide in two ways.
+\revision{In evaluating Landslide, we sought to answer seven overall questions.
+
+\begin{itemize}
+	\item Is Landslide an effective use of time by a developer seeking to find and understand race conditions in their code? How can Landslide be improved to improve the time tradeoff?
+	\item When testing kernel code with a systematic testing framework, what overall use patterns are most effective for getting meaningful results?
+How can future work employ such patterns to build more sophisticated systematic testing tools for kernels?
+\end{itemize}
+
+To answer these questions, we evaluated Landslide in two ways.}
 First, we conducted a user study with students of 15-410 during the spring 2012 semester, to understand the ease and difficulty of Landslide's user interface process, to understand how to polish it into a more useful debugging tool.
@@
 
-We wished to measure which phases of using Landslide are most time-consuming, so we asked students approximately every half-hour to record the amount of time they had spent doing each phase:
+We wished to measure which phases of using Landslide are most time-consuming, so we asked students \revision{in intervals between 30 and 60 minutes} to record the amount of time they had spent doing each phase:
 
@@
 
-We also wanted descriptive feedback, to evaluate what specific parts of the process of using Landslide need improved.
+We also wanted descriptive feedback, to evaluate which specific parts of the process of using Landslide need improved.
 We asked students to write brief remarks whenever they recorded amounts of time spent, and also after they were finished to answer the following questions.
@@
 One student reported the output was easy to follow by ``tracing the life of the buggy thread'', which is an insight into the learning process we should capitalise on.
-	\item {\bf Ease of configuring decision sets.} Most students seemed to find the process of adding more decision points intuitive. One reported, ``quick and effective at detecting basic bugs/races.''
+	\item {\bf Ease of configuring decision sets.} Most students seemed to find the process of adding more decision points intuitive.
+One reported, ``quick and effective at detecting basic bugs/races.'' \revision{It could always be improved, but of all the parts of the user interface, this process seems to need it least.}
 	\item {\bf General feature wishlist.} The following features, of varying difficulty to implement, would have improved user experience overall.
@@
 		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the allocator that 15-410 provides.
+
+		\revision{The idle requirement is an inherent part of Landslide's model, whereas the dynamic allocator requirement could be replaced with a richer set of optional annotations.}
 \end{itemize}
@@
 \begin{itemize}
-	\item One student bypassed the \texttt{kern\_ready\_for\_timer\_interrupt} annotation by disabling their scheduler's preemption-disabling mechanism and replacing it with explicit disabling of interrupts, thereby making every timer interrupt cause a context switch.
+	\item One student bypassed the \texttt{kern\_ready\_for\_timer\_interrupt} annotation by disabling their scheduler's preemption-disabling mechanism and replacing it with explicit disabling of interrupts.
+\revision{This caused the kernel to context switch every timer interrupt, so Landslide did not need to be aware of extra conditions that would cause the scheduler to ignore timer ticks.}
 	\item One ex-TA was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
+\revision{We did not advertise this method because we suspected students might make more mistakes instrumenting this way, but it might be worth advocating anyway for the sake of making the interface simpler.}
 \end{itemize}
@@
 They found one race and a problem with idle.
-		After fixing this bug, they were able to completely explore both test cases with decision points on both \texttt{mutex\_lock} and \texttt{mutex\_unlock}, which we consider a strong negative result (i.e., no bug found with fine-grained interleavings).
+		After fixing these bugs, they were able to completely explore both test cases with decision points on both \texttt{mutex\_lock} and \texttt{mutex\_unlock}, which we consider a strong negative result (i.e., no bug found in those two test cases with fine-grained interleavings).
 	\item % margaret/echo
@@
 They found a problem with idle, and after fixing it, found a vanish-related race condition in the fix.
-They also found a bug in an assertion which they added during the process, but that bug was deterministic, so we consider this part of the ``fixing incorrect instrumentation'' process.
+They also found a bug in an assertion which they had added during the process, but that bug was deterministic, so we consider this part of the ``fixing incorrect instrumentation'' process.
 	\item % tpassaro/?
@@
 \begin{itemize}
-	\item {\bf Too many waiters allowed.} Using the \texttt{double\_wait} test case, group 1 found a bug in which more threads invoking \texttt{wait} would be allowed to block than the number of child processes that could ever be reaped, and some of the waiters would end up blocking forever.
-The fix for this was two lines of code, and it could have been found with an in-depth code review.
+	\item {\bf Too many waiters allowed.} Using the \texttt{double\_wait} test case, Group 1 found a bug in which more threads invoking \texttt{wait} would be allowed to block than the number of child processes that could ever be reaped, and some of the waiters would end up blocking forever.
+The fix for this was two lines of code.
+\revision{The group claimed it could have been found with an ``in-depth'' code review, though we claim Landslide's contribution here was still non-trivial, because (as exhibited here) such bugs exist anyway for lack of sufficient review.}
 	\item {\bf Accidental return to a vanished thread.} Group 2 had a problem with the scheduler running idle when it was not supposed to; when they fixed it, however, their implementation would return to a thread that had already vanished in the case when no thread was left to run.
@@
 
-We also note that though group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a subtle race in the reparenting section of their \texttt{vanish} code using the \texttt{vanish\_vanish} test.
+We also note that, though Group 2 stopped using Landslide after getting it to explore the minimal decision tree, if they had configured decision points around \texttt{mutex\_lock} (which was the recommended next step in the process), they would have found a subtle race in the reparenting section of their \texttt{vanish} code using the \texttt{vanish\_vanish} test.
 We verified this by having an ex-TA re-instrument their kernel and add the decision points: from the time of exploring a minimal decision tree to finding the bug with the new decision set took half an hour of work, and fully understanding the nature of the bug took an hour on top of that.
@@
 \begin{itemize}
-	\item {\bf Inappropriate idling.} Two groups had errors in their keyboard handling code, where when a newline was received when the kernel was idle, a thread blocked on \texttt{readline} would be added to the runqueue and but not switched directly to, so the kernel would continue idling even if there was useful work to do.
+	\item {\bf Inappropriate idling.} Two groups had errors in their keyboard handling code, where when a newline was received when the kernel was idle, a thread blocked on \texttt{readline} would be added to the runqueue and but not switched directly to, so the kernel would continue idling even though there was useful work to do.
 Another group left idle on the scheduler runqueue, and had no special-case code to skip over it, so would interleave idle arbitrarily with other threads.
@@
 	\item {\bf Deterministic use-after-free.} One group had a deterministic use-after-free bug in their \texttt{wait} implementation, in which they freed something and immediately dereferenced it.
-This hadn't presented a problem because the block was never reallocated during the critical window, which would have taken a complicated and fine-grained interleaving, but Landslide detected the use-after-free immediately, without the need for any interleaving.
+This hadn't presented a problem because the block was never reallocated during the critical window, which would have taken a complicated and fine-grained interleaving, but Landslide's heap checker detected the use-after-free immediately, without the need for any interleaving.
 \end{itemize}
@@
 When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{cond\_queue\_next} pointer.
-However, when woken, the condition variable code did not clear the value of this pointer.
+However, when awakened, the condition variable code did not clear the value of this pointer.
 Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
-Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing - if the woken thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
+Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing - if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
 Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}
@@
 This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple\footnote{
-Although many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the bug may never be noticed when using such libraries.
-(This caused the author some trouble.)},
+Many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the kernel bug may never be noticed when using such libraries (this caused the author some trouble).},
 and the fix is only two lines long; the dereference should simply occur before making the child thread runnable.

### 2012-05-13 evaluation.tex
more

@@
 
-\revision{In evaluating Landslide, we sought to answer seven overall questions.
+\revision{In evaluating Landslide, we sought to answer two overall questions.
 
@@
 This results in init infinitely looping around failing calls to \texttt{wait}.
-To catch this bug, we added an assert statement in the \texttt{wait} failure case that the current thread was not init (because init should always have children to reap or wait for), which Landslide then caused to trip.
+To catch this bug \revision{immediately, rather than relying on infinite loop detection,} we added an assert statement in the \texttt{wait} failure case that the current thread was not init (because init should always have children to reap or wait for), which Landslide then caused to trip.
 
@@
 	\item All Landslide trial times include the Simics start-up and kernel boot-up time (time between issuing the command and the test case beginning to run), roughly 15 seconds for POBBLES and 10 seconds for LudicrOS.
-	\item All Landslide trials were run on the Gates-Hillman cluster machines (2.6 GHz Xeon).
+	\item All Landslide trials were run on the Gates-Hillman cluster machines (2.6 GHz Xeon; \revision{four cores, though only one was used; 8 GB RAM}).
 	\item All Landslide trials were run with ``backwards exploration'' enabled (Section~\ref{sec:using-search}).
@@
 	\item {\bf Finding a bug is fast, if it exists.} As especially exhibited in the POBBLES vanish/vanish (b) case, if two sets of decision points generate trees of approximately equal size, but one tree contains a bug and the other doesn't, then searching the bugful tree will likely terminate much more quickly.
-	Of course, it is always possible that a bug may only exist in the very last branch of a tree, but we found that in general bugs show up early during exploration.
+	Of course, it is always possible that a bug may exist only in the very last branch of a tree, but we found that in general bugs show up early during exploration.
 \end{enumerate}

### 2012-05-13 futurework.tex
more

@@
 We believe that a tool such as Landslide could change the way students learn such skills for the better, beyond simply increasing their likelihood of finding bugs during testing.
-The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving''.
-Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.\footnote{
+The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving.'' Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.\footnote{
 \revision{In 15-410, before the kernel project, students also implement a user-space thread library.
@@
 \begin{enumerate}
-	\item It should be aware of the types and interfaces associated with the kernel's synchronisation primitives (mutexes, condvars, semaphores, etc), and be able to treat operations on those as ``guaranteed to work'' (similar to Section~\ref{sec:por-independence}).
+	\item It should be aware of the types and interfaces associated with the kernel's synchronisation primitives (mutexes, condvars, semaphores, etc), and be able to treat operations on those as ``guaranteed to work'' (similar to Section~\ref{sec:por-independence}).\footnote{
+	\revision{If such operations were instead themselves incorrect, it would mean that data race detection might fail to identify certain racy accesses.
+The technique would then still identify some racy accesses, but might miss others.}}
 	\item Next, it should recognise acquire and release (or sleep and wake) operations on synchronisation primitives, and use that information to track lockset-type information.
@@
 Major benefit could be achieved by redesigning Landslide to work in a virtualised environment instead of a simulated one.
-This would pose several challenges, because Landslide would no longer be running in ``god mode'', having immediate access to every single instruction and memory access executed.
+This would pose several challenges, because Landslide would no longer be running in ``\revision{omniscient} mode'', having immediate access to every single instruction and memory access executed.
 It would, however, still be able to modify the kernel's memory and address space, which would be important to harness for increased control over the kernel's execution.
@@
 Finally, Landslide will need to find new ways to control non-deterministic events in the system.
-Since virtualised kernels use the same timer interrupt as the host they run on, Landslide is unlikely to be able to intercept it.
-Instead, we may replace a guest kernel's timer-driven context switcher with a ``Landslide-driven'' context switcher, leaving the timer handler only to do other work.
+Since virtualised kernels use the same timer interrupt as the host they run on, Landslide \revision{will need to intercept it, and perhaps control the guest kernel's context switching with a custom ``Landslide-driven'' context switcher instead.}
 
@@
 
-One member of group 1 from the user study fully explored decision trees from \texttt{double\_wait} and \texttt{vanish\_vanish} (Section~\ref{sec:eval-victory}), and found an oddity in the exploration time for a particular decision set.
+One member of Group 1 from the user study fully explored decision trees from \texttt{double\_wait} and \texttt{vanish\_vanish} (Section~\ref{sec:eval-victory}), and found an oddity in the exploration time for a particular decision set.
 

### 2012-05-13 testcases.tex
more

@@
 
-The test cases that use the \texttt{thread\_fork} system call are written in assembly, because thread creation can only safely be done from C using a userspace thread library.
-However, using a userspace thread library for thread creation might forcibly serialise certain operations that would expose kernel bugs only if concurrent, and also necessitates many more system calls, which dramatically increases the size of the decision tree.
+The test cases that use the \texttt{thread\_fork} system call are written in assembly, because thread creation can only safely be done from C using a user-space thread library.
+However, using a user-space thread library for thread creation might forcibly serialise certain operations that would expose kernel bugs only if concurrent, and also necessitates many more system calls, which dramatically increases the size of the decision tree.
 

### 2012-05-13 evaluation.tex
more

@@
 	\item {\bf Ease of configuring decision sets.} Most students seemed to find the process of adding more decision points intuitive.
-One reported, ``quick and effective at detecting basic bugs/races.'' \revision{It could always be improved, but of all the parts of the user interface, this process seems to need it least.}
+One reported, ``quick and effective at detecting basic bugs/races.'' \revision{While this interface could always be improved, of all the parts of the user experience, this process seems to need it least.}
 	\item {\bf General feature wishlist.} The following features, of varying difficulty to implement, would have improved user experience overall.

### 2012-05-13 thesis.tex
more
add changebars
write intro

@@
 \usepackage{mjz-titlepage}
+\usepackage{framed}
 \usepackage{setspace}
+\setstretch{1.05}
 \usepackage{subfig}
-\setstretch{1.05}
+\usepackage{changebar}
 
@@
 
+\setcounter{changebargrey}{35}
+\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
+%\newcommand\revision[1]{#1}
+
 \begin{abstract}
@@
 
+\tableofcontents
 \input{acknowledgements}
@@
 
+\revision{
+Many techniques exist for dynamic testing of concurrent systems for race conditions.
+Systematic exploration\cite{chess}, the strategy we focus on in this work, involves making educated guesses at what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
+Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
+Compared to other dynamic analyses, such as data race detection\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
+
 In kernel space, race condition debugging becomes even more difficult.
-Many aspects of the concurrency implementation itself are part of the system being tested, and may themselves have race-inducing flaws.
+Many aspects of the concurrency implementation itself are part of the system being tested.
+While user-space testing frameworks rely on the underlying kernel to provide common concurrency abstractions which the framework can use to drive the test, systematic exploration in the kernel itself necessarily exists underneath the level of such abstractions.
+Furthermore, a tool that supports testing multiple kernels must support many different possible scheduler designs.
 
-% TODO: talk about systematic exploration
-%It provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
-
-% TODO: make sure to talk about TAs
+\begin{quote}
+\bf
+Systematic exploration in kernel space is made possible by instrumenting the kernel's internal abstractions to understand its concurrency, and made efficient by relying on the user to control the scope of the test.
+\end{quote}
+}
 
@@
 Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make easier the process of debugging kernel-space races.
+is an effort to make possible the use of systematic exploration in kernel-space.
 It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}, the x86 simulator that students use to run their Pebbles kernels.
@@
 
-% TODO: this paragraph is incongruous in light of the above three; what audience are you writing for, anyway?
-Many techniques exist for dynamic testing of concurrent systems for race conditions.
-Systematic exploration, the strategy that Landslide uses, involves modelling a concurrent test case as a series of {\em transitions} by each of a set of {\em agents}
-\footnote{In this work, we equate agents with a kernel's scheduler threads.}
-between {\em decision points} (i.e., actions which the tool and/or the user determine to be relevant to the concurrent behaviour of the test), and attempting to exercise all possible orderings of the transitions in hope that one or more ordering will expose a bug.
-
-In this work, we present Landslide as a tool and framework for finding race conditions in kernels using systematic exploration, implemented as a module for the Simics simulator and geared towards Pebbles, the specification implemented by students in 15-410 at CMU.
-We study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications, and present techniques (some sound, and some heuristic) for addressing them.
-% We study the relationship between the developer and the tool during the debugging process, in terms of what information the tool needs the developer to provide beforehand in order to provide meaningful test results.
-% Compared to conventional race condition debugging, we show that the relationship Landslide provides makes it a better strategy.
-We evaluate Landslide's usefulness in helping students of 15-410 better understand and solve the concurrency problems while implementing a Pebbles kernel, and also as a testing framework for TAs to use while grading student submissions.
-We show that Landslide is an effective tool for both purposes.
-
-% TODO: contribution and organisation sections
+\revision{
+In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
+We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).
+Finally, we discuss possible avenues for future work in education, use of additional testing techniques, application to general purpose kernels, long-running testing approaches, and theoretical study of systematic exploration (Chapter~\ref{sec:future}).
+}
 

### 2012-05-14 design.tex
write intro
rewrite soundness

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:design}
 
@@
 
-\subsection{Landslide Scheduling}
+\subsection{Scheduling}
 \label{sec:components-sched}
@@
 
-\subsection{Coarse Granularity}
+\subsection{Soundness}
 
-\cite{dpor} and \cite{dbug-retreat} present a partial order reduction algorithm in which only the first ``evil ancestor'' (term defined in \cite{dbug-retreat}) of each transition need be considered when identifying which alternate interleavings need to be explored.
+\revision{
+Because we recommend using Landslide with coarse-grained decision sets, as opposed to mandating a decision point between every pair of shared memory accesses, it is possible for Landslide to overlook race conditions that require finer-grained interleavings to expose.
+In this way, Landslide's search is not ``sound'', but instead false-negative-oriented.
+
+We do, however, claim that Landslide's search is sound in a different way: if the provided decision points are sufficient to express an interleaving that would expose a race, then Landslide will find it.
+That is, if Landslide is exploring an execution tree with bugs in some branches, it will eventually find a bug.
+
+Dynamic Partial Order Reduction works by identifying ``evil ancestors'' for each transition in a branch after executing that branch.
+Intuitively, a transition's evil ancestor is another transition such that executing the test with the order of those two transitions reversed could cause different concurrent behaviour to arise.}
+\cite{dpor} and \cite{dbug-retreat} present a partial order reduction algorithm in which only the first evil ancestor of each transition need be considered when identifying which alternate interleavings need to be explored.
 
@@
 			\cline{2-4}
-			$C$ & & & \texttt{x=42;} \\
+			$A$ & \texttt{x=42;} & & \\
 			$B$ & & \texttt{y=31337;} & \\
-			\multirow{2}{*}{$A_1$} & \texttt{if (y==31337)} & & \\
-			& \texttt{assert(x==42);} & & \\
+			\multirow{2}{*}{$C_1$} & & & \texttt{if (y==31337)} \\
+			& & & \texttt{assert(x==42);} \\
 			\cline{2-4}
@@
 			\cline{2-4}
-			$C$ & & & \texttt{x=42;} \\
-			$A_2$ & \texttt{if (y==31337)} & & \\
+			$A$ & \texttt{x=42;} & & \\
+			$C_2$ & & & \texttt{if (y==31337)} \\
 			$B$ & & \texttt{y=31337;} & \\
@@
 			$B$ & & \texttt{y=31337;} & \\
-			\multirow{2}{*}{$A_3$} & \texttt{if (y==31337)} & & \\
-			& \texttt{assert(x==42);} & & \\
-			$C$ & & & \texttt{x=42;} \\
+			\multirow{2}{*}{$C_3$} & & & \texttt{if (y==31337)} \\
+			& & & \texttt{assert(x==42);} \\
+			$A$ & \texttt{x=42;} & & \\
 			\cline{2-4}
@@
 	\end{tabular}
-	\caption{This sample code defeats a DPOR implementation that uses only the first evil ancestor if a decision point is not defined between the two lines of thread 1's code.
-The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next.
-In the second interleaving, $C$ is no longer a second evil ancestor of $A_2$, so the third interleaving is never explored, and the assertion failure is missed.}
+	\caption{This sample code defeats a DPOR implementation that uses only the first evil ancestor if a decision point is not defined between the two lines of thread 3's code.
+The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $C_1$, and runs the second interleaving next, reordering them.
+In the second interleaving, $A$ is no longer a second evil ancestor of $C_2$, so the third interleaving is never explored, and the assertion failure is missed.}
 	\label{fig:evil-ancestors}
@@
 
-We found that implementing DPOR in this way was not sound in our setup, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
-Because Landslide builds its decision trees with coarser granularity, i.e.
-multiple different shared memory conflicts may happen in each transition, Landslide must consider every evil ancestor of each transition when performing DPOR.
+We found that implementing DPOR in this way violated our second notion of soundness, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
+\revision{However, a key point of Landslide's recommended usage pattern is to sacrifice maximally-fine-grained decision points for performance, and so multiple different shared memory conflicts may happen in each transition.}
+Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given our usage pattern.
 
-Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.
-
-While DPOR is a proven-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment.
-We advocate the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave investigatin soundness to future work.
+\revision{Instead, Landslide uses considers every evil ancestor of each transition when performing DPOR.
+This modification enables Landslide to avoid the problem presented in the figure, because it will identify both alternate interleavings after exploring the first one.
+We claim without certainty that this change restores the second notion of soundness to Landslide's search, and leave proving this property to future work.}
 

### 2012-05-14 evaluation.tex
write intro
more

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:evaluation}
 
@@
 \revision{This caused the kernel to context switch every timer interrupt, so Landslide did not need to be aware of extra conditions that would cause the scheduler to ignore timer ticks.}
-	\item One ex-TA was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
+	\item One former 15-410 Teaching Assistant (TA) was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
 \revision{We did not advertise this method because we suspected students might make more mistakes instrumenting this way, but it might be worth advocating anyway for the sake of making the interface simpler.}

### 2012-05-14 design.tex
get rid of dashes

@@
 In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
-When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
+When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace}: a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear.
 Figure~\ref{fig:found_a_bug} depicts a simple decision trace for a use-after-free bug.

### 2012-05-14 evaluation.tex
get rid of dashes

@@
 \revision{(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.)} Of those groups, two found race conditions in their kernel.
-Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
+Additionally, Landslide helped all four groups find deterministic bugs: three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
 
@@
 Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
-Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing - if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
+Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing: if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
 Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}
@@
 This bug is moderately complicated, requiring a preemption during that specific window of the first thread's execution during which the second thread must exit entirely and have its TCB freed.
-The fix is moderately complicated - some protection mechanism must be designed so that threads cannot exit after being looked-up by TID.
+The fix is moderately complicated: some protection mechanism must be designed so that threads cannot exit after being looked-up by TID.
 
@@
 
-	We believe it is reasonable to test for these bugs in this way - using the minimal set of system calls to be paid attention to as necessary to find the bug - because it follows the recommended workflow of using Landslide, which is to start with what the user judges to be the ``smallest relevant set'' of decision points.
+	We believe it is reasonable to test for these bugs in this way (using the minimal set of system calls to be paid attention to as necessary to find the bug) because it follows the recommended workflow of using Landslide, which is to start with what the user judges to be the ``smallest relevant set'' of decision points.
 The configuration using \texttt{within\_function} was as follows.
@@
 not missing any bugs that exist in the resulting tree).\footnote{
-When running LudicrOS yield/vanish with decision points on \texttt{mutex\_unlock} but not on \texttt{mutex\_lock}, we found that the ordering invariant failed - ``backwards'' exploration found no bug, while ``forwards'' exploration did.
+When running LudicrOS yield/vanish with decision points on \texttt{mutex\_unlock} but not on \texttt{mutex\_lock}, we found that the ordering invariant failed: ``backwards'' exploration found no bug, while ``forwards'' exploration did.
 We attribute this to a bug in Landslide itself, and present the results for the backwards exploration as usual, in which Landslide found no bug.}
@@
 	\item {\bf Fewer is faster.} While it is theoretically possible that a tree built of finer-grained interleavings might encounter a bug after fewer overall backtracks, we found that this did not happen in practice.\footnote{
-		The one suspicious case is the LudicrOS vanish/vanish bug - exploring the ``both'' tree found the bug faster than exploring the ``unlock'' tree, despite the latter's decision points being a subset of the former's.
+		The one suspicious case is the LudicrOS vanish/vanish bug.
+Exploring the ``both'' tree found the bug faster than exploring the ``unlock'' tree, despite the latter's decision points being a subset of the former's.
 However, we see that exploring the ``lock'' tree found the bug faster than either, so overall the ``both'' tree did not outperform the fastest of the smaller trees.}

### 2012-05-14 futurework.tex
get rid of dashes

@@
 Most notably among the information Landslide has at its disposal is the collection of conflicting shared memory accesses among transitions.
-Theoretically, to identify a decision point immediately before and after each such access would generate an execution tree with perfect granularity (i.e., exactly one shared memory access per transition), and thence find every possible race condition - this is, of course, the opposite extreme to the current setup.
+Theoretically, to identify a decision point immediately before and after each such access would generate an execution tree with perfect granularity (i.e., exactly one shared memory access per transition), and thence find every possible race condition, which would be the opposite extreme to the current setup.
 
@@
 
-The event-based nature of the \texttt{tell\_landslide} annotations can still be harnessed in a virtualised environment - they would need to be converted into hypercalls, which would briefly transfer control to Landslide at key execution points to enable it to update its state machine and potentially influence the kernel's execution.
+The event-based nature of the \texttt{tell\_landslide} annotations can still be harnessed in a virtualised environment.
+They would need to be converted into hypercalls, which would briefly transfer control to Landslide at key execution points to enable it to update its state machine and potentially influence the kernel's execution.
 

### 2012-05-14 related.tex
get rid of dashes

@@
 Finally, DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
-Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) --- it provides the user with options to constrain decision point identification to certain subsets of the kernel.
+Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.
 

### 2012-05-14 thesis.tex
get rid of dashes

@@
 \bf
-Systematic exploration in kernel space is made possible by instrumenting the kernel's internal abstractions to understand its concurrency, and made efficient by relying on the user to control the scope of the test.
+Systematic exploration in kernel space is possible by understanding a kernel's concurrency through instrumentation of its internal abstractions, and can be made efficient by relying on the user to control the scope of the test.
 \end{quote}
@@
 
-Landslide\footnote{
-Landslide {\em (n)} - A phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-is an effort to make possible the use of systematic exploration in kernel-space.
+We present Landslide,\footnote{
+Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+a tool for applying systematic race detection techniques in to kernel-level code.
 It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}, the x86 simulator that students use to run their Pebbles kernels.

### 2012-05-14 using.tex
get rid of dashes

@@
 	\small
-	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in the context switcher, when a new thread is switched to.
-	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when the scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations.
-	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
+	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)}: Call this in the context switcher, when a new thread is switched to.
+	\item \texttt{tell\_landslide\_sched\_init\_done()}: Call this when the scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations.
+	\item \texttt{tell\_landslide\_forking()}: Call this when a new thread is being forked.
 		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)
 
-		Call it ``just before'' the action which makes the new thread runnable - whether the kernel just adds it to the runqueue (in which case the next annotation called would be \texttt{thread\_on\_rq}) or begins running it immediately (in which case \texttt{thread\_switch}); Landslide handles both cases.
-	\item \texttt{tell\_landslide\_vanishing()} - Call this when a thread is about to go away.
+		Call it ``just before'' the action which makes the new thread runnable.
+It does not matter whether the kernel just adds it to the runqueue (in which case the next annotation called would be \texttt{thread\_on\_rq}) or begins running it immediately (in which case \texttt{thread\_switch}); Landslide handles both cases.
+	\item \texttt{tell\_landslide\_vanishing()}: Call this when a thread is about to go away.
 		Call it ``just before'' the relevant context switch; i.e., the one that should never return.
-	\item \texttt{tell\_landslide\_sleeping()} - Call this when a thread is about to go to sleep (that's \texttt{sleep()} the system call).
+	\item \texttt{tell\_landslide\_sleeping()}: Call this when a thread is about to go to sleep (that's \texttt{sleep()} the system call).
 		Call it ``just before'' the relevant context switch; i.e., the one that won't return for as long as the thread requested to sleep.
-	\item \texttt{tell\_landslide\_thread\_on\_rq(int tid)} - Call this when a thread is about to be added to the runqueue.
+	\item \texttt{tell\_landslide\_thread\_on\_rq(int tid)}: Call this when a thread is about to be added to the runqueue.
 		(Make sure this call is done within whatever protection is also used for the runqueue modification itself.)
-	\item \texttt{tell\_landslide\_thread\_off\_rq(int tid)} - Call this when a thread is about to be removed from the runqueue. (Same protection clause as above applies.)
-	\item Mutex annotations - These are only necessary if the kernel uses mutexes that leave blocked threads on the runqueue, rather than explicitly descheduling them.
+	\item \texttt{tell\_landslide\_thread\_off\_rq(int tid)}: Call this when a thread is about to be removed from the runqueue. (Same protection clause as above applies.)
+	\item Mutex annotations. These are only necessary if the kernel uses mutexes that leave blocked threads on the runqueue, rather than explicitly descheduling them.
 	\begin{itemize}
-		\item \texttt{tell\_landslide\_mutex\_locking(void *mutex\_addr)} - Call this when a thread is about to start locking a mutex.
+		\item \texttt{tell\_landslide\_mutex\_locking(void *mutex\_addr)}: Call this when a thread is about to start locking a mutex.
 Landslide needs the mutex address for deadlock detection, to track which mutex each thread is blocked on.
-		\item \texttt{tell\_landslide\_mutex\_locking\_done()} - Call this when a thread finishes locking a mutex (and now owns it).
-		\item \texttt{tell\_landslide\_mutex\_unlocking(void *mutex\_addr)} - Call this when a thread is about to start unlocking a mutex.
-		\item \texttt{tell\_landslide\_mutex\_unlocking\_done()} - Call this when a thread finishes unlocking a mutex (and no longer owns it).
-		\item \texttt{tell\_landslide\_mutex\_blocking(int owner\_tid)} - Call this when a thread is contending on a locked mutex, and hence is no longer logically ``runnable'' yet still on the runqueue.
+		\item \texttt{tell\_landslide\_mutex\_locking\_done()}: Call this when a thread finishes locking a mutex (and now owns it).
+		\item \texttt{tell\_landslide\_mutex\_unlocking(void *mutex\_addr)}: Call this when a thread is about to start unlocking a mutex.
+		\item \texttt{tell\_landslide\_mutex\_unlocking\_done()}: Call this when a thread finishes unlocking a mutex (and no longer owns it).
+		\item \texttt{tell\_landslide\_mutex\_blocking(int owner\_tid)}: Call this when a thread is contending on a locked mutex, and hence is no longer logically ``runnable'' yet still on the runqueue.
 Landslide needs to know who owns the mutex for deadlock detection.
@@
 	\small
-	\item \texttt{tell\_landslide\_decide()} - Call this to define additional decision points. See section~\ref{sec:using-decision}.
+	\item \texttt{tell\_landslide\_decide()}: Call this to define additional decision points. See section~\ref{sec:using-decision}.
 	\item \texttt{assert()} or \texttt{panic()} -
-		The more important invariants for which a kernel has \texttt{assert}s, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
+		The more important invariants for which a kernel has \texttt{assert}s, the more likely Landslide is to find bugs that would trigger them; otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
 \end{itemize}
@@
 	\small
-        \item \texttt{bool kern\_current\_extra\_runnable(conf\_object\_t *cpu)} - See Section~\ref{sec:using-runqueue}.
+        \item \texttt{bool kern\_current\_extra\_runnable(conf\_object\_t *cpu)}: See Section~\ref{sec:using-runqueue}.
 This says whether the current thread is ``logically runnable'' despite not being on the runqueue itself (that is, \texttt{tell\_landslide\_on\_rq} won't have been called for it).
-        \item \texttt{bool kern\_ready\_for\_timer\_interrupt(conf\_object\_t *cpu)} - This function can be used to express when the scheduler is ``locked'' in a way not involving disabled interrupts.
+        \item \texttt{bool kern\_ready\_for\_timer\_interrupt(conf\_object\_t *cpu)}: This function can be used to express when the scheduler is ``locked'' in a way not involving disabled interrupts.
 Landslide will avoid trying to preempt the kernel whenever this is false.
@@
 	\small
-	\item \texttt{sched\_func} - When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them.
+	\item \texttt{sched\_func}: When functions that make up the kernel's scheduler are identified, Landslide knows to ignore shared memory accesses from them.
 Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.
-	\item \texttt{ignore\_sym} - The user may wish to ignore memory accesses from other global data structures as well.
+	\item \texttt{ignore\_sym}: The user may wish to ignore memory accesses from other global data structures as well.
 For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
-	\item \texttt{within\_function} - If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{
+	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{
 For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.}
 This configuration option allows the user to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.
-	\item \texttt{without\_function} - Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
+	\item \texttt{without\_function}: Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
 \end{itemize}
@@
 	\small
-	\item \texttt{EXPLORE\_BACKWARDS} - In which order to explore the branches?
+	\item \texttt{EXPLORE\_BACKWARDS}: In which order to explore the branches?
 Backwards means with more preemptions first; forwards means tending to let the current thread keep running more often.
@@
 (Section~\ref{sec:future-backwards})
-	\item \texttt{DECISION\_INFO\_ONLY} - This option makes Landslide stop after one branch, and output the list of decision points it identified using the current config.
+	\item \texttt{DECISION\_INFO\_ONLY}: This option makes Landslide stop after one branch, and output the list of decision points it identified using the current config.
 Useful to see all decision points, and tweak settings to get rid of frivolous ones.
@@
 \begin{itemize}
-        \item \texttt{vanish\_vanish} - Tests when a parent and child process \texttt{vanish()} simultaneously.
-        \item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child.
-        \item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in one process vanishing.
-        \item \texttt{yield\_vanish} - Tests for interactions between \texttt{yield()} and \texttt{vanish()}.
+        \item \texttt{vanish\_vanish}: Tests when a parent and child process \texttt{vanish()} simultaneously.
+        \item \texttt{double\_wait}: Tests interactions of multiple waiters on a single child.
+        \item \texttt{double\_thread\_fork}: Tests for interactions of multiple threads in one process vanishing.
+        \item \texttt{yield\_vanish}: Tests for interactions between \texttt{yield()} and \texttt{vanish()}.
 \end{itemize}

### 2012-05-14 thesis.tex
add a citation

@@
 a tool for applying systematic race detection techniques in to kernel-level code.
-It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}, the x86 simulator that students use to run their Pebbles kernels.
+It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
 During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.

### 2012-05-14 challenges.tex
write intro
add model section

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:challenges}
 
@@
 \end{itemize}
-
-In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts.
-In future work (Section~\ref{sec:future-drivers}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
 
@@
 	\item {\bf Recognising blocked threads.} It is difficult to detect when threads are ready to do useful work.
-Even if notionally runnable, they may be waiting in a yield loop (Sections~\ref{sec:using-annotations} and \ref{sec:future-linux}).
+Even if notionally runnable, they may be waiting in a yield loop (Sections~\ref{sec:using-annotations} and \ref{sec:future-linux}), which can be difficult to discern without an understanding of whatever shared resource the thread is waiting for.
 A tool that did not recognise this might decide to attempt to run that thread over and over, and get stuck because the other threads would never be allowed to make progress.
 	\item {\bf Thread lifecycle tracking.} It is also difficult to demarcate threads' lifecycles: when a new thread gets created, when is it available to run?
-and when a thread is exiting, at what point does it stop running code (Section~\ref{sec:using-annotations})?
+When a thread is exiting, at what point does it stop running code (Section~\ref{sec:using-annotations})?
 In user-space, these boundaries are defined by the system call instructions, but in kernel-space, the boundary is fuzzier.
 	\item {\bf Use of virtual memory.} The kernel's complete control over the machine's virtual memory system means a system for tracking memory accesses must be aware of changing virtual memory mappings (Section~\ref{sec:using-requirements}).
+A tool not aware of virtual memory might falsely identify whether or not two transitions' memory accesses conflict.
 \end{itemize}
 
-\section{Kernel Design}
+\section{Kernel Design Independence}
 \label{sec:challenges-design}
@@
 \end{itemize}
-
-Conversely, Landslide needs to make some assumptions about scheduling behaviour, in order for its control over scheduling to work:
-
-\begin{itemize}
-	\item {\bf Timer ticks control ``runnable'' threads.} With the exception of a non-preemptible ``scheduler lock'', and yield-looping mutexes (both of which must be instrumented by the user), a thread's presence on the runqueue indicates that a finite number of timer interrupts in succession will eventually cause it to run.
-	\item {\bf No idling when progress can be made.} The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
-Landslide uses the idle loop to detect when a test begins/ends and to detect when all threads in a test are wedged.
-\end{itemize}

### 2012-05-14 design.tex
add model section

@@
 
+In this chapter we describe Landslide's implementation in detail.
+Section~\ref{sec:model} outlines the testing model on top of which Landslide's exploration mechanisms are built.
+Section~\ref{sec:components} describes the individual components within Landslide.
+Section~\ref{sec:techniques-bugs} describes Landslide's metrics for identifying bugs.
+Section~\ref{sec:techniques-por} describes how Landslide achieves effective state space reduction.
+Section~\ref{sec:techniques-feedback} describes the debugging feedback Landslide provides to the user when bugs are encountered.
+
+\revision{
+\section{Landslide's View of the World}
+\label{sec:model}
+
+Landslide's model for applying systematic exploration in kernel-space is comprised of four key points, which we overview here.
+
+\subsection{Simulated Execution}
+\label{sec:model-simics}
+
+Landslide is implemented as a module for Simics\cite{simics}, a full-system x86 emulator.
+When running the kernel, Simics calls into Landslide once every time the kernel executes an instruction or performs a memory read or write.
+Landslide uses this information, in conjunction with the user-provided instrumentation, to maintain its internal representations of the state of the guest kernel.
+
+Landslide makes use of Simics ``bookmarks'', a feature which enables checkpointing and restoring the execution state of the guest kernel, to implement backtracking when the end of each branch in the decision tree is reached.
+
+\subsection{Timer-Driven Scheduling}
+\label{sec:model-timer}
+
+In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts.
+Landslide assumes that timer interrupts are the only source of nondeterminism for the guest kernel, so controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
+In future work (Section~\ref{sec:future-drivers}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
+
+Landslide needs to impose some requirements on the guest kernel's scheduling behaviour, in order for its control over scheduling to work:
+
+\begin{itemize}
+	\item {\bf Timer ticks control ``runnable'' threads.} With the exception of a non-preemptible ``scheduler lock'', and yield-looping mutexes (both of which must be instrumented by the user), a thread's presence on the runqueue indicates that a finite number of timer interrupts in succession will eventually cause it to run.
+
+	Landslide treats the sleep queue no differently from the runqueue when deciding which threads are runnable.
+The guest kernel will treat it differently, but since sleeping for a predetermined amount of time is never an appropriate way to solve race conditions, Landslide treats sleeping threads as notionally runnable.
+This fits directly into its model that runnable threads are ones that can be caused to run with a finite number of timer ticks in succession.
+	\item {\bf No idling when progress can be made.} The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
+Landslide uses the idle loop to detect when a test begins/ends (test lifecycle tracking) and to detect when all threads in a test are wedged (bug detection).
+\end{itemize}
+
+Section~\ref{sec:components-inflight} gives more detail about how Landslide abstracts raw timer interrupts into the ability to cause arbitrary threads to run.
+
+\subsection{False-Negative-Oriented Bug Detection}
+\label{sec:model-bugs}
+
+% A healthy dose of snark.
+Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
+If Landslide does report a bug, though, it is almost certainly correct.
+Section~\ref{sec:techniques-bugs} details the conditions Landslide uses to identify when something has gone wrong.
+
+\subsection{User-Assisted State Space Reduction}
+\label{sec:model-user}
+
+In addition to Dynamic Partial Order Reduction (DPOR) for automated state space reduction, Landslide relies on the user's guidance to help mitigate the possible combinatorial explosion of thread interleaving possibilities.
+This manifests in two ways:
+
+\begin{itemize}
+	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby only generating thread interleavings around points within those modules.
+	\item {\bf Memory conflict selection.} Landslide also provides a mechanism for ignoring shared memory conflicts on certain common data structures (especially those accessed in every thread transition, such as the scheduler queues), to enable DPOR to achieve greater reduction.
+Intuitively, this represents sacrificing the ability to find races around those data structures to make searching other components more feasible.
+\end{itemize}
+
+We advocate relying on the user for such information because the user need only understand the basics of their kernel design (which is difficult for Landslide to guess) while not knowing in advance what bugs are being looked for (which Landslide is able to provide).
+We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user ``steers'' Landslide towards focused search spaces that are more likely to find bugs.
+
+The interfaces by which the user provides this information are documented in Section~\ref{sec:using-customise}.
+}
+
 \section{Components of Landslide}
+\label{sec:components}
 
@@
 It maintains a ``mirror image'' of the guest kernel's scheduler state in the form of three queues, a pointer to the currently-running thread, and a pointer to the previously-running thread.
-The queues are the {\em runqueue}, containing the runnable threads, the {\em sleep queue}, containing threads which become runnable after a certain number of timer ticks\footnote{
-Landslide treats the sleep queue no differently from the runqueue when deciding which threads are runnable.
-The guest kernel will treat it differently, but since sleeping for a predetermined amount of time is never an appropriate way to solve race conditions, Landslide treats sleeping threads as notionally runnable.
-This fits directly into its model that runnable threads are ones that can be caused to run with a finite number of timer ticks in succession.},
-and the {\em deschedule queue}, which might not correspond to a data structure in the guest kernel, but contains all other threads that exist on the system, which are not runnable for whatever reason.
+The queues are the {\em runqueue}, containing the runnable threads, the {\em sleep queue}, containing threads which become runnable after a certain number of timer ticks, and the {\em deschedule queue}, which might not correspond to a data structure in the guest kernel, but contains all other threads that exist on the system, which are not runnable for whatever reason.
 
@@
 
-In Landslide's current model, timer interrupts are the only source of nondeterminism for the guest kernel, and controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-(We discuss expanding this limited model in Section~\ref{sec:future-drivers}.) However, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
+Though we define timer interrupts as the only source of non-determinism in our environment, it is more useful to view the concurrent behaviour with a higher-level abstraction, in terms of the set of runnable threads and the ability to preempt the currently-running thread with any different runnable one.
 
@@
 For these cases an additional check is needed.
-	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.\footnote{
-		If the kernel is idling and the shell is not waiting for keyboard input, it is likely that the test case has deadlocked.}
+	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.
 	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
@@
 
-\section{On Simics}
-
-Landslide is implemented as a module for Simics\cite{simics}.
-When running the kernel, Simics calls into Landslide once every time the kernel executes an instruction or performs a memory read or write.
-Landslide uses this information, in conjunction with the user-provided instrumentation, to maintain its internal representations of the state of the guest kernel.
-
-Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the guest kernel's execution state at decision points during the test.
-
 \section{Identifying Bugs}
@@
 
-% A healthy dose of snark.
-Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
 In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.
 
-Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
-If Landslide does report a bug, though, it is almost certainly correct (except in cases as listed below).
-
-\subsection{Accurate Bug-Detection Conditions}
+\subsection{Definite Bug-Detection Conditions}
 
@@
 
-\subsection{Uncertain Bug-Detection Conditions}
+\subsection{Probable Bug-Detection Conditions}
 
@@
 We found that implementing DPOR in this way violated our second notion of soundness, because it relies on the invariant that each transition contains only one inter-thread communication event (in kernel-space, this means a shared memory access).
-\revision{However, a key point of Landslide's recommended usage pattern is to sacrifice maximally-fine-grained decision points for performance, and so multiple different shared memory conflicts may happen in each transition.}
-Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given our usage pattern.
+\revision{However, a key point of Landslide's recommended usage pattern is to sacrifice maximally-fine-grained decision points for performance, and so multiple different shared memory conflicts may happen in each transition.} Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given our usage pattern.
 
@@
 \section{Debugging Feedback}
+\label{sec:techniques-feedback}
 

### 2012-05-14 evaluation.tex
add one space

@@
 This bug was also found during grading, and took substantially less time to verify by hand than the previous bug.
-When a thread does \texttt{yield} to a specific other TID, the other thread's thread control block (TCB)is not protected between lookup-by-TID and yield-to.
+When a thread does \texttt{yield} to a specific other TID, the other thread's thread control block (TCB) is not protected between lookup-by-TID and yield-to.
 If that thread exits during that time, a use-after-free (and worse, a completely invalid schedule to a nonexistent thread) results.

### 2012-05-14 challenges.tex
more copyedit

@@
 
-In certain avenues of future work (Section~\ref{sec:future-linux}), remaining agnostic of most implementation details becomes less of an issue.
+% In certain avenues of future work (Section~\ref{sec:future-linux}), remaining agnostic of most implementation details becomes less of an issue.
 For example, if we target Landslide to Linux, it is unlikely that core scheduler or interrupt-handling behaviour will change for as long as Landslide remains useful for finding bugs.
@@
 \begin{itemize}
-	\item {\bf Runqueues}: Does the kernel store the currently-running thread on the runqueue, or is it removed and stored separately?
-	\item {\bf Mutexes}: When a thread blocks on a mutex, is it left on the runqueue in a yield-loop, or is it explicitly descheduled?
+	\item {\bf Runqueues}: Does the kernel store the currently-running thread on the runqueue, or is it ``checked out'' while running and stored separately?
+	\item {\bf Mutexes}: When a thread blocks on a mutex, is it left on the runqueue in a yield loop, or is it explicitly descheduled (e.g., moved from the runqueue to another queue)?
 In yield-looping mutexes, when do blocked threads become ``unblocked'', notionally?
@@
 Do just-forked threads begin life through the usual context-switch-return path, or is there a special path for that?
-	\item {\bf Test Lifecycle}: Even apart from running on a particular input that the user wishes to test, the kernel may need to perform extra work irrelevant to the test, such as initialisation (e.g.
-the bootup process) and housekeeping (e.g.
-cleaning up dead processes).
-Which parts of this work should be included or excluded in the test?
+	\item {\bf Test Lifecycle}: Apart from running on a particular input that the user wishes to test, the kernel may need to perform extra work irrelevant to the test, such as initialisation (e.g., the bootup process) and housekeeping (e.g., cleaning up dead processes).
+Which parts of this work should be included in or excluded from the test?
 \end{itemize}

### 2012-05-14 related.tex
more copyedit

@@
 
-Landslide distinguishes itself from related systematic testing tools in its field as more of an exploratory work, presenting techniques to make systematic testing compatible with kernel-level code; less so as a development on the core testing approach itself.
+Landslide is distinguished from related systematic testing tools in its field as more of an exploratory work, presenting techniques to make systematic testing compatible with kernel-level code; less so as a development on the core testing approach itself.
 

### 2012-05-14 thesis.tex
add model section
more copyedit

--- 2012-05-14 thesis.tex
@@
 \setcounter{changebargrey}{35}
-\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
+%\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
+\newcommand\revision[1]{\cbstart#1\cbend}
 %\newcommand\revision[1]{#1}
@@
 We discuss the techniques Landslide uses to address the general challenges of kernel-level concurrency, and we evaluate its effectiveness and usability as a debugging aid.
-We show that our techniques make systematic testing in kernel-space feasible, and that Landslide is a useful tool for doing so in the context of 15-410.
+We show that our techniques make systematic testing in kernel-space feasible and that Landslide is a useful tool for doing so in the context of 15-410.
 \end{abstract}
@@
 a tool for applying systematic race detection techniques in to kernel-level code.
-It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and implemented as a module for Wind River Simics\textsuperscript{\texttrademark}, the x86 simulator that students use to run their Pebbles kernels.
+It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
 During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.

### 2012-05-15 evaluation.tex
break it down now

@@
 
-Finally, we briefly studied each bug that the students found, which we describe below.
+Finally, we briefly studied each bug that the students found: whether they were deterministic or race conditions, whether they were deep unsolved concurrency problems or simple oversights, and the complexity of the necessary fixes.
+
+\revision{Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree.
+(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.)}
 
@@
 
-This section will be written after the draft is submitted. Hang on tight, my friends.
-% TODO
+\newcommand\asdf[1]{\hspace{-0.05in}\footnotesize\bf{#1}\hspace{-0.05in}}
+\revision{
+\begin{figure}[t]
+	\begin{center}
+	\small
+	\begin{tabular}{|l||c|c|c|c|c|c||c|c|}
+		\hline
+		\multirow{2}{*}{\bf Group} & \multicolumn{6}{|c||}{\bf Minutes spent doing\dots} & \multicolumn{2}{|c|}{\bf Total time spent} \\
+		\cline{2-9}
+		& \asdf{annotating} & \asdf{config} & \asdf{student.c} & \asdf{fixing} & \asdf{customising} & \asdf{found races} & \asdf{required} & \asdf{refinement} \\
+		\hline \hline
+		% ntaha/jmchow
+		\multirow{2}{*}{Group 1} & 35 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 5 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} \\
+		\cline{2-9}
+		& 5 & 30 & 10 & 100 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 145 & {\footnotesize \em N/A} \\
+		\hline
+		% echo/margaret
+		\multirow{2}{*}{Group 2} & 20 & 30 & 10 & 80 & {\footnotesize \em N/A} & 10 & 140 & 10 \\
+		\cline{2-9}
+		& 15 & 45 & 30 & 60 & {\footnotesize \em N/A} & 30 & 150 & 30 \\
+		\hline
+		% tpassaro
+		\multirow{2}{*}{Group 3} & 15 & 10 & 5 & 60 & 30 & 30 & 90 & 60 \\
+		\cline{2-9}
+		& 15 & 10 & 5 & 60 & 30 & 30 & 90 & 60 \\
+		\hline
+		% pjumde
+		\multirow{2}{*}{Group 4} & 10 & 42 & 3 & 5 & 30 & {\footnotesize \em N/A} & 60 & 30 \\
+		\cline{2-9}
+		& 15 & 35 & {\footnotesize \em N/A} & 5 & 35 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 35 \\
+		\hline
+		% apodolsk
+		Group 5 & 50 & 15 & {\footnotesize \em N/A} & 15 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} \\
+		\hline
+		Former TA & 40 & 20 & 3 & 95 & {\footnotesize \em N/A} & 30 & 158 & 30 \\
+		\hline
+		\hline
+		{\bf Average } & {\bf 22 } & {\bf 26.33 } & {\bf 9.43 } & {\bf 48.5 } & {\bf 31.25 } & {\bf 26 } & {\bf 119 } & {\bf 36.43} \\
+		\hline
+	\end{tabular}
+	\end{center}
+	\caption{Self-reported times spent by users on each phase while using Landslide.
+``N/A'' indicates a student reported spending no time on a particular phase.
+``Required'' indicates the sum of the times from the first four phases; ``refinement'' inducates the sum from the last two phases.
+	``N/A'' indicates that a user did not report spending any time on a phase; the ``required'' field is ``N/A'' if any of its represented phases are, and ``N/A'' fields are not reflected in the averages.}
+	\label{fig:student-times}
+\end{figure}
+
+While the students were using Landslide, we asked them at intervals from 30 to 60 minutes to record how much time they had spent on each phase of instrumenting and using Landslide.
+
+\begin{itemize}
+	\item {\bf Required instrumentation.} These phases represent work necessary to getting Landslide to work with a kernel at all.
+	\begin{itemize}
+		\item {\bf Annotating the kernel.} In this phase, the student modifies their kernel to use the provided annotations (Section~\ref{sec:using-annotations}).
+		\item {\bf Writing the configuration file.} In this phase, the student fills out certain information about their kernel in \texttt{config.landslide} (Section~\ref{sec:using-config-landslide}).
+		\item {\bf Instrumenting within Landslide.} In this phase, the student implements the two instrumentation functions in \texttt{student.c} (Section~\ref{sec:using-student-c}).
+		\item {\bf Fixing instrumentation problems/crashing.} In this phase, the user repairs problems that arose due to incorrect instrumentation (or bugs in Landslide itself, in some cases).
+	\end{itemize}
+	\item {\bf Extra refinement.} These phases represent additional work to be done after successfully exploring a minimal decision tree.
+When such minimal exploration finds no bugs, such additional work is necessary to provide meaningful results.
+	\begin{itemize}
+		\item {\bf Customising decision points.} In this phase, the user configures Landslide to use decision sets more refined than the minimal one, in search of hard-to-find bugs (Section~\ref{sec:using-customise}).
+		\item {\bf Investigating found race conditions.} In this phase, the user attempts to understand and fix bugs that Landslide found and reported.
+	\end{itemize}
+\end{itemize}
+
+Table~\ref{fig:student-times} presents the time breakdown among phases that students experienced while using Landslide.
+On average, students spent two hours getting Landslide to work with their kernels, and slightly more than half an hour customising the search and/or studying found bugs on top of that.
+
+We consider the absolute values of these numbers largely artifacts of the user interface's quality (i.e., they could be improved with extra interface polishing, though they also currently present an upper bound).
+The relative values among phases, however, serve to give a notion of which phases were most difficult.
+
+We do not think it useful to consider these times amortised over how many bugs were found for each group, because in general we do not know whether each group would have found more bugs with marginal extra effort, having already paid the initial cost of instrumenting.
+(The vanish/vanish bug we describe finding ourselves in Section~\ref{sec:eval-victory-races} serves as an example of this perennial possibility.)
+Instead, we claim that the fact that Landslide found bugs for all groups who completed the required instrumentation proves its baseline effectivness as a debugging tool, and we note that the time tradeoff will become more worthwhile the more the user interface improves.
+}
 
@@
 
-Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree.
-\revision{(A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.)} Of those groups, two found race conditions in their kernel.
+Of the four groups that completed the required instrumentation, two found race conditions in their kernel.
 Additionally, Landslide helped all four groups find deterministic bugs: three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.
@@
 \subsubsection{Race Conditions Found}
+\label{sec:eval-victory-races}
 

### 2012-05-15 evaluation.tex
ta edit

@@
 To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project in spring 2012, and had them use Landslide with their own kernels.
+\revision{Additionally, one former 15-410 Teaching Assistant (TA) participated in the study.}
+
 In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.\footnote{
@@
 \revision{This caused the kernel to context switch every timer interrupt, so Landslide did not need to be aware of extra conditions that would cause the scheduler to ignore timer ticks.}
-	\item One former 15-410 Teaching Assistant (TA) was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
+	\item The former TA was able to avoid the \texttt{kern\_current\_extra\_runnable} annotation by using the \texttt{on\_rq} and \texttt{off\_rq} annotations (Section~\ref{sec:using-annotations}) to express the ``abstract set of runnable threads'' (which included the currently-running thread, even though the runqueue didn't) instead of the literal runqueue itself.
 \revision{We did not advertise this method because we suspected students might make more mistakes instrumenting this way, but it might be worth advocating anyway for the sake of making the interface simpler.}

### 2012-05-16 design.tex

@@
 
-A voluntary reschedule is identified when:
-\begin{itemize}
-	\item The previously-running thread entered the context switcher without first entering the timer interrupt handler,
-\end{itemize}
+In Landslide, it is always necessary to identify a decision point during a voluntary reschedule, to maintain the invariant that each transition between decision points is comprised of the execution of only one thread.
+\revision{In future work, we may also investigate the implications of allowing some transitions to consist of multiple threads' executions, which would have applications for heuristic state space reduction.}
 
-and {\em any one} of the following:
-\begin{itemize}
-	\item The currently-running thread was handling a timer interrupt, and exits the {\em timer interrupt handler}.
-	\item The currently-running thread was not handling a timer interrupt, and exits the {\em context switcher}.
-	\item The currently-running thread was {\em just forked} (i.e, it has never run before), and hence may never exit either the timer handler or context switcher.
-\end{itemize}
+\revision{
+Because voluntary reschedules happen of the guest kernel's own volition, when a thread begins one, it will necessarily not have been preempted by the timer handler.
+Hence, we identify voluntary reschedules when a thread enters the context switcher without having entered the timer handler.\footnote{
+Landslide actually identifies voluntary reschedules at the point when the reschedule ends, so the {\em previous} thread would not have entered the timer handler, and the current thread may or may not have.
+We do this to avoid the complexity of dealing with the possibility that the guest kernel would have interrupts disabled at the beginning of the reschedule and not re-enable them until after the thread switch.}
+}
 
-It is always necessary to identify a decision point during a voluntary reschedule, to maintain the invariant that each transition between decision points is comprised of the execution of only one thread.
+%\begin{itemize}
+%	\item The previously-running thread entered the context switcher without first entering the timer interrupt handler,
+%\end{itemize}
+%
+%and {\em any one} of the following:
+%\begin{itemize}
+%	\item The currently-running thread was handling a timer interrupt, and exits the {\em timer interrupt handler}.
+%	\item The currently-running thread was not handling a timer interrupt, and exits the {\em context switcher}.
+%	\item The currently-running thread was {\em just forked} (i.e, it has never run before), and hence may never exit either the timer handler or context switcher.
+%\end{itemize}
 

### 2012-05-16 conclusion.tex
write conclusion; fix intro

@@
 
+\revision{
+Systematic exploration is a powerful technique for exposing race conditions in concurrent systems.
+%It is more reliable than stress testing, and able to find more diverse types of bugs than data race detection.
+In previous research, systematic exploration has proven useful in user-level and distributed systems, yet a gap stands between the technique and the complex world of kernel-level concurrency debugging.
+In this thesis, we present techniques for bridging this gap.
+
+We have built Landslide, a tool for performing systematic exploration in kernel space with a focus on Pebbles kernels written for 15-410.
+We showed that Landslide's techniques make systematic exploration in kernel space both possible and efficient, and that Landslide can help students find bugs in their own kernels.
+
+With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case, and second, in the user's configuration of Landslide to express which parts of the kernel are ``interesting'' and which are irrelevant.
+Finding and understanding race conditions exposed by a given test becomes a joint effort between the user and Landslide, combining the user's knowledge about the kernel's design and Landslide's ability to explore many interleavings efficiently.
+
+Our work on Landslide has indicated many possible avenues for continued work on systematic exploration in kernel space.
+We see potential for Landslide to be a useful debugging tool both for students of 15-410 and for developers of general purpose operating systems.
+Application to mainstream kernels also opens up the possibility to apply systematic testing techniques to device driver code, known for being prone to concurrency errors.
+Finally, the current emphasis on user-guided steering of test parameters suggests long-running testing approaches which could automate a user's intuition for how to find meaningful results quickly.
+
+We hope that the ideas presented here serve as stepping stones to develop more sophisticated systematic debugging techniques for kernel-level race conditions.
+}
+
 \newcommand{\s}{\fontfamily{jkpvos}\selectfont{}s}
 
-% TODO write this for real
+\begin{quote}
+\em
 And lo, the Author did{\s}t pre{\s}ent Land{\s}lide to yon re{\s}earch Community, and to the {\s}tudents of XV-CDX as well.
@@
 The {\s}oftware then{\s}eforth worked as Intended, cra{\s}hing nevermore, and there was much rejoicing.
+\end{quote}

### 2012-05-16 thesis.tex
write conclusion; fix intro

@@
 It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
-During execution of a kernel, Landslide records important actions performed by the kernel, attempts to decide at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
+\revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.
+%During a kernel's execution, Landslide records important actions performed by the kernel, decides at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
+Using this instrumentation and configuration, Landslide is able to enumerate all possible interleavings of kernel threads at a granularity both fine-grained enough to expose interesting concurrency behaviour and coarse-grained enough to produce state spaces that are feasible to explore.
+Landslide then forces the kernel to exercise each of these interleavings, and checks for race conditions in each of them.}
 When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.
-
-With Landslide, we see testing a kernel as a process of manipulating test parameters in two ways: first, in the choice of test case (the user-space program that exercises a specific set of system calls), and second, in the configuration of Landslide in regard to which parts of the kernel are ``interesting'' in the behaviour of the test case and which are irrelevant.
-Searching for and understanding race conditions exposed by a given test becomes a joint effort between the programmer and Landslide, combining the programmer's specific knowledge about the design of the kernel and Landslide's ability to explore many interleavings efficiently.
 

### 2012-05-16 design.tex
fix the infinite loop thing; add figures

@@
 
+\begin{figure}[ht]
+	\centering
+	\begin{tabular}{cc}
+	\subfloat[Infinite loop without decision points. Disproportionately many instructions have been executed since the red state, so the kernel is probably stuck in a ``tight'' loop.]{
+	\hspace{0.12\textwidth}
+	\includegraphics[width=0.2\textwidth]{tightloop.png}
+	\hspace{0.12\textwidth}
+	} &
+	\subfloat[Infinite loop around decision points.
+Disproportionately many decision points have been encountered on the red branch, so the kernel is probably stuck in a loop around certain decision points.]{
+	\hspace{0.12\textwidth}
+	\includegraphics[width=0.2\textwidth]{livelock.png}
+	\hspace{0.12\textwidth}
+	}
+	\end{tabular}
+	\caption{Landslide judges whether the kernel has gotten stuck by analysing the structure of the execution tree.
+The red branches indicate non-terminating thread interleavings, which Landslide would identify by comparing them with their respective alternate interleavings.}
+	\label{fig:infiniteloop}
+\end{figure}
+
 \resume{enumerate}
@@
 		(Some kernel designs may legitimately behave this way, so this bug-check may be disabled when testing such kernels.)\footnote{There is, of course, much room for improvement in this metric, but it is not part of the research contribution.}
-	\item {\bf Infinite loop bugs.} Landslide judges whether the kernel has entered an infinite loop by comparing the current state of the decision tree to past executions of the same test case.
+	\item {\bf Infinite loop bugs.} Landslide judges whether the kernel has entered an infinite loop by comparing the current branch of the decision tree to past executions of the same test case.
+	Figure~\ref{fig:infiniteloop} depicts such tree structures in abstract.
 	\begin{itemize}
-		\item While exploring the decision tree, Landslide computes the average number of instructions executed between two consecutive decision points.
+		\item {\bf Infinite loops without decision points.} While exploring the decision tree, Landslide computes the average number of instructions executed between two consecutive decision points.
 If at any point the current number of instructions executed since the most recent decision point exceeds this average times a constant factor (arbitrarily chosen to be 2000), Landslide assumes the kernel must have gotten stuck in an infinite loop.
-		\item Landslide also computes the average number of decision points in each branch of the decision tree (the average ``branch depth'').
+		\item {\bf Infinite loops around decision points.} Landslide also computes the average number of decision points in each branch of the decision tree (the average ``branch depth'').
 If the depth of the current branch ever exceeds this average times a constant factor (arbitrarily chosen to be 20), Landslide assumes the kernel must have gotten stuck in an infinite loop.
 	\end{itemize}
+	In our experience, these heuristics have never falsely identified an infinite loop when the kernel was making real progress.
+	However, both of these heuristics require a minimum number of branches to be explored before Landslide considers the already-explored tree structure to be statistically significant (arbitrarily chosen to be 20).
+	As such, these heuristics occasionally fail to trigger on a real infinite loop in cases when the loop occurs before enough ``safe'' branches were explored.
+In the future we might improve this heuristic by scaling the associated cutoff factor depending on how many branches were explored: the fewer branches explored so far, the less reliable the comparison, and hence the higher the cutoff factor to be used.
 \end{enumerate}

### 2012-05-16 related.tex
cite verisoft
fix the infinite loop thing; add figures

@@
 
-Systematic exploration is a relatively young approach to concurrency verification.
+Systematic exploration is a relatively young approach to concurrency verification\cite{verisoft}.
 Recent work has focused on theoretical aspects of the technique, on distributed systems, and in userspace in general.
@@
 
-Dynamic Partial Order Reduction (DPOR)\hspace{0in}\cite{dpor,sdpor,dbug-retreat,distributed-dpor} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
+Dynamic Partial Order Reduction (DPOR)\hspace{0in}\cite{dpor,distributed-dpor,sdpor,dbug-retreat} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
 Landslide makes generous use of the DPOR algorithm.

### 2012-05-16 challenges.tex
fix the infinite loop thing; add figures
buncha minor changes + landslide-friendly spec changes

@@
 	\begin{center}
-	\includegraphics[width=0.75\textwidth]{vip1066720.jpg}
+	\includegraphics[width=0.65\textwidth]{vip1066720.jpg}
 	\end{center}
@@
 One challenge particular to the context of 15-410 is that the kernels Landslide must be able to test may all use slightly different implementations to achieve the same goals.
+\revision{While 15-410's kernel project mandates that students implement the Pebbles specification, students are free to make their own design decisions under the hood, which can result in significantly different scheduler behaviour among multiple kernels.}
 For Landslide to be generally applicable to (almost) any kernel that students may submit, it must make some abstract assumptions about the kernel design which are compatible with many different implementations.

### 2012-05-16 design.tex
buncha minor changes + landslide-friendly spec changes

@@
 The Landslide scheduler also tracks which important actions each thread is performing.
-These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing, and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
+These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section \revision{(note that ``vanish'' is simply Pebbles's name for ``exit'')}, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing, and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
 There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}
@@
 
+\revision{
 \begin{figure}[ht]
@@
 	\caption{Landslide judges whether the kernel has gotten stuck by analysing the structure of the execution tree.
-The red branches indicate non-terminating thread interleavings, which Landslide would identify by comparing them with their respective alternate interleavings.}
+The red branches indicate non-terminating thread interleavings, which Landslide would identify by comparing them with other branches in their respective trees.}
 	\label{fig:infiniteloop}
 \end{figure}
+}
 
@@
 	\end{itemize}
-	In our experience, these heuristics have never falsely identified an infinite loop when the kernel was making real progress.
+	\revision{In our experience, these heuristics have never falsely identified an infinite loop when the kernel was making real progress.
 	However, both of these heuristics require a minimum number of branches to be explored before Landslide considers the already-explored tree structure to be statistically significant (arbitrarily chosen to be 20).
 	As such, these heuristics occasionally fail to trigger on a real infinite loop in cases when the loop occurs before enough ``safe'' branches were explored.
-In the future we might improve this heuristic by scaling the associated cutoff factor depending on how many branches were explored: the fewer branches explored so far, the less reliable the comparison, and hence the higher the cutoff factor to be used.
+In the future we might improve this heuristic by scaling the associated cutoff factor depending on how many branches were explored: the fewer branches explored so far, the less reliable the comparison, and hence the higher the cutoff factor to be used.}
 \end{enumerate}
@@
 		A use-after-free bug happens whenever one thread accesses an address within an allocated block that another thread previously freed.
-Even if no other code makes a conflicting access to the same address after the \texttt{free}, the C idiom says that accessing dynamically-allocated memory after it has been freed is illegal no matter what.
+Even if no other code makes a conflicting access to the same address after the \texttt{free}, accessing dynamically-allocated memory after it has been freed is illegal no matter what.
 

### 2012-05-16 evaluation.tex
buncha minor changes + landslide-friendly spec changes

@@
 	\item % apodolsk
-		Group 5 did not get Landslide to explore a minimal decision tree.
+		Group 5 did not invest enough time to get Landslide to explore a minimal decision tree.
 \end{enumerate}
@@
 While building Landslide, we worked primarily with two kernels: one, the kernel the author wrote as a student in 15-410 (Fall 2008), and another, a kernel the author graded as a TA for 15-410 (Fall 2011).
-We refer to the first kernel as ``POBBLES'' and the second as ``LudicrOS''.
+With permission, we refer to the first kernel as ``POBBLES'' and the second as ``LudicrOS''.
 

### 2012-05-16 futurework.tex
buncha minor changes + landslide-friendly spec changes

@@
 While far from being able to replace talented humans for grading, we believe that, after some improvement, Landslide could augment or replace Fritz as a tool for automatically finding several common patterns of concurrency errors that students have (especially races involving exiting threads, which we have shown Landslide is effective at finding).
+
+\subsection{Making Pebbles Landslide-Friendly}
+
+\revision{
+If Landslide becomes integrated into an undergraduate course curriculum, it will be important to minimise the amount of effort students have to go through to instrument their kernels.
+This could be achieved in part by changing the project definition in ways that make Landslide better able to automatically instrument kernels.
+For Pebbles, we recommend the following changes.
+
+\begin{enumerate}
+	\item {\bf Timer ticks control ``runnable'' threads.} As discussed in Section~\ref{sec:model-timer}.
+Additionally, to prevent the users from having to instrument non-descheduling mutexes, the project could mandate that all concurrency primitives actually remove blocked threads from the runqueue.
+	\item {\bf No idling when progress can be made.} Also as discussed in Section~\ref{sec:model-timer}.
+With the current specification, inappropriate idling is merely a performance bug, though it is incompatible with Landslide.
+If the project gave stricter requirements regarding this, fewer students would need to fix problems with their scheduler idling during the instrumentation process.
+	\item {\bf Mandated function names for scheduling events.} As part of the instrumentation process, students tell Landslide where in their schedulers notable events happen.
+Landslide could automatically detect some of these if the project mandated the names of certain functions, such as the timer handler, context switcher, and runqueu manipulation routines.
+\end{enumerate}
+}
 

### 2012-05-16 related.tex
buncha minor changes + landslide-friendly spec changes

@@
 
-Landslide is distinguished from related systematic testing tools in its field as more of an exploratory work, presenting techniques to make systematic testing compatible with kernel-level code; less so as a development on the core testing approach itself.
+\revision{Compared to related systematic testing tools in its field, Landslide is more of an exploratory work, presenting techniques to make systematic testing compatible with kernel-level code, rather than a development on the core testing approach itself.}
 
@@
 There are also several tools, such as MaceMC\cite{macemc} and MODIST\cite{modist}, which provide systematic testing frameworks for networked and distributed applications.
+\revision{MODIST is especially notable in its ability to test unmodified distributed systems by using event interposition.}
 
@@
 Because Haskell is type-safe (strict constraints on legal program behaviour that make it easier to reason about) and pure (most code cannot have stateful side-effects), the developers of seL4 were able to formally prove the specification's correctness and the implementation's adherence to the spec.
-Developing kernel code in languages and environments that enable formal complete verification, even in small components of a larger codebase, represents an admirable step towards writing correct code to begin with and avoiding many classes of bugs entirely.
+Developing kernel code in languages and environments that enable formal complete verification, even in small components of a larger codebase \revision{(generally more tractable than seL4's extreme approach)}, represents an admirable step towards writing correct code to begin with and avoiding many classes of bugs entirely.
 

### 2012-05-16 terminology.tex
write intro
buncha minor changes + landslide-friendly spec changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\label{sec:terminology}
 
@@
 Practically, a user-space program which runs on top of the guest kernel to execute a particular set of system calls, while Landslide performs systematic exploration.
+		\revision{The test case's expected results of its system call invocation pattern informally define the kernel specification.}
 \suspend{enumerate}
@@
 		Performed at the end of each branch of the tree, when the test case has finished running.
-Landslide identifies a decision point from the current branch at which it wants to have chosen a different thread to run, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
+Landslide identifies a decision point from the current branch at which the \revision{next interleaving to explore diverges from the current one}, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
 	\item {\bf Decision trace}:

### 2012-05-16 futurework.tex
bit more future

@@
 \subsection{Symbolic Execution}
+\label{sec:future-dawson}
 
@@
 
-Many components of Linux contain ad-hoc synchronisation which would be difficult to reason about.
+Many components of Linux contain ad-hoc synchronisation, often done for performance reasons, which would be difficult to reason about.
 One common pattern is attempting some operation, checking later if it got interfered with, and rewinding and trying again if so.
@@
 
-% FIXME
-%\subsection{Other Kernel Environments}
-
-%Some other production kernels would also make good targets for systematic exploration.
+\revision{Because many driver bugs result from interactions with error-handling paths, it will also help to understand the difference between normal device input and faulty input in order to maximise code path coverage.
+Symbolic execution (Section~\ref{sec:future-dawson}) may help address this.}
 
@@
 
-%\subsection{Iterating Decision Sets}
 In Section~\ref{sec:discussion-strategies}, we recommend an iteration strategy to effectively explore trees with multiple different sets of decision points, while not knowing ahead of time which, if any, may contain a bug.
@@
 
-% TODO talk about heuristic-guided tree construction
-
-% FIXME
-%In Section~\ref{Sec:future-nadim}, we discuss an interesting property of certain decision trees, and suggest a heuristic
+\revision{One possible application of such a strategy would be in searching for bugs given only a stack trace and/or kernel log from a crash report.
+The components of the kernel mentioned in the trace or log could be used as the initial focus for the test, and the framework could gradually expand its scope to search for an interleaving granularity which would reproduce the bug.}
 

### 2012-05-16 using.tex
improve sample code

@@
 
+\newcommand\hilight[2]{\color{#1}#2\color{white}}
 \begin{figure}[h]
 	\centering
-	\includegraphics[width=\textwidth]{tell_landslide.png}
+	\colorbox{black}{\color{white}
+	{\small
+	\begin{tabular}{l}
+	\texttt{\hilight{green}{int}~thread\_fork()} \\
+	\texttt{\{} \\
+	\texttt{~~~~\hilight{green}{thread\_t}~*child = construct\_new\_thread();} \\
+	\texttt{~~~~\hilight{cyan}{tell\_landslide\_forking}();} \\
+	\texttt{~~~~add\_to\_runqueue(child);} \\
+	\texttt{~~~~\hilight{cyan}{tell\_landslide\_decide}(); \hilight{purple}{/* Interrupt me here! */}} \\
+	\texttt{~~~~\hilight{brown}{return}~child->tid;} \\
+	\texttt{\}} \\
+	\end{tabular}
+	}
+	}
 	\caption{Example annotated code.
@@
 	\centering
-	\begin{framed}
-	{\small \begin{verbatim}
-	bool kern_current_extra_runnable(conf_object_t *cpu)
-	{
-	    int tcb = READ_MEMORY(cpu, GUEST_CURRENT_TCB_POINTER);
-	    int state_flag = READ_MEMORY(cpu, tcb + GUEST_TCB_STATE_FLAG_OFFSET);
-	    return state_flag == GUEST_TCB_RUNNABLE;
+	\colorbox{black}{\color{white}
+	{\small
+	\begin{tabular}{l}
+	\texttt{\hilight{green}{bool}~kern\_current\_extra\_runnable(\hilight{green}{conf\_object\_t}~*cpu)} \\
+	\texttt{\{} \\
+	\texttt{~~~~\hilight{green}{int}~tcb = \hilight{cyan}{READ\_MEMORY}(cpu, GUEST\_CURRENT\_TCB\_POINTER);} \\
+	\texttt{~~~~\hilight{green}{int}~state\_flag = \hilight{cyan}{READ\_MEMORY}(cpu, tcb + GUEST\_TCB\_STATE\_FLAG\_OFFSET);} \\
+	\texttt{~~~~\hilight{brown}{return}~state\_flag == GUEST\_TCB\_RUNNABLE;} \\
+	\texttt{\}} \\
+	\texttt{\hilight{green}{bool}~kern\_ready\_for\_timer\_interrupt(\hilight{green}{conf\_object\_t}~*cpu)} \\
+	\texttt{\{} \\
+	\texttt{~~~~\hilight{green}{int}~x = \hilight{cyan}{READ\_MEMORY}(cpu, GUEST\_PREEMPTION\_FLAG);} \\
+	\texttt{~~~~\hilight{brown}{return}~x == GUEST\_PREEMPTION\_ENABLED;} \\
+	\texttt{\}} \\
+	\end{tabular}
 	}
-	bool kern_ready_for_timer_interrupt(conf_object_t *cpu)
-	{
-	    int x = READ_MEMORY(cpu, GUEST_PREEMPTION_FLAG);
-	    return x == GUEST_PREEMPTION_ENABLED;
 	}
-	\end{verbatim}
-	}
-	\end{framed}
 	\caption{Example implementations of the two functions in \texttt{student.c}.}

### 2012-05-16 design.tex
cite valgrind
found_a_bug in text

@@
 This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
+\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), such as in Valgrind\cite{valgrind}.}
 
@@
 
+\newcommand\hilight[2]{\color{#1}#2\color{white}}
 \begin{figure}[h]
-	\begin{center}
-		\includegraphics[width=0.95\textwidth]{found_a_bug.png}
-	\end{center}
+	%\includegraphics[width=0.95\textwidth]{found_a_bug.png}
+	\centering
+	\colorbox{black}{\color{white}
+	{\small
+	\begin{tabular}{l}
+	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~thread 4 vanished} \\
+	\texttt{\hilight{green}{[SCHEDULE]}~~~~~~switched threads 4 -> 3} \\
+	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~\hilight{red}{USE AFTER FREE - read from 0x0015a8f0 at eip 0x00104209}} \\
+	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~Heap contents:~\{...\}} \\
+	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~[0x15a8f0 | 4136] was allocated by TID3 at (...)} \\
+	\texttt{\hilight{brown}{[MEMORY]}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~and freed by TID4 at (...)} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{****~~~~~A bug was found!~~~~****}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{red}{**** Decision trace follows.~****}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{1:~~1347079 instructions, old 3 new 4, current 4}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID3 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x001041f4 in \hilight{cyan}{thread\_fork},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~\hilight{yellow}{2:~~1350725 instructions, old 4 new 3, current 3}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~TID4 at 0x00105a10 in \hilight{cyan}{context\_switch},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104681 in \hilight{cyan}{yield},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00104570 in \hilight{cyan}{vanish},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~0x00103708 in \hilight{cyan}{vanish\_wrapper}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Stack:~TID3 at 0x00104209 in \hilight{cyan}{thread\_fork},} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~~~~~~~~~~~~~~~~0x0010362b in \hilight{cyan}{thread\_fork\_wrapper}} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Total decision points 24, total backtracks 5} \\
+	\texttt{\hilight{red}{[BUG!]}~~~~~~~~~~Average instrs/decision 16155, average branch depth 5} \\
+	\end{tabular}
+	}
+	}
 	\caption{A sample decision trace (with extraneous text trimmed) that Landslide generated using the \texttt{double\_thread\_fork} test case (Section~\ref{sec:using-tests}).}

### 2012-05-16 challenges.tex
several minor

--- 2012-05-14 challenges.tex
@@
 \label{sec:challenges}
+
+\revision{In this chapter, we discuss several problems that stand in the way of systematic exploration's applicability to kernel space.
+We address these problems in part with certain aspects of Landslide's design and in part by relying on the user to provide important information, which we discuss in Chapter~\ref{sec:design} and Chapter~\ref{sec:using} respectively.}
 
@@
 One challenge particular to the context of 15-410 is that the kernels Landslide must be able to test may all use slightly different implementations to achieve the same goals.
-For Landslide to be generally applicable to (almost) any kernel that students may submit, it must make some abstract assumptions about the kernel design which are compatible with many different implementations.
+\revision{While 15-410's kernel project mandates that students implement the Pebbles specification, students are free to make their own design decisions under the hood, which can result in significantly different scheduler behaviour among multiple kernels.
+In general, Landslide may rely on concrete details provided by kernel specifications, but must be compatible with arbitrary implementation details.}
+For Landslide to be generally applicable to many different kernels, even only multiple ones that implement the same specification (such as Pebbles), it must make some abstract assumptions about the kernel design which are compatible with many different implementations.
 

### 2012-05-16 evaluation.tex
several minor

@@
 While building Landslide, we worked primarily with two kernels: one, the kernel the author wrote as a student in 15-410 (Fall 2008), and another, a kernel the author graded as a TA for 15-410 (Fall 2011).
-With permission, we refer to the first kernel as ``POBBLES'' and the second as ``LudicrOS''.
+We refer to the first kernel as ``POBBLES'' and to the second, with permission, as ``LudicrOS''.
 
@@
 In short, even though Landslide may give false negatives from using imperfect sets of decision points, the exploration itself must be sound (i.e.
-not missing any bugs that exist in the resulting tree).\footnote{
+not missing any bugs that exist in the resulting tree).
+
 When running LudicrOS yield/vanish with decision points on \texttt{mutex\_unlock} but not on \texttt{mutex\_lock}, we found that the ordering invariant failed: ``backwards'' exploration found no bug, while ``forwards'' exploration did.
-We attribute this to a bug in Landslide itself, and present the results for the backwards exploration as usual, in which Landslide found no bug.}
+We attribute this to a bug in Landslide itself, and present the results for the backwards exploration as usual, in which Landslide found no bug.
 

### 2012-05-16 terminology.tex
several minor
'fix' figure in terminology section

@@
 Practically, a user-space program which runs on top of the guest kernel to execute a particular set of system calls, while Landslide performs systematic exploration.
-		\revision{The test case's expected results of its system call invocation pattern informally define the kernel specification.}
+		\revision{The expected results of the test case's system call invocation pattern informally define the kernel specification.}
 \suspend{enumerate}
@@
 
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.75\textwidth]{threadfork.png}
-	% TODO: make some other subfigs
-	\caption{A simple decision tree, showing possible interleavings of a buggy \texttt{thread\_fork} implementation. (This describes a bug studied in Section~\ref{sec:eval-thread-fork}.)}
-	\label{fig:threadfork}
-\end{figure}
-
 \revision{The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.}
@@
 		A decision tree is created/discovered during an exploration using a particular decision point set; the set is what the user configures, and the tree is what arises as a result.
+
+		Figure~\ref{fig:threadfork} depicts a decision tree for a simple race with three possible interleavings.
 	\item {\bf Branch}:
@@
 
+\definecolor{thread1}{RGB}{87,172,255}
+\definecolor{thread2}{RGB}{255,201,102}
+\newcommand\hilight[2]{\color{#1}#2\color{white}}
+\begin{figure}[h]
+	\centering
+	\begin{framed}
+	\begin{tabular}{cc}
+	\multicolumn{2}{c}{\colorbox{black}{\color{white}
+	{\small
+	\begin{tabular}{l}
+	\texttt{\hilight{green}{int}~thread\_fork()} \\
+	\texttt{\{} \\
+	\texttt{~~~~\hilight{green}{thread\_t}~*child = construct\_new\_thread();} \\
+	\texttt{~~~~add\_to\_runqueue(child);} \\
+	\texttt{~~~~\hilight{brown}{return}~child->tid; \hilight{magenta}{/* Hopefully child hasn't exited yet!~*/}} \\
+	\texttt{\}} \\
+	\end{tabular}
+	}
+	}
+	}
+	\vspace{0.2in}
+	\\
+	\scriptsize
+	\begin{tabular}{|c|c|}
+	\hline
+	\cellcolor{thread1}{\bf Thread 1} & \cellcolor{thread2}{\bf Thread 2} \\
+	\hline
+	\texttt{construct\_new\_thread();} & \\
+	\hline
+	\texttt{add\_to\_runqueue(child);} & \\
+	\hline
+	\texttt{return child->tid;} & \\
+	\hline
+	& \texttt{exit();} \\
+	\hline
+	& \texttt{(free TCB)} \\
+	\hline
+	\end{tabular} \vspace{0.1in} &
+	\multirow{3}{*}{\includegraphics[width=0.48\textwidth]{threadfork.png}\vspace{0.5in}} \\
+	\scriptsize
+	\vspace{0.1in}
+	\begin{tabular}{|c|c|}
+	\hline
+	\cellcolor{thread1}{\bf Thread 1} & \cellcolor{thread2}{\bf Thread 2} \\
+	\hline
+	\texttt{construct\_new\_thread();} & \\
+	\hline
+	\texttt{add\_to\_runqueue(child);} & \\
+	\hline
+	& \texttt{exit();} \\
+	\hline
+	\texttt{return child->tid;} & \\
+	\hline
+	& \texttt{(free TCB)} \\
+	\hline
+	\end{tabular} \vspace{0.1in} & \\
+	\scriptsize
+	\begin{tabular}{|c|c|}
+	\hline
+	\cellcolor{thread1}{\bf Thread 1} & \cellcolor{thread2}{\bf Thread 2} \\
+	\hline
+	\texttt{construct\_new\_thread();} & \\
+	\hline
+	\texttt{add\_to\_runqueue(child);} & \\
+	\hline
+	& \texttt{exit();} \\
+	\hline
+	& \texttt{(free TCB)} \\
+	\hline
+	\texttt{return child->tid;} & \\
+	\hline
+	\end{tabular} & \\
+
+	\end{tabular}
+	\end{framed}
+	\caption{This \texttt{thread\_fork} implementation is prone to a use-after-free bug.
+Three different possible interleavings with the newly-forked thread (left) can be viewed together as a single decision tree (right).
+	(Section~\ref{sec:eval-thread-fork} studies this particular bug in detail.)}
+	\label{fig:threadfork}
+\end{figure}
+
 % vim: ft=tex

### 2012-05-16 challenges.tex
more minor

@@
 Do just-forked threads begin life through the usual context-switch-return path, or is there a special path for that?
-	\item {\bf Test Lifecycle}: Apart from running on a particular input that the user wishes to test, the kernel may need to perform extra work irrelevant to the test, such as initialisation (e.g., the bootup process) and housekeeping (e.g., cleaning up dead processes).
+	\item {\bf Test Lifecycle}: Apart from running on a particular input that the user wishes to test, the kernel may need to perform extra work, such as initialisation (e.g., the bootup process) and housekeeping (e.g., cleaning up dead processes).
+	\revision{Though the entire kernel is notionally the system under test, the user-space test case expresses what the user is actually interested in, and some of this extra work is irrelevant.}
 	Which parts of this work should be included in or excluded from the test?

### 2012-05-16 acknowledgements.tex
add title page and convert to book format
fix the infinite loop thing; add figures
acknowledge pdl

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
+\chapter*{Acknowledgements}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Save this for last :)
+
+Thanks to Deb Cavlovich, for holding on to this thesis for a few days more for Ben to finish it properly.
+
+Thanks to my advisor, too, I guess, and maybe also to the people I worked with and my friends as well.
+
+Thanks to the members and companies of the PDL Consortium (including Actifo,
+APC, EMC, Emulex, Facebook, Fusion-io, Google, Hewlett-Packard, Hitachi, Huawei
+Technologies, Intel, Microsoft, NEC, NetApp, Oracle, Panasas, Riverbed,
+Samsung, Seagate, STEC, Symantec, VMware, and Western Digital) for their
+interest, insights, feedback, and support.

### 2012-05-16 design.tex
'fix' figure in terminology section
more minor
fix some footnotes

@@
 
-%\subsection{Stack Tracing}
-% FIXME
-
 \section{Identifying Bugs}
@@
 	\item {\bf Always-accessed memory locations.}
-		In kernel-space, every thread switch goes through common scheduler and context-switcher routines. These routines inevitably access scheduler data structures, such as the runqueue.\footnote{
-		Another example is the timer tick counter: this global variable is not necessarily accessed every transition, because voluntary reschedules do not involve timer interrupts, but every involuntary preemption will cause a write to this counter.}
+		In kernel-space, every thread switch goes through common scheduler and context-switcher routines.
+These routines inevitably access scheduler data structures, such as the runqueue.
+The timer tick counter is another example: this global variable is not necessarily accessed every transition, because voluntary reschedules do not involve timer interrupts, but every involuntary preemption will cause a write to this counter.
+
 		If we include such memory accesses in the independence relation, it will result that all transitions conflict, and DPOR won't be able to achieve any reduction.
@@
 
-\newcommand\hilight[2]{\color{#1}#2\color{white}}
 \begin{figure}[h]

### 2012-05-16 evaluation.tex
fix some footnotes

@@
 However, when awakened, the condition variable code did not clear the value of this pointer.
-Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.\footnote{
-Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing: if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable cause a spurious wakeup of the second thread.
-Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in a use-after-free.}
+Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.
+
+Even with this assert, extremely mysterious behaviour could have arisen in other use cases, such as during stress testing: if the awakened thread had gone to sleep on another condition variable, its \texttt{cond\_queue\_next} pointer could point to a thread asleep on a different condition variable, in which case a broadcast on the new condition variable would cause a spurious wakeup of the second thread.
+Worse yet, if the second thread had also been awoken, and thence exited, a broadcast would result in scheduler data structure corruption.
 
@@
 
-This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple\footnote{
-Many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the kernel bug may never be noticed when using such libraries (this caused the author some trouble).},
+This bug is exposed with the \texttt{double\_thread\_fork} test case. The complexity is relatively simple,\footnote{
+Many userspace thread libraries forcibly serialise invocations of \texttt{thread\_fork} and \texttt{vanish} with their own synchronisation barriers, so the kernel bug may never be noticed when using such libraries (this caused the author some trouble).}
 and the fix is only two lines long; the dereference should simply occur before making the child thread runnable.

### 2012-05-16 terminology.tex
fix some footnotes

@@
 	\item {\bf Voluntary reschedule}:
-		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.\footnote{Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not).
-Involuntary preemptions are never necessary for correct kernel behaviour.}
+		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.
+
+		Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not).
+Involuntary preemptions, however, are never necessary for correct kernel behaviour.
 		See Section~\ref{sec:components-arbiter}.

### 2012-05-16 using.tex
found_a_bug in text
fix some footnotes

@@
 	\texttt{~~~~add\_to\_runqueue(child);} \\
-	\texttt{~~~~\hilight{cyan}{tell\_landslide\_decide}(); \hilight{purple}{/* Interrupt me here! */}} \\
+	\texttt{~~~~\hilight{cyan}{tell\_landslide\_decide}(); \hilight{magenta}{/* Interrupt me here! */}} \\
 	\texttt{~~~~\hilight{brown}{return}~child->tid;} \\
@@
 For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
-	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.\footnote{
-For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.}
+	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.
+For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.
+
 This configuration option allows the user to whitelist functions so that Landslide will only decide if the current thread is ``within'' one of those functions.

### 2012-05-16 challenges.tex
phrasing

@@
 \revision{In this chapter, we discuss several problems that stand in the way of systematic exploration's applicability to kernel space.
-We address these problems in part with certain aspects of Landslide's design and in part by relying on the user to provide important information, which we discuss in Chapter~\ref{sec:design} and Chapter~\ref{sec:using} respectively.}
+We address these problems partly with certain aspects of Landslide's design and partly by relying on the user to provide important information, which we discuss in Chapter~\ref{sec:design} and Chapter~\ref{sec:using} respectively.}
 

### 2012-05-16 terminology.tex
restore a footnote

@@
 	\item {\bf Voluntary reschedule}:
-		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.
-
+		A context switch from one thread to another which the guest kernel performed automatically, not triggered by nondeterministic events and/or Landslide but rather as part of its normal execution.\footnote{
 		Voluntary reschedules may or may not be necessary for the kernel to behave correctly at all (for example: switching away from an exiting thread or from a thread blocked on \texttt{wait} is necessary, but switching to a newly-forked thread is not).
-Involuntary preemptions, however, are never necessary for correct kernel behaviour.
+Involuntary preemptions, however, are never necessary for correct kernel behaviour.}
 		See Section~\ref{sec:components-arbiter}.

### 2012-05-17 related.tex
racetrack
simtester

@@
 Due to the interrupt-driven nature of many drivers, they are also prone to concurrency bugs, which Carburizer does not address.
+
+SimTester\cite{simtester} is another tool for testing kernel device drivers.
+Like Landslide, SimTester uses Simics as its execution environment, and identifies key decision points during execution at which to inject interrupts.
+SimTester focuses on races involving device interrupt handling code, whereas Landslide is geared towards non-driver inter-thread races.
+Additionally, SimTester's testing model involves forcing only one interrupt per test run, which is a lighter-weight approach than systematic exploration.
+
 In Section~\ref{sec:future-drivers} we discuss the potential applicability of systematic exploration to concurrency bugs in kernel device drivers.
@@
 Such accesses, called ``data races'', might not necessarily cause incorrect behaviour, but are usually warning indicators of such, and represent bad coding practice in any case.
+\revision{RaceTrack\cite{racetrack} is another data race detection tool notable for its ability to adaptively adjust its testing granularity after some execution analysis.}
 

### 2012-05-17 futurework.tex
add graphs for bwd/fwd

@@
 
+\revision{
+\renewcommand\bugnum[2]{#1 {\scriptsize \em (#2)}}
+\begin{figure}[t]
+	\centering
+	\begin{tabular}{cc}
+	\multicolumn{2}{c}{\footnotesize
+	\centering
+	\begin{tabular}{|l|l||c|c|c|}
+	\hline
+	{\bf Bug} & {\bf Direction} & {\bf Time} & {\bf Backtracks} & {\bf Trace Length} \\
+	\hline\hline
+	\multirow{2}{*}{POBBLES vanish/vanish (a)} & Forwards & \bugnum{124.8}{4.9} & 1264 & 7 \\
+	& Backwards & \bugnum{57.1}{1.7} & 376 & 11 \\
+	\hline
+	\multirow{2}{*}{POBBLES vanish/vanish (b)} & Forwards & \bugnum{106.3}{5.7} & 1030 & 7 \\
+	& Backwards & \bugnum{51.5}{2.1} & 376 & 10 \\
+	\hline
+	\multirow{2}{*}{POBBLES wait/wait} & Forwards & \bugnum{62.6}{3.4} & 132 & 8 \\
+	& Backwards & \bugnum{27.9}{0.8} & 17 & 13 \\
+	\hline
+	\multirow{2}{*}{POBBLES thread\_fork/vanish} & Forwards & \bugnum{21.4}{0.7} & 0 & 6 \\
+	& Backwards & \bugnum{37.4}{1.1} & 70 & 6 \\
+	\hline
+	\multirow{2}{*}{LudicrOS vanish/vanish} & Forwards & \bugnum{14.7}{0.9} & 4 & 3 \\
+	& Backwards & \bugnum{13.7}{0.7} & 3 & 7 \\
+	\hline
+	\multirow{2}{*}{LudicrOS yield/vanish} & Forwards & \bugnum{13}{0.9} & 3 & 3 \\
+	& Backwards & \bugnum{11.4}{0.4} & 0 & 5 \\
+	\hline
+	\end{tabular}
+	}
+	\vspace{0.1in}
+	\\
+	\includegraphics[width=0.48\textwidth]{graphs/backwards-time.pdf} &
+	\includegraphics[width=0.48\textwidth]{graphs/backwards-trace.pdf}
+	\end{tabular}
+	\caption{Comparison of forwards and backwards explorations for the six case study bugs (using decision points on \texttt{mutex\_lock} but not \texttt{mutex\_unlock}).
+In general, backwards exploration finds bugs faster, but forwards exploration produces shorter (more understandable) decision traces.}
+	\label{fig:backwards}
+\end{figure}
+}
+
 We configured Landslide to be able to explore trees in two different orderings, and we found that the two orderings behaved substantially differently in terms of time taken to find bugs and length of decision traces produced.
@@
 \end{itemize}
-% TODO: provide results here
-% FIXME: study diffs in trees with no bugs at all
+
+\revision{Figure~\ref{fig:backwards} compares the tree explorations and resulting decision traces.
+The one exception is the POBBLES thread\_fork/vanish bug, which Landslide found immediately during forwards exploration, because the race actually required no forced preemptions to expose.}
 
@@
 
-\begin{figure}[h]
+\begin{table}[h]
 	\small
@@
 	\label{fig:nadim-stats}
-\end{figure}
+\end{table}
 

### 2012-05-17 futurework.tex
change bwd/fwd layout

@@
 \renewcommand\bugnum[2]{#1 {\scriptsize \em (#2)}}
-\begin{figure}[t]
+\begin{table}[t]
 	\centering
-	\begin{tabular}{cc}
-	\multicolumn{2}{c}{\footnotesize
-	\centering
+	\footnotesize
 	\begin{tabular}{|l|l||c|c|c|}
@@
 	\end{tabular}
-	}
-	\vspace{0.1in}
-	\\
-	\includegraphics[width=0.48\textwidth]{graphs/backwards-time.pdf} &
+	\caption{Comparison of forwards and backwards explorations for the six case study bugs (using decision points on \texttt{mutex\_lock} but not \texttt{mutex\_unlock}).}
+	\label{fig:backwards}
+\end{table}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.47\textwidth]{graphs/backwards-time.pdf}
+	\hspace{0.02\textwidth}
 	\includegraphics[width=0.48\textwidth]{graphs/backwards-trace.pdf}
-	\end{tabular}
-	\caption{Comparison of forwards and backwards explorations for the six case study bugs (using decision points on \texttt{mutex\_lock} but not \texttt{mutex\_unlock}).
-In general, backwards exploration finds bugs faster, but forwards exploration produces shorter (more understandable) decision traces.}
-	\label{fig:backwards}
+	\caption{In general, backwards exploration finds bugs faster, but forwards exploration produces shorter (more understandable) decision traces.}
+	\label{fig:backwards-graphs}
 \end{figure}
@@
 
-\revision{Figure~\ref{fig:backwards} compares the tree explorations and resulting decision traces.
+\revision{Table~\ref{fig:backwards} and Figure~\ref{fig:backwards-graphs} compare the tree explorations and resulting decision traces.
 The one exception is the POBBLES thread\_fork/vanish bug, which Landslide found immediately during forwards exploration, because the race actually required no forced preemptions to expose.}

### 2012-05-17 evaluation.tex
eval numbers bw print support
change figs to tables
add eval graphs

@@
 \revision{
-\begin{figure}[t]
+\begin{table}[t]
 	\begin{center}
@@
 	\label{fig:student-times}
+\end{table}
+\begin{figure}[h]
+	\center
+	\includegraphics[width=0.5\textwidth]{graphs/studence-pie.pdf}
+	\caption{Average time breakdown for each phase of instrumenting a kernel with Landslide, based on student self-reported times.}
+	\label{fig:student-pie}
 \end{figure}
@@
 
-Table~\ref{fig:student-times} presents the time breakdown among phases that students experienced while using Landslide.
+Table~\ref{fig:student-times} presents the time breakdown among phases that students experienced while using Landslide, visualised in Figure~\ref{fig:student-pie}.
 On average, students spent two hours getting Landslide to work with their kernels, and slightly more than half an hour customising the search and/or studying found bugs on top of that.
@@
 
-Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points, and using conventional stress testing.
+Table~\ref{fig:numbers} shows the time it takes to find each of these bugs using Landslide, configured with several different sets of decision points.
+\revision{Figure~\ref{fig:number-graph} compares the total exploration time among each different decision set for the trials that did find bugs.}
 
 \newcommand\bugnum[2]{\textcolor{BrickRed}{{\bf #1} {\scriptsize \em (#2)}}}
-\newcommand\nobugnum[2]{\textcolor{Blue}{{\em #1} {\scriptsize \em (#2)}}}
+\newcommand\nobugnum[2]{\textcolor{Blue}{[{\em #1 {\scriptsize (#2)}}]}}
 
-\begin{figure*}[t!]
+\begin{table}[t!]
 	\begin{center}
-	\small
-	\begin{tabular}{|l||c|c|c|c||c|}
+	\footnotesize
+	\begin{tabular}{|l||c|c|c|c|}
 		\hline
-		\multirow{2}{*}{\bf Kernel and test case} & \multicolumn{4}{c||}{\bf Decision points used} & \multirow{2}{*}{\bf Stress test} \\
+		\multirow{2}{*}{\bf Kernel and test case} & \multicolumn{4}{c|}{\bf Decision points used} \\ % & \multirow{2}{*}{\bf Stress test} \\
 		\cline{2-5}
-		& \bf default & \bf lock & \bf unlock & \bf both & \\
+		& {\bf default} & {\bf lock} & {\bf unlock} & {\bf both} \\
 		\hline\hline
-		POBBLES vanish/vanish (a) & \nobugnum{31.8}{0.6} & \bugnum{57.1}{1.7} & $\infty$ & $\infty$ & \\
+		POBBLES vanish/vanish (a) & \nobugnum{31.8}{0.6} & \bugnum{57.1}{1.7} & $\infty$ & $\infty$ \\
 		\hline
-		POBBLES vanish/vanish (b) & \nobugnum{32.0}{0.4} & \bugnum{51.5}{2.1} & \nobugnum{8057.9}{336.9} & $\infty$ & \\
+		POBBLES vanish/vanish (b) & \nobugnum{32.0}{0.4} & \bugnum{51.5}{2.1} & \nobugnum{8057.9}{336.9} & $\infty$ \\
 		\hline
-		POBBLES wait/wait & \bugnum{23.3}{0.7} & \bugnum{27.9}{0.8} & \bugnum{27.9}{2.1} & \bugnum{41.6}{1.4} & \\
+		POBBLES wait/wait & \bugnum{23.3}{0.7} & \bugnum{27.9}{0.8} & \bugnum{27.9}{2.1} & \bugnum{41.6}{1.4} \\
 		\hline
-		POBBLES thread\_fork/vanish & \bugnum{22.0}{0.6} & \bugnum{37.4}{1.1} & \bugnum{27.6}{0.5} & \bugnum{72.0}{2.6} & \\
+		POBBLES thread\_fork/vanish & \bugnum{22.0}{0.6} & \bugnum{37.4}{1.1} & \bugnum{27.6}{0.5} & \bugnum{72.0}{2.6} \\
 		\hline
-		LudicrOS vanish/vanish & \nobugnum{13.2}{0.2} & \bugnum{13.7}{0.7} & \bugnum{34.6}{1.1} & \bugnum{17.1}{0.3} & \\
+		LudicrOS vanish/vanish & \nobugnum{13.2}{0.2} & \bugnum{13.7}{0.7} & \bugnum{34.6}{1.1} & \bugnum{17.1}{0.3} \\
 		\hline
-		LudicrOS yield/vanish & \nobugnum{12.3}{0.3} & \bugnum{11.4}{0.4} & \nobugnum{27.4}{0.8} & \bugnum{11.7}{0.4} & \\
+		LudicrOS yield/vanish & \nobugnum{12.3}{0.3} & \bugnum{11.4}{0.4} & \nobugnum{27.4}{0.8} & \bugnum{11.7}{0.4} \\
 		\hline
-		\multicolumn{6}{c}{Key: \bugnum{seconds}{stddev} indicates bug found; \nobugnum{seconds}{stddev} indicates whole tree explored with no bug.} \\
+		\multicolumn{5}{c}{Key: \bugnum{seconds}{stddev} indicates bug found; \nobugnum{seconds}{stddev} indicates whole tree explored with no bug.} \\
 	\end{tabular}
 	\end{center}
-	\caption{Comparison of time taken (in seconds) to find bugs using Landslide and using conventional stress testing.
-	Landslide's times are given for several different sets of decision points: the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
-	All numbers represent the average from 5 trials, with the standard deviations given in parentheses.
-``$\infty$'' indicates the bug was not found with stress testing (after 1 hour), or that Landslide's search did not finish (after 8 hours).
-% TODO: what timeout for stress testing?
+	\caption{Comparison of time taken (in seconds) to find bugs using Landslide with various decision sets:
+	the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
+	All numbers represent the average from 5 trials, with the standard deviations given in parentheses. ``$\infty$'' indicates that Landslide's search did not finish (after 8 hours).
 	}
 	\label{fig:numbers}
-\end{figure*}
+\end{table}
+\revision{
+\begin{figure}[h]
+	\center
+	\includegraphics[width=0.7\textwidth]{graphs/decisions.pdf}
+	\caption{Bar graph visualisation of the exploration time in trials that did find bugs. The decision trees that did not expose bugs, and the ones that timed out, are not shown.}
+	\label{fig:number-graph}
+\end{figure}
+}
 
-% TODO: have a graph
-
-The Landslide experimental set-up is as follows:
+The experimental set-up is as follows:
 
@@
 %\end{itemize}
-% TODO: say specs of crash machine
 
@@
 \newcommand\bugtree[1]{\textcolor{BrickRed}{\bf #1}}
-\newcommand\nobugtree[1]{\textcolor{Blue}{\em #1}}
-\begin{figure*}[t!]
+\newcommand\nobugtree[1]{\textcolor{Blue}{[{\em #1}]}}
+\begin{table}[t!]
 	\begin{center}
-	\small
+	\footnotesize
 	\begin{tabular}{|l|l||c|c|c|c|}
@@
 	\label{fig:trees}
-\end{figure*}
+\end{table}
 

### 2012-05-17 futurework.tex
appease dave re VM

@@
 
+\revision{
 \subsection{Making Pebbles Landslide-Friendly}
 
-\revision{
 If Landslide becomes integrated into an undergraduate course curriculum, it will be important to minimise the amount of effort students have to go through to instrument their kernels.
@@
 \end{enumerate}
+
+\subsection{Virtual Memory Bug-Finding}
+
+In this work, we focused especially on races related to the thread lifecycle routines.
+We believe Landslide would be immediately applicable to certain varieties of virtual memory bugs, but other types might not be easily exposed with Landslide's current model.
+
+For example, one of the groups in the user study said they were looking for a race that only arose in out of memory conditions.
+A test case that could expose that race, by exhausting the kernel's memory resources, would necessarily have a very large state space that would be impractical to explore systematically.
+
+We suggested to the group to configure their kernel to fail memory allocations every so often, so Landslide would be better able to find the race while still using a small test case.
+To make Landslide more applicable to such bugs in general, we might further investigate this debugging strategy in a more structured way.
 }

### 2012-05-17 challenges.tex
fix cite fmt

@@
 
-In user space, a systematic exploration tool may cooperate with the underlying kernel to help control the concurrent behaviour of the system\cite{dbug-ssv}.
+In user space, a systematic exploration tool may cooperate with the underlying kernel to help control the concurrent behaviour of the system~\cite{dbug-ssv}.
 Simple system call invocations can cause a particular thread to run at a particular time, or to block while another thread runs first.
@@
 	\end{center}
-	\caption{{\em !`Cuidado! !`Las llamas son muy peligrosas!}\cite{de0u}}
+	\caption{{\em !`Cuidado! !`Las llamas son muy peligrosas!}~\cite{de0u}}
 \end{figure*}
@@
 
-Inter-thread communication can be much more ad-hoc in kernel-space than in user-space\hspace{0in}\cite{datacollider}, and a systematic testing framework needs to be aware of all the kernel's synchronisation idioms.
+Inter-thread communication can be much more ad-hoc in kernel-space than in user-space \cite{datacollider}, and a systematic testing framework needs to be aware of all the kernel's synchronisation idioms.
 
 \begin{itemize}
-	\item {\bf Avoiding reliance on message-passing.} Some systems for user-space systematic testing require that threads communicate only by message-passing, to better track the concurrency relationships between thread transitions\cite{dbug-ssv}.
-With few exceptions\cite{barrelfish}, kernels do not rely on message-passing as a primary communication mechanism, and to be compatible with the kernels of today, a testing framework must allow for less idealised state-sharing (Section~\ref{sec:components-memory}).
+	\item {\bf Avoiding reliance on message-passing.} Some systems for user-space systematic testing require that threads communicate only by message-passing, to better track the concurrency relationships between thread transitions~\cite{dbug-ssv}.
+With few exceptions~\cite{barrelfish}, kernels do not rely on message-passing as a primary communication mechanism, and to be compatible with the kernels of today, a testing framework must allow for less idealised state-sharing (Section~\ref{sec:components-memory}).
 	\item {\bf Recognising blocked threads.} It is difficult to detect when threads are ready to do useful work.

### 2012-05-17 design.tex
fix cite fmt

@@
 
-Landslide is implemented as a module for Simics\cite{simics}, a full-system x86 emulator.
+Landslide is implemented as a module for Simics~\cite{simics}, a full-system x86 emulator.
 When running the kernel, Simics calls into Landslide once every time the kernel executes an instruction or performs a memory read or write.
@@
 % A healthy dose of snark.
-Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
+Without a formal specification of the internals of the guest kernel's implementation~\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
 Landslide's bug reporting is false-negative oriented, meaning that it does not check for suspicious behaviours that might indicate underlying bugs, so it may report that it found no bugs even if some existed.
@@
 This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
-\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), such as in Valgrind\cite{valgrind}.}
+\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), such as in Valgrind~\cite{valgrind}.}
 

### 2012-05-17 futurework.tex
fix cite fmt

@@
 
-Landslide could make use of data race detection techniques\cite{datacollider}, however, to strike a middle ground in which shared memory accesses it chooses.
+Landslide could make use of data race detection techniques~\cite{datacollider}, however, to strike a middle ground in which shared memory accesses it chooses.
 \begin{enumerate}
@@
 However, because DPOR's approach of tagging which sibling branches should be explored next generally follows a workqueue-based depth-first-search structure, it should not be too difficult to parallelise.
-Prior work exists for this technique\cite{distributed-dpor}, so it would not be a research contribution, but would substantially improve Landslide's effectiveness regardless.
+Prior work exists for this technique~\cite{distributed-dpor}, so it would not be a research contribution, but would substantially improve Landslide's effectiveness regardless.
 
@@
 One interesting phenomenon that would occur from combining symbolic execution and systematic exploration would be a new notion of decision points.
-Symbolic execution involves its own state space exploration, in which control flow branch points correspond to decision points; hence, to combine it with systematic exploration would create a hybrid state space in which a ``decision point'' could be a place where either different threads could preempt each other or multiple control flow paths could be taken\cite{dawson}.
+Symbolic execution involves its own state space exploration, in which control flow branch points correspond to decision points; hence, to combine it with systematic exploration would create a hybrid state space in which a ``decision point'' could be a place where either different threads could preempt each other or multiple control flow paths could be taken~\cite{dawson}.
 
@@
 
-Also in Section~\ref{sec:related-orthogonal}, we mentioned related research that attempts to {\em minimise} the set of conditions that are known to be associated with a bug\cite{dag-mining}.
+Also in Section~\ref{sec:related-orthogonal}, we mentioned related research that attempts to {\em minimise} the set of conditions that are known to be associated with a bug~\cite{dag-mining}.
 In the context of Landslide, this corresponds to how easy the decision trace is to understand.
@@
 
-Compared to Iterative Context Bounding\cite{chess}, which is based on the insight that bugs need fewer forced preemptions to expose, we believe that our insight is orthogonal.
+Compared to Iterative Context Bounding~\cite{chess}, which is based on the insight that bugs need fewer forced preemptions to expose, we believe that our insight is orthogonal.
 The ICB insight results in smaller decision trees overall needing to be explored before bugs appear, while our insight determines exploration ordering once the tree is constructed.

### 2012-05-17 related.tex
fix cite fmt

@@
 
-Systematic exploration is a relatively young approach to concurrency verification\cite{verisoft}.
+Compared to conventional stress testing, systematic exploration is a relatively young approach to concurrency verification~\cite{verisoft}.
 Recent work has focused on theoretical aspects of the technique, on distributed systems, and in userspace in general.
@@
 
-Dynamic Partial Order Reduction (DPOR)\hspace{0in}\cite{dpor,distributed-dpor,sdpor,dbug-retreat} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
+Dynamic Partial Order Reduction (DPOR)\hspace{0in}~\cite{dpor,distributed-dpor,sdpor,dbug-retreat} is an algorithm for reducing the state space of ``all possible thread interleavings'' by identifying independent thread transitions and pruning the redundant state spaces that result from interleaving those transitions.
 Landslide makes generous use of the DPOR algorithm.
 
-dBug\cite{dbug-ssv} is a systematic testing tool which makes use of DPOR.
+dBug~\cite{dbug-ssv} is a systematic testing tool which makes use of DPOR.
 It focuses on multithreaded userspace applications, identifying when to preempt by interposing on \revision{\texttt{libc} and \texttt{pthreads}} library calls.
@@
 
-CHESS\cite{chess} is another userspace systematic testing tool.
+CHESS~\cite{chess} is another userspace systematic testing tool.
 The authors explore the insight that many races require very few forced preemptions to uncover, and develop a search strategy which prioritises thread interleavings with fewer preemptions.
@@
 
-There are also several tools, such as MaceMC\cite{macemc} and MODIST\cite{modist}, which provide systematic testing frameworks for networked and distributed applications.
+There are also several tools, such as MaceMC~\cite{macemc} and MODIST~\cite{modist}, which provide systematic testing frameworks for networked and distributed applications.
 \revision{MODIST is especially notable in its ability to test unmodified distributed systems by using event interposition.}
 
-RacePRO\cite{racepro} is a tool based in kernel space which uncovers ``process races'' \revision{(races involving interleaving system call patterns between multiple processes)} using systematic exploration of multi-process interactions in user-space.
+RacePRO~\cite{racepro} is a tool based in kernel space which uncovers ``process races'' \revision{(races involving interleaving system call patterns between multiple processes)} using systematic exploration of multi-process interactions in user-space.
 They use system calls, such as filesystem accesses, as points to explore different interleavings of concurrent code, and detect bugs which can corrupt persistent system resources.
 
-Finally, DeMeter\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
+Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
 Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.
@@
 
-DataCollider\cite{datacollider} is a tool for detecting data races, a special kind of concurrency bug which involve concurrent unprotected accesses to memory that can interleave unsafely on a fine-grained level.
+DataCollider~\cite{datacollider} is a tool for detecting data races, a special kind of concurrency bug which involve concurrent unprotected accesses to memory that can interleave unsafely on a fine-grained level.
 DataCollider uses memory access sampling, and remains largely agnostic of synchronisation operations, to find data races in the Windows 7 kernel.
@@
 
-Carburizer\cite{carburizer} is an effort to make imperfect kernel-level drivers work more reliably in the presence of faulty peripheral devices.
+Carburizer~\cite{carburizer} is an effort to make imperfect kernel-level drivers work more reliably in the presence of faulty peripheral devices.
 They use static analysis to identify variables ``tainted'' by device input and unsafe uses thereof.
@@
 
-SimTester\cite{simtester} is another tool for testing kernel device drivers.
+SimTester~\cite{simtester} is another tool for testing kernel device drivers.
 Like Landslide, SimTester uses Simics as its execution environment, and identifies key decision points during execution at which to inject interrupts.
@@
 
-On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4\hspace{0in}\cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
+On the far end of the ``formality and correctness'' spectrum for kernel verification lies seL4 \cite{sel4}, a microkernel fully designed and specified in Haskell and translated directly into C.
 Because Haskell is type-safe (strict constraints on legal program behaviour that make it easier to reason about) and pure (most code cannot have stateful side-effects), the developers of seL4 were able to formally prove the specification's correctness and the implementation's adherence to the spec.
@@
 
-Barrelfish\cite{barrelfish} is another novel kernel architecture designed with concurrency in mind.
+Barrelfish~\cite{barrelfish} is another novel kernel architecture designed with concurrency in mind.
 The Barrelfish kernel relies exclusively on message-passing for inter-thread communication, for increased performance on massively multi-core systems.
@@
 Data race detection is the canonical runtime concurrency verification technique.
-Tools such as ThreadSanitizer\cite{tsan} track synchronisation operations and happens-before relationships to identify unsafe concurrent memory accesses.
+Tools such as ThreadSanitizer~\cite{tsan} track synchronisation operations and happens-before relationships to identify unsafe concurrent memory accesses.
 Such accesses, called ``data races'', might not necessarily cause incorrect behaviour, but are usually warning indicators of such, and represent bad coding practice in any case.
-\revision{RaceTrack\cite{racetrack} is another data race detection tool notable for its ability to adaptively adjust its testing granularity after some execution analysis.}
+\revision{RaceTrack~\cite{racetrack} is another data race detection tool notable for its ability to adaptively adjust its testing granularity after some execution analysis.}
 
-Symbolic execution\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpreting symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
-Tools such as KLEE\cite{klee} and projects such as Automated Exploit Generation\cite{aeg} demonstrate the effectiveness of symbolic execution in dynamic verification.
+Symbolic execution~\cite{symbolic,symbolic-disks} is a technique for testing programs by abstractly interpreting symbols and operations within a program, rather than directly executing its compiled code, which grants the ability to analyse conditional statements and cause-effect relationships among certain code paths.
+Tools such as KLEE~\cite{klee} and projects such as Automated Exploit Generation~\cite{aeg} demonstrate the effectiveness of symbolic execution in dynamic verification.
 
-There has also been recent research in searching for which conditions, among a known set of conditions associated with a buggy behaviour, are actually necessary to cause bugs to arise\cite{dag-mining}.
+There has also been recent research in searching for which conditions, among a known set of conditions associated with a buggy behaviour, are actually necessary to cause bugs to arise~\cite{dag-mining}.
 Such strategies are helpful because even once a bug is exposed, it can be difficult to determine what specifically went wrong.
@@
 Deterministic multithreading is a different technique for dealing with concurrency bugs which focuses on avoiding them when running production code rather than trying to expose them during testing.
-Deterministic multithreading tools, such as Kendo\cite{kendo}, Peregrine\hspace{0in}\cite{peregrine}, and DThreads\cite{dthreads}, analyse the execution of a concurrent system, compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
+Deterministic multithreading tools, such as Kendo~\cite{kendo}, Peregrine \cite{peregrine}, and DThreads~\cite{dthreads}, analyse the execution of a concurrent system, compute particular scheduling patterns which will not produce buggy behaviours, and force the system to follow those schedules while maintaining good parallel performance.
 Deterministic multithreading serves a different purpose than systematic exploration: it aims to ensure that code already running in the real world does not encounter concurrency bugs even though they might exist, while our goals are instead to uncover such bugs and help fix them beforehand.

### 2012-05-17 thesis.tex
'fix' figure in terminology section
fix cite fmt

@@
 \usepackage{changebar}
+\usepackage{colortbl}
 
@@
 Many techniques exist for dynamic testing of concurrent systems for race conditions.
-Systematic exploration\cite{chess}, the strategy we focus on in this work, involves making educated guesses at what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
+Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses at what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
 Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
-Compared to other dynamic analyses, such as data race detection\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
+Compared to other dynamic analyses, such as data race detection~\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
 
@@
 a tool for applying systematic race detection techniques in to kernel-level code.
-It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
+It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
 \revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.

### 2012-05-17 design.tex
insert science here

@@
 One alternative simpler method would be, when triggering a timer interrupt, to tell the guest kernel explicitly which thread should be run next.
-This would require the kernel programmer to write extra code in their timer handler and/or context switcher.
-We chose the ``in-flight'' approach instead, because it does not require additional kernel modifications.
+This approach could potentially save the cost of extra context switches, especially in test cases with very many threads.
+However, it would also require the kernel programmer to write extra code in their timer handler and/or context switcher.
+Because we mostly focus on test cases with few threads, we judged that ease of use was more important than the marginal performance increase, so we chose the ``in-flight'' approach instead.
 

### 2012-05-17 design.tex
more dave revisions

@@
 \begin{itemize}
+	\item {\bf Iterative refinement of interleaving granularity.} The recommended usage pattern is to begin by searching state spaces with coarse-grained interleavings, and then refining the decision set afterwards until bugs are found.
+Landslide automatically identifies a ``minimal decision set'', comprising only voluntary reschedules (Section~\ref{sec:components-arbiter}), to use as the coarsest granularity.
 	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby only generating thread interleavings around points within those modules.

### 2012-05-17 futurework.tex
more dave revisions

@@
 One possibility for backtracking would be to record-and-replay the events of each branch (up to the point at which Landslide chooses to have decided differently), restarting the system's execution from the beginning each time.
-Another option might be to snapshot the state of the kernel at each decision point, and use those to resume without re-executing everything from the beginning (kernel boot-up and initialisation is much more expensive than individual system calls, after all).
+Another option might be to snapshot the state of the kernel (and associated devices) at each decision point, and use those to resume without re-executing everything from the beginning (kernel boot-up and initialisation is much more expensive than individual system calls, after all).
 

### 2012-05-17 using.tex
more dave revisions

@@
 
+The recommended ``attack plan'' for using Landslide corresponds directly to the sections in this chapter.
+Section~\ref{sec:using-requirements} lists the requirements a kernel must fulfill to be compatible with Landslide.
+Section~\ref{sec:using-instrumenting} describes the process of instrumenting a kernel, which involves annotating the kernel, filling out a configuration file, and instrumenting within Landslide itself.
+Section~\ref{sec:using-customise} describes ways to configure Landslide's behaviour to help focus its search for bugs.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:using-requirements}
+
+In addition to meeting the Pebbles specifcation for 15-410, there are some additional restrictions that Landslide imposes, some inherent to its testing model, and others as artifacts of its interface/implementation.
 

### 2012-05-18 evaluation.tex
more dave revisions
fuzz things for dave & remove fixmes

@@
 Students also described compatibility issues they faced between their kernels and Landslide, which required them to make certain changes to their kernels.
-% FIXME: Say how we will address these issues?
 
@@
 In version (a) of this bug, we introduce a mutex for each process's parent pointer.
-When reparenting, the parent takes its own process control block (PCB) lock, iterates over its child-list, and takes its children's parent pointer locks.
-When signalling, the child takes its parent pointer lock, then takes the parent's PCB lock to ensure safe wake-up of waiters.
-This results in deadlock.
+\revision{While holding the lock around all related process structure accesses prevents data races, the circular nature of the lock acquisitions between parent and child results in deadlock.}
 
@@
 However, when awakened, the condition variable code did not clear the value of this pointer.
-Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been, and the kernel panicked.
+Fortunately, an assert statement in the thread destruction code caught that the pointer was non-\texttt{NULL} when it should have been \texttt{NULL}, and the kernel panicked.
 
@@
 This bug was also found during grading, and took substantially less time to verify by hand than the previous bug.
-When a thread does \texttt{yield} to a specific other TID, the other thread's thread control block (TCB) is not protected between lookup-by-TID and yield-to.
+When a thread does \texttt{yield} to a specific other TID, the other thread's thread control block (TCB) is not protected \revision{after being returned from the look-up function}.
 If that thread exits during that time, a use-after-free (and worse, a completely invalid schedule to a nonexistent thread) results.
@@
 This bug is moderately complicated, requiring a preemption during that specific window of the first thread's execution during which the second thread must exit entirely and have its TCB freed.
-The fix is moderately complicated: some protection mechanism must be designed so that threads cannot exit after being looked-up by TID.
+The fix is moderately complicated \revision{as well; since simply reordering lines of code won't offer the necessary protection.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Summary of Bugs Found}
-
-% FIXME
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 
-While evaluating Landslide on these bugs, we determined two invariants that should hold for multiple explorations on the same test case.
+While evaluating Landslide on these bugs, we determined two invariants that must hold for multiple explorations on the same test case.
 

### 2012-05-18 evaluation.tex
fix some figs in eval

@@
 \end{table}
-\begin{figure}[h]
-	\center
-	\includegraphics[width=0.5\textwidth]{graphs/studence-pie.pdf}
-	\caption{Average time breakdown for each phase of instrumenting a kernel with Landslide, based on student self-reported times.}
-	\label{fig:student-pie}
-\end{figure}
 
@@
 \end{itemize}
+
+\begin{wrapfigure}{r}{0.4\textwidth}
+	\center
+	\includegraphics[width=0.4\textwidth]{graphs/studence-pie.pdf}
+	\caption{Average time breakdown for each phase of instrumenting a kernel with Landslide, based on student self-reported times.}
+	\label{fig:student-pie}
+\end{wrapfigure}
 
@@
 	\begin{center}
-	\footnotesize
+	\small
 	\begin{tabular}{|l||c|c|c|c|}
@@
 		\hline
-		\multicolumn{5}{c}{Key: \bugnum{seconds}{stddev} indicates bug found; \nobugnum{seconds}{stddev} indicates whole tree explored with no bug.} \\
 	\end{tabular}
@@
 	the default set, consisting only of voluntary reschedules (Section~\ref{sec:components-arbiter}); and using custom decision points in addition to the default set: calls to \texttt{mutex\_lock}, calls to \texttt{mutex\_unlock}, and both.
-	All numbers represent the average from 5 trials, with the standard deviations given in parentheses. ``$\infty$'' indicates that Landslide's search did not finish (after 8 hours).
+	All numbers represent the average from 5 trials, with the standard deviations given in parentheses. \\
+	Key: \bugnum{seconds}{\footnotesize stddev} indicates bug found; \nobugnum{seconds}{\footnotesize stddev} indicates whole tree explored with no bug.
+	``$\infty$'' indicates that Landslide's search did not finish (after 8 hours).
 	}
@@
 	\center
-	\includegraphics[width=0.7\textwidth]{graphs/decisions.pdf}
-	\caption{Bar graph visualisation of the exploration time in trials that did find bugs. The decision trees that did not expose bugs, and the ones that timed out, are not shown.}
+	\includegraphics[width=0.8\textwidth]{graphs/decisions.pdf}
+	\caption{Bar graph visualisation of the exploration time in trials that did find bugs. The decision trees that did not expose bugs and the ones that timed out are not shown.}
 	\label{fig:number-graph}
@@
 	\begin{center}
-	\footnotesize
+	\small
 	\begin{tabular}{|l|l||c|c|c|c|}

### 2012-05-18 futurework.tex
fuzz things for dave & remove fixmes
add nadim's picture

@@
 	\centering
-	\footnotesize
+	\small
 	\begin{tabular}{|l|l||c|c|c|}
@@
 	\centering
-	\includegraphics[width=0.47\textwidth]{graphs/backwards-time.pdf}
-	\hspace{0.02\textwidth}
+	\includegraphics[width=0.48\textwidth]{graphs/backwards-time.pdf}
+	\hspace{0.01\textwidth}
 	\includegraphics[width=0.48\textwidth]{graphs/backwards-trace.pdf}
-	\caption{In general, backwards exploration finds bugs faster, but forwards exploration produces shorter (more understandable) decision traces.}
+	\caption{In general, backwards exploration finds bugs faster, but forwards exploration produces shorter (more understandable) decision traces. (Lower is better in both graphs.)}
 	\label{fig:backwards-graphs}
@@
 \revision{Table~\ref{fig:backwards} and Figure~\ref{fig:backwards-graphs} compare the tree explorations and resulting decision traces.
-The one exception is the POBBLES thread\_fork/vanish bug, which Landslide found immediately during forwards exploration, because the race actually required no forced preemptions to expose.}
+The one exception to backwards being faster is the POBBLES thread\_fork/vanish bug, which Landslide found immediately during forwards exploration, because the race actually required no forced preemptions to expose.}
 
@@
 However, when using decision points on both mutex functions, \texttt{double\_wait} took substantially longer, despite the tree structure suggesting they should have been comparable, with \texttt{double\_wait} even being perhaps slightly faster.
-Figure~\ref{fig:nadim-stats} shows a comparison of the tree anatomy for these two test cases.
+Table~\ref{fig:nadim-stats} shows a comparison of the tree anatomy for these two test cases.
 
@@
 		\hline
-		Average instructions/branch& 1,975,033 & 1,984,249 \\
+		Average instructions/branch& {\bf 1,975,033} & {\bf 1,984,249} \\
 		\hline
@@
 		\hline
-		% FIXME: add stddev?
 	\end{tabular}
@@
 The fact that Simics's backtracking avoids the cost of execution that happened before the backtrack target is to credit for this discrepancy.}
-% as visualised in figure\ldots TODO
+We show an exaggerated visualisation of this tree structure in Figure~\ref{fig:nadim}
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\textwidth]{nadim.pdf}
+	\caption{Though the execution trees in \texttt{double\_wait} and in \texttt{vanish\_vanish} had the same number of branches, with each approximately as many instructions deep, the backtracks in \texttt{double\_wait} were much longer, resulting in more {\em total} instructions executed during the exploration.}
+	\label{fig:nadim}
+\end{figure}
 

### 2012-05-18 related.tex
add nadim's picture

@@
 Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
-Landslide makes use of the same general concept in its user interface (Section~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.
+Landslide makes use of the same general concept in its user interface (Chapter~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.
 

### 2012-05-18 futurework.tex
credit vail photo

@@
 	\end{center}
-	\caption{Former members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
+	\caption{Former members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.\textsuperscript{1}}
 \end{figure*}
+\footnotetext{Joshua Wise, Ben Blum, and Michael Sullivan at Vail, Colorado. Included with permission.}
 

### 2012-05-18 using.tex
fuzz things for dave & remove fixmes
more syntax hilight

@@
 \begin{figure}[h]
-	% FIXME: replace with an image
 	\centering
@@
 	\texttt{\{} \\
-	\texttt{~~~~\hilight{green}{int}~tcb = \hilight{cyan}{READ\_MEMORY}(cpu, GUEST\_CURRENT\_TCB\_POINTER);} \\
-	\texttt{~~~~\hilight{green}{int}~state\_flag = \hilight{cyan}{READ\_MEMORY}(cpu, tcb + GUEST\_TCB\_STATE\_FLAG\_OFFSET);} \\
-	\texttt{~~~~\hilight{brown}{return}~state\_flag == GUEST\_TCB\_RUNNABLE;} \\
+	\texttt{~~~~\hilight{green}{int}~tcb = \hilight{cyan}{READ\_MEMORY}(cpu, \hilight{Salmon}{GUEST\_CURRENT\_TCB\_POINTER});} \\
+	\texttt{~~~~\hilight{green}{int}~state\_flag = \hilight{cyan}{READ\_MEMORY}(cpu, tcb + \hilight{Salmon}{GUEST\_TCB\_STATE\_FLAG\_OFFSET});} \\
+	\texttt{~~~~\hilight{brown}{return}~state\_flag == \hilight{Salmon}{GUEST\_TCB\_RUNNABLE};} \\
 	\texttt{\}} \\
@@
 	\texttt{\{} \\
-	\texttt{~~~~\hilight{green}{int}~x = \hilight{cyan}{READ\_MEMORY}(cpu, GUEST\_PREEMPTION\_FLAG);} \\
-	\texttt{~~~~\hilight{brown}{return}~x == GUEST\_PREEMPTION\_ENABLED;} \\
+	\texttt{~~~~\hilight{green}{int}~x = \hilight{cyan}{READ\_MEMORY}(cpu, \hilight{Salmon}{GUEST\_PREEMPTION\_FLAG});} \\
+	\texttt{~~~~\hilight{brown}{return}~x == \hilight{Salmon}{GUEST\_PREEMPTION\_ENABLED;}} \\
 	\texttt{\}} \\

### 2012-05-18 using.tex
appease dave more

@@
 
+\revision{
 \begin{itemize}
-        \item Use \texttt{tell\_landslide\_decide()} to indicate decision points.\footnote{
-	We recommend at the start of \texttt{mutex\_lock} and at the end of \texttt{mutex\_unlock}, though it is also important to use one's own intuition, depending on which system call(s) are being tested.}
-        \item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.
-        \item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of decision points, and figure out which ones are irrelevant.
-        \item Use \texttt{within\_func} and \texttt{without\_func} to make Landslide only use relevant decision points.
+        \item {\bf Indicating decision points.} Insert calls to \texttt{tell\_landslide\_decide()} in the guest kernel where Landslide should identify decision points.
+        We recommend at the start of \texttt{mutex\_lock} and at the end of \texttt{mutex\_unlock}, though it is also important to use one's own intuition, depending on which system call(s) are being tested.
+        \item {\bf Ignore always-conflicting memory locations.} Use \texttt{sched\_func} and \texttt{ignore\_sym} to identify shared memory locations that Landslide should ignore, to enable better pruning.
+        \item {\bf Examine decision points.} Set \texttt{DECISION\_INFO\_ONLY} to make Landslide print the decision points instead of exploring.
+Examine the decision set, and figure out which ones are irrelevant to the test.
+        \item {\bf Focus decision points.} Use \texttt{within\_func} and \texttt{without\_func} to specify which components of the kernel Landslide should identify decision points in, to reduce the state space and focus only on relevant interleavings.
 \end{itemize}
+}
 
@@
 For example, ignoring a mutex which is locked very frequently, but irrelevant to the actual test case, will result in an independence relation that \revision{enables much better state space reduction without losing the ability to find bugs in the parts of the kernel the user is interested in testing}.
+\end{itemize}
+
+\revision{These configuration options enable the user to whitelist or blacklist certain components of the kernel for the purposes of identifying decision points.}
+
+\begin{itemize}
 	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.

### 2012-05-18 conclusion.tex
minor updates

@@
 
-We hope that the ideas presented here serve as stepping stones to develop more sophisticated systematic debugging techniques for kernel-level race conditions.
+We hope that the ideas presented here serve as stepping stones for future development of more sophisticated systematic debugging techniques for kernel-level race conditions.
 }
@@
 
+%\begin{figure}[p]
 \begin{quote}
@@
 \end{quote}
+%\end{figure}

### 2012-05-18 futurework.tex
remove some todos and fixmes
minor updates

@@
 
-% TODO: make visualisations for both of these
 \subsection{``Backwards'' Exploration}
@@
 The fact that Simics's backtracking avoids the cost of execution that happened before the backtrack target is to credit for this discrepancy.}
-We show an exaggerated visualisation of this tree structure in Figure~\ref{fig:nadim}
+\revision{We show an exaggerated visualisation of this tree structure in Figure~\ref{fig:nadim}.
 
@@
 \end{figure}
+}
 

### 2012-05-18 design.tex
patch up design

@@
 	\centering
-	\includegraphics[width=0.75\textwidth]{landslide.png}
+	\includegraphics[width=0.65\textwidth]{landslide.png}
 	\caption{A simple visualisation of Landslide and its components, and how they interact with the guest kernel.}
@@
 The Landslide scheduler also tracks which important actions each thread is performing.
-These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section \revision{(note that ``vanish'' is simply Pebbles's name for ``exit'')}, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing, and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
+These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section \revision{(note that ``vanish'' is simply Pebbles's name for ``exit'')}, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing (voluntary reschedule, etc), and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
 There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}
@@
 
-The Landslide scheduler relies on the guest kernel informing it about certain important events in the thread lifecycle.
+The Landslide scheduler relies on the guest kernel to inform it about certain important events in the thread lifecycle.
 The guest kernel does this by means of annotations (which are described in Section~\ref{sec:using-annotations}).
@@
 
-Landslide's scheduling technique, called the {\em schedule in-flight}, involves successive triggering of timer interrupts to trigger context switches until the desired thread begins to run.
+Landslide's scheduling technique, called the {\em schedule in-flight}, involves successively injecting timer interrupts to trigger context switches until the desired thread begins to run.
 Algorithm~\ref{alg:inflight} shows how Landslide makes this happen.

### 2012-05-18 design.tex
copyedit

@@
 Landslide assumes that timer interrupts are the only source of nondeterminism for the guest kernel, so controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case.
-In future work (Section~\ref{sec:future-drivers}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
+In future work (Section~\ref{sec:future-drivers}), we also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.
 
@@
 Landslide automatically identifies a ``minimal decision set'', comprising only voluntary reschedules (Section~\ref{sec:components-arbiter}), to use as the coarsest granularity.
-	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby only generating thread interleavings around points within those modules.
+	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby generating thread interleavings only around points within those modules.
 	\item {\bf Memory conflict selection.} Landslide also provides a mechanism for ignoring shared memory conflicts on certain common data structures (especially those accessed in every thread transition, such as the scheduler queues), to enable DPOR to achieve greater reduction.
@@
 
-We advocate relying on the user for such information because the user need only understand the basics of their kernel design (which is difficult for Landslide to guess) while not knowing in advance what bugs are being looked for (which Landslide is able to provide).
+We advocate relying on the user for such information because the user need only understand the basics of their kernel design (which is difficult for Landslide to guess) while not knowing in advance which bugs are being looked for (which Landslide is able to provide).
 We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user ``steers'' Landslide towards focused search spaces that are more likely to find bugs.
@@
 	\centering
-	\includegraphics[width=0.65\textwidth]{landslide.png}
+	\includegraphics[width=0.6\textwidth]{landslide.png}
 	\caption{A simple visualisation of Landslide and its components, and how they interact with the guest kernel.}
@@
 The Landslide scheduler also tracks which important actions each thread is performing.
-These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section \revision{(note that ``vanish'' is simply Pebbles's name for ``exit'')}, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing (voluntary reschedule, etc), and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
+These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section,\footnote{\revision{Note that ``vanish'' is simply Pebbles's name for ``exit''.}} and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing (voluntary reschedule, etc), and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
 There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}
@@
 This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
-\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), such as in Valgrind~\cite{valgrind}.}
+\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), in a similar fashion to Valgrind~\cite{valgrind}.}
 
@@
 
-In Landslide, it is always necessary to identify a decision point during a voluntary reschedule, to maintain the invariant that each transition between decision points is comprised of the execution of only one thread.
+In Landslide, it is always necessary to identify a decision point during a voluntary reschedule, to maintain an invariant that each transition between decision points is comprised of the execution of only one thread.
 \revision{In future work, we may also investigate the implications of allowing some transitions to consist of multiple threads' executions, which would have applications for heuristic state space reduction.}
@@
 Because voluntary reschedules happen of the guest kernel's own volition, when a thread begins one, it will necessarily not have been preempted by the timer handler.
-Hence, we identify voluntary reschedules when a thread enters the context switcher without having entered the timer handler.\footnote{
+Hence, we identify voluntary reschedules by determining when a thread enters the context switcher without having entered the timer handler.\footnote{
 Landslide actually identifies voluntary reschedules at the point when the reschedule ends, so the {\em previous} thread would not have entered the timer handler, and the current thread may or may not have.
@@
 \begin{itemize}
-	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
+	\item {\bf Is the shell waiting for keyboard input?} In general, the shell waits for keyboard input only when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
 		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process might not yet have reaped a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well.
@@
 The states in the test lifecycle, and the rules for transitioning between them, are as follows.
+Figure~\ref{fig:teststate} depicts the states and transitions visually.
 
@@
 \end{enumerate}
-
-Figure~\ref{fig:teststate} depicts the states and transitions visually.
 
@@
 \subsection{Probable Bug-Detection Conditions}
+
+\resume{enumerate}
+	\item {\bf Memory leak bugs.} Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
+If memory allocated during the test was not freed, Landslide assumes that it was leaked.
+		(Some kernel designs may legitimately behave this way, so this bug-check may be disabled when testing such kernels.)\footnote{There is, of course, much room for improvement in this metric, but it is not part of the research contribution.}
+	\item {\bf Infinite loop bugs.} Landslide judges whether the kernel has entered an infinite loop by comparing the current branch of the decision tree to past executions of the same test case.
+	Figure~\ref{fig:infiniteloop} depicts such tree structures in abstract.
+	\begin{itemize}
+		\item {\bf Infinite loops without decision points.} While exploring the decision tree, Landslide computes the average number of instructions executed between two consecutive decision points.
+If at any point the current number of instructions executed since the most recent decision point exceeds this average times a constant factor (arbitrarily chosen to be 2000), Landslide assumes the kernel must have gotten stuck in an infinite loop.
+		\item {\bf Infinite loops around decision points.} Landslide also computes the average number of decision points in each branch of the decision tree (the average ``branch depth'').
+If the depth of the current branch ever exceeds this average times a constant factor (arbitrarily chosen to be 20), Landslide assumes the kernel must have gotten stuck in an infinite loop.
+	\end{itemize}
+	\revision{In our experience, these heuristics have never falsely identified an infinite loop when the kernel was making real progress.
+	However, both of these heuristics require a minimum number of branches to be explored before Landslide considers the already-explored tree structure to be statistically significant (arbitrarily chosen to be 20).
+	As such, these heuristics occasionally fail to trigger on a real infinite loop in cases when the loop occurs before enough ``safe'' branches were explored.
+In the future we might improve this heuristic by scaling the associated cutoff factor depending on how many branches were explored: the fewer branches explored so far, the less reliable the comparison, and hence the higher the cutoff factor to be used.}
+\end{enumerate}
 
@@
 }
-
-\resume{enumerate}
-	\item {\bf Memory leak bugs.} Landslide records the state of the heap before the test case begins, and compares it to the state of the heap after the test case ends.
-If memory allocated during the test was not freed, Landslide assumes that it was leaked.
-		(Some kernel designs may legitimately behave this way, so this bug-check may be disabled when testing such kernels.)\footnote{There is, of course, much room for improvement in this metric, but it is not part of the research contribution.}
-	\item {\bf Infinite loop bugs.} Landslide judges whether the kernel has entered an infinite loop by comparing the current branch of the decision tree to past executions of the same test case.
-	Figure~\ref{fig:infiniteloop} depicts such tree structures in abstract.
-	\begin{itemize}
-		\item {\bf Infinite loops without decision points.} While exploring the decision tree, Landslide computes the average number of instructions executed between two consecutive decision points.
-If at any point the current number of instructions executed since the most recent decision point exceeds this average times a constant factor (arbitrarily chosen to be 2000), Landslide assumes the kernel must have gotten stuck in an infinite loop.
-		\item {\bf Infinite loops around decision points.} Landslide also computes the average number of decision points in each branch of the decision tree (the average ``branch depth'').
-If the depth of the current branch ever exceeds this average times a constant factor (arbitrarily chosen to be 20), Landslide assumes the kernel must have gotten stuck in an infinite loop.
-	\end{itemize}
-	\revision{In our experience, these heuristics have never falsely identified an infinite loop when the kernel was making real progress.
-	However, both of these heuristics require a minimum number of branches to be explored before Landslide considers the already-explored tree structure to be statistically significant (arbitrarily chosen to be 20).
-	As such, these heuristics occasionally fail to trigger on a real infinite loop in cases when the loop occurs before enough ``safe'' branches were explored.
-In the future we might improve this heuristic by scaling the associated cutoff factor depending on how many branches were explored: the fewer branches explored so far, the less reliable the comparison, and hence the higher the cutoff factor to be used.}
-\end{enumerate}
 
@@
 
-\revision{Instead, Landslide uses considers every evil ancestor of each transition when performing DPOR.
+\revision{Instead, Landslide considers every evil ancestor of each transition when performing DPOR.
 This modification enables Landslide to avoid the problem presented in the figure, because it will identify both alternate interleavings after exploring the first one.
@@
 	\item If a use-after-free bug is found, Landslide prints information about the most-recently-freed chunk containing the accessed address: the stack trace and TID both for when it was allocated and freed.
-	\item If deadlock is detected, Landslide prints the cycle of TIDs that are blocked on each other.\footnote{This currently only works when using the annotations for yielding mutexes, though it is not difficult to implement more generally.}
+	\item If deadlock is detected, Landslide prints the cycle of TIDs that are blocked on each other.\footnote{This currently works only when using the annotations for yielding mutexes, though it is not difficult to implement more generally.}
 	\item If a memory leak is suspected, Landslide prints how many bytes bigger the heap is after the test ended than when the test began.\footnote{Future feature: Printing when-allocated stack traces for each suspicious heap chunk.}

### 2012-05-18 related.tex
remove some todos and fixmes
minor updates
copyedit

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% FIXME: Give another pass for dad's feedback - another sentence about each thing saying what landslide does better
 
@@
 
-Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time.
+Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique to only one component of the program at a time.
 Landslide makes use of the same general concept in its user interface (Chapter~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.
@@
 
+\revision{
 SimTester~\cite{simtester} is another tool for testing kernel device drivers.
@@
 Additionally, SimTester's testing model involves forcing only one interrupt per test run, which is a lighter-weight approach than systematic exploration.
+}
 

### 2012-05-18 terminology.tex
copyedit

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:terminology}
 
@@
 
-\revision{The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.}
+\revision{The following terms refer to parts of Landslide's analysis, regardless of the implementation of the kernel under test.}
 
@@
 		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause.
-Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace may also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
+Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace could also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
 \end{enumerate}

### 2012-05-18 thesis.tex
fix some figs in eval
copyedit

@@
 \usepackage{colortbl}
+\usepackage{wrapfig}
 
@@
 Many techniques exist for dynamic testing of concurrent systems for race conditions.
-Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses at what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
+Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses as to what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
 Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
@@
 Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
-a tool for applying systematic race detection techniques in to kernel-level code.
+a tool for applying systematic race detection techniques to kernel-level code.
 It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
@@
 \revision{
-In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
+In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), define common terms used in the rest of this document (Chapter~\ref{sec:terminology}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
 We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).

### 2012-05-18 conclusion.tex
more copyedit

@@
 %It is more reliable than stress testing, and able to find more diverse types of bugs than data race detection.
-In previous research, systematic exploration has proven useful in user-level and distributed systems, yet a gap stands between the technique and the complex world of kernel-level concurrency debugging.
+In previous research, systematic exploration has proven useful in user-level code and distributed systems, yet a gap stands between the technique and the complex world of kernel-level concurrency debugging.
 In this thesis, we present techniques for bridging this gap.

### 2012-05-18 evaluation.tex
remove some todos and fixmes
more copyedit

@@
 	\item Implementing instrumentation within Landslide (Section~\ref{sec:using-student-c}).
-	\item Fixing problems encountered getting Landslide to run, such as editing their kernel to meet Landslide's requirements and debugging incorrect instrumentation.
+	\item Fixing problems encountered getting Landslide to run, such as revising their kernel to meet Landslide's requirements and debugging incorrect instrumentation.
 	\item Customising Landslide's search parameters such as decision points (Section~\ref{sec:using-customise}).
@@
 		% ntaha/jmchow
-		\multirow{2}{*}{Group 1} & 35 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 5 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} \\
+		\multirow{2}{*}{Group 1} & 35 & {\footnotesize \em --} & {\footnotesize \em --} & 5 & {\footnotesize \em --} & {\footnotesize \em --} & {\footnotesize \em --} & {\footnotesize \em --} \\
 		\cline{2-9}
-		& 5 & 30 & 10 & 100 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 145 & {\footnotesize \em N/A} \\
+		& 5 & 30 & 10 & 100 & {\footnotesize \em --} & {\footnotesize \em --} & 145 & {\footnotesize \em --} \\
 		\hline
 		% echo/margaret
-		\multirow{2}{*}{Group 2} & 20 & 30 & 10 & 80 & {\footnotesize \em N/A} & 10 & 140 & 10 \\
+		\multirow{2}{*}{Group 2} & 20 & 30 & 10 & 80 & {\footnotesize \em --} & 10 & 140 & 10 \\
 		\cline{2-9}
-		& 15 & 45 & 30 & 60 & {\footnotesize \em N/A} & 30 & 150 & 30 \\
+		& 15 & 45 & 30 & 60 & {\footnotesize \em --} & 30 & 150 & 30 \\
 		\hline
@@
 		% pjumde
-		\multirow{2}{*}{Group 4} & 10 & 42 & 3 & 5 & 30 & {\footnotesize \em N/A} & 60 & 30 \\
+		\multirow{2}{*}{Group 4} & 10 & 42 & 3 & 5 & 30 & {\footnotesize \em --} & 60 & 30 \\
 		\cline{2-9}
-		& 15 & 35 & {\footnotesize \em N/A} & 5 & 35 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & 35 \\
+		& 15 & 35 & {\footnotesize \em --} & 5 & 35 & {\footnotesize \em --} & {\footnotesize \em --} & 35 \\
 		\hline
 		% apodolsk
-		Group 5 & 50 & 15 & {\footnotesize \em N/A} & 15 & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} & {\footnotesize \em N/A} \\
+		Group 5 & 50 & 15 & {\footnotesize \em --} & 15 & {\footnotesize \em --} & {\footnotesize \em --} & {\footnotesize \em --} & {\footnotesize \em --} \\
 		\hline
-		Former TA & 40 & 20 & 3 & 95 & {\footnotesize \em N/A} & 30 & 158 & 30 \\
+		Former TA & 40 & 20 & 3 & 95 & {\footnotesize \em --} & 30 & 158 & 30 \\
 		\hline
@@
 	\caption{Self-reported times spent by users on each phase while using Landslide.
-``N/A'' indicates a student reported spending no time on a particular phase.
+``--'' indicates a student reported spending no time on a particular phase.
 ``Required'' indicates the sum of the times from the first four phases; ``refinement'' inducates the sum from the last two phases.
-	``N/A'' indicates that a user did not report spending any time on a phase; the ``required'' field is ``N/A'' if any of its represented phases are, and ``N/A'' fields are not reflected in the averages.}
+	``--'' indicates that a user did not report spending any time on a phase; the ``required'' field is ``--'' if any of its represented phases are, and ``--'' fields are not reflected in the averages.}
 	\label{fig:student-times}
@@
 
-% TODO: Investigate the tree size that bugs did exist in. Investigate how many buggy branches there were.
-
 In light of these, we recommend several principles to govern an overall strategy to automatically iterate through different sets of decision points in search of a bug.\footnote{

### 2012-05-18 futurework.tex
more copyedit

@@
 
-If Landslide becomes integrated into an undergraduate course curriculum, it will be important to minimise the amount of effort students have to go through to instrument their kernels.
+If Landslide becomes integrated into an undergraduate course curriculum, it will be important to reduce the amount of effort students must go through to instrument their kernels.
 This could be achieved in part by changing the project definition in ways that make Landslide better able to automatically instrument kernels.
@@
 With the current specification, inappropriate idling is merely a performance bug, though it is incompatible with Landslide.
-If the project gave stricter requirements regarding this, fewer students would need to fix problems with their scheduler idling during the instrumentation process.
+%If the project gave stricter requirements regarding this, fewer students would need to fix problems with their scheduler idling during the instrumentation process.
 	\item {\bf Mandated function names for scheduling events.} As part of the instrumentation process, students tell Landslide where in their schedulers notable events happen.
-Landslide could automatically detect some of these if the project mandated the names of certain functions, such as the timer handler, context switcher, and runqueu manipulation routines.
+Landslide could automatically detect some of these if the project mandated the names of certain functions, such as the timer handler, context switcher, and runqueue manipulation routines.
 \end{enumerate}
@@
 
-For example, one of the groups in the user study said they were looking for a race that only arose in out of memory conditions.
+For example, one of the groups in the user study said they were looking for a race that only arose in out-of-memory conditions.
 A test case that could expose that race, by exhausting the kernel's memory resources, would necessarily have a very large state space that would be impractical to explore systematically.
 
-We suggested to the group to configure their kernel to fail memory allocations every so often, so Landslide would be better able to find the race while still using a small test case.
+We suggested that the group configure their kernel to fail memory allocations every so often, so Landslide would be better able to find the race while still using a small test case.
 To make Landslide more applicable to such bugs in general, we might further investigate this debugging strategy in a more structured way.
@@
 
-Compared to executing the code being tested on real hardware (which tools such as \cite{dbug-ssv} are able to do for user-space programs), Simics is {\em slow}, and even slower with Landslide hooked up to it.
+Compared to executing the code being tested on real hardware (which tools such as \cite{dbug-ssv} are able to do for user-space programs), simulation in Simics is slow, and even slower with Landslide hooked up to it.
 

### 2012-05-18 design.tex
move test diagram

@@
 
-\begin{figure}[h]
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{teststate.png}
-	\end{center}
-	\caption{The test lifecycle state machine.}
-	\label{fig:teststate}
-\end{figure}
 Landslide knows the state of the test case through a simple state machine, which is updated with information about the number of threads currently on the system, the number of threads on the runqueue, the state of the idle thread (if it exists), and whether or not the shell is waiting for keyboard input.
@@
 
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{teststate.png}
+	\end{center}
+	\caption{The test lifecycle state machine.}
+	\label{fig:teststate}
+\end{figure}
 The states in the test lifecycle, and the rules for transitioning between them, are as follows.

### 2012-05-18 design.tex
ryan's copyedit

@@
 
-Landslide's model for applying systematic exploration in kernel-space is comprised of four key points, which we overview here.
+Landslide's model for applying systematic exploration in kernel space is comprised of four key points, which we overview here.
 
@@
 This fits directly into its model that runnable threads are ones that can be caused to run with a finite number of timer ticks in succession.
-	\item {\bf No idling when progress can be made.} The kernel must not enter its idle loop (whether in an explicit idle thread or no) when the kernel is not truly idle.
+	\item {\bf No idling when progress can be made.} The kernel must not enter its idle loop (whether in an explicit idle thread or not) when the kernel is not truly idle.
 Landslide uses the idle loop to detect when a test begins/ends (test lifecycle tracking) and to detect when all threads in a test are wedged (bug detection).
@@
 	\item {\bf Test cleanup.} The test processes have exited, but some housekeeping may yet remain. When the the kernel is idling and the shell is waiting for keyboard input, advance state.
-	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack (to step 3).
+	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide where to rewind to, and backtrack (to step 3).
 \end{enumerate}

### 2012-05-18 futurework.tex
ryan's copyedit

@@
 \begin{itemize}
-	\item Landslide currently does not support ability to replay a particular choice sequence.
+	\item {\bf Replay.} Landslide currently does not support ability to replay a particular choice sequence.
 If the user finds a bug, and wants to re-execute the interleaving that led to it, they have no choice but to re-explore the tree.
-	\item Although Simics can pause execution and give the user a debug prompt, Landslide currently does not support being interrupted during exploration.
+	\item {\bf Debug prompt support.} Although Simics can pause execution and give the user a debug prompt, Landslide currently does not support being interrupted during exploration.
 		This is because the current implementation uses a wrapper script around the Simics prompt which Landslide communicates with to perform backtracking.
 		Though it would be difficult to redesign, it would be good to work around, to enable the user to interrupt, use the Simics debug prompt arbitrarily, and resume Landslide when ready.
-	\item Landslide could print its decision traces in a better format.
+	\item {\bf Decision trace formatting.} Landslide could print its decision traces in a better format.
 The current format is basically a ``raw data dump''.
 Raw text is probably the wrong interface; a graphical/tabular format with distinct columns for the execution of each different thread would be easier to understand.
-	\item Landslide could attempt to detect when something has gone wrong that is not its fault (such as incorrect instrumentation, or the kernel violating some of its requirements), and print a useful error message rather than behaving mysteriously.
+	\item {\bf Feedback for incorrect instrumentation.} Landslide could attempt to detect when something has gone wrong that is not its fault (such as incorrect instrumentation, or the kernel violating some of its requirements), and print a useful error message rather than behaving mysteriously.
 \end{itemize}

### 2012-05-18 related.tex
ryan's copyedit

@@
 
-DataCollider~\cite{datacollider} is a tool for detecting data races, a special kind of concurrency bug which involve concurrent unprotected accesses to memory that can interleave unsafely on a fine-grained level.
+DataCollider~\cite{datacollider} is a tool for detecting data races, a special kind of concurrency bug which involves concurrent unprotected accesses to memory that can interleave unsafely on a fine-grained level.
 DataCollider uses memory access sampling, and remains largely agnostic of synchronisation operations, to find data races in the Windows 7 kernel.
@@
 Carburizer is able to transform such badly-written code to be more robust.
-Kernel device drivers are a ripe breeding ground for bugs, because they are frequently written by less careful developers, and the sheer volume of code makes it difficult to rigorously review.
+Kernel device drivers are a ripe breeding ground for bugs, because they are frequently written by less careful developers, and the sheer volume of code makes them difficult to rigorously review.
 Due to the interrupt-driven nature of many drivers, they are also prone to concurrency bugs, which Carburizer does not address.

### 2012-05-18 terminology.tex
ryan's copyedit

@@
 		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause.
-Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace could also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
+The trace contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace could also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
 \end{enumerate}

### 2012-05-18 thesis.tex
ryan's copyedit

@@
 Many aspects of the concurrency implementation itself are part of the system being tested.
-While user-space testing frameworks rely on the underlying kernel to provide common concurrency abstractions which the framework can use to drive the test, systematic exploration in the kernel itself necessarily exists underneath the level of such abstractions.
+While user-space testing frameworks rely on the underlying kernel to provide common concurrency abstractions which the framework can use to drive the test, systematic exploration in the kernel itself necessarily exists underneath such abstractions.
 Furthermore, a tool that supports testing multiple kernels must support many different possible scheduler designs.

### 2012-05-18 using.tex
more copyedit
more

@@
 
-\revision{Both of these functions are predicates on the guest kernel's execution state (usually on the contents of its scheduler's data structures).
+\revision{Both of these functions are predicates on the guest kernel's execution state, usually on the contents of its scheduler's data structures.
 Example implementations are given in Figure~\ref{fig:student-c}.}
@@
 	\item \texttt{within\_function}: If the user configures decision points to happen in common code paths, such as \texttt{mutex\_lock}, it will likely generate more branching than needed.
-For example, if testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.
+For example, if testing \texttt{vanish}, the user may judge there to be no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.
 

### 2012-05-18 evaluation.tex
add nobe and fix virt timer

@@
 Finding this bug was somewhat of a surprise.
-When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{cond\_queue\_next} pointer.
+When two threads in the same process wait for a child to exit, they both block on the same condition variable, and one of them has a pointer to the other in its \texttt{nobe->cond\_queue\_next} pointer.
 However, when awakened, the condition variable code did not clear the value of this pointer.

### 2012-05-18 futurework.tex
add nobe and fix virt timer

@@
 Finally, Landslide will need to find new ways to control non-deterministic events in the system.
-Since virtualised kernels use the same timer interrupt as the host they run on, Landslide \revision{will need to intercept it, and perhaps control the guest kernel's context switching with a custom ``Landslide-driven'' context switcher instead.}
+Since timer ticks in virtualised kernels are mediated by the hypervisor, Landslide \revision{may need to intercept the timer, and perhaps control the guest kernel's context switching with a custom ``Landslide-driven'' context switcher instead.}
 

### 2012-05-18 acknowledgements.tex
some ack

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Save this for last :)
 
-Thanks to Deb Cavlovich, for holding on to this thesis for a few days more for Ben to finish it properly.
+This has been an immensely fun and fulfilling project, and it wouldn't have been possible without the contributions, guidance, and support of many people.
 
-Thanks to my advisor, too, I guess, and maybe also to the people I worked with and my friends as well.
+\subsubsection{Big Thanks}
+
+First of all, I owe the lion's share of my gratitude equally to Garth Gibson, Jiri Simsa, and Dave Eckhardt.
+
+Garth's advisorship over the past year has been excellent, from showing me how to keep my research both useful and personally satisfying, to teaching me how to share ideas with fellow researchers, to helping me think through far-off future work directions.
+Not working directly with his main research groups was sometimes unsettling, but validation was not in short supply: when Garth emailed me unprompted asking for my input on the future of my research field was definitely one of the high points.
+
+Jiri's mentorship was invaluable, both in explanations of complicated algorithms and in general advice on good researcher habits.
+I would have had a much harder time understanding Dynamic Parial Order Reduction without him, and I suspect my posters and talks would have been much less put-together as well.
+
+Working with Dave to share my project with 15-410 really made the thesis come together.
+I knew I would have a rough time making Landslide available to 15-410 without compromising the class's integrity policies, and Dave helped me make it work.
+Dave's advice was also crucial when designing the lecture I gave to 15-410 to explain Landslide and recruit students for the user study; I learned more about lecture design during those two weeks than I ever thought there was to know.
+
+\subsubsection{Landslide's Users}
+
+I'm also highly grateful for all the patient work put in by the kernel programmers who used Landslide.
+
+- user study folks, and TAs, esp nwf, also jboning acrichto rpearl
+
+\subsubsection{Support}
+
+- deb and cathy
+
+- revisers and support - carlo, wolf, sully, joshua, maurer, ghanners
+
+\subsubsection{Collaboration}
+
+A big thanks to Wind River for Simics, the simulation software for which Landslide is built.
+While developing code to fit into larger infrastructure always has its hairy points, Simics's interfaces were generally pleasant, and the expressive power of operating in its simulation environment was invaluable.
 
@@
 interest, insights, feedback, and support.
+
+I hope you enjoy reading some or all of this.

### 2012-05-18 acknowledgements.tex
final ack

@@
 
-First of all, I owe the lion's share of my gratitude equally to Garth Gibson, Jiri Simsa, and Dave Eckhardt.
+First of all, I owe the lion's share of my gratitude equally to Garth Gibson, Jiri Simsa, and David Eckhardt.
 
 Garth's advisorship over the past year has been excellent, from showing me how to keep my research both useful and personally satisfying, to teaching me how to share ideas with fellow researchers, to helping me think through far-off future work directions.
-Not working directly with his main research groups was sometimes unsettling, but validation was not in short supply: when Garth emailed me unprompted asking for my input on the future of my research field was definitely one of the high points.
+Not working directly with his main research groups was sometimes unsettling, but validation was not in short supply: for example, it was definitely a high point when Garth emailed me unprompted asking for my input on the future of my research field.
 
@@
 
-Working with Dave to share my project with 15-410 really made the thesis come together.
-I knew I would have a rough time making Landslide available to 15-410 without compromising the class's integrity policies, and Dave helped me make it work.
-Dave's advice was also crucial when designing the lecture I gave to 15-410 to explain Landslide and recruit students for the user study; I learned more about lecture design during those two weeks than I ever thought there was to know.
+Working with David to share my project with the students of 15-410 really made the thesis come together.
+I knew I would have a rough time making Landslide available without compromising the class's integrity policies, and David helped me make it work.
+His advice was also crucial when designing the lecture I gave to the class to explain Landslide and recruit students for the user study; I learned more about lecture design during those two weeks than I ever thought there was to know.
 
@@
 
-- user study folks, and TAs, esp nwf, also jboning acrichto rpearl
+Nathaniel Filardo, as the first person to use Landslide besides myself, ran into several compatibility issues (and outright bugs) in Landslide when instrumenting his kernel.
+For all Landslide gave him a hard time, though, he stayed with me, was patient while I debugged, and even debugged some on his own when I was too busy myself.
+It was when he got Landslide working with his kernel that I finally found confidence that it'd be able to work on arbitrary student kernels.
 
-\subsubsection{Support}
+Thanks also to the other members (and former members) of 15-410 staff who used Landslide on their kernels to help me prepare for working with the students: Josiah Boning, Alex Crichton, and Michael Sullivan.
 
-- deb and cathy
+A special form of thanks to the students I worked with, who not only had to navigate Landslide's interfaces but also had to learn the underlying concepts as well, all while under time pressure to finish their kernel projects at the same time.
+To Nadim Taha, Margaret Schervish, Timothy Passaro, Joon-Sup Han, Pranjal Jumde, and others who did not ask to be mentioned by name: I could not have evaluated Landslide without you.
 
-- revisers and support - carlo, wolf, sully, joshua, maurer, ghanners
+Another piece of thanks goes to Nadim Taha for his above-and-beyond investigation of the decision trees Landslide explored using his kernel, which resulted in the insights presented in Section~\ref{sec:future-nadim}.
 
@@
 
-A big thanks to Wind River for Simics, the simulation software for which Landslide is built.
+Thanks to Wind River for Simics\textsuperscript{\texttrademark}, the simulation software for which Landslide is built.
 While developing code to fit into larger infrastructure always has its hairy points, Simics's interfaces were generally pleasant, and the expressive power of operating in its simulation environment was invaluable.
@@
 interest, insights, feedback, and support.
+It has been a pleasure to work in this research group.
 
-I hope you enjoy reading some or all of this.
+\subsubsection{Support}
+
+More gratitude goes to the people who helped me revise my thesis document and defence talk.
+While Garth and David provided the most detailed criticism, many of my friends helped me polish my work: Wolfgang Richter, Carlo Angiuli, Joshua Wise, Ryan Pearl, and Michael Sullivan critiqued this document, and Jiri Simsa, Matthew Maurer, Greg Hanneman, and Carlo Angiuli critiqued my defence presentation.
+
+Finally, huge thanks to Deborah Cavlovich and Catherine Copetas, whose organisational efforts carried me through the Master's program. These are the people who make the gears turn under the hood.
+
+\vspace{0.5in}
+
+\noindent I hope you enjoy reading some or all of this.

### 2012-05-31 rv/paper.tex
start writing for RV

--- 2012-05-18 thesis.tex
@@
-\documentclass[12pt]{book}
+\documentclass{llncs}
 %\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
-\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
+%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
 \usepackage{upgreek}
 %\usepackage{times}
-\usepackage[charter]{mathdesign}
+%\usepackage[charter]{mathdesign}
 \usepackage{hyperref}
-\usepackage{draftwatermark}
 \usepackage{algorithm}
@@
 \usepackage[bottom]{footmisc}
-\usepackage{mjz-titlepage}
+%\usepackage{mjz-titlepage}
 \usepackage{framed}
@@
 
-\input{titlepage}
+%%%%%%%%%%%%%%%%%%%%%%%%
+\title{\bf Landslide: Systematic Exploration for Kernel-Space Race Detection}
+\author{Ben Blum \and Garth Gibson}
+\institute{Computer Science Department, Carnegie Mellon University, Pittsburgh}
+\date{May 2012}
 
+%\support{This research was sponsored by the U.S. Army Research Office under grant number W911NF0910273.}
+
+\maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%%
 \newcommand\true{\;\textit{true}}
@@
 %\newcommand\revision[1]{\begin{samepage}\cbstart#1\cbend\end{samepage}}
-\newcommand\revision[1]{\cbstart#1\cbend}
-%\newcommand\revision[1]{#1}
+%\newcommand\revision[1]{\cbstart#1\cbend}
+\newcommand\revision[1]{#1}
 
 \begin{abstract}
-Systematic exploration is an approach to finding race conditions by deterministically executing every possible interleaving of thread transitions and identifying which ones expose bugs.
-Current systematic exploration techniques are suitable for testing user-space programs, but are inadequate for testing kernels, where the testing framework's control over concurrency is more complicated.
+Systematic exploration is an approach to finding race conditions by deterministically executing many thread interleavings and identifying which ones expose bugs.
+Current techniques are suitable for testing user-space programs, but are inadequate for testing kernels.
+Testing kernel-level code necessitates understanding the kernel's design in order to effectively control nondeterminism and achieve reasonable state-space reduction.
 
 We present Landslide, a systematic exploration tool for finding races in kernels.
-Landslide targets Pebbles, the kernel specification that students implement in the undergraduate Operating Systems course at Carnegie Mellon University (15-410).
-We discuss the techniques Landslide uses to address the general challenges of kernel-level concurrency, and we evaluate its effectiveness and usability as a debugging aid.
-We show that our techniques make systematic testing in kernel-space feasible and that Landslide is a useful tool for doing so in the context of 15-410.
+Landslide relies on user-provided configuration to enable efficient exploration and testing of meaningful interleavings.
+The user instruments their kernel to inform Landslide of important concurrency events, and configures Landslide to focus its search on relevant kernel components.
+This combines the user's design knowledge with Landslide's ability to explore large state spaces, and encourages the user to steer Landslide towards focused searches that are more likely to find bugs.
+We claim that this steering is an effective approach for kernel-space race detection.
+
+%Landslide targets Pebbles, the kernel specification that students implement in the undergraduate Operating Systems course at Carnegie Mellon University (15-410).
+%We discuss the techniques Landslide uses to address the general challenges of kernel-level concurrency, and we evaluate its effectiveness and usability as a debugging aid.
+%We show that our techniques make systematic testing in kernel-space feasible and that Landslide is a useful tool for doing so in the context of 15-410.
+%%%%
+
+\noindent {\bf Keywords:} concurrency, kernel debugging, race conditions, runtime verification
 \end{abstract}
 
-\tableofcontents
-\input{acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\chapter{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% put this somewhere
+%We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user steers Landslide towards focused search spaces that are more likely to find bugs.
 
@@
 
+The claim of this dissertation is as follows.
 \begin{quote}
@@
 
-\input{related} % COMPLETE!
-\input{challenges} % COMPLETE!
-\input{terminology} % almost complete
-\input{design}
-\input{using}
-\input{evaluation}
-\input{futurework}
-\input{conclusion}
-
-% TODO: read presubmit.txt and check everything off
+%\input{related} % COMPLETE!
+%\input{challenges} % COMPLETE!
+%\input{terminology} % almost complete
+%\input{design}
+%\input{using}
+%\input{evaluation}
+%\input{futurework}
+%\input{conclusion}
 
 \bibliography{citations}{}
-\bibliographystyle{alpha}
-
-\appendix
-\input{testcases}
+\bibliographystyle{splncs03}
 

### 2012-06-02 rv/paper.tex
outline paper

@@
 %\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
+\usepackage{amsmath,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
 \usepackage{upgreek}
@@
 \date{May 2012}
-
-%\support{This research was sponsored by the U.S. Army Research Office under grant number W911NF0910273.}
 
@@
 The user instruments their kernel to inform Landslide of important concurrency events, and configures Landslide to focus its search on relevant kernel components.
-This combines the user's design knowledge with Landslide's ability to explore large state spaces, and encourages the user to steer Landslide towards focused searches that are more likely to find bugs.
-We claim that this steering is an effective approach for kernel-space race detection.
+This combines the user's design knowledge with Landslide's ability to explore large state spaces.
+Our experience with Landslide shows that a tool built for this usage pattern is capable of identifying otherwise-overlooked kernel-space races.
+
+%, and encourages the user to steer Landslide towards focused searches that are more likely to find bugs.
+%We claim that this steering is an effective approach for kernel-space race detection.
 
@@
 \end{abstract}
+
+%%%% Outline %%%%
+
+% challenges of kernel space
+% = causes of concurrency
+% = ad-hoc
+% - design independence (merge into model/scope)
+%
+% implementation
+% + model (maybe separate into a scope section, add talk of 15410)
+% - components (keep only challenge-addressing stuff)
+% + bug identification
+% = por (-soundness)
+%
+% user interface
+% - requirements (timer-driven should be in model)
+% = instrumenting (cut very short)
+% + configuring
+% = test cases (merge into model maybe -- probably not though?)
+%
+% future work
+% - education
+% = new techniques
+% + linux
+% = virtualisation (merge into linux?)
+% ? shaping
+% - oddities
 

### 2012-06-05 rv/paper.tex
Write introduction

@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-\title{\bf Landslide: Systematic Exploration for Kernel-Space Race Detection}
+\title{\bf Landslide: \\ Systematic Exploration for Kernel-Space Race Detection}
 \author{Ben Blum \and Garth Gibson}
@@
 
+%%%%% better outline %%%%
+% intro
+% related
+% challenges of kernel-space (short section, 2-3 paragraphs)
+% - causes of concurrency
+% - ad-hoc memory sharing
+% scope and model
+% - summary of 410 & kernel requirements
+% - simulated execution
+% - timer-driven scheduling/concurrency
+% - orientation for false-negative bugs & coarse-granularity search
+% implementation
+% - components (very briefly; only keep challenge-solving stuff)
+% - bug identification techniques
+% - making DPOR work
+% user interface
+% - instrumenting process
+% - iterative configuring + focusing of search
+% - types of test cases to use
+% future work
+% - new techniques to incorporate
+% - virtualisation and linux
+% - maybe shaping? (not sure how i could say it quickly.)
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% put this somewhere
-%We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user steers Landslide towards focused search spaces that are more likely to find bugs.
-
 Race conditions are notoriously difficult to debug.
@@
 
-\revision{
-Many techniques exist for dynamic testing of concurrent systems for race conditions.
-Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses as to what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
-Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
+Systematic exploration~\cite{verisoft} is a dynamic testing technique that aims to find race conditions or verify their absence.
+It involves enumerating many different interleavings of multiple threads and forcing a system to execute all such interleavings to check if any of them expose bugs.
+In comparison with conventional stress testing, systematic exploration is less likely to overlook bugs due to its deterministic testing approach.
+While other dynamic analysis techniques exist for finding race conditions, such as data race detection, % TODO cite here
+systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
+
+%\revision{
+%Many techniques exist for dynamic testing of concurrent systems for race conditions.
+%Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses as to what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
+%Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information.
 Compared to other dynamic analyses, such as data race detection~\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
 
-In kernel space, race condition debugging becomes even more difficult.
-Many aspects of the concurrency implementation itself are part of the system being tested.
-While user-space testing frameworks rely on the underlying kernel to provide common concurrency abstractions which the framework can use to drive the test, systematic exploration in the kernel itself necessarily exists underneath such abstractions.
-Furthermore, a tool that supports testing multiple kernels must support many different possible scheduler designs.
+In kernel space, race condition debugging becomes even more difficult. Many aspects of the concurrency implementation itself are part of the system being tested.
+While user-space testing frameworks operate above such abstractions, a systematic exporation tool for kernels cannot rely on them.
+Furthermore, execution in the kernel's scheduler and other similar components is common to all concurrent threads, which threatens our ability to achieve state-space reduction.
 
-The claim of this dissertation is as follows.
-\begin{quote}
-\bf
-Systematic exploration in kernel space is possible by understanding a kernel's concurrency through instrumentation of its internal abstractions, and can be made efficient by relying on the user to control the scope of the test.
-\end{quote}
-}
+%While user-space testing frameworks rely on the underlying kernel to provide common concurrency abstractions which the framework can use to drive the test, systematic exploration in the kernel itself necessarily exists underneath such abstractions.
+%Furthermore, a tool that supports testing multiple kernels must support many different possible scheduler designs.
 
-We present Landslide,\footnote{
-Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
+% The claim of this dissertation is as follows.
+% \begin{quote}
+% \bf
+% Systematic exploration in kernel space is possible by understanding a kernel's concurrency through instrumentation of its internal abstractions, and can be made efficient by relying on the user to control the scope of the test.
+% \end{quote}
+% }
+
+We present Landslide,
+%\footnote{Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
 a tool for applying systematic race detection techniques to kernel-level code.
-It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
-\revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.
+Landslide is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, and targets Pebbles, a fully preemptible UNIX-like kernel specification that students implement in Carnegie Mellon University's undergraduate operating systems class (15-410).
+To use Landslide, a programmer must instrument their kernel to inform Landslide of important concurrency events during the execution.
+In order to make Landslide's search efficient, the programmer should also configure Landslide to focus on the most relevant parts of the kernel.
+This leads to a usage dynamic that combines the user's design knowledge and Landslide's exploration mechanisms.
+Intuitively, the user ``steers'' Landslide towards focused search spaces that are more likely to quickly find bugs.
+
+%It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
+%\revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.
 %During a kernel's execution, Landslide records important actions performed by the kernel, decides at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
-Using this instrumentation and configuration, Landslide is able to enumerate all possible interleavings of kernel threads at a granularity both fine-grained enough to expose interesting concurrency behaviour and coarse-grained enough to produce state spaces that are feasible to explore.
+%Using this instrumentation and configuration, Landslide is able to enumerate all possible interleavings of kernel threads at a granularity both fine-grained enough to expose interesting concurrency behaviour and coarse-grained enough to produce state spaces that are feasible to explore.
 Landslide then forces the kernel to exercise each of these interleavings, and checks for race conditions in each of them.}
-When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.
+%When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.
 
-\revision{
-In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), define common terms used in the rest of this document (Chapter~\ref{sec:terminology}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
-We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).
-Finally, we discuss possible avenues for future work in education, use of additional testing techniques, application to general purpose kernels, long-running testing approaches, and theoretical study of systematic exploration (Chapter~\ref{sec:future}).
-}
+We found that, with modest instrumentation effort, Landslide was able to find race conditions in several different Pebbles kernels that had previously been overlooked by stress testing and/or expert manual inspection.
+Many of these bugs required complicated thread interleavings to expose, and could not have been found with simpler testing techniques such as data race detection or static analysis.
 
-%\input{related} % COMPLETE!
-%\input{challenges} % COMPLETE!
-%\input{terminology} % almost complete
-%\input{design}
-%\input{using}
-%\input{evaluation}
-%\input{futurework}
-%\input{conclusion}
+%\revision{
+%In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), define common terms used in the rest of this document (Chapter~\ref{sec:terminology}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
+%We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).
+%Finally, we discuss possible avenues for future work in education, use of additional testing techniques, application to general purpose kernels, long-running testing approaches, and theoretical study of systematic exploration (Chapter~\ref{sec:future}).
+%}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Challenges of Kernel Space}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Scope}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Environment}
+% 410, requirements, simulated
+\subsection{Testing Approach}
+% timer driven, false negative bugs, coarse granularity search
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Design}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%% solving challenges %%%%
+% timer-driven scheduling
+% tracking which threads
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{User Interface}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% decision trace
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Evaluation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%We evaluated Landslide in two ways: first, by instrumenting two kernels the authors were familiar with and experimenting with many different configurations for Landslide's search, and second, by meeting with students
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-06-07 rv/paper.tex
write challenges

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Challenges of Kernel Space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Scope}
@@
 \subsection{Testing Approach}
+\label{sec:approach}
 % timer driven, false negative bugs, coarse granularity search
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Challenges of Kernel Space}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% FIXME: maybe say in each paragraph how we address the issue?
+
+Compared to systematic exploration in user space, Landslide faces several challenges when applying the same techniques in kernel space.
+
+{\bf Causes of concurrency.} In user space, a testing framework may rely on system calls to the underlying kernel to deterministically control the concurrent execution of the system under test.
+However, since Landslide tests the kernel itself, it must control the underlying causes of nondeterminism, such as interrupts from the timer, disk, or network.
+(In this work we focus only on timer events, as discussed in \S\ref{sec:approach}.)
+Landslide contains its own internal scheduler, which manages timer interrupts to provide a mechanism for causing an arbitrarily-chosen thread to preempt execution of the current one (see \S\ref{sec:design}).
+
+{\bf Understanding the kernel's design.} In order to track the state of the kernel's scheduler, such as which threads exist or are runnable at any given time, a testing framework must know when certain important concurrency events happen during the kernel's execution.
+Landslide relies on in-kernel annotations to provide this information (see \S\ref{sec:instrument}).
+
+{\bf Shared memory conflicts.} Since the kernel contains its own concurrency implementation, every thread switch will execute through a common path that modifies shared scheduler data structures.
+Because Dynamic Partial Order Reduction relies on a memory independence relation between thread transitions, this abundance of shared memory conflicts threatens Landslide's ability to achieve good state space reduction.
+We provide an interface with which the user can configure Landslide to ignore such conflicts to more efficiently test unrelated components of the kernel (see \S\ref{sec:config}).
+
+% Talk about frame_lock in user interface as a "side benefit" of addressing this.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:design}
 %%%% solving challenges %%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Instrumenting Kernels with Landslide}
+\label{sec:instrument}
+
+\subsection{Configuring Landslide}
+\label{sec:config}
 % decision trace

### 2012-06-07 rv/paper.tex
write scope

@@
 \subsection{Environment}
+\label{sec:environment}
 % 410, requirements, simulated
+
+% FIXME awkward paragraph
+Landslide's ability to test different types of kernels in different execution environments is currently very limited.
+We built Landslide to test Pebbles kernels, which students implement from the ground up in a six-week project in 15-410 at CMU.
+The Pebbles specification is a set of UNIX-like system calls which includes routines for thread lifecycle management and virtual memory management.
+
+Landslide is implemented as a module for Simics, a full-system x86 simulator which students in 15-410 use as an execution environment for their kernels.
+Simics can load Landslide while booting a kernel, and Landslide will perform systematic exploration on that kernel using Simics interface calls during the simulated execution.
+
+Currently, Landslide assumes that timer interrupts are the only nondeterministic events that control a kernel's concurrent execution.
+While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
+
 \subsection{Testing Approach}
 \label{sec:approach}
-% timer driven, false negative bugs, coarse granularity search
+% false negative bugs, coarse granularity search
+
+Landslide's testing approach is false-negative-oriented both in terms of identifying bugs and in terms of constructing the state space to search. 
+We identify bugs only when something definitely wrong happens, rather than flagging suspicious behaviours which might indicate underlying bugs, and we rely on the user to define a granularity for thread interleavings which will produce meaningful results.
 
@@
 However, since Landslide tests the kernel itself, it must control the underlying causes of nondeterminism, such as interrupts from the timer, disk, or network.
-(In this work we focus only on timer events, as discussed in \S\ref{sec:approach}.)
+(In this work we focus only on timer events, as discussed in \S\ref{sec:environment}.)
 Landslide contains its own internal scheduler, which manages timer interrupts to provide a mechanism for causing an arbitrarily-chosen thread to preempt execution of the current one (see \S\ref{sec:design}).
@@
 
+% bugs
+% With false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it. However, when Landslide does identify a bug, the user can be sure that a race exists.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % decision trace
+
+% Likewise, with user-defined state spaces to search, Landslide might not explore thread interleavings at a granularity fine enough to cause a bug to arise.
 

### 2012-06-08 rv/paper.tex
add related and eval and citations

@@
 
-We present Landslide,
+We present Landslide \cite{landslide},
 %\footnote{Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Among related dynamic race condition detection tools, some are systematic exploration tools, and others are kernel race-finding tools.
+To our knowledge, Landslide is the first to combine the former with the latter.
+
+% remove dbug-retreat citation if need be
+{\bf Systematic exploration.} Many techniques have been developed for efficient systematic exploration of concurrent systems.
+Dynamic Partial Order Reduction (DPOR) \cite{dpor,dbug-retreat} is an algorithm for achieving state space reduction by identifying ``independent'' thread transitions.
+dBug \cite{dbug-ssv} is a tool that makes use of DPOR in user-space; Landslide applies it to test kernels.
+
+% Among other systematic testing tools are CHESS \cite{chess}, which heuristically orders its search by number of preemptions; MaceMC \cite{macemc} and MODIST \cite{modist}, which provide systematic testing frameworks for networked and distributed applications; RacePRO \cite{racepro}, which uncovers races between system calls made by multiple processes; and DeMeter \cite{demeter}, which focuses its search on only one component of a system at a time.
+
+{\bf Kernel-level verification.}
+% Kernel-level code has seen many dynamic testing techniques apart from systematic exploration.
+DataCollider \cite{datacollider} uses memory access sampling to find data races in Windows 7.
+Compared to systematic exploration, data race detection is less expensive, but is limited in which types of races it can find.
+% Carburizer \cite{carburizer} is a tool for enhancing resiliency of imperfect kernel drivers in the presence of faulty peripheral devices.
+SimTester \cite{simtester} is another driver testing tool; like Landslide, it runs in Simics, and injects interrupts to control the system's concurrency.
+SimTester focuses on races involving device interrupts, whereas Landslide focuses on non-driver inter-thread races.
+Additionally, SimTester only injects one interrupt per test run, which is both less expensive and less powerful than systematic exploration.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%We evaluated Landslide in two ways: first, by instrumenting two kernels the authors were familiar with and experimenting with many different configurations for Landslide's search, and second, by meeting with students
+
+We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find many different races, and second, by meeting with students of 15-410 to see if they could find bugs on their own with Landslide.
+
+We instrumented one kernel the first author wrote when he was a student and also one kernel he had manually graded as a teaching assistant.
+We configured Landslide to search for five complicated already-known race conditions.
+In addition to finding all five races, Landslide also found a sixth previously unknown race in the author's own kernel.
+Using decision points only on calls to \texttt{mutex\_lock}, Landslide found each of the six bugs in between 11 and 57 seconds, executing between 1 and 377 distinct interleavings for each.
+% FIXME: also say 'on voluntary rescheds'?
+
+In the user study, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
+Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two race conditions and two deterministic errors.
+These bugs manifested as infinite loops, a kernel panic, and a use-after-free.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work}
+\section{Future Work and Conclusion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-06-09 rv/paper.tex
begin write design and iface; begin remove challenges

@@
 
-% remove dbug-retreat citation if need be
 {\bf Systematic exploration.} Many techniques have been developed for efficient systematic exploration of concurrent systems.
-Dynamic Partial Order Reduction (DPOR) \cite{dpor,dbug-retreat} is an algorithm for achieving state space reduction by identifying ``independent'' thread transitions.
+Dynamic Partial Order Reduction (DPOR) \cite{dpor} is an algorithm for achieving state space reduction by identifying ``independent'' thread transitions.
 dBug \cite{dbug-ssv} is a tool that makes use of DPOR in user-space; Landslide applies it to test kernels.
 
-% Among other systematic testing tools are CHESS \cite{chess}, which heuristically orders its search by number of preemptions; MaceMC \cite{macemc} and MODIST \cite{modist}, which provide systematic testing frameworks for networked and distributed applications; RacePRO \cite{racepro}, which uncovers races between system calls made by multiple processes; and DeMeter \cite{demeter}, which focuses its search on only one component of a system at a time.
+Among other systematic testing tools are CHESS \cite{chess}, with a heuristic for search ordering; MaceMC \cite{macemc} and MODIST \cite{modist}, which test networked and distributed applications; RacePRO \cite{racepro}, which uncovers inter-process races; and DeMeter \cite{demeter}, which focuses its search on only one component of a system at a time.
 
@@
 % FIXME: maybe say in each paragraph how we address the issue?
+% TODO: get rid of this section
 
@@
 
-{\bf Causes of concurrency.} In user space, a testing framework may rely on system calls to the underlying kernel to deterministically control the concurrent execution of the system under test.
-However, since Landslide tests the kernel itself, it must control the underlying causes of nondeterminism, such as interrupts from the timer, disk, or network.
+% {\bf Causes of concurrency.} In user space, a testing framework may rely on system calls to the underlying kernel to deterministically control the concurrent execution of the system under test.
+% However, since Landslide tests the kernel itself, it must control the underlying causes of nondeterminism, such as interrupts from the timer, disk, or network.
 (In this work we focus only on timer events, as discussed in \S\ref{sec:environment}.)
-Landslide contains its own internal scheduler, which manages timer interrupts to provide a mechanism for causing an arbitrarily-chosen thread to preempt execution of the current one (see \S\ref{sec:design}).
+% Landslide contains its own internal scheduler, which manages timer interrupts to provide a mechanism for causing an arbitrarily-chosen thread to preempt execution of the current one (see \S\ref{sec:design}).
 
@@
 Landslide relies on in-kernel annotations to provide this information (see \S\ref{sec:instrument}).
-
-{\bf Shared memory conflicts.} Since the kernel contains its own concurrency implementation, every thread switch will execute through a common path that modifies shared scheduler data structures.
-Because Dynamic Partial Order Reduction relies on a memory independence relation between thread transitions, this abundance of shared memory conflicts threatens Landslide's ability to achieve good state space reduction.
-We provide an interface with which the user can configure Landslide to ignore such conflicts to more efficiently test unrelated components of the kernel (see \S\ref{sec:config}).
 
@@
 \label{sec:design}
-%%%% solving challenges %%%%
-% timer-driven scheduling
-% tracking which threads
+% prelude: define decision point and maybe voluntary reschedule
 
-% bugs
+The design of Landslide's exploration is based around {\em decision points}.
+These are places during the execution at which we force preemptions and cause different threads to interleave execution around.
+Searching with few decision points results in coarser-grained interleavings, smaller state spaces, and less likelihood of finding bugs; searching with more results in the opposite.
+
+Landslide automatically identifies a minimal set of decision points at voluntary reschedules between threads (e.g.
+\texttt{yield}).
+We also provide an interface for configuring additional decision points (see \S\ref{sec:config}).
+
+\subsection{Exploration Mechanism}
+
+Landslide's exploration mechanism has two parts.
+First, at each decision point, it is able to force an arbitary thread to run.
+Second, upon termination of the test case, it can revert the system to an earlier state to explore additional interleavings.
+
+{\bf Thread scheduling.} Compared with user space testing frameworks, Landslide cannot rely on kernel-provided system calls to control the concurrent execution of the system under test.
+Instead, Landslide has an internal abstraction for scheduling arbitrary threads built upon injecting timer interrupts.
+
+If thread $X$ is running and Landslide decides to preempt with thread $Z$, it fires a timer interrupt and waits for the kernel to finish the triggered context switch.
+Then if the kernel is running thread $Y$, Landslide repeats until thread $Z$ begins running, at which point it lets the kernel continue execution.
+
+{\bf Backtracking.} Landslide makes use of Simics ``bookmarks'', a feature
+which enables checkpointing and restoring the kernel's state, to implement
+backtracking when the end of the test case is reached.
+
+\subsection{Identifying Bugs}
+% TODO
 % With false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it. However, when Landslide does identify a bug, the user can be sure that a race exists.
+
+% decision trace
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{sec:interface}
+
+% Likewise, with user-defined state spaces to search, Landslide might not explore thread interleavings at a granularity fine enough to cause a bug to arise. 
+% TODO
 
@@
 
+% TODO
+
+Finally, there are two two-line functions within Landslide itself that the user must implement.
+These are predicates on the kernel's scheduler state, and express potentially nontrivial conditions: whether the current thread is runnable but not on the runqueue, and whether preemption is disabled while interrupts are on.
+
 \subsection{Configuring Landslide}
 \label{sec:config}
-% decision trace
 
-% Likewise, with user-defined state spaces to search, Landslide might not explore thread interleavings at a granularity fine enough to cause a bug to arise.
+{\bf Configuring decision points.} % TODO
+
+{\bf Configuring Dynamic Partial Order Reduction.} Performing DPOR requires establishing a memory independence relation between transitions.
+Unfortunately, since the kernel contains its own concurrency implementation, every thread switch will execute through a common path that modifies shared scheduler data structures.
+Unchecked, this would cause every transition to conflict with each other and result in no reduction.
+
+We provide an interface with which the user can configure Landslide to ignore such conflicts to more efficiently test unrelated components of the kernel.
 
@@
 In addition to finding all five races, Landslide also found a sixth previously unknown race in the author's own kernel.
-Using decision points only on calls to \texttt{mutex\_lock}, Landslide found each of the six bugs in between 11 and 57 seconds, executing between 1 and 377 distinct interleavings for each.
-% FIXME: also say 'on voluntary rescheds'?
+Using decision points only on calls to \texttt{mutex\_lock} and on voluntary reschedules, Landslide found each of the six bugs in between 11 and 57 seconds, executing between 1 and 377 distinct interleavings for each.
 
@@
 
+% prob. don't actually say future work here; say it inline
+% TODO
+
 \bibliography{citations}{}

### 2012-06-10 rv/paper.tex
write design and interface

@@
 Dynamic Partial Order Reduction (DPOR) \cite{dpor} is an algorithm for achieving state space reduction by identifying ``independent'' thread transitions.
-dBug \cite{dbug-ssv} is a tool that makes use of DPOR in user-space; Landslide applies it to test kernels.
+dBug \cite{dbug-ssv} is a user-space tool that uses DPOR in user-space; Landslide applies it to test kernels.
 
@@
 SimTester focuses on races involving device interrupts, whereas Landslide focuses on non-driver inter-thread races.
-Additionally, SimTester only injects one interrupt per test run, which is both less expensive and less powerful than systematic exploration.
+Additionally, SimTester injects only one interrupt per test run, which is both less expensive and less powerful than systematic exploration.
 
@@
 Landslide is implemented as a module for Simics, a full-system x86 simulator which students in 15-410 use as an execution environment for their kernels.
-Simics can load Landslide while booting a kernel, and Landslide will perform systematic exploration on that kernel using Simics interface calls during the simulated execution.
+Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and memory access of the simulated execution.
+Landslide will then perform systematic exploration on that kernel using Simics interface calls during the simulated execution.
 
@@
 While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
+Landslide's model is also restricted to uniprocessor execution. Supporting multiprocessor environments, device driver testing, and production kernels such as Linux is left to future work.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Challenges of Kernel Space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Challenges of Kernel Space}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % FIXME: maybe say in each paragraph how we address the issue?
@@
 
-Compared to systematic exploration in user space, Landslide faces several challenges when applying the same techniques in kernel space.
+% Compared to systematic exploration in user space, Landslide faces several challenges when applying the same techniques in kernel space.
 
@@
 
-{\bf Understanding the kernel's design.} In order to track the state of the kernel's scheduler, such as which threads exist or are runnable at any given time, a testing framework must know when certain important concurrency events happen during the kernel's execution.
-Landslide relies on in-kernel annotations to provide this information (see \S\ref{sec:instrument}).
-
-% Talk about frame_lock in user interface as a "side benefit" of addressing this.
+% {\bf Understanding the kernel's design.} In order to track the state of the kernel's scheduler, such as which threads exist or are runnable at any given time, a testing framework must know when certain important concurrency events happen during the kernel's execution.
+% Landslide relies on in-kernel annotations to provide this information (see \S\ref{sec:instrument}).
 
@@
 The design of Landslide's exploration is based around {\em decision points}.
-These are places during the execution at which we force preemptions and cause different threads to interleave execution around.
+These are places during the execution at which we force preemptions and around which we cause different threads to interleave.
 Searching with few decision points results in coarser-grained interleavings, smaller state spaces, and less likelihood of finding bugs; searching with more results in the opposite.
@@
 Landslide's exploration mechanism has two parts.
-First, at each decision point, it is able to force an arbitary thread to run.
-Second, upon termination of the test case, it can revert the system to an earlier state to explore additional interleavings.
+First, at each decision point, it can force an arbitary thread to run in place of the current one.
+Second, upon termination of the test case, it reverts the system to an earlier state to explore additional interleavings.
 
@@
 Instead, Landslide has an internal abstraction for scheduling arbitrary threads built upon injecting timer interrupts.
-
 If thread $X$ is running and Landslide decides to preempt with thread $Z$, it fires a timer interrupt and waits for the kernel to finish the triggered context switch.
@@
 \subsection{Identifying Bugs}
-% TODO
-% With false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it. However, when Landslide does identify a bug, the user can be sure that a race exists.
+
+%\begin{wrapfigure}{r}{0.2\textwidth}
+%	\center
+%	\includegraphics[width=0.2\textwidth]{livelock.pdf}
+%	\caption{Infinite loop detection.
+Each circle represents a decision point; each branch represents one interleaving.
+Disproportionately many decision points have been encountered in the highlighted interleaving, which indicates the kernel is likely stuck.}
+%\end{wrapfigure}
+
+Landslide has several checks that detect when a bug arises. It can identify kernel panics, use-after-free accesses (and other memory errors such as double-free and memory leaks), and deadlocks.
+
+Landslide can optionally also check heuristically for infinite loops and livelock.
+This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
+We have never found false positives with this technique, although false negatives have arisen when too few previous interleavings have been tested to make a reliable comparison.
+
+When Landslide identifies a bug, it aborts execution and generates a {\em decision trace}.
+The trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, its stack trace when it was switched away from, and the thread that we caused to preempt it.
+With this information, the user can better understand the concurrent execution that led to the race.
+
+In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.
+However, when Landslide does identify a bug, the user can be sure that a race exists.
 
@@
 
-% Likewise, with user-defined state spaces to search, Landslide might not explore thread interleavings at a granularity fine enough to cause a bug to arise. 
-% TODO
-
 \subsection{Instrumenting Kernels with Landslide}
@@
 
-% TODO
-
+Users must annotate their kernels to inform Landslide of certain important concurrency events during execution.
+We provide a set of annotation functions, called \texttt{tell\_landslide}, for this purpose.
+The annotations denote when a thread is doing \texttt{fork}, \texttt{sleep}, or \texttt{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
+There is also a configuration file, \texttt{config.landslide}, in which the user must specify constant information such as the names of the timer handler and context switcher, which threads exist when the kernel boots up, and which userspace test program to run.
 Finally, there are two two-line functions within Landslide itself that the user must implement.
@@
 
-{\bf Configuring decision points.} % TODO
+{\bf Configuring decision points.} While Landslide automatically identifies decision points on voluntary reschedules, these will result in coarse-grained interleavings likely to overlook bugs.
+After successfully instrumenting their kernel and exploring a minimal state space, the user should then add more decision points for a finer-grained search.
+We provide an extra annotation, \texttt{tell\_landslide\_decide()}, for this purpose.
+We recommend using it in concurrency primitives, such as at the start of \texttt{mutex\_lock()} and at the end of \texttt{mutex\_unlock()}.
+
+This strategy may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated system calls.
+We provide an interface for the user to limit decision points to within relevant functions: the user may write ``\texttt{within\_function foo}'' to whitelist function \texttt{foo}, and ``\texttt{without\_function bar}'' to blacklist function \texttt{bar}.
+For example, if testing thread death and reaping, the user should write \texttt{within\_function exit} and \texttt{within\_function wait}.
 
@@
 Unchecked, this would cause every transition to conflict with each other and result in no reduction.
+Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested.
+If testing for races in thread livecycle routines, the user likely does not care about accesses to the frame allocator mutex.
 
-We provide an interface with which the user can configure Landslide to ignore such conflicts to more efficiently test unrelated components of the kernel.
+We provide an interface with which the user can configure Landslide to ignore such unrelated conflicts to more efficiently test components of the kernel they care about.
+In \texttt{config.landslide}, the user writes ``\texttt{ignore\_sym}'' followed by the name of the global variable to ignore and its type size, and Landslide may then identify transitions as independent even if they conflict on that memory.
 

### 2012-06-10 rv/paper.tex
garth's review; maurer's and rpearl's review; joshua's intro review

@@
 \author{Ben Blum \and Garth Gibson}
-\institute{Computer Science Department, Carnegie Mellon University, Pittsburgh}
+\institute{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA}
 \date{May 2012}
@@
 Systematic exploration is an approach to finding race conditions by deterministically executing many thread interleavings and identifying which ones expose bugs.
-Current techniques are suitable for testing user-space programs, but are inadequate for testing kernels.
+Current techniques are suitable for testing user-space programs, but are inadequate for testing operating system kernels.
 Testing kernel-level code necessitates understanding the kernel's design in order to effectively control nondeterminism and achieve reasonable state-space reduction.
-
+% paragraph split would go here, but abstrax
 We present Landslide, a systematic exploration tool for finding races in kernels.
-Landslide relies on user-provided configuration to enable efficient exploration and testing of meaningful interleavings.
-The user instruments their kernel to inform Landslide of important concurrency events, and configures Landslide to focus its search on relevant kernel components.
+Landslide makes use of user-provided configuration to enable efficient exploration and testing of meaningful interleavings.
+The user instruments the kernel to inform Landslide of important concurrency events, and configures Landslide's search to de-emphasize irrelevant kernel components.
 This combines the user's design knowledge with Landslide's ability to explore large state spaces.
@@
 
-Systematic exploration~\cite{verisoft} is a dynamic testing technique that aims to find race conditions or verify their absence.
-It involves enumerating many different interleavings of multiple threads and forcing a system to execute all such interleavings to check if any of them expose bugs.
-In comparison with conventional stress testing, systematic exploration is less likely to overlook bugs due to its deterministic testing approach.
+Systematic exploration~\cite{verisoft} is a dynamic testing technique that aims to find race conditions.
+It involves enumerating many different interleavings of multiple threads and forcing a system to execute all such interleavings to check if any expose bugs.
+% In comparison with conventional stress testing, systematic exploration is less likely to overlook bugs, due to its deterministic testing approach.
 While other dynamic analysis techniques exist for finding race conditions, such as data race detection, % TODO cite here
@@
 In kernel space, race condition debugging becomes even more difficult. Many aspects of the concurrency implementation itself are part of the system being tested.
-While user-space testing frameworks operate above such abstractions, a systematic exporation tool for kernels cannot rely on them.
-Furthermore, execution in the kernel's scheduler and other similar components is common to all concurrent threads, which threatens our ability to achieve state-space reduction.
+While user-space testing frameworks operate above clear abstractions and interfaces, a systematic exporation tool for kernels cannot rely on such abstractions and interfaces.
+Furthermore, execution in the scheduler and in similar components is common to all concurrent threads, even otherwise independent ones; this threatens our ability to achieve state-space reduction.
 
@@
 a tool for applying systematic race detection techniques to kernel-level code.
-Landslide is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, and targets Pebbles, a fully preemptible UNIX-like kernel specification that students implement in Carnegie Mellon University's undergraduate operating systems class (15-410).
-To use Landslide, a programmer must instrument their kernel to inform Landslide of important concurrency events during the execution.
-In order to make Landslide's search efficient, the programmer should also configure Landslide to focus on the most relevant parts of the kernel.
-This leads to a usage dynamic that combines the user's design knowledge and Landslide's exploration mechanisms.
-Intuitively, the user ``steers'' Landslide towards focused search spaces that are more likely to quickly find bugs.
+Landslide is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, and targets Pebbles \cite{kspec}, a fully preemptible UNIX-like kernel specification that students implement in Carnegie Mellon University's senior level operating systems class (15-410).
+% To use Landslide efficiently, a programmer must instrument their kernel to inform Landslide of important concurrency events.
+In order to make Landslide's search efficient, the programmer should also configure Landslide to focus on the most relevant parts of the test at hand.
+Landslide's interface allows the user to configure its search to focus on the most relevant parts of the test at hand, which leads to a usage dynamic that combines the user's design knowledge and Landslide's exploration mechanisms.
+Intuitively, the user ``steers'' Landslide towards search spaces that are more likely to quickly find suspected bugs.
 
@@
 
-We found that, with modest instrumentation effort, Landslide was able to find race conditions in several different Pebbles kernels that had previously been overlooked by stress testing and/or expert manual inspection.
+We found that, with modest instrumentation effort, Landslide was able to find race conditions in several different Pebbles kernels.
+These races had previously been overlooked by stress testing and/or expert manual inspection.
 Many of these bugs required complicated thread interleavings to expose, and could not have been found with simpler testing techniques such as data race detection or static analysis.
+% TODO: decide whether to prove it?
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Environment}
-\label{sec:environment}
+% \subsection{Environment}
 % 410, requirements, simulated
 
-% FIXME awkward paragraph
-Landslide's ability to test different types of kernels in different execution environments is currently very limited.
-We built Landslide to test Pebbles kernels, which students implement from the ground up in a six-week project in 15-410 at CMU.
-The Pebbles specification is a set of UNIX-like system calls which includes routines for thread lifecycle management and virtual memory management.
+% Landslide's ability to test different types of kernels in different execution environments is currently very limited.
+% We built Landslide to test Pebbles kernels, which students implement from the ground up in a six-week project in 15-410 at CMU.
 
-Landslide is implemented as a module for Simics, a full-system x86 simulator which students in 15-410 use as an execution environment for their kernels.
-Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and memory access of the simulated execution.
+In this first attempt at kernel-space systematic testing, we aimed to use a less complex kernel than production kernels such as Linux and Windows.
+We built Landslide to test Pebbles kernels, which students at CMU develop from scratch in 6 weeks at CMU.
+These kernels are necessarily less complex, yet still rife with race conditions that are debugged furiously every semester.
+The Pebbles specification \cite{kspec} is a set of UNIX-like system calls which includes routines for thread lifecycle management and virtual memory management.
+Among these are \texttt{fork}, \texttt{exec}, \texttt{exit}, \texttt{wait}, and \texttt{yield}.
+
+{\bf Testing environment.} Landslide is implemented as a module for Simics, a full-system x86 simulator which students in 15-410 use as an execution environment for their kernels.
+Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and once per memory access of the simulated execution.
 Landslide will then perform systematic exploration on that kernel using Simics interface calls during the simulated execution.
@@
 While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
-Landslide's model is also restricted to uniprocessor execution. Supporting multiprocessor environments, device driver testing, and production kernels such as Linux is left to future work.
+Landslide's model is also restricted to uniprocessor execution. Supporting multiprocessor environments, device driver testing, and production kernels is left to future work.
 
-\subsection{Testing Approach}
-\label{sec:approach}
-% false negative bugs, coarse granularity search
+% \subsection{Testing Approach}
 
-Landslide's testing approach is false-negative-oriented both in terms of identifying bugs and in terms of constructing the state space to search. 
-We identify bugs only when something definitely wrong happens, rather than flagging suspicious behaviours which might indicate underlying bugs, and we rely on the user to define a granularity for thread interleavings which will produce meaningful results.
+{\bf Testing approach.} Landslide's testing approach is false-negative-oriented both in terms of identifying bugs and in terms of constructing the state space to search. 
+We identify bugs only when something definitely wrong happens, rather than flagging suspicious behaviours which might indicate underlying bugs, and we allow the user to manually define a granularity for thread interleavings which will produce the most meaningful results.
 
@@
 The design of Landslide's exploration is based around {\em decision points}.
-These are places during the execution at which we force preemptions and around which we cause different threads to interleave.
-Searching with few decision points results in coarser-grained interleavings, smaller state spaces, and less likelihood of finding bugs; searching with more results in the opposite.
+These are places in the kernel under test at which we force preemptions and around which we cause different threads to interleave.
+Searching with few decision points results in coarser-grained interleavings, shorter test execution, and less likelihood of finding bugs; searching with more results in the opposite.
 
@@
 
+In order to know which threads are runnable at each decision point, Landslide maintains a mirror image of the kernel's scheduler state.
+This consists of a runqueue (all immediately runnable threads), a sleep queue (all threads blocked on \texttt{sleep} -- we treat these as runnable, because a finite number of timer ticks can cause these to run), and a descheduled queue (all other existing threads).
+Landslide uses the user-provided scheduler annotations (see \S\ref{sec:instrument}) to keep this state up-to-date.
+
 {\bf Backtracking.} Landslide makes use of Simics ``bookmarks'', a feature
-which enables checkpointing and restoring the kernel's state, to implement
+which enables checkpointing and restoring the kernel's execution state, to implement
 backtracking when the end of the test case is reached.
@@
 In general, with false-negative bug detection, the kernel might execute a buggy behaviour yet Landslide would miss it.
-However, when Landslide does identify a bug, the user can be sure that a race exists.
+However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the user can be sure that a race exists.
 
@@
 
-Users must annotate their kernels to inform Landslide of certain important concurrency events during execution.
+Users annotate their kernels to inform Landslide of certain important concurrency events during execution.
 We provide a set of annotation functions, called \texttt{tell\_landslide}, for this purpose.
@@
 There is also a configuration file, \texttt{config.landslide}, in which the user must specify constant information such as the names of the timer handler and context switcher, which threads exist when the kernel boots up, and which userspace test program to run.
-Finally, there are two two-line functions within Landslide itself that the user must implement.
+Finally, there are two two-line functions used within Landslide itself that the user must implement.
 These are predicates on the kernel's scheduler state, and express potentially nontrivial conditions: whether the current thread is runnable but not on the runqueue, and whether preemption is disabled while interrupts are on.
@@
 
+After successfully instrumenting their kernel and exploring a minimal state space, the user should then configure Landslide's search to be more likely to find suspected bugs.
+For brevity, we omit discussion of some conceptually unimportant configuration options.
+
 {\bf Configuring decision points.} While Landslide automatically identifies decision points on voluntary reschedules, these will result in coarse-grained interleavings likely to overlook bugs.
-After successfully instrumenting their kernel and exploring a minimal state space, the user should then add more decision points for a finer-grained search.
-We provide an extra annotation, \texttt{tell\_landslide\_decide()}, for this purpose.
+We provide an extra annotation, \texttt{tell\_landslide\_decide()}, for the user to add more decision points for a finer-grained search.
 We recommend using it in concurrency primitives, such as at the start of \texttt{mutex\_lock()} and at the end of \texttt{mutex\_unlock()}.
@@
 This strategy may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated system calls.
-We provide an interface for the user to limit decision points to within relevant functions: the user may write ``\texttt{within\_function foo}'' to whitelist function \texttt{foo}, and ``\texttt{without\_function bar}'' to blacklist function \texttt{bar}.
-For example, if testing thread death and reaping, the user should write \texttt{within\_function exit} and \texttt{within\_function wait}.
+We provide interface options for the user to view currently identified decision points and to selectively limit them.
+First, the user writes ``\texttt{DECISION\_INFO\_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
+Then, the user writes ``\texttt{within\_function foo}'' to whitelist function \texttt{foo}, and ``\texttt{without\_function bar}'' to blacklist function \texttt{bar}.
+For example, if testing thread death and reaping, the user should write \texttt{within\_function exit}, and \texttt{within\_function wait}, and perhaps \texttt{without\_function destroy\_address\_space}.
 
-{\bf Configuring Dynamic Partial Order Reduction.} Performing DPOR requires establishing a memory independence relation between transitions.
+{\bf Configuring Dynamic Partial Order Reduction.} Performing DPOR \cite{dpor} requires establishing a memory independence relation between transitions.
 Unfortunately, since the kernel contains its own concurrency implementation, every thread switch will execute through a common path that modifies shared scheduler data structures.
-Unchecked, this would cause every transition to conflict with each other and result in no reduction.
+Unchecked, this would cause every transition to conflict with each other and result in no reduction in execution time from DPOR.
 Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested.
-If testing for races in thread livecycle routines, the user likely does not care about accesses to the frame allocator mutex.
+For example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
 
@@
 
-We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find many different races, and second, by meeting with students of 15-410 to see if they could find bugs on their own with Landslide.
+We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with this year's students before they submitted their kernel for grading to see if they could find bugs on their own with Landslide.
 
-We instrumented one kernel the first author wrote when he was a student and also one kernel he had manually graded as a teaching assistant.
+We instrumented one kernel written by an author in a previous year, and also one kernel that author manually graded as a teaching assistant.
 We configured Landslide to search for five complicated already-known race conditions.
 In addition to finding all five races, Landslide also found a sixth previously unknown race in the author's own kernel.
-Using decision points only on calls to \texttt{mutex\_lock} and on voluntary reschedules, Landslide found each of the six bugs in between 11 and 57 seconds, executing between 1 and 377 distinct interleavings for each.
+Using decision points only on calls to \texttt{mutex\_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.
 
-In the user study, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
+In the current student user study, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
 Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two race conditions and two deterministic errors.

### 2012-06-10 rv/paper.tex
finish joshua's copyedit

@@
 
-In this first attempt at kernel-space systematic testing, we aimed to use a less complex kernel than production kernels such as Linux and Windows.
-We built Landslide to test Pebbles kernels, which students at CMU develop from scratch in 6 weeks at CMU.
+To constrain the problem space some, we aimed to use a less complex kernel than production kernels such as Linux and Windows.
+We built Landslide to test Pebbles kernels, which students at CMU develop from scratch in a 6 week project.
 These kernels are necessarily less complex, yet still rife with race conditions that are debugged furiously every semester.
 The Pebbles specification \cite{kspec} is a set of UNIX-like system calls which includes routines for thread lifecycle management and virtual memory management.
-Among these are \texttt{fork}, \texttt{exec}, \texttt{exit}, \texttt{wait}, and \texttt{yield}.
+Among these are \texttt{fork} and \texttt{exec}, \texttt{exit} (known as \texttt{vanish}) and \texttt{wait}, and \texttt{yield}.
 
@@
 
-Currently, Landslide assumes that timer interrupts are the only nondeterministic events that control a kernel's concurrent execution.
+Landslide assumes that timer interrupts are the only nondeterministic events that control a kernel's concurrent execution.
 While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
@@
 
-The design of Landslide's exploration is based around {\em decision points}.
+The driving concept of Landslide's exploration is the idea of {\em decision points}.
 These are places in the kernel under test at which we force preemptions and around which we cause different threads to interleave.
@@
 \texttt{yield}).
-We also provide an interface for configuring additional decision points (see \S\ref{sec:config}).
+We provide an interface for configuring additional decision points (described in \S\ref{sec:config}).
 
@@
 Instead, Landslide has an internal abstraction for scheduling arbitrary threads built upon injecting timer interrupts.
-If thread $X$ is running and Landslide decides to preempt with thread $Z$, it fires a timer interrupt and waits for the kernel to finish the triggered context switch.
-Then if the kernel is running thread $Y$, Landslide repeats until thread $Z$ begins running, at which point it lets the kernel continue execution.
+If thread $M$ is running and Landslide decides to preempt with thread $N$, it fires a timer interrupt and waits for the kernel to finish the triggered context switch.
+Then if the kernel is running a thread that is not $N$, Landslide repeats until thread $N$ begins running, at which point it lets the kernel continue execution.
 
@@
 This consists of a runqueue (all immediately runnable threads), a sleep queue (all threads blocked on \texttt{sleep} -- we treat these as runnable, because a finite number of timer ticks can cause these to run), and a descheduled queue (all other existing threads).
-Landslide uses the user-provided scheduler annotations (see \S\ref{sec:instrument}) to keep this state up-to-date.
+Landslide uses the user-provided scheduler annotations (described in \S\ref{sec:instrument}) to keep this mirror image up-to-date.
 
@@
 This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
-We have never found false positives with this technique, although false negatives have arisen when too few previous interleavings have been tested to make a reliable comparison.
+We have never found this technique to erroneously report bugs (false positives), although it may fail to trigger when it should if too few previous interleavings have been tested to make a reliable comparison.
 
-When Landslide identifies a bug, it aborts execution and generates a {\em decision trace}.
+When Landslide identifies a bug, it halts execution and generates a {\em decision trace}.
 The trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, its stack trace when it was switched away from, and the thread that we caused to preempt it.
@@
 We provide a set of annotation functions, called \texttt{tell\_landslide}, for this purpose.
-The annotations denote when a thread is doing \texttt{fork}, \texttt{sleep}, or \texttt{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
-There is also a configuration file, \texttt{config.landslide}, in which the user must specify constant information such as the names of the timer handler and context switcher, which threads exist when the kernel boots up, and which userspace test program to run.
-Finally, there are two two-line functions used within Landslide itself that the user must implement.
+The annotations denote when a thread runs \texttt{fork}, \texttt{sleep}, or \texttt{exit}, when a thread is added to or removed from the runqueue, and when threads become blocked on mutexes.
+There is also a configuration file, \texttt{config.landslide}, in which the user must specify constant information such as the function names of the timer handler and context switcher, which threads exist when the kernel boots, and which userspace test program Landslide should invoke.
+Finally, there are two short (nominally two-line) functions used within Landslide itself that the user must implement.
 These are predicates on the kernel's scheduler state, and express potentially nontrivial conditions: whether the current thread is runnable but not on the runqueue, and whether preemption is disabled while interrupts are on.
@@
 
-After successfully instrumenting their kernel and exploring a minimal state space, the user should then configure Landslide's search to be more likely to find suspected bugs.
+After the user successfully instruments their kernel and explores a minimal state space, they should then configure Landslide's search to be more likely to find suspected bugs.
 For brevity, we omit discussion of some conceptually unimportant configuration options.
 
-{\bf Configuring decision points.} While Landslide automatically identifies decision points on voluntary reschedules, these will result in coarse-grained interleavings likely to overlook bugs.
+{\bf Configuring decision points.} Using only the decision points automatically identified on voluntary reschedules will result in coarse-grained interleavings likely to overlook bugs.
 We provide an extra annotation, \texttt{tell\_landslide\_decide()}, for the user to add more decision points for a finer-grained search.
@@
 This strategy may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated system calls.
-We provide interface options for the user to view currently identified decision points and to selectively limit them.
+We provide interface options in \texttt{config.landslide} for the user to view currently identified decision points and to selectively limit them.
 First, the user writes ``\texttt{DECISION\_INFO\_ONLY}'' to make Landslide run only one interleaving and then print all decision points instead of continuing to explore.
@@
 
-We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with this year's students before they submitted their kernel for grading to see if they could find bugs on their own with Landslide.
+We evaluated Landslide in two ways: first, by instrumenting two kernels we were familiar with to measure time spent to find different races, and second, by meeting with this year's 15-410 students before they submitted their kernel for grading to see if they could find bugs on their own with Landslide.
 
-We instrumented one kernel written by an author in a previous year, and also one kernel that author manually graded as a teaching assistant.
+We instrumented one kernel written by an author in a previous year, and also one kernel the same author manually graded as a teaching assistant.
 We configured Landslide to search for five complicated already-known race conditions.
-In addition to finding all five races, Landslide also found a sixth previously unknown race in the author's own kernel.
+In addition to finding all five races, Landslide also found a sixth previously unknown race in said author's own kernel.
 Using decision points only on calls to \texttt{mutex\_lock} and on voluntary reschedules, Landslide found each of the six bugs in 11 to 57 seconds, executing between 1 and 377 distinct interleavings for each.
 
-In the current student user study, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
-Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two race conditions and two deterministic errors.
+In the user study of current students, we found that students spent on average 119 minutes (60 to 158) on the required instrumentation, and a further 36 minutes (10 to 60) refining Landslide's search.
+Of the four groups who finished the required part, all four found previously unknown bugs with Landslide: two races and two deterministic errors.
 These bugs manifested as infinite loops, a kernel panic, and a use-after-free.

### 2012-06-10 rv/paper.tex
cite dc again

@@
 % In comparison with conventional stress testing, systematic exploration is less likely to overlook bugs, due to its deterministic testing approach.
-While other dynamic analysis techniques exist for finding race conditions, such as data race detection, % TODO cite here
+While other dynamic analysis techniques exist for finding race conditions, such as data race detection \cite{datacollider},
 systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work and Conclusion}
+\section{Conclusion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-06-10 rv/paper.tex
final version of paper

@@
 
+% \newcommand\squish{\vspace{-8pt}}
+\newcommand\squish{}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Introduction}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In kernel space, race condition debugging becomes even more difficult. Many aspects of the concurrency implementation itself are part of the system being tested.
-While user-space testing frameworks operate above clear abstractions and interfaces, a systematic exporation tool for kernels cannot rely on such abstractions and interfaces.
+While user-space testing frameworks operate above clear abstractions and interfaces, a systematic exporation tool for kernels cannot rely on such easy means of understanding the system under test.
 Furthermore, execution in the scheduler and in similar components is common to all concurrent threads, even otherwise independent ones; this threatens our ability to achieve state-space reduction.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Related Work}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Among other systematic testing tools are CHESS \cite{chess}, with a heuristic for search ordering; MaceMC \cite{macemc} and MODIST \cite{modist}, which test networked and distributed applications; RacePRO \cite{racepro}, which uncovers inter-process races; and DeMeter \cite{demeter}, which focuses its search on only one component of a system at a time.
+Among other user-space systematic testing tools are CHESS \cite{chess}, with a heuristic for search ordering based on number of preemptions; MaceMC \cite{macemc} and MODIST \cite{modist}, which test networked and distributed applications; RacePRO \cite{racepro}, which uncovers inter-process races around system call invocations; and DeMeter \cite{demeter}, which focuses its search on only one component of a system at a time.
 
@@
 % Carburizer \cite{carburizer} is a tool for enhancing resiliency of imperfect kernel drivers in the presence of faulty peripheral devices.
-SimTester \cite{simtester} is another driver testing tool; like Landslide, it runs in Simics, and injects interrupts to control the system's concurrency.
+SimTester \cite{simtester} is a tool for testing kernel drivers; like Landslide, it runs in Simics, and injects interrupts to control the system's concurrency.
 SimTester focuses on races involving device interrupts, whereas Landslide focuses on non-driver inter-thread races.
-Additionally, SimTester injects only one interrupt per test run, which is both less expensive and less powerful than systematic exploration.
+Additionally, SimTester injects only one interrupt per test run, which is less expensive than systematic exploration, but less powerful as well.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Scope}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-To constrain the problem space some, we aimed to use a less complex kernel than production kernels such as Linux and Windows.
+To constrain the problem space some, we aimed to use a less complex kernel than production kernels such as Linux or Windows.
 We built Landslide to test Pebbles kernels, which students at CMU develop from scratch in a 6 week project.
@@
 Simics can load Landslide while booting a kernel, and calls into Landslide once per instruction and once per memory access of the simulated execution.
-Landslide will then perform systematic exploration on that kernel using Simics interface calls during the simulated execution.
+Landslide will then perform systematic exploration on that kernel using Simics interface calls to control the kernel's execution.
 
-Landslide assumes that timer interrupts are the only nondeterministic events that control a kernel's concurrent execution.
+Landslide assumes that timer interrupts are the only nondeterministic events that influence a kernel's concurrent execution.
 While this prevents us from finding races related to external input such as disk or network I/O, Landslide is already able to find many types of complicated races by controlling timer-driven thread scheduling.
@@
 % FIXME: maybe say in each paragraph how we address the issue?
-% TODO: get rid of this section
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Design}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \texttt{yield}).
-We provide an interface for configuring additional decision points (described in \S\ref{sec:config}).
+We provide an interface for configuring additional decision points, described in \S\ref{sec:config}.
 
+\squish
 \subsection{Exploration Mechanism}
+\squish
 
@@
 
-{\bf Thread scheduling.} Compared with user space testing frameworks, Landslide cannot rely on kernel-provided system calls to control the concurrent execution of the system under test.
+{\bf Thread scheduling.} Compared with user-space testing frameworks, Landslide cannot rely on kernel-provided system calls to control the concurrent execution of the system under test.
 Instead, Landslide has an internal abstraction for scheduling arbitrary threads built upon injecting timer interrupts.
@@
 
+\squish
 \subsection{Identifying Bugs}
+\squish
 
@@
 This is done by comparing the length of the current interleaving against previously-explored ones: if the current interleaving has been running disproportionately longer (by a large arbitrary constant factor), it indicates the kernel is likely stuck.
-We have never found this technique to erroneously report bugs (false positives), although it may fail to trigger when it should if too few previous interleavings have been tested to make a reliable comparison.
+We have never found this technique to erroneously report bugs (false positives), although it may fail to trigger when it ought to if too few previous interleavings have been tested to make a reliable comparison.
 
 When Landslide identifies a bug, it halts execution and generates a {\em decision trace}.
-The trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, its stack trace when it was switched away from, and the thread that we caused to preempt it.
+This trace reports what kind of bug was detected (for uses-after-free, it also prints stack traces for when the block was allocated and freed), and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that we caused to preempt it.
 With this information, the user can better understand the concurrent execution that led to the race.
@@
 
+%\begin{figure}
+%	\centering
+%	{\small
+%	\begin{tabular}{l}
+%	\texttt{void add\_to\_runqueue(tcb\_t *thread) \{} \\
+%	\texttt{~~~~tell\_landslide\_on\_rq(thread->id);} \\
+%	\texttt{~~~~list\_insert(\&runqueue, thread);} \\
+%	\texttt{\}} \\
+%	\end{tabular}
+%	}
+%	\caption{Example annotated code.}
+%\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{User Interface}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Instrumenting Kernels with Landslide}
+\squish
 \label{sec:instrument}
@@
 
+\squish
 \subsection{Configuring Landslide}
+\squish
 \label{sec:config}
@@
 Then, the user writes ``\texttt{within\_function foo}'' to whitelist function \texttt{foo}, and ``\texttt{without\_function bar}'' to blacklist function \texttt{bar}.
-For example, if testing thread death and reaping, the user should write \texttt{within\_function exit}, and \texttt{within\_function wait}, and perhaps \texttt{without\_function destroy\_address\_space}.
+For example, if testing thread death and reaping, the user should write \texttt{within\_function exit}, and \texttt{within\_function wait}, and perhaps \texttt{without\_function destroy\_address\_space} (if they don't care about the virtual memory operations associated with \texttt{exit}).
 
@@
 Unchecked, this would cause every transition to conflict with each other and result in no reduction in execution time from DPOR.
-Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested.
-For example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
+Certain other shared memory conflicts also arise that are irrelevant to whichever system calls are being tested; for example, if testing for races in thread lifecycle routines, the user likely does not care about accesses to the frame allocator mutex.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Evaluation}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\squish
 \section{Conclusion}
+\squish
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% prob. don't actually say future work here; say it inline
-% TODO
+We have built Landslide, a tool for systematic exploration on kernel-level code.
+Landslide runs as a Simics module during simulated execution of Pebbles kernels.
+It uses user-provided instrumentation to keep track of the kernel's state, such as the set of runnable threads, and it provides the user configuration options for making its search more effective.
+It identifies decision points in the kernel around which to interleave thread executions, and it controls how threads preempt each other by injecting timer interrupts.
+We found that Landslide identified previously unknown bugs in kernels we were already familiar with, and also that novice kernel programmers were able to find otherwise-overlooked bugs with Landslide without investing large amounts of time.
 

### 2012-06-10 rv/paper.tex
merge in garth's conclusion

@@
 \squish
-\section{Conclusion}
+\section{Concluding Remarks}
 \squish
@@
 We have built Landslide, a tool for systematic exploration on kernel-level code.
-Landslide runs as a Simics module during simulated execution of Pebbles kernels.
-It uses user-provided instrumentation to keep track of the kernel's state, such as the set of runnable threads, and it provides the user configuration options for making its search more effective.
-It identifies decision points in the kernel around which to interleave thread executions, and it controls how threads preempt each other by injecting timer interrupts.
-We found that Landslide identified previously unknown bugs in kernels we were already familiar with, and also that novice kernel programmers were able to find otherwise-overlooked bugs with Landslide without investing large amounts of time.
+% Landslide runs as a Simics module during simulated execution of Pebbles kernels.
+It uses user-provided instrumentation to keep track of the kernel's state, such as the set of runnable threads, and it provides configuration options for making its search more effective.
+%It identifies decision points in the kernel around which to interleave thread executions, and it controls how threads preempt each other by injecting timer interrupts.
+We found that it was able to identify previously unknown bugs in familiar kernels, and also that novice kernel programmers were able to find otherwise-overlooked bugs with Landslide without investing large amounts of time.
+
+Our experience with Landslide to date indicates that systematic exploration is feasible in environments that lack clean standardized interfaces like message passing.
+% Operating systems kernels, tied strongly to hardware details, external events, timing, efficient execution and shared memory, are such environments.
+We believe our techniques can be applied to mainstream production kernels, especially Linux.
+% Reflection on the evolution of the Linux operating system indicates that kernel developers extensively define and discuss internal interfaces, even as they change them, so requiring kernel developers to expose a little about these interfaces seems feasible.
+Moreover, our experience with Landslide differs from user-level systematic testing in that we emphasize user guided testing rather than exhaustive testing.
+This is suggestive of tools in which bug finding for races can be effective for complex concurrent codes long before we have the computing resources or state space reduction techniques to allow exhaustive testing to be feasible.
 

### 2014-04-03 userguide/guide-short.tex
Add shorter landslide guide.

\documentclass{article}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{multirow}
\usepackage{upgreek}
\usepackage{epsfig}
\usepackage[bottom]{footmisc}
\usepackage[charter]{mathdesign}

\begin{document}
\captionsetup{width=.75\textwidth,font=small,labelfont=bf}
\title{\bf Landslide User Guide}
\author{Ben Blum (\texttt{bblum@cs.cmu.edu})}
\maketitle

\newcommand\true{\;\textit{true}}
\newcommand\false{\;\textit{false}}

\newcommand\alpher\alpha
\newcommand\beter\beta
\newcommand\gammer\gamma
\newcommand\delter\delta
\newcommand\zeter\zeta
\newcommand\Sigmer\Sigma

\newcommand\NN{\mathbb{N}}
\newcommand\QQ{\mathbb{Q}}
\newcommand\RR{\mathbb{R}}
\newcommand\ZZ{\mathbb{Z}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

This document is a guide to instrumenting a Pebbles kernel for Landslide. If you have not seen the associated lecture, please review the slides on the course website (\texttt{http://www.cs.cmu.edu/{\textasciitilde}410/}).
A longer, more comprehensive, and slightly outdated version of this guide can be found at \texttt{http://bblum.net/landslide-guide.pdf}.
This version is intended to get you started as soon as possible.
Nevertheless, I must bore you one final time with the words of warning.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Perspective}
\label{sec:expect}

When working with Landslide, please keep realistic expectations about the nature of the tool.

Landslide is not a stress tester or a fuzzer, like the \texttt{cho} family of test cases.
Tests such as \texttt{cho} attempt to break the kernel by making many system calls in as many different calling patterns as possible, and hoping that some of them will fail.
Landslide attempts to break the kernel by exploring all possible interleavings of kernel threads in one particular pattern of a few system calls.
Systematic and stress testing are orthogonal, incompatible testing strategies.

Landslide is not a ``race condition oracle''. If you run Landslide with a certain test case and a certain set of preemption points, and it finishes exploring the preemption tree and found no bugs, it does not mean that the system calls that got executed are free of race conditions.
It does, however, mean that the interactions caused by the test case's pattern of system calls do not race at the granularity of the specific preemption points chosen.
It is up to your human intuition to choose an appropriate set of preemption points, and if Landslide says ``I didn't find a race'', it is up to you to judge whether that assurance is strong enough.

%Landslide is a {\em framework} that enables you, the programmer, to explore properties of a concurrent test case that you previously could not. It relies on you to configure it accurately in accordance with your goals: whether you are seeking to find a race that you suspect exists, or seeking to show that your kernel doesn't race on a particular test, it is your responsibility to choose a set of preemption points that is relevant, granular enough to yield a meaningful exploration, yet minimal enough that Landslide can finish in a reasonable amount of time. More on this in section~\ref{sec:choice}.

%While instrumenting your kernel, it may seem tempting to look for ``simpler'' ways of circumventing some of the more complicated parts of the instrumentation process. Please do not do this - it will make your instrumentation brittle, and quite likely to break in more mysterious ways. If you are confused about what some part of the process is asking you to do, do not guess randomly (it will lead to frustrating crashes!); ask Ben for help instead.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Roadmap}
\label{sec:stepbystep}

I recommend you work on the Andrew Linux machines, as they are supported by the 410 environment, and Landslide has only been tested with Simics installation on AFS.

%As a way of keeping the attack plan steps small, we recommend not adding/configuring custom decision points until after getting Landslide working with the automatic minimal set of decision points (i.e., it explores the tree to completion).

\begin{itemize}
	\item Get the repository: \texttt{git clone git://github.com/bblum/landslide.git}
		\footnote{All directory paths henceforth are relative to the base directory of the repository.}
	\item Read section~\ref{sec:requirements} to make sure your kernel is Landslide-compatible.
	\item Build your kernel for Landslide\dots
	\begin{itemize}
		\item Run \texttt{prepare-workspace.sh}
		\item Copy your kernel sources/build directory into \texttt{pebsim/}.
		\item Add \texttt{tell\_landslide.c} and \texttt{tell\_landslide.h} to your kernel, and add \texttt{tell\_landslide.o} to your kernel's \texttt{config.mk}.
		\item Copy in the test program(s) you want to run, from \texttt{tests/}, and add them to your kernel's \texttt{config.mk}. (section~\ref{sec:testcase})
		\item Add calls to the annotation functions in your kernel. (section~\ref{sec:tell})
		\item Build your kernel, and put \texttt{bootfd.img} and \texttt{kernel} in \texttt{pebsim/}.
	\end{itemize}
	\item Fill out \texttt{config.landslide} in \texttt{pebsim/}. (section~\ref{sec:config})
	\item In \texttt{pebsim/}, run \texttt{./landslide}. This will build everything (if necessary), and then run Landslide.
	\item Iterate reconfiguring Landslide until satisfied with the results. (section~\ref{sec:choice})
	\item Fill out feedback. (appendix~\ref{sec:30mins})
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Kernel Requirements}
\label{sec:requirements}

\subsection{Scheduler Functionality}
\label{sec:schedfunc}

Landslide assumes that timer interrupts will always trigger context switches, and that if a thread is ``runnable'', at any point where interrupts/preemption is enabled, a finite number of timer preemptions will eventually cause that thread to run.

The kernel must not spin-wait anywhere, especially in \texttt{readline} or \texttt{sleep}. The kernel must also never run the idle loop when there are other runnable threads.

\subsection{VM}

The kernel should direct-map kernel memory (i.e. below \texttt{USER\_MEM\_START}). It should also use the LMM allocator (i.e. the \texttt{malloc} package provided in the base code).

\subsection{System Calls}

The following system calls must be implemented and working in at least a rudimentary way:
\texttt{fork}, \texttt{exec}, \texttt{vanish}, \texttt{wait}, \texttt{readline}.

The following system calls are exercised in some, but not all, of the included test programs, so will be helpful to have:
\texttt{thread\_fork}, \texttt{yield}.

Apart from the system calls which init and shell execute, the rest are irrelevant and do not need to be implemented to use Landslide.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Instrumenting Your Kernel with Landslide}

\subsection{Annotations}
\label{sec:tell}

The kernel should call certain annotation functions, provided in \texttt{tell\_landslide.c}, to communicate what interesting events are happening.

{\bf Important note about runqueues}.
Two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, track the state of the scheduler's runqueue at each point during execution.
Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not your kernel keeps the current thread on the runqueue. Use \texttt{CURRENT\_THREAD\_LIVES\_ON\_RQ} in the config file (next section) to deal with this if necessary.

%\subsubsection{Required Annotations}
\begin{itemize}
	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in your context switcher, when a new thread is switched to.
	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when your scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations. (This corresponds to \texttt{starting\_threads} in section~\ref{sec:config}.)
	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)

		Call it ``just before'' the action which makes the new thread runnable - whether your kernel just adds it to the runqueue (in which case the next annotation called would be \texttt{thread\_on\_rq}) or begins running it immediately (in which case \texttt{thread\_switch}); Landslide handles both cases.\footnote{It may help to think of this as setting a ``currently forking'' state flag in the calling thread: when the next annotation is called, Landslide will see the flag is set and act accordingly.
		%It is okay if interrupts are on between this call and the subsequent event.
		}
	\item \texttt{tell\_landslide\_vanishing()} - Call this when a thread is about to go away.
		Call it ``just before'' the relevant context switch; i.e., the one that should never return.
	\item \texttt{tell\_landslide\_sleeping()} - Call this when a thread is about to go to sleep (that's \texttt{sleep()} the system call).
		Call it ``just before'' the relevant context switch; i.e., the one that won't return for as long as the thread requested to sleep.
	\item \texttt{tell\_landslide\_thread\_on\_rq(int tid)} - Call this when a thread is about to be added to the runqueue.\footnote{If there are multiple such places, which is likely if you use \texttt{variable\_queue}, I recommend wrapping them in a helper function which contains the annotation.}
		%(Make sure this call is done within whatever protection is also used for the runqueue modification itself.)
	\item \texttt{tell\_landslide\_thread\_off\_rq(int tid)} - Call this when a thread is about to be removed from the runqueue.
		%(Same protection clause as above applies.)
	%\item Mutex annotations - These are only necessary if your kernel uses mutexes that leave blocked threads on the runqueue, rather than explicitly descheduling them.
	%\begin{itemize}
	%	\item \texttt{tell\_landslide\_mutex\_locking(void *mutex\_addr)} - Call this when a thread is about to start locking a mutex. Landslide needs the mutex address for deadlock detection, to track which mutex each thread is blocked on.
	%	\item \texttt{tell\_landslide\_mutex\_locking\_done()} - Call this when a thread finishes locking a mutex (and now owns it).
	%	\item \texttt{tell\_landslide\_mutex\_unlocking(void *mutex\_addr)} - Call this when a thread is about to start unlocking a mutex.
	%	\item \texttt{tell\_landslide\_mutex\_unlocking\_done()} - Call this when a thread finishes unlocking a mutex (and no longer owns it).
	%	\item \texttt{tell\_landslide\_mutex\_blocking(int owner\_tid)} - Call this when a thread is contending on a locked mutex, and hence is no longer logically ``runnable'' yet still on the runqueue. Landslide needs to know who owns the mutex for deadlock detection.
	%\end{itemize}
\end{itemize}

If your kernel uses mutexes that leave blocked threads on the runqueue but blocked in some implicit sense, rather than explicitly descheduling them, you will also need to use the \texttt{tell\_landslide\_mutex\_*} annotations.

%\subsubsection{Optional Annotations}
%\begin{itemize}
%	\item \texttt{tell\_landslide\_decide()} - Call this to define additional decision points. See section~\ref{sec:choice}.
%	\item \texttt{assert()} or \texttt{panic()} - You should be using these already, but they are extra useful in Landslide.
%		The more important invariants for which you have \texttt{assert}s in your kernel, the more likely Landslide is to find bugs that would trigger them - otherwise, even if they do happen, Landslide might never know (just like in conventional stress testing).
%		\footnote{Relatedly, a useful trick is to use poison values for memory that should not be read from until it is written again, and to use canary members in data structures that should not get scribbled over. Functions that operate on such memory can assert that poison values don't appear and that canary values don't change.}
%\end{itemize}

\subsection{Config File}
\label{sec:config}

You will also need to fill out the file \texttt{config.landslide}. Some of the fields are required information, and some are tweaks that change the way Landslide behaves. They are documented in more detail in the comments in the file.

\subsubsection{Required Fields}
\begin{itemize}
	\item \texttt{KERNEL\_IMG} - The build scripts use \texttt{objdump} on this file to extract code and data addresses.
	\item \texttt{KERNEL\_SOURCE\_DIR} - Used for file and line number information in stack traces.
	\item \texttt{TEST\_CASE} - What userspace program should Landslide explore the state space of?
	\item \texttt{TIMER\_WRAPPER} - Used for controlling timer interrupts and tracking whether a thread is handling an interrupt.
	\item \texttt{CONTEXT\_SWITCH} - Used for detecting voluntary reschedules.
	\item \texttt{READLINE} - Used for detecting when the kernel is ready to run a test, and when the test finishes.
	\item \texttt{INIT\_TID}, \texttt{SHELL\_TID}, \texttt{IDLE\_TID}, \texttt{FIRST\_TID} - Used for initialization, and tracking of test state.
	\item \texttt{starting\_threads} - This lets you tell Landslide what threads exist at the point where \\ \texttt{tell\_landslide\_sched\_init\_done} will be called. You should probably use it for each of init and (if it exists in your kernel) idle (but probably not shell).
	\item \texttt{CURRENT\_THREAD\_LIVES\_ON\_RQ} - Set to 1 or 0 to indicate whether or not your kernel keeps the currently-running thread on the runqueue.
\end{itemize}

{\bf Important note about preemptions.} Some schedulers are designed in such a way that sometimes, interrupts may be enabled but the kernel is ``not ready'' for a timer interrupt to cause a context switch. If your kernel uses such a design, ask Ben about what extra config options should be set to deal with this.

\subsubsection{Optimizations}

Landslide prunes redundant thread interleavings out of the state space with an algorithm called {\em partial order reduction}. This algorithm determines which interleavings are redundant by identifying
{\em independent} transitions between preemption points - that is, they that have no conflicting memory accesses.

However, as this is the kernel, every such transition will inevitably touch some scheduler state (such as the runqueue and the timer tick count) that we don't care about.
\texttt{config.landslide} provides some options for ignoring irrelevant memory accesses when it comes to pruning the state space. {\bf Directing Landslide to ignore certain memory accesses will prune the state space better, but sacrifice the ability to find races related to those accesses.}

\begin{itemize}
	\item \texttt{sched\_func} - Landslide will ignore shared memory accesses from scheduler functions.\footnote{Accesses to scheduler data structures happen during {\em every} transition, so ignoring these is necessary for achieving any reduction at all.}
	\item \texttt{ignore\_sym} - You may wish to ignore memory accesses from other global data structures as well, such as certain irrelevant global mutexes or counters.
\end{itemize}

If you configure preemption points to happen in common code paths, such as \texttt{mutex\_lock}, you may have way too many preemption points on your hands. (For example, if you're testing \texttt{vanish}, there's no need to branch on a \texttt{mutex\_lock} in \texttt{exec}.) \texttt{within\_function} and \texttt{without\_function} allow you to configure which parts of your kernel preemption points are allowed to show up in.

\begin{itemize}
	\item \texttt{within\_function} - Whitelist functions so that Landslide will only preempt if the current thread is ``within'' one of the functions listed.
	\item \texttt{without\_function} - Similar to above, but a blacklist. These two can be used together; later invocations take precedence over earlier ones.
\end{itemize}

\subsubsection{Behaviour Tweaks}
\begin{itemize}
	\item \texttt{BUG\_ON\_THREADS\_WEDGED} - Landslide's deadlock detection is not perfect. The default is to let the kernel keep running, but if it seems like you have an ``all threads are wedged'' bug, turn this on.
	\item \texttt{EXPLORE\_BACKWARDS} - In which order to explore the branches? Backwards means with more preemptions first; forwards means tending to let the current thread keep running more often. Backwards tends to find bugs more quickly, but produces longer (more difficult to understand) preemption traces.
	\item \texttt{DONT\_EXPLORE} - This option makes Landslide stop after one branch, and output the list of preemption points it identified using the current config. Useful if you want to see all preemption points, and tweak settings to get rid of frivolous ones.
	\item \texttt{BREAK\_ON\_BUG} - If you want to drop into the simics debug prompt when a bug is found, turn this on. By default Landslide will instead quit simics entirely.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Userspace Test Cases}
\label{sec:testcase}

Landslide comes with a suite of small test cases designed to expose several common races in Pebbles kernels.\footnote{You may wonder why some of these are written in assembly.
This is because using your userspace thread library for thread creation (a) is not guaranteed to work, (b) might forcibly serialise certain operations that would expose kernel bugs only if concurrent, and (c) necessitates many more system calls, which dramatically increases the size of the state space.}

\begin{itemize}
	\item \texttt{vanish\_vanish} - Tests when a parent and child process \texttt{vanish()} simultaneously. (This test is identical to \texttt{fork\_test1} from the P3 hurdle suite.)
	\item \texttt{fork\_wait} - Tests basic interaction of \texttt{fork()} and \texttt{wait()}. (This test is from the P3 hurdle suite.)
	%\item \texttt{wild\_test2} - Tests simultaneous thread death by unnatural causes. (This test is an upgrade of \texttt{wild\_test1} from the P3 hurdle suite.)
	\item \texttt{double\_wait} - Tests interactions of multiple waiters on a single child. (new)
	\item \texttt{double\_thread\_fork} - Tests for interactions of multiple threads in one process vanishing. (new)\footnote{This test tests for the use-after-free bug presented in lecture - if your \texttt{thread\_fork} implementation has ``\texttt{return child->tid;}'' as its last line of code.}
	\item \texttt{yield\_vanish} - Tests for interactions between \texttt{yield()} and \texttt{vanish()}. (new)
\end{itemize}

It may help your intuition to note why all of these test cases are ``Landslide-friendly'': they all perform very little work on a single run, enabling Landslide to completely explore the state space. They also run several, but not too many, threads at once, producing potentially interesting interleavings.

To warm up, I recommend you test \texttt{double\_thread\_fork} (and edit your kernel explicitly to introduce the associated bug - see the footnote in the test's description above).
After that, I recommend trying \texttt{vanish\_vanish}, which is the most likely to expose many different types of bugs in many different kernel designs.
But of course all of them are interesting -- with \texttt{double\_wait} I once found an obscure panic in my student kernel several years later, that neither I, the TA who graded me, or the stress tests knew about.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Configuring Preemption Points}
\label{sec:choice}

Getting a good set of preemption points is important to being able both to explore reasonably quickly and to produce meaningful interleavings that are likely to find bugs. By default, Landslide only preempts during ``voluntary reschedules'' -- that is, whenever your kernel does \texttt{yield()} of its own accord. After you get Landslide working with this default set of preemption points, it's time to get creative.

In no particular order, think about the following things when building your set of preemption points.

\begin{itemize}
	\item Use \texttt{tell\_landslide\_preempt()} to indicate preemption points. Recommended places are at the start of \texttt{mutex\_lock}, and if that doesn't find bugs, at the end of \texttt{mutex\_unlock} as well. But use your own intuition, depending on what system call(s) you're testing.
	\item Use \texttt{sched\_func} and \texttt{ignore\_sym} to make Landslide ignore global memory accesses, to enable better pruning.
	\item Use \texttt{DECISION\_INFO\_ONLY} to examine the set of preemption points, and figure out which ones you want to get rid of.
	\item Use \texttt{within\_func} and \texttt{without\_func} to make Landslide only pay attention to parts of your kernel that you care about.
\end{itemize}

%Example tip: If you \texttt{tell\_landslide\_preempt()} in \texttt{mutex\_lock}, you could surround the call with an if statement that makes sure the mutex is not some global mutex you don't care about. You could also (or instead) make Landslide ignore it by using \texttt{without\_func} on whatever function takes the mutex you don't care about.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Interpreting Landslide's Results}
%
%Landslide prints a lot of information. Most of it is only useful for development, but you may find some of it interesting anyway.
%
%The most important thing that Landslide prints is the preemption trace, when it finds a bug. Landslide prints the following information:
%
%\begin{itemize}
%	\item Before proclaiming ``A bug was found!'', there should be a message in bright red explaining what the bug is.
%	\item The trace itself consists of a list of the preemption points that were encountered in the buggy interleaving. For each preemption point, Landslide says what thread used to be running, what thread it chose to run next, and the stack trace of the old thread at the point it was switched away from.
%	\item At the end, there will be a stack trace for the current point in time, which will be where the bug was detected, and not a preemption point itself.
%	\item Also at the end there will be some statistics about the exploration: how many preemption points were visited, how many backtracks were taken (i.e., how many interleavings were explored), the average number of preemption points per branch, etc.
%\end{itemize}
%
%We ship Landslide with a somewhat high verbosity setting. If you want it to be quieter, you can edit it in \texttt{work/modules/landslide/common.h} - \texttt{CHOICE} is the recommended minimum level (\texttt{INFO} is highest, \texttt{DEV} is default).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Feedback}
%\label{sec:feedback}
%
%Ben needs feedback about your experience using Landslide to help him present his thesis. He wants to find answers to the following questions:
%
%\begin{itemize}
%	\item What ways can Landslide be improved to make it a better race-finding tool? (This means in terms of the high-level process of finding useful preemption points, less so in terms of nasty implementation details - the latter would be good for development and bug reports, but has little research value.)
%	\item Does using Landslide make you think differently about race conditions, in a way that makes you better at debugging them, even at times when you're not using Landslide?
%	\item Is Landslide effective at finding bugs, when properly configured? Also, is it effective at saying it found no bug when using a reasonably-thorough set of preemption points?
%\end{itemize}
%
%Of course, you shouldn't try to answer those questions directly. But keep them in mind as the goal of the study, when answering the following questions.
%
%\subsection{While using Landslide}
%
%While you're using Landslide, please try to take 5 minutes approximately every half-hour to give short updates. For convenience, you can print out appendix~\ref{sec:30mins} and fill it in as you go.
%
%\subsection{After using Landslide}
%
%After you've finished with Landslide, please write some conclusions about your experience:
%
%\begin{itemize}
%	\item Were you able to get Landslide to work with your kernel (i.e., run through a minimal exploration tree completely)? If so, how long did it take, from sitting down to getting it to work? If not, did your kernel do something that was incompatible with Landslide, or were you unable to get the instrumentation right for some other reason?
%	\item Did you find bugs while using Landslide, that you imagine would have been very hard to find otherwise? Did Landslide's output help you understand what caused them?
%	\item Were you able to get Landslide to say ``you survived!'' with custom preemption points? Did you think the set of preemption points you used for this provide a strong guarantee about the absence of races?
%	\item Describe your experience configuring the set of preemption points. What was intuitive, obvious to do? Did you feel stuck at any point, not knowing where to go next?
%	\item Was there anything you wanted to make Landslide do that it didn't support?
%\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
% TODO A section for debuggng landslide problems

\newpage
\section{While-You-Work Feedback Form}
\label{sec:30mins}

Questions to be thinking about while you're working:
\begin{itemize}
	\item How much time did you spend on each phase of the process? (annotating your kernel, doing \texttt{config.landslide}, fixing instrumentation problems/crashes, customizing preemption points for more efficient search, or trying to understand/solve a race that was found)
	\item What problems were you fighting, getting things to work? (conceptual understanding, fixing instrumentation, finding good preemption points)
	\item Did you have to change your kernel at all to make it compatible?
	\item Is there a feature you wish Landslide had to help you out, that it doesn't offer?
	\item Did you realize anything interesting or surprising about Landslide, or about your kernel?
\end{itemize}
Please fill out a row after every 30-60 minutes spent. Approximate time estimates are totally fine.

%\addtolength{\oddsidemargin}{-.5in} 
\newcommand\asdf[1]{\hspace{-0.05in}\small{#1}\hspace{-0.05in}}
\begin{center}
\begin{tabular}{|c||c|c|c|c|c||c|}
	\hline
	\multirow{2}{*}{Time} & \multicolumn{5}{|c||}{Minutes spent doing\dots} & Remarks \\
	\cline{2-7}
	& \asdf{annotate} & \asdf{config} & \asdf{fixing} & \asdf{customizing} & \asdf{found races} & \hspace{0.7in} (problems, realizations, features) \hspace{0.7in} \\
	\hline \hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
	& & & & & & \\
	& & & & & & \\
	& & & & & & \\
	\hline
\end{tabular}
\end{center}

\end{document}

### 2014-04-04 userguide/guide-short.tex
incorporate dave's edits to guide-short

@@
 		\footnote{All directory paths henceforth are relative to the base directory of the repository.}
-	\item Read section~\ref{sec:requirements} to make sure your kernel is Landslide-compatible.
+	\item Read Section~\ref{sec:requirements} to make sure your kernel is Landslide-compatible.
 	\item Build your kernel for Landslide\dots
@@
 		\item Add \texttt{tell\_landslide.c} and \texttt{tell\_landslide.h} to your kernel, and add \texttt{tell\_landslide.o} to your kernel's \texttt{config.mk}.
-		\item Copy in the test program(s) you want to run, from \texttt{tests/}, and add them to your kernel's \texttt{config.mk}. (section~\ref{sec:testcase})
-		\item Add calls to the annotation functions in your kernel. (section~\ref{sec:tell})
+		\item Copy in the test program(s) you want to run, from \texttt{tests/}, and add them to your kernel's \texttt{config.mk}. (Section~\ref{sec:testcase})
+		\item Add calls to the annotation functions in your kernel. (Section~\ref{sec:tell})
 		\item Build your kernel, and put \texttt{bootfd.img} and \texttt{kernel} in \texttt{pebsim/}.
 	\end{itemize}
-	\item Fill out \texttt{config.landslide} in \texttt{pebsim/}. (section~\ref{sec:config})
-	\item In \texttt{pebsim/}, run \texttt{./landslide}. This will build everything (if necessary), and then run Landslide.
-	\item Iterate reconfiguring Landslide until satisfied with the results. (section~\ref{sec:choice})
+	\item Fill out \texttt{config.landslide} in \texttt{pebsim/}. (Section~\ref{sec:config})
+	\item In \texttt{pebsim/}, run \texttt{./landslide}. This will recompile Landslide's source (if necessary), and then run Landslide.
+	\item Iterate reconfiguring Landslide until satisfied with the results. (Section~\ref{sec:choice})
 	\item Fill out feedback. (appendix~\ref{sec:30mins})
@@
 Until this is called, Landslide will ignore all other annotations.
-(This corresponds to \texttt{starting\_threads} in section~\ref{sec:config}.)
+(This corresponds to \texttt{starting\_threads} in Section~\ref{sec:config}.)
 	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
@@
 
-If your kernel uses mutexes that leave blocked threads on the runqueue but blocked in some implicit sense, rather than explicitly descheduling them, you will also need to use the \texttt{tell\_landslide\_mutex\_*} annotations.
+If your kernel uses mutexes that leave blocked threads on the runqueue but blocked in some implicit sense, instead of explicitly descheduling them, you will also need to use the \texttt{tell\_landslide\_mutex\_*} annotations.
 
@@
 	\item \texttt{TEST\_CASE} - What userspace program should Landslide explore the state space of?
-	\item \texttt{TIMER\_WRAPPER} - Used for controlling timer interrupts and tracking whether a thread is handling an interrupt.
-	\item \texttt{CONTEXT\_SWITCH} - Used for detecting voluntary reschedules.
+	\item \texttt{TIMER\_WRAPPER} - The name of the timer interrupt wrapper registered in the IDT. Used for controlling timer interrupts and tracking whether a thread is handling an interrupt.
+	\item \texttt{CONTEXT\_SWITCH} - The name of the context switch function (does not need to be the innermost and/or assembly function). Used for detecting voluntary reschedules.
 	\item \texttt{READLINE} - Used for detecting when the kernel is ready to run a test, and when the test finishes.
@@
 Useful if you want to see all preemption points, and tweak settings to get rid of frivolous ones.
-	\item \texttt{BREAK\_ON\_BUG} - If you want to drop into the simics debug prompt when a bug is found, turn this on. By default Landslide will instead quit simics entirely.
+	\item \texttt{BREAK\_ON\_BUG} - If you want to drop into the Simics debug prompt when a bug is found, turn this on. By default Landslide will instead quit Simics entirely.
 \end{itemize}
@@
 Getting a good set of preemption points is important to being able both to explore reasonably quickly and to produce meaningful interleavings that are likely to find bugs.
-By default, Landslide only preempts during ``voluntary reschedules'' -- that is, whenever your kernel does \texttt{yield()} of its own accord.
+By default, Landslide preempts only during ``voluntary reschedules'' -- that is, whenever your kernel does \texttt{yield()} of its own accord.
 After you get Landslide working with this default set of preemption points, it's time to get creative.

### 2014-04-04 userguide/guide.tex
apparently some old guide.tex edits were still lying around

@@
 
-We recommend you work on the Andrew Linux machines, as they are supported by the 410 environment, and Landslide has only been tested with Simics installation on AFS.
+We recommend you work on the Andrew Linux machines, as they are supported by the 410 environment, and Landslide has only been tested with the Simics installation on AFS.
 
@@
 \begin{itemize}
-	\item Get the repository: \texttt{http://www.cs.cmu.edu/{\textasciitilde}410/landslide.tar.bz2}
+	\item Get the repository: \texttt{git clone git://github.com/bblum/landslide.git}
 		\footnote{All directory paths henceforth are relative to the base directory of the repository.}
